---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      # Figures must live under man/figures/ -- pkgdown reads
                      # images only from there and from vignettes/, so a figure
                      # written to the github_document default of
                      # README_files/figure-gfm/ renders on GitHub but is
                      # missing from the package website.
                      fig.path = "man/figures/README-")
# Pin duckdb's extension/secret home so it does not announce the location on
# every connection; duckdbfs re-opens a connection per exit finalizer, and those
# messages escape message = FALSE.
options(duckdb.home = "~/.duckdb")
```

# obus


<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


{obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for the ordinary user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little, but hopefully well.

The package code resides on [GitHub](https://github.com/einarhjorleifsson/obus).

For purists, one regrets to inform that the functionality of {obus} rests on a quite substantial number of dependencies (see [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/main/DESCRIPTION) for the full list). Some of them may possibly be trimmed, but never all.

Access to the parquet datafiles is, however, independent of {obus} and, for that matter, of any software platform used. The current **temporary** path to the data (with the new header lingo) used in {obus} is:

```
# The exchange files:
https://heima.hafro.is/~einarhj/datras/raw/HH.parquet
https://heima.hafro.is/~einarhj/datras/raw/HL.parquet
https://heima.hafro.is/~einarhj/datras/raw/CA.parquet
# Auxiliary files:
https://heima.hafro.is/~einarhj/datras/species.parquet
...
# Experimentation with standardization of the HL data:
https://heima.hafro.is/~einarhj/datras/HL_length.parquet
https://heima.hafro.is/~einarhj/datras/HL_summary.parquet
```

One way to access these is just using {duckdb}/{duckdbfs}:

```{r}
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HH.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HL.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/CA.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/species.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_length.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_summary.parquet") |> dplyr::glimpse()
```

## What it looks like

The same few steps recur: **get** a table, **add** what you need, **summarise**
it, and **draw** it. Every step returns an ordinary table, so you can stop and
look at it, join it, or draw it yourself.

```{r setup-example}
library(obus)
library(dplyr)
library(ggplot2)

hh <- dr_con("HH") |>
  filter(Survey == "NS-IBTS", Quarter == 1, Gear == "GOV",
         Year >= 1984, HaulValidity == "V")
```

Mean catch per hour by year, with a 95% interval, for four species. Every haul
counts, the ones that caught nothing included, and the species are named by
their Latin names (or `Valid_Aphia` codes):

```{r simple-plot, fig.alt = "Mean numbers caught per hour by year, with 95 percent intervals, for four species in separate panels. Atlantic cod falls steadily from the mid-1980s to the 2020s. Haddock arrives in pulses. Whiting and herring fluctuate with no comparable trend."}
hh |>
  dr_summarise_cpue(c("Gadus morhua", "Melanogrammus aeglefinus",
                      "Merlangius merlangus", "Clupea harengus"), Year) |>
  autoplot()
```

The same hauls, grouped by ICES rectangle instead, give the share of hauls that
caught thornback ray:

```{r map-plot, fig.alt = "Map of the North Sea and the English Channel in one-degree by half-degree cells, coloured by the share of valid hauls that caught thornback ray. The share is highest in the southern North Sea and low elsewhere."}
hh |>
  filter(Year >= 2015) |>
  dr_add_cell() |>
  dr_summarise_occurrence("Raja clavata", cell_lon, cell_lat) |>
  autoplot()
```

The pieces:

| step | functions |
|---|---|
| get | `dr_con()`, `dr_con_raw()`, `dr_get()` |
| add | `dr_add_id()`, `dr_add_length_cm()`, `dr_add_length_mid()`, `dr_add_catch()`, `dr_add_cell()` |
| build the catch tables | `dr_hl_length()`, `dr_hl_summary()`, `dr_hl_collapse()` |
| summarise | `dr_summarise_hauls()`, `dr_summarise_occurrence()`, `dr_summarise_cpue()`, `dr_summarise_length()` |
| draw | `autoplot()` |
| weight from length | `dr_add_length_tl()`, `dr_add_predicted_weight()`, `dr_compare_length_weight()` |
| use with {DATRAS} | `dr_get_datras()` |

The summaries run on the lazy table and collect only the small result, so they
give the same answer whether the hauls are on the server or already in memory.
{ggplot2} cannot summarise a lazy table (it quietly collects every row first),
which is why the work is done by the verbs and the drawing is the last step.
