---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# obus

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/imbus)](https://CRAN.R-project.org/package=imbus)
<!-- badges: end -->

{obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for everyday user. Some of this may be taken up in a more offical package.

## Installation

You can install the development version of imbus from [GitHub](https://github.com/einarhjorleifsson/obus) with:

``` r
devtools::install_github("einarhjorleifsson/obus")
```

## Table connections

You can generate a connection to the main DATRAS tables by:

```{r}
library(obus)
hh <- dr_con("HH")
hl <- dr_con("HL")
ca <- dr_con("CA")
```

## A little peek

```{r}
#| warning: false
#| message: false
hh
hl
ca
```
