---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  error = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# obus

## Preamble

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The aim of {obus} is to provide users fast and efficient access to DATRAS tables with non-ambiguous variables.

That said, {obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for ordinary user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little.

The package code resides on [GitHub](https://github.com/einarhjorleifsson/obus) and there is also a [package website](https://heima.hafro.is/~einarhj/pkg/obus/).

For purists, one regrets to inform that the functionality of {obus} has quite some number of dependencies (56, see also [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/master/DESCRIPTION)). Some of them may possibly be trimmed, but never all.

Access to the parquet datafiles is though independent of {obus} and for that matter **any software platform** used. The current **temporary** path to the exchange files (with the new header lingo) used in {obus} is:

```
https://heima.hafro.is/~einarhj/datras/HH.parquet
https://heima.hafro.is/~einarhj/datras/HL.parquet
https://heima.hafro.is/~einarhj/datras/CA.parquet
```



## Installation

You can install {obus} using one of:

```{r, eval = FALSE}
remotes::install_github("einarhjorleifsson/obus")
pak::pak("einarhjorleifsson/obus")
```

There are two ways to access the DATRAS data, either by importing the whole datasets into R or by making an in-process DuckDB database connection.

```{r}
library(obus)
library(dplyr)
```

## Importing

The fastest way to **import** the full DATRAS data into R is:

```{r}
system.time({
  hh <- dr_get("HH", source = "parquet")
  hl <- dr_get("HL", source = "parquet")
  ca <- dr_get("CA", source = "parquet")
})
```

So we are talking about around 5 seconds if you're sitting on the optic fiber. If you are connected via poor wifi this may take on the order of a minute.

The dr_get is just a thin wrapper around the path stated above. User can thus access the data without the {obus} as middle man via e.g.:

```
arrow::read_parquet("https://heima.hafro.is/~einarhj/datras/HH.parquet")
```

And if Python is your preferred platform:

```
import pandas as pd
pd.read_parquet("https://heima.hafro.is/~einarhj/datras/HH.parquet")
```

Whatever the case one can assume that nobody will complain about the speed of access, given that the dimensions just imported are as follows:

```{r, echo = FALSE}
tibble(type = c("HH", "HL", "CA"),
              rows = c(nrow(hh), nrow(hl), nrow(ca)),
              cols = c(ncol(hh), ncol(hl), ncol(ca))) |> 
  knitr::kable(caption = "Number of records and variables")
```

The above fast importing is achieved by importing from parquet files that are hosted on a conventional https-server. At this stage it is unclear how much simultaneous traffic it can handle. The way things have been setup so far is just a proof of concept. By the same nature, do not expect that the datasets accessed are the latest mirror of the data residing in ICES Datras Database.

Those familiar with DATRAS data will notice when viewing the data that the variable names are based on the new lingo. If your downstream code-flow depends on the old lingo one can easily revert back via:

```{r}
hl |> dr_translate(from = "new", to = "old")
```

The old faithful (`icesDatras::getDATRAS`) is wrapped within `dr_get()` (specify `source = "xml"`), with the addition that variable types are set, -9 values are turned to NA, and one can retrieve more than one survey at a time. Here we only dare to call one year of HL-data:

```{r}
system.time({
  hl_xml <- dr_get(recordtype = "HL", years = 2026, source = "xml")
})
```

In store we now have:

```{r}
hl_xml |> count(Survey)
```

The above demonstrates the following:

* Both approaches return R data frames with standard column names and correct variable types.
* Hosting the full DATRAS dataset as parquet files on a https-server provides the fastest download and import time.
* For all practical purposes the parquet source could be embedded in `icesDatras::getDatras` — the user would simply observe faster responses and upstream type handling.

## Connecting

Although the DATRAS data can not be considered big data, one can use techniques developed for such datasets. So instead of importing the full dataset into R one can generate a connection to the **same** web-hosted parquet files as above using in-process DuckDB database. 


```{r}
system.time({
  hh <- dr_con("HH")
  hl <- dr_con("HL")
  ca <- dr_con("CA")
})
class(hl) ; nrow(hl)
hl |> glimpse()
```

Note here that beside the above function accessing the **same parquet files** as when using `dr_get`:

* the time is almost instantaneous
* the objects are some kind of a tibble of variant type sql, more specifically a duckdb_connection
* the nrow returned is NA (one can get that number via `hl |> count()`
* the data feel like R-dataframe(s)

Let's look at the hl object from another angle:

```{r}
hl |> show_query()
```

So the object hl is actually some kind of an SQL-query. What happens behind the scene when using dr_con is:

* A tiny database engine called DuckDB is started silently from R — no separate server or installation needed (The feature is installed initially on the computer when {obus} is installed).
* DuckDB opens an HTTP connection to the parquet file on the remote server and reads only its *index* (column names, types, and where each chunk of rows lives in the file). This is why the connection is nearly instantaneous — only limited number of records have travelled over the wire, just to give the user a whiff of the data.

Now we can do the usual:

```{r}
q <- 
  hl |> 
  filter(Survey == "NS-IBTS", Year == 2026, Quarter == 1)
```

What we now have in store is:

```{r}
q |> show_query()
```

Ergo: 

* We have added to the original SQL query a filter using dplyr-verbs
* We no longer have preset argument within a function (like icesDatras::getDATRAS) but supply that via the dplyr-filter function

The magic is that we now have a system were we can supply any filter to any of the variables. And any other magic one may spin up using the {dplyr} verbs. E.g.:

```{r}
q <- 
  hl |> 
  mutate(length_cm = dplyr::case_when(
          LengthCode == "-9" ~ NA_real_,          # Invalid length codes marked as NA
          LengthCode %in% c(".", "0") ~ LengthClass / 10,  # Divide by 10
          LengthCode %in% c("1", "2", "5") ~ LengthClass,  # Direct mapping
          TRUE ~ NA_real_                                  # Any other case is NA
        )) |> 
  filter(Gear == "GOV", length_cm > 50)
```

```{r}
q |> show_query()
```

We now have to decide how much code to hide from the average user. E.g. instead of the mutate above to convert all length classes to centimeters we could use an experimental function:

```{r}
q <- 
  hl |> 
  dr_add_length_cm() |> 
  filter(Gear == "GOV", length_cm > 50)
```

We can also join two or more tables:

```{r}
species <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/species.parquet")
q <- 
  hl |> 
  left_join(hh |> select(.id, DataType, HaulDuration, HaulValidity,
                         lon = ShootLongitude, lat = ShootLatitude),
            by = join_by(.id)) |> 
  left_join(species,
            by = join_by(Valid_Aphia == aphia)) |>  # Note: Valid_Aphia is the parquet column name; canonical standardisation pending ICES Datacenter confirmation
  dr_add_length_cm() |> 
  dr_add_n_and_cpue() |> 
  filter(HaulValidity == "V",
         !is.na(LengthCode),
         latin == "Gadus morhua",
         Survey %in% c("NS-IBTS", "SCOWCGFS"),
         Quarter == 1) |> 
  select(.id, Survey, Year, lon, lat, species, length_cm, n_haul, n_hour)
```

We have now built quite a substantive SQL query (without knowing any SQL):

```{r}
q |> show_query()
```

Let's now import the data:

```{r}
system.time(
  data <- q |> collect()
)
```

So we basically have obtained some ~150 thousand cod measurements from two surveys (1965 onwards) where we have:

* standardized the length
* obtained the number of cod in the haul
* standardized the number to an hour

And this less than 1 second!

It needs to be mentioned that the same code can be used if the original source (hh and hl objects) were just dataframes already imported into R, either from the parquet files or via icesDatras::getDATRAS, the latter with only some minor additional twist.

It also should be highlighted than any local bookkeeping of Datras data files is kind of obsolete, unless one expects to be off-line. In these days and ages even vessels on the high seas are rarely without decent internet connection.

## Specs

```{r, echo = FALSE}
devtools::session_info()
```
