---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

# obus

## Preamble

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


{obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for the ordinary user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little, but hopefully well.

The package code resides on [GitHub](https://github.com/einarhjorleifsson/obus).

For purists, one regrets to inform that the functionality of {obus} rests on a quite substantial number of dependencies (see [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/main/DESCRIPTION) for the full list). Some of them may possibly be trimmed, but never all.

Access to the parquet datafiles is, however, independent of {obus} and, for that matter, of any software platform used. The current **temporary** path to the data (with the new header lingo) used in {obus} is:

```
# The exchange files:
https://heima.hafro.is/~einarhj/datras/raw/HH.parquet
https://heima.hafro.is/~einarhj/datras/raw/HL.parquet
https://heima.hafro.is/~einarhj/datras/raw/CA.parquet
# Auxiliary files:
https://heima.hafro.is/~einarhj/datras/species.parquet
...
# Experimentation with standardization of the HL data:
https://heima.hafro.is/~einarhj/datras/HL_length.parquet
https://heima.hafro.is/~einarhj/datras/HL_summary.parquet
```

One way to access these is just using {duckdb}/{duckdbfs}:

```{r}
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HH.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HL.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/CA.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/species.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_length.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_summary.parquet") |> dplyr::glimpse()
```

