---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
# Pin duckdb's extension/secret home so it does not announce the location on
# every connection; duckdbfs re-opens a connection per exit finalizer, and those
# messages escape message = FALSE.
options(duckdb.home = "~/.duckdb")
```

# obus


<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


{obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for the ordinary user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little, but hopefully well.

The package code resides on [GitHub](https://github.com/einarhjorleifsson/obus).

For purists, one regrets to inform that the functionality of {obus} rests on a quite substantial number of dependencies (see [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/main/DESCRIPTION) for the full list). Some of them may possibly be trimmed, but never all.

Access to the parquet datafiles is, however, independent of {obus} and, for that matter, of any software platform used. The current **temporary** path to the data (with the new header lingo) used in {obus} is:

```
# The exchange files:
https://heima.hafro.is/~einarhj/datras/raw/HH.parquet
https://heima.hafro.is/~einarhj/datras/raw/HL.parquet
https://heima.hafro.is/~einarhj/datras/raw/CA.parquet
# Auxiliary files:
https://heima.hafro.is/~einarhj/datras/species.parquet
...
# Experimentation with standardization of the HL data:
https://heima.hafro.is/~einarhj/datras/HL_length.parquet
https://heima.hafro.is/~einarhj/datras/HL_summary.parquet
```

One way to access these is just using {duckdb}/{duckdbfs}:

```{r}
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HH.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/HL.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/raw/CA.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/species.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_length.parquet") |> dplyr::glimpse()
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL_summary.parquet") |> dplyr::glimpse()
```

One could do:

```{r}
system.time({
  library(obus)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  
  options(width = 100)
  
  focus <- c("Atlantic cod", "haddock", "whiting", "Atlantic herring")
  
  hh <- dr_con_raw("HH") |>
    dr_add_id() |> 
    filter(Survey == "NS-IBTS", Quarter == 1, Gear == "GOV",
           Year >= 1984, HaulValidity == "V") |>
    select(.id, Year) |>
    collect()
  
  catch <- dr_con("HL_summary") |>
    filter(Survey == "NS-IBTS", Quarter == 1, species %in% focus) |>
    group_by(.id, species) |>
    summarise(n_hour = sum(n_totalnumber_hour, na.rm = TRUE), .groups = "drop") |>
    collect()
  
  # every haul x every species, so that "not caught" becomes a zero
  # rather than a missing row - the next section is about why
  abundance <- crossing(hh, species = focus) |>
    left_join(catch, by = c(".id", "species")) |>
    mutate(n_hour = coalesce(n_hour, 0))
})
```

```{r simple-plot}
ggplot(abundance, aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot") +
  facet_wrap(~species, scales = "free_y") +
  labs(x = NULL, y = "Numbers per hour")
```

