---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# obus

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/imbus)](https://CRAN.R-project.org/package=imbus)
<!-- badges: end -->

The aim of {obus} to provide users with tidy DATRAS tables with non-ambiguous variables.

That said, {obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for everyday user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little.

For purist, one regrets to inform that this package has quite some number of dependencies (see [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/master/DESCRIPTION)). It should however be possible to trim down that fat.

For more information, check the [README.md](https://github.com/einarhjorleifsson/obus).

## Installation

You can install the development version of {obus} from [GitHub](https://github.com/einarhjorleifsson/obus) running:

```{r, eval = FALSE}
remotes::install_github("einarhjorleifsson/obus")
```

In some cases {obus} uses wrapper functions depending on {icesDatras} features that have, as of yet, not been taken up in the official ICES version (issues pending) install that package via:

```{r, eval = FALSE}
remotes::install_github("einarhjorleifsson/icesDatras", force = TRUE)
```

There are two ways to connect to the DATRAS data, either by importing the whole datasets into R or by making an in-process DuckDB database connection.

```{r}
library(obus)
```

## Importing

The fastest way to **import** the full DATRAS data into R is:

```{r}
system.time({
  hh <- dr_get("HH", from = "parquet")
  hl <- dr_get("HL", from = "parquet")
  ca <- dr_get("CA", from = "parquet")
})
```

So we are talking about less than 5 seconds if you sitting on the optic fiber. If you are connected via poor wifi this may take closer to a minute. Whatever the case one can assume that nobody will complain given that the dimension just imported are as follows:

```{r, echo = FALSE}
tibble::tibble(type = c("HH", "HL", "CA"),
               rows = c(nrow(hh), nrow(hl), nrow(ca)),
               cols = c(ncol(hh), ncol(hl), ncol(ca))) |> 
  knitr::kable(caption = "Number of records and variables")
```

The above fast access is achieved by importing from parquet files that are hosted on a conventional https-server. The parquet data source is as of now **not** a mirror of the data residing at ICES, but a recent copy. ICES datacenter is currently exploring ways to serve a mirror of the DATRAS data via parquet files hosted on a cloud service.

The old faithful (icesDatras::getDATRAS) is wrapped within the above function, with the addition that variable type is set and -9 are turned to NA. Here we only dare to call one year of HL-data:

```{r}
system.time({
  hl_old <- dr_get(recordtype = "HL", years = 2026, from = "old", quiet = TRUE)
})
```

In store we now have:

```{r}
hl_old |> dplyr::count(Survey)
```

A little faster approach is also in experimental phase (icesDatras::get_datras_unaggregated_data) but it is not yet a mirror of the most up to date data (hence here year 2025 is used)

```{r}
system.time({
  hl_new <- dr_get("HL", years = 2025, from = "new", quiet = TRUE)
})
```

In store we now have:

```{r}
hl_new |> dplyr::count(Survey)
```


## Connecting

Although the DATRAS data can not be considered big data, one can use techniques developed for such datasets. So instead of importing the full dataset into R one can generate a connection to the url-hosted parquet files (these are not fully up-to-date data as of yet) using in-process DuckDB database.

### HH data

```{r}
hh <- dr_con("HH")
hh |> dplyr::glimpse()
```

### HL data

```{r}
hl <- dr_con("HL", trim = FALSE)
hl |> dplyr::glimpse()
```

### CA data

```{r}
ca <- dr_con("CA", trim = FALSE)
ca |> dplyr::glimpse()
```

### Data processing using a connection

For those familiar with using dplyr-verbs to process data most of those function as well as many base-R functions can be used to process the data. E.g. one can get all survey stations in 2024 and add to that the number of cod observed using the following script:

```{r}
system.time({
  q <-
    data <-
    # Process the data in DuckDB
    dr_con("HH") |> 
    dplyr::filter(Year == 2024,
                  Quarter %in% 1:4) |> 
    dplyr::select(.id, sur, lon = ShootLongitude, lat = ShootLatitude) |> 
    dplyr::left_join(dr_con("HL", trim = TRUE) |> 
                       dplyr::filter(latin == "Gadus morhua") |> 
                       dplyr::group_by(.id) |> 
                       dplyr::summarise(n_haul = sum(n_haul, na.rm = TRUE)),
                     by = dplyr::join_by(.id)) |> 
    dplyr::mutate(n_haul = coalesce(n_haul, 0))
})
q |> dplyr::glimpse()
```

Here all the code steps are automatically translated to SQL and passed to the in-process DuckDB. We can view the sql via:

```{r}
q |> dplyr::show_query()
```

The data is not yet in R, it only at the "collect"-step that this happens:

```{r}
system.time({
  d <- q |> dplyr::collect()
})
d |> dplyr::glimpse()
```

More importantly, only the variables .id (unique station id), sur, lon, lat, latin, and n_haul are ever passed over the web. In addition only certain chunks of the parquet files (read: rows), those chunks that "fall within the range" of the filtered values are passed over the web.

### What is it good for?

If the approach illustrated above is adopted as the primary DATRAS datasource, gone are all the tedious maintenance of up-to-date local copies because the instantaneous access makes that redundant. Only case where one may want a full local copy is in cases when internet access is unstable/slow or non-existent. But those days are all but behind us.

## Small print

This stuff is in development, thus bugs, snags and errors are expected. {obus} still has some experimental hangover functions that need to be pruned or removed.

Actually, as of now {obus} does not do very much, the initial focus has been on experimenting with fast access to DATRAS data. For all practical purposes one can totally live without it. Importing the full HH, HL and CA data into R can be achieved by (anticipated that ICES datacenter with maintain an official path):

```
hh <- arrow::read_parquet("https://heima.hafro.is/~einarhj/datras/raw/HH.parquet")
hl <- arrow::read_parquet("https://heima.hafro.is/~einarhj/datras/raw/HL.parquet")
ca <- arrow::read_parquet("https://heima.hafro.is/~einarhj/datras/raw/CA.parquet")
```

And a duckdb connection can be achieved by (here we have slightly augmented dataframes, using some simple obus wrapper functions):

```
hh <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HH.parquet")
hl <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL.parquet")
ca <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CA.parquet")
```

## Specs

```{r, echo = FALSE}
devtools::session_info()
```

