---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  error = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# obus

## Preamble

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The aim of {obus} is to provide users fast and efficient access to DATRAS tables with non-ambiguous variables.

That said, {obus} is a temporary experimental package used to explore various DATRAS data connections and wrapper functions to make life a little easier for ordinary user. Some of that may be taken up in a more official package. Or possibly not. So far {obus} does actually very little.

The package code resides on [GitHub](https://github.com/einarhjorleifsson/obus) and there is also a [package website](https://einarhjorleifsson.github.io/obus/).

For purists, one regrets to inform that the functionality of {obus} has quite some number of dependencies (56, see also [DESCRIPTION](https://raw.githubusercontent.com/einarhjorleifsson/obus/refs/heads/master/DESCRIPTION)). Some of them may possibly be trimmed, but never all.

Access to the parquet datafiles is though independent of {obus} and for that matter **any software platform** used. The current **temporary** path to the exchange files (with the new header lingo) used in {obus} is:

```
https://heima.hafro.is/~einarhj/datras/HH.parquet
https://heima.hafro.is/~einarhj/datras/HL.parquet
https://heima.hafro.is/~einarhj/datras/CA.parquet
```



## Installation

You can install {obus} using one of:

```{r, eval = FALSE}
remotes::install_github("einarhjorleifsson/obus")
pak::pak("einarhjorleifsson/obus")
```

```{r}
library(obus)
library(dplyr)
```

## Importing

The fastest way to import the full DATRAS data into R memory:

```{r}
system.time({
  hh <- dr_get("HH")
  hl <- dr_get("HL")
  ca <- dr_get("CA")
})
```

```{r, echo = FALSE}
tibble(type = c("HH", "HL", "CA"),
       rows = c(nrow(hh), nrow(hl), nrow(ca)),
       cols = c(ncol(hh), ncol(hl), ncol(ca))) |>
  knitr::kable(caption = "Number of records and variables")
```

All tables use standard column names and correct variable types. If your downstream code depends on the old DATRAS column names, revert with:

```{r}
hl |> dr_translate(from = "new", to = "old")
```

The XML API (`icesDatras::getDATRAS`) is also available via `source = "xml"` — useful for targeted queries by survey, year, and quarter without downloading the full table.

## Connecting

Rather than importing, you can open a lazy DuckDB connection to the same parquet files. Nothing is downloaded until you explicitly call `collect()`:

```{r}
system.time({
  hh <- dr_con("HH")
  hl <- dr_con("HL")
})
```

The connection is nearly instant — DuckDB reads only the file index, not the data. You then use ordinary `{dplyr}` verbs to filter, join, and compute; DuckDB translates them to SQL and executes against the remote file. Only the rows and columns you actually need travel over the wire:

```{r}
system.time(
  cod <- hl |>
    left_join(hh |> select(.id, DataType, HaulDuration, HaulValidity,
                           lon = ShootLongitude, lat = ShootLatitude)) |>
    left_join(dr_con("species"), by = join_by(aphia)) |>
    dr_add_length_cm() |>
    dr_add_n_and_cpue() |>
    filter(HaulValidity == "V", latin == "Gadus morhua",
           Survey %in% c("NS-IBTS", "SCOWCGFS"), Quarter == 1) |>
    select(.id, Survey, Year, lon, lat, latin, length_cm, n_haul, n_hour) |>
    collect()
)
glimpse(cod)
```

**The same pipeline works unchanged whether `hh` and `hl` are DuckDB connections or plain R data frames.** You can mix and match — switch to `dr_get()` when you need the full table in memory, use `dr_con()` when you want lazy filtering. The `dr_add_*()` functions handle both transparently.

For local parquet files, skip `dr_con()` and use `duckdbfs::open_dataset()` directly — the result is identical:

```{r, eval = FALSE}
hl <- duckdbfs::open_dataset("~/datras/HL.parquet")
```

See the [Parquet and DuckDB](articles/parquet_and_duckdb.html) article for a full explanation of how Parquet, DuckDB, and dbplyr fit together, including why the file format matters and a step-by-step walkthrough of lazy query building.

## Specs

```{r, echo = FALSE}
devtools::session_info()
```
