---
title: "DATRAS conventions, and what obus does about them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DATRAS conventions, and what obus does about them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Working notes on the parts of the DATRAS exchange format that shape how
`dr_HL_length()` and `dr_HL_summary()` are built. Each entry states what the
format says, what the archive actually contains, and what obus does — in that
order, because the order matters.

## What counts as authoritative here

**Only ICES's own published documents.** For this package that means the
primary sources kept in `~/R/Pakkar/imbus/DATRAS/external/` — principally
`DATRAS_Field_descriptions_and_example_file_December2025.xlsx` (the field
descriptions, with its `General Notes` sheet), the SISP survey manuals, and
the ICES workshop and working-group reports.

Derived commentary — including the authored `.qmd` documents one directory up,
and everything in `obus_retired` — is a useful pointer to *where to look*, and
is not evidence on its own. Several claims checked during this work turned out
to differ from the primary source. Where this article cites a figure, it was
measured against the full archive on 2026-08-31.

## The HL record key

The field descriptions define both `TotalNumber` and `SpeciesCategoryWeight`
as totals

> per category (unique combination of cruise haul, species, sex, devstage and
> subsampling category identifier)

so an HL record is keyed by **haul × species × sex × DevelopmentStage ×
SpeciesCategory**, and the length rows add length to that.

**obus:** both tables carry the full key. `HL_length` is exactly unique at
`.id × aphia × length_mm × length_cm × accuracy × LengthType × sex ×
DevelopmentStage × SpeciesValidity`; `HL_summary` at
`.id × aphia × SpeciesValidity`. Verified: zero duplicated groups in either.

`DevelopmentStage` is easy to overlook — it is `NA` for 99.8% of rows — but
5,476 `HL_length` groups genuinely split on it, and dropping it would sum a
berried or egg-bearing record together with an unstaged one.

## `SpeciesValidity` is a record type, not a quality grade

The field descriptions call it the

> Validity of species records, e.g. to specify if species has only been
> counted, measured, weighed, etc.

The codes describe *what data is present*, not how good it is: `1` is lengths,
numbers and weights; `4` is total number only; `5` presence/absence; `6` weight
only; `7` numbers and weights but no lengths. There is no ordering, so there is
no "most valid" code to collapse to.

**obus:** `SpeciesValidity` is part of `HL_summary`'s grain. 1,219 of its
groups (0.05%) carry more than one record type; the commonest pairs are
`{1,5}` and `{4,7}`. Filtering to one code makes `.id × aphia` exactly unique.

```{r}
dr_con("HL_summary") |>
  dplyr::filter(SpeciesValidity == "1")   # the usual starting point
```

**Can-Mar is the exception.** It puts real `LengthClass`/`NumberAtLength` data
on `"5"` rows, where every other survey uses `"5"` for a length-free presence
marker. Filtering to `"1"` there discards genuine length data. Treat Can-Mar
separately; ICES documents the situation as unresolved.

## `-9` means "no information", and it is already gone

The `General Notes` sheet instructs submitters:

> For the fields with no information but header, please submit -9

So the format has no concept of "blank" — an absent value is submitted as
`-9`. The code carries a per-field meaning where the field has a vocabulary:

| field | `-9` |
|---|---|
| `SpeciesSex` | "not sampled" (distinct from `U`, "Unidentified") |
| `DevelopmentStage` | "No information" |
| `DataType` | "Invalid hauls" — a real classification |

**The archive:** zero `-9` values anywhere in HL. opus resolves them when
building the archive, nulling any whose vocabulary label means absence and
keeping the three that assert something (`Tickler`, `AgePlusGroup`,
`DataType`). That line — keep a sentinel that states a fact, null one that
states nothing — is why an `NA` `sex` in obus can be read as "not sampled".

**obus:** never re-introduces a sentinel. It cannot tell which `NA`s were
`-9`, and a sentinel placed back into a numeric column would corrupt any sum
taken over it.

## `SubsamplingFactor` by `DataType`

The field is **mandatory**, and the rule is per `DataType`:

> If DataType is C, it should be reported as 1. If DataType is S, it is always
> >1. If DataType is R, the SubFactor is = 1 or >1 for different species
> depending on whether they were subsampled.

So `1` is a specific claim — *this species was not subsampled* — and it is not
the same statement as "no factor was supplied", which the format expects as
`-9`.

**obus:** a missing `SubsamplingFactor` propagates to `NA` for every
`DataType`, including `"R"`.

This is a deliberate departure from the DATRAS R package, which treats `"R"`
plus a missing factor as 1. That convention asserts a fact the submission does
not contain: it silently converts "the raising factor is unknown" into "the
catch was fully measured", understating `n_haul` whenever the true factor was
greater than 1. Following the format instead costs little — 24,298 length rows
(0.17%), worth 0.003% of the archive's total `n_haul`, and 98% of them
Can-Mar.

## `DataType` `P` versus `R`

`R` is one sub-sampling rate per species; `P` is size-stratified, with several
`SpeciesCategory` grades each sub-sampled separately. The haul total is
computed the same way for both — sum `NumberAtLength × SubsamplingFactor` over
the rows — so the distinction is descriptive rather than arithmetic.

It is also widely got wrong, and ICES has acknowledged why: IBTSWG 2024 §3.3
records a 2023 intersessional group flagging `DataType` value **`P` as missing
from the format documentation**. Submitters cannot use a code that is not
documented, so size-stratified hauls are submitted as `R`. Measured here,
34,531 `R` groups carry more than one `SpeciesCategory` each with its own
rate — the shape `P` describes — across many surveys and back to at least
2009.

**obus:** unaffected, because it uses the row-level formula for every
`DataType` and never assumes one category per species.

## `TotalNumber` and the raised length sum are two different routes

The field descriptions define `TotalNo = SUM(HLNoAtLngt)` — so the reported
total and the length-frequency sum are meant to be the same number.

**obus keeps both, in different tables, and names them apart:**

- `HL_summary$n_totalnumber` — the reported `TotalNumber`
- `sum(HL_length$n_haul)` — the raised length frequencies

Summing `HL_length` reproduces `Σ(NumberAtLength × SubsamplingFactor)` exactly
(1,925,444 groups, zero differing), so nothing is lost by not duplicating it
into `HL_summary`:

```{r}
dr_con("HL_summary") |>
  dplyr::left_join(
    dr_con("HL_length") |>
      dplyr::group_by(.id, aphia, SpeciesValidity) |>
      dplyr::summarise(n_length = sum(n_haul), .groups = "drop"),
    by = c(".id", "aphia", "SpeciesValidity")
  )
```

`HL_summary` still needs `TotalNumber` because 16% of its rows are species
with no length data at all, where no length-derived figure exists.

The two disagree for **3.37%** of comparable groups. That is a real property of
the submitted data, not a defect: it is concentrated in `DataType == "C"`
(where the reported rate and a raised sample need not match exactly) and in
Can-Mar's documented 2021–22 historical conversion.

## What the 3.37% residual actually is

"Residual" here means one specific thing: the haul x species groups where the
two independent routes to a catch total disagree.

- `HL_summary$n_totalnumber` — the total DATRAS *reported* (`TotalNumber`)
- `sum(HL_length$n_haul)` — the same total *reconstructed* by raising the
  measured length frequencies

The format says these are the same number (`TotalNo = SUM(HLNoAtLngt)`).
Measured on the full archive, **64,539 of 1,912,556 comparable groups (3.37%)**
differ by more than 0.51. Groups where either side is unavailable — a species
never length-measured, a missing `SubsamplingFactor`, a non-positive
`HaulDuration` — are not "residual"; they are simply not comparable and are
excluded.

The residual is not noise, and it is not one thing. Classified:

| what it is | groups | share |
|---|---|---|
| Can-Mar: raising factor lost, length under-raised | 26,056 | 40.4% |
| `DataType == "C"`: small difference, intrinsic | 19,501 | 30.2% |
| Non-integer `SubsamplingFactor` rounding | 7,048 | 10.9% |
| Small difference, integer factor — **open** | 4,367 | 6.8% |
| Can-Mar: other | 4,164 | 6.5% |
| `DataType == "C"`: large difference | 2,311 | 3.6% |
| Unexplained | 1,091 | 1.7% |

**Can-Mar (46.8%) — a known historical conversion.** A bulk conversion of
Can-Mar's historical data set `SubsamplingFactor` to 1 on records whose catch
had actually been subsampled. `TotalNumber` kept its correct raised value; the
length rows lost the factor needed to raise them. So the length sum
under-counts, and it under-counts in one direction only: **29,668 of Can-Mar's
30,220 mismatches (98%) have the reported total higher than the length sum**.
The ratio between them is the factor that went missing — median 1.55, upper
quartile 2.5. It spans both 1970-1996 (13,358) and 1997 onward (12,698). Trust
`n_totalnumber` here, not the length sum.

**`DataType == "C"` (33.8%, mostly small) — not an error at all.** For a "C"
submission `TotalNumber` is a directly reported catch rate, and the length
table is a separately measured and raised sample. They are two independent
observations of a similar quantity, not one computed from the other, so they
need not agree exactly. The signature fits: 69% of C-type differences are
under 1% relative, 89% under 5%, and hand-checked examples sit a couple of
individuals apart *in either direction*. Nothing to fix, and no "true" value
to converge on.

**Rounding (10.9%).** A non-integer `SubsamplingFactor` (2.91, say) applied to
integer counts gives a non-integer raised sum, against an integer
`TotalNumber`. Sub-unit differences, mechanically expected.

**Still open (6.8%).** Small differences where `SubsamplingFactor` is an
integer — and in 4,330 of 4,367 cases **exactly 1**, where the arithmetic is
pure integer multiplication and rounding cannot occur. Spread across BITS
(1,443), SCOWCGFS (865), BTS (536), NS-IBTS (531), SCOROC (511) and EVHOE
(466). An earlier explanation of this bucket as rounding was checked and did
not hold. One observation worth chasing: a SCOROC haul showed the same small
offset across five different species at once, which points at a haul-level
rather than species-level cause.

**Unexplained (1.7%).** 1,091 groups with a large difference, integer factor,
neither Can-Mar nor "C" — mostly BTS (580) and BITS (264).

So roughly **88% has a confirmed mechanism, 10% a partial one, and under 2%
none.** The practical consequence for a user: the disagreement is
concentrated, not spread evenly, so check which bucket your data falls in
before treating either number as wrong. Can-Mar and `DataType == "C"` between
them account for four-fifths of it.

## `HaulDuration`

An hourly rate needs a duration. 217 hauls report `HaulDuration == 0` and 2
report a negative value.

**obus:** where the duration is not positive, any figure that needs it is `NA`.
The direction depends on `DataType`, because `"C"` reports a rate and derives
the haul total, while everything else reports the haul total and derives the
rate:

| | non-`C` | `C` |
|---|---|---|
| hourly rate | derived — `NA` | reported — kept |
| per-haul total | reported — kept | derived — `NA` |

Without this the arithmetic produces `Inf` on one side and a plausible-looking
`0` on the other.

## An engineering note: R and SQL disagree about `NA`

Not a DATRAS matter, but the single largest source of defects found while
building these tables. obus runs the same functions eagerly (data frames, in
the tests) and lazily (DuckDB, in the build), and the two backends differ in
three separate ways:

1. **Joining.** SQL treats `NULL = NULL` as unknown, so a lazy join drops
   every row with an `NA` key — `sex` is `NA` in 54% of HL. Fixed with
   `na_matches = "na"`, which renders as `IS NOT DISTINCT FROM`.
2. **Aggregating.** `sum(x, na.rm = TRUE)` over an all-`NA` group is `0` in R
   and `NULL` in SQL. Every aggregation here carries an explicit non-`NA`
   counter so both return `NA`.
3. **Self-reference in `summarise()`.** dplyr evaluates expressions in order,
   so `n_ok = sum(!is.na(n_haul))` written after `n_haul = sum(n_haul)` sees
   the new scalar; SQL resolves every aggregate against the source column.
   Renaming before aggregating removes the ambiguity.

Each of these produced a real, silent, wrong number. Keeping both an eager
test suite and a lazy build is what catches them — neither does alone.
