---
title: "CPUE products: `dr_cpue_by_length` and `dr_cpue_by_haul`"
format:
  html:
    footnotes-hover: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Two DATRAS exchange tables — the haul header (HH) and the length table (HL) —
contain everything needed to derive the CPUE products that the ICES Datacenter
publishes via the `icesDatras` package. This article documents the obus
functions that compute those products from first principles, and explains where
and why results may diverge from the ICES pre-computed versions.

| obus function | Level | icesDatras equivalent |
|---|---|---|
| `dr_cpue_by_length()` | CPUE per haul × species × length class | `getCPUELength()` |
| `dr_cpue_by_haul()` | CPUE + weight per haul × species | `getCatchWgt()` |

Both functions accept HH and HL tables with new-style column names — the
format returned by `dr_get()` (default `from = "parquet"`) or `dr_con()` —
and apply the same filters and DataType-aware arithmetic. All of the relevant
calculation choices are made in the open: which hauls to accept (`haulval`),
which species records to include (`specval`), whether to add explicit
zero-catch rows (`zerofill`).

---

## Setup

```{r}
library(dplyr)
library(obus)
hh <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HH.parquet") |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1)
hl <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/HL.parquet") |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1)
```

---

## `dr_cpue_by_length`

### Basic use

```{r}
cpue_len <- dr_cpue_by_length(hh, hl)
cpue_len |> collect() |> head(5)
```

One row per haul × species × length class. The key output column is `n_hour`:
estimated numbers caught per hour of hauling at each length class.

### CPUE formula

The `n_hour` value depends on the haul-level `DataType` field (joined from HH):

| `DataType` | Meaning | `n_haul` | `n_hour` |
|---|---|---|---|
| `R` / `S` / `P` | Raw, subsampled, or pseudocategory | `NumberAtLength × SubsamplingFactor` | `n_haul / HaulDuration × 60` |
| `C` | Already raised to 1 hr | `NumberAtLength × SubsamplingFactor × HaulDuration / 60` | `NumberAtLength × SubsamplingFactor` |

`NumberAtLength` is the count in the measured sub-sample, `SubsamplingFactor`
raises it to the full catch, and `HaulDuration` (minutes) standardises to one
hour. All length classes within a haul × species group are then summed across
`SpeciesSex` and `SpeciesCategory`, because the ICES CPUE per-length product
does not disaggregate by sex.

### Zero-fill {#sec-zerofill}

The ICES CPUE product inserts explicit zero rows for species that were observed
in at least one haul of a survey × year × quarter but were absent from a
particular haul. This makes it possible to compute mean CPUE across hauls
without losing the zero-catch information.

```{r}
cpue_len_zf <- dr_cpue_by_length(hh, hl, zerofill = TRUE)
```

```{r}
#| echo: false
non_zero <- dr_cpue_by_length(hh, hl) |> collect() |> nrow()
with_zero <- nrow(cpue_len_zf)
```

For NS-IBTS 2023 Q1, zero-fill adds `r format(with_zero - non_zero, big.mark = ",")`
rows (`r format(with_zero, big.mark = ",")` total vs
`r format(non_zero, big.mark = ",")` non-zero). Zero rows carry `length_mm = NA`
and `n_hour = 0`.

::: {.callout-note}
When `zerofill = TRUE` the result is collected into memory and returned as a
data frame rather than a lazy DuckDB table. For multi-year or multi-survey
queries, collect before calling `dr_cpue_by_length` or filter HH/HL first.
:::

### Diagnostic mode

`diag = TRUE` skips the SpeciesSex/SpeciesCategory aggregation and returns
one row per raw HL record, exposing the intermediate columns used for QC:

```{r}
dr_cpue_by_length(hh, hl, diag = TRUE) |>
  filter(ValidAphiaID == 126417L) |>   # herring
  select(.id, ValidAphiaID, SpeciesSex, SpeciesCategory,
         NumberAtLength, SubsamplingFactor, DataType, HaulDuration,
         n_haul, n_hour) |>
  collect() |>
  head(5)
```

Use this to detect hauls where the same length class appears under both
`SpeciesSex = NA` (total) and named sex codes — these create inflated
`n_hour` when summed, as discussed in @sec-cpuel-diff below.

---

## Comparison with `getCPUELength` {#sec-cpuel-diff}

### SYQ-level vs haul-level coverage gaps

Not all obus-only or ICES-only records arise from the same cause. A useful
first cut is to separate:

- **Case A** — an entire Survey × Year × Quarter (SYQ) is present in one source
  but wholly absent from the other. A systematic gap: a year-quarter of a
  survey that was never submitted to or published by the Datacenter.
- **Case B** — the SYQ appears in both sources but individual hauls differ.
  A partial gap from data revisions, HaulVal filtering differences, or version
  drift at the haul level.

```{r}
#| eval: false
# ── Load sources ─────────────────────────────────────────────────────────────
# Note: CPUEL parquet uses old-style names; HH/HL parquets use new-style names
cpue_ices <- duckdbfs::open_dataset(
  "https://heima.hafro.is/~einarhj/datras/old/CPUEL.parquet"
) |>
  mutate(.id2 = paste(Survey, Year, Quarter, Ship, Gear, HaulNo, sep = ":"))

hh_all <- duckdbfs::open_dataset(
  "https://heima.hafro.is/~einarhj/datras/HH.parquet"
)
sur_cpue <- cpue_ices |> distinct(Survey) |> pull(Survey)
hh_all   <- hh_all   |> filter(Survey %in% sur_cpue)

# ── Step 1: SYQ presence ──────────────────────────────────────────────────────
sqy_ices <- cpue_ices |> distinct(Survey, Year, Quarter) |> collect()
sqy_hh   <- hh_all   |> filter(HaulValidity == "V") |>
              distinct(Survey, Year, Quarter) |> collect()

sqy_status <- full_join(
  sqy_hh   |> mutate(in_hh   = TRUE),
  sqy_ices |> mutate(in_ices = TRUE),
  by = c("Survey", "Year", "Quarter")
) |>
  mutate(across(c(in_hh, in_ices), \(x) coalesce(x, FALSE)))

sqy_status |> count(in_hh, in_ices)
#> # A tibble: 3 × 3
#>   in_hh in_ices     n
#>   <lgl> <lgl>   <int>
#> 1 FALSE TRUE        9   # SYQs in CPUEL only
#> 2 TRUE  FALSE     339   # SYQs in HH only (most are surveys absent from CPUEL entirely)
#> 3 TRUE  TRUE      339   # SYQs shared by both

# Of the 339 HH-only SYQs, only 20 belong to surveys that DO appear in CPUEL;
# the rest (319) are surveys entirely absent from the CPUEL product (DYFS,
# Can-Mar, BTS, SNS, etc.) and are therefore outside the comparison scope.

sqy_shared <- sqy_status |> filter(in_hh, in_ices) |> select(Survey, Year, Quarter)

# ── Step 2: Record-level full join (as in steps 1–5 in the section below) ────
# ... (run the full comparison first to obtain `comparison`) ...

# ── Step 3: Tag each unmatched row by case ────────────────────────────────────
obus_only <- comparison |>
  filter(!is.na(n_hour), is.na(n_hour_ices)) |>
  left_join(sqy_shared |> mutate(shared = TRUE), by = c("Survey", "Year", "Quarter"))

ices_only <- comparison |>
  filter(is.na(n_hour), !is.na(n_hour_ices)) |>
  left_join(sqy_shared |> mutate(shared = TRUE), by = c("Survey", "Year", "Quarter"))

obus_only |>
  mutate(case = if_else(coalesce(shared, FALSE),
                        "B: SYQ shared — haul missing from CPUEL",
                        "A: entire SYQ absent from CPUEL")) |>
  count(case)
#> # A tibble: 2 × 2
#>   case                                          n
#>   <chr>                                     <int>
#> 1 A: entire SYQ absent from CPUEL          294371
#> 2 B: SYQ shared — haul missing from CPUEL  331316

ices_only |>
  mutate(case = if_else(coalesce(shared, FALSE),
                        "B: SYQ shared — haul missing from HH",
                        "A: entire SYQ absent from HH")) |>
  count(case)
#> # A tibble: 2 × 2
#>   case                                    n
#>   <chr>                               <int>
#> 1 A: entire SYQ absent from HH       77233
#> 2 B: SYQ shared — haul missing from HH 107474
```

The 294k Case A obus-only records are driven almost entirely by IE-IGFS (197k):
the CPUEL product was not updated for this survey after 2017, so all hauls from
2018 onward are absent. SCOROC (39k), SCOWCGFS (34k), and EVHOE (23k) make up
the rest.

The 331k Case B obus-only records reflect haul-level discrepancies within SYQs
that appear in both sources: NS-IBTS (137k), BITS (56k), SWC-IBTS (55k). Note
that for SWC-IBTS the CPUEL product contains *more* hauls than the current HH
parquet for years 1986–2010, meaning some historical hauls were removed from
the exchange tables after the CPUEL snapshot was generated. The CPUEL product
and the exchange tables are not always ordered by recency.

### Full record-level comparison

`icesDatras::getCPUELength()` returns the pre-computed CPUE per-length product
from the ICES Datacenter. The API call takes approximately 30 seconds per
survey × year × quarter combination and requires internet access:

```{r}
#| eval: false
cpue_ices <- icesDatras::getCPUELength("NS-IBTS", 2023, 1)
```

The result carries station metadata alongside the CPUE values. The column
mapping to `dr_cpue_by_length` output is:

| `getCPUELength` | `dr_cpue_by_length` | Notes |
|---|---|---|
| `Survey`, `Year`, `Quarter` | same | |
| `Ship`, `Gear`, `HaulNo` | part of `.id2` | `.id2 = Survey:Year:Quarter:Platform:Gear:HaulNumber` (new-style field names) |
| `AphiaID` | `ValidAphiaID` | both are WoRMS AphiaIDs |
| `LngtClas` | `length_mm` | both in millimetres |
| `CPUE_number_per_hour` | `n_hour` | |
| `Sex` | *(aggregated away)* | always `NA` in CPUEL output |
| `HaulDur`, `ShootLat`, `ShootLon`, `Depth`, `Area`, `SubArea`, `DayNight` | not in obus output | join from HH if needed |

A full-dataset comparison across all surveys in the CPUEL parquet can be run
as follows. The CPUEL parquet is a recent snapshot of the ICES Datacenter
product; the HH and HL parquets are the up-to-date exchange tables.

```{r}
#| eval: false

# ── 1. ICES CPUEL product ───────────────────────────────────────────────────
# Note: CPUEL parquet uses old-style names (Ship, HaulNo); build .id2 to match
cpue_ices <- duckdbfs::open_dataset(
  "https://heima.hafro.is/~einarhj/datras/old/CPUEL.parquet"
) |>
  mutate(.id2 = paste(Survey, Year, Quarter, Ship, Gear, HaulNo, sep = ":"))

# ── 2. HH and HL restricted to surveys present in CPUEL ────────────────────
sur_cpue <- cpue_ices |> distinct(Survey) |> pull(Survey)

# New-style parquets (no /old/ in path)
hh_all <- duckdbfs::open_dataset(
  "https://heima.hafro.is/~einarhj/datras/HH.parquet"
) |> filter(Survey %in% sur_cpue)

hl_all <- duckdbfs::open_dataset(
  "https://heima.hafro.is/~einarhj/datras/HL.parquet"
) |> filter(Survey %in% sur_cpue)

# ── 3. Compute obus CPUE (all surveys, non-zero only) ──────────────────────
cpue_obus <- dr_cpue_by_length(hh_all, hl_all) |>
  rename(AphiaID = ValidAphiaID)

# ── 4. Prepare ICES product for join (non-zero records only) ───────────────
# CPUEL uses old-style Ship/HaulNo in .id2; obus now uses Platform/HaulNumber.
# These should match when Ship == Platform and HaulNo == HaulNumber (usually true).
cpue_ices_nz <- cpue_ices |>
  filter(CPUE_number_per_hour > 0) |>
  select(.id2, Survey, Year, Quarter,
         AphiaID, length_mm = LngtClas, n_hour_ices = CPUE_number_per_hour)

# ── 5. Full outer join on haul × species × length ──────────────────────────
comparison <- cpue_obus |>
  full_join(cpue_ices_nz,
            by = c(".id2", "Survey", "Year", "Quarter", "AphiaID", "length_mm")) |>
  collect()

# ── 6. Coverage summary ─────────────────────────────────────────────────────
comparison |>
  summarise(
    matched   = sum(!is.na(n_hour) & !is.na(n_hour_ices)),
    ices_only = sum( is.na(n_hour) & !is.na(n_hour_ices)),
    obus_only = sum(!is.na(n_hour) &  is.na(n_hour_ices))
  )
#> # A tibble: 1 × 3
#>    matched ices_only obus_only
#>      <int>     <int>     <int>
#> 1  6941482    184659    625378

# ── 7. Distribution of relative differences for matched records ─────────────
matched <- comparison |>
  filter(!is.na(n_hour), !is.na(n_hour_ices)) |>
  mutate(rel_diff = (n_hour - n_hour_ices) / pmax(n_hour_ices, 1e-9))

quantile(matched$rel_diff, c(0, .01, .25, .50, .75, .99, 1))
#>           0%           1%          25%          50%          75%
#> -9.961e-01   -4.135e-01    0.000e+00    0.000e+00    0.000e+00
#>          99%         100%
#>  4.475e+00    2.409e+05

mean(abs(matched$rel_diff) > 1)   # fraction differing by > 100 %
#> [1] 0.07665
```

The exact matches dominate (median relative difference = 0). The sources of
disagreement are:

**Snapshot lag.** The CPUEL product is generated from a specific HH/HL snapshot
and is not continuously updated. Of the 625k obus-only records, about half
(294k, Case A) come from entire Survey × Year × Quarter combinations absent
from the CPUEL product — mainly IE-IGFS years 2018 onward. The remaining 331k
(Case B) are haul-level discrepancies within SYQs present in both sources,
spread across NS-IBTS, BITS, and SWC-IBTS. For SWC-IBTS the CPUEL product
actually contains *more* hauls than the current HH parquet for years 1986–2010,
showing that the exchange tables can be revised in either direction relative to
the published product. The full breakdown is shown in @sec-cpuel-diff.

**Duplicate SpeciesSex records.** Some hauls in BITS and NS-IBTS submit the
same species both as `SpeciesSex = NA` (total) and as named sex codes
(`U` / `M` / `F`). `dr_cpue_by_length` sums all SpeciesSex rows for each
length class, which double-counts hauls where the total-row and the sex-coded
rows both carry non-zero `NumberAtLength`. The ICES product appears to use only
the `SpeciesSex = NA` rows in these cases. Restricting to `SpeciesSex = NA` in
`diag` mode before re-aggregating reduces the large-difference rate from 7.5 %
to 3.6 % but also reduces coverage.

**Version mismatch in older data.** For hauls prior to ~1990, some length
classes appear in the CPUEL product at different `LngtClas` values than those
in the current HL parquet, indicating the historical data has been revised
since the product was generated.

The join key difference is also worth noting: `getCPUELength` lacks `Country`
and `StNo`, so its haul identifier (`Survey:Year:Quarter:Ship:Gear:HaulNo`)
is not unique when multiple countries record hauls with the same ship and haul
number in the same quarter. `dr_cpue_by_length` uses the full 8-field `.id`
internally and provides `.id2` as a convenience column for comparison.

---

## `dr_cpue_by_haul`

### Basic use

```{r}
cpue_haul <- dr_cpue_by_haul(hh, hl)
cpue_haul |> collect() |> head(5)
```

One row per haul × species. Counts (`n_haul`, `n_hour`) are derived from
`TotalNumber` — the total estimated catch in the haul already accounting for
subsampling — and weights (`w_haul`, `w_hour`) from `SpeciesCategoryWeight`
(grams). The function deduplicates HL rows to one record per haul × species ×
sex × SpeciesCategory group before scaling, so that length-level repetition of
`TotalNumber` and `SpeciesCategoryWeight` does not inflate the totals.

### DataType scaling

| `DataType` | `n_haul` | `n_hour` | `w_haul` | `w_hour` |
|---|---|---|---|---|
| `R` / `S` / `P` | `TotalNumber` | `TotalNumber / HaulDuration × 60` | `SpeciesCategoryWeight` | `SpeciesCategoryWeight / HaulDuration × 60` |
| `C` | `TotalNumber × HaulDuration / 60` | `TotalNumber` | `SpeciesCategoryWeight × HaulDuration / 60` | `SpeciesCategoryWeight` |

### Relationship to `dr_cpue_by_length`

`dr_cpue_by_haul` and `dr_cpue_by_length` use different source fields for
counts: `TotalNumber` vs `sum(NumberAtLength × SubsamplingFactor)`. For
well-formed data these are equal, but they can diverge slightly due to rounding
in submitted records. The two functions therefore make independent calculations
rather than one wrapping the other.

`n_hour` from `dr_cpue_by_haul` should equal the sum of `n_hour` across all
length classes in `dr_cpue_by_length` for the same haul × species:

```{r}
by_len_summed <- dr_cpue_by_length(hh, hl) |>
  group_by(.id, ValidAphiaID) |>
  summarise(n_hour_len = sum(n_hour, na.rm = TRUE), .groups = "drop") |>
  collect()

by_haul <- dr_cpue_by_haul(hh, hl) |>
  select(.id, ValidAphiaID, n_hour_haul = n_hour) |>
  collect()

inner_join(by_len_summed, by_haul, by = c(".id", "ValidAphiaID")) |>
  summarise(max_abs_diff = max(abs(n_hour_len - n_hour_haul), na.rm = TRUE),
            n_exact      = sum(abs(n_hour_len - n_hour_haul) < 1e-9, na.rm = TRUE),
            n_total      = n())
```

The small floating-point differences (< 10^-9) reflect independent arithmetic
paths; they are not meaningful.

### Diagnostic mode

`diag = TRUE` returns the deduplicated, scaled table before aggregation across
`SpeciesSex` and `SpeciesCategory`:

```{r}
dr_cpue_by_haul(hh, hl, diag = TRUE) |>
  filter(ValidAphiaID == 126417L) |>
  select(.id, ValidAphiaID, SpeciesSex, SpeciesCategory,
         TotalNumber, SpeciesCategoryWeight, DataType, HaulDuration,
         n_haul, n_hour, w_haul, w_hour) |>
  collect() |>
  head(5)
```

---

## Comparison with `getCatchWgt` {#sec-cw-diff}

`icesDatras::getCatchWgt()` returns catch weight per haul for a specified list
of AphiaIDs. Unlike `dr_cpue_by_haul`, it requires species to be named
explicitly and makes one API call per survey:

```{r}
#| eval: false
# getCatchWgt requires species — returns NA in CatchWgt for absent hauls
cw <- icesDatras::getCatchWgt("NS-IBTS", 2023, 1, aphia = c(126417L, 126436L))
```

The result is a wide table that contains all HH columns plus `CatchWgt`
(grams, `NA` = species absent from haul) and `Valid_Aphia`. Column mapping:

| `getCatchWgt` | `dr_cpue_by_haul` | Notes |
|---|---|---|
| `Valid_Aphia` | `ValidAphiaID` | both are WoRMS AphiaIDs |
| `CatchWgt` | `w_haul` | both in grams; `NA` vs `0` for absent hauls |
| *(absent)* | `n_haul`, `n_hour`, `w_hour` | `getCatchWgt` provides no count or hourly weight |
| `HaulDur`, `ShootLat`, … | not in obus output | join from HH if needed |

**`NA` vs `0` for absent hauls.** `getCatchWgt` uses `NA` to signal that a
species was not caught in a given haul. `dr_cpue_by_haul` by default omits
absent-species rows entirely; setting `zerofill = TRUE` adds them with
`w_haul = 0`. The semantics differ: `getCatchWgt` returns every haul in the
survey regardless of presence, while `dr_cpue_by_haul(zerofill = TRUE)` inserts
zeros only for species observed somewhere in that survey × year × quarter.

**Weight agreement.** For hauls where both sources report a catch, `w_haul` and
`CatchWgt` are typically identical. A notable exception arises when a species is
submitted with both `Sex = NA` (total) and named sex codes (`M`, `F`, `U`) all
carrying the same `CatCatchWgt` value. In that case `dr_cpue_by_haul`
incorrectly sums the weight once per sex group, inflating the total:

```{r}
#| echo: false
# Reproduce the example from NS-IBTS 2023 Q1 ship 58G2 haul 27, herring
hl |>
  filter(Platform == "58G2", Gear == "GOV", HaulNumber == 27L,
         ValidAphiaID == 126417L, SpeciesValidity == "1") |>
  distinct(Platform, HaulNumber, ValidAphiaID, SpeciesSex, SpeciesCategory,
           TotalNumber, SpeciesCategoryWeight) |>
  collect()
```

Here `SpeciesCategoryWeight = 3019 g` is the total catch weight for herring,
repeated across three sex rows (`M`, `F`, `NA`). `dr_cpue_by_haul` sums to
`9057 g`; `getCatchWgt` returns the correct `3019 g`. When this pattern is
present, use the `diag = TRUE` output to identify the problematic hauls and
apply a sex-deduplication step before using the weight columns:

```{r}
#| eval: false
# Correct weight by taking one SpeciesCategoryWeight per (haul, species, SpeciesCategory)
dr_cpue_by_haul(hh, hl, diag = TRUE) |>
  distinct(.id, .id2, Survey, Year, Quarter, ValidAphiaID,
           SpeciesCategory, DataType, HaulDuration, SpeciesCategoryWeight) |>
  mutate(
    w_haul = if_else(DataType == "C",
                     SpeciesCategoryWeight * HaulDuration / 60,
                     SpeciesCategoryWeight),
    w_hour = if_else(DataType == "C",
                     SpeciesCategoryWeight,
                     SpeciesCategoryWeight / HaulDuration * 60)
  ) |>
  group_by(.id, .id2, Survey, Year, Quarter, ValidAphiaID) |>
  summarise(w_haul = sum(w_haul, na.rm = TRUE),
            w_hour = sum(w_hour, na.rm = TRUE),
            .groups = "drop")
```

---

## Summary

| | `dr_cpue_by_length` | `dr_cpue_by_haul` | `getCPUELength` | `getCatchWgt` |
|---|---|---|---|---|
| **Source** | HH + HL (parquet) | HH + HL (parquet) | ICES API product | ICES API product |
| **Input** | arbitrary HH/HL | arbitrary HH/HL | survey/year/quarter | survey/year/quarter + AphiaID |
| **Species** | all at once | all at once | all at once | specified only |
| **Count source** | `NumberAtLength` | `TotalNumber` | ICES calculation | not returned |
| **Weight** | — | `SpeciesCategoryWeight` (g) | — | `CatchWgt` (g) |
| **Length breakdown** | yes | no | yes | no |
| **Zero-fill** | optional | optional | always | always (as `NA`) |
| **Speed** | fast (parquet) | fast (parquet) | ~30 s / call | ~5 s / call |
| **Currency** | matches raw data | matches raw data | lags raw data | lags raw data |

The principal advantage of the obus functions is currency: they work directly
from the HH and HL exchange data, so any corrections or new hauls submitted
since the last CPUE product generation are immediately reflected. The ICES API
products are authoritative for historical comparisons but cannot be assumed to
match the current exchange tables.
