---
title: "Issues"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(obus)
```

## Preamble


## 2025-01-24 API HH variable names

Use this medium until API set up on github

```{r, eval = FALSE}
icesDatras::get_datras_unaggregated_data("HH", "NS-IBTS", 2025, 3) |> 
  colnames() |> dput()
```

```
c("RecordHeader", "Survey", "Quarter", "Country", "Platform", 
"Gear", "SweepLength", "GearExceptions", "DoorType", "StationName", 
"HaulNumber", "Year", "Month", "Day", "StartTime", "DepthStratum", 
"HaulDuration", "DayNight", "ShootLatitude", "ShootLongitude", 
"HaulLatitude", "HaulLongitude", "StatisticalRectangle", "BottomDepth", 
"HaulValidity", "HydrographicStationID", "StandardSpeciesCode", 
"BycatchSpeciesCode", "DataType", "NetOpening", "Rigging", "Tickler", 
"Distance", "WarpLength", "WarpDiameter", "WarpDensity", "DoorSurface", 
"DoorWeight", "DoorSpread", "WingSpread", "Buoyancy", "KiteArea", 
"GroundRopeWeight", "TowDirection", "SpeedGround", "SpeedWater", 
"SurfaceCurrentDirection", "SurfaceCurrentSpeed", "BottomCurrentDirection", 
"BottomCurrentSpeed", "WindDirection", "WindSpeed", "SwellDirection", 
"SwellHeight", "SurfaceTemperature", "BottomTemperature", "SurfaceSalinity", 
"BottomSalinity", "ThermoCline", "ThermoClineDepth", "CodendMesh", 
"SecchiDepth", "Turbidity", "TidePhase", "TideSpeed", "PelagicSamplingType", 
"MinTrawlDepth", "MaxTrawlDepth", "SurveyIndexArea", "EDOM", 
"ReasonHaulDisruption", "DateofCalculation")
```

* Variable name **EDOM** is specified as **EDMO** in [this document](https://www.ices.dk/data/Documents/DATRAS/DATRAS_NewHeaders_Lookup_Dec2024.xlsx).
* [Here](https://datras.ices.dk/Data_products/ReportingFormat.aspx) it is **EDOM**

Will assume that **EDOM** is correct.

Action: None needed, except possibly in the 2024-12 excel file.

## 2025-01-24 API CA variable names

Use this medium until API set up on github

```{r, eval = FALSE}
icesDatras::get_datras_unaggregated_data("CA", "NS-IBTS", 2025, 3) |> 
  colnames() |> dput()
```

```
c("RecordHeader", "Survey", "Quarter", "Country", "Platform", 
"Gear", "SweepLength", "GearExceptions", "DoorType", "StationName", 
"HaulNumber", "Year", "SpeciesCodeType", "SpeciesCode", "AreaType", 
"AreaCode", "LengthCode", "LengthClass", "IndividualSex", "IndividualMaturity", 
"AgePlusGroup", "IndividualAge", "CANoAtLngt", "IndividualWeight", 
"FishID", "GeneticSamplingFlag", "StomachSamplingFlag", "AgeSource", 
"AgePreparationMethod", "OtolithGrading", "ParasiteSamplingFlag", 
"MaturityScale", "LiverWeight", "AphiaID", "Species", "DateofCalculation")
```

Variable name CANoAtLngt is the old variable name: should [according to this](https://www.ices.dk/data/Documents/DATRAS/DATRAS_NewHeaders_Lookup_Dec2024.xlsx) be NumberAtLength.

Action: Change variable name **CANoAtLngt** to **NumberAtLength** at server end.

## 2025-01-25 API HH variable names mixups

When running this:

```{r, eval = FALSE}
icesDatras::get_datras_unaggregated_data("HH", "NS-IBTS", 2025, 3) |> 
  dplyr::glimpse()
```

I get this:

```
Rows: 358
Columns: 72
$ RecordHeader            <chr> "﻿HH", "HH", "HH", "HH", "HH", "HH", "HH", "HH", "HH", "HH", "HH", "HH"…
$ Survey                  <chr> "NS-IBTS", "NS-IBTS", "NS-IBTS", "NS-IBTS", "NS-IBTS", "NS-IBTS", "NS-I…
$ Quarter                 <int> 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3,…
$ Country                 <chr> "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE",…
$ Platform                <chr> "77SE", "77SE", "77SE", "77SE", "77SE", "77SE", "77SE", "77SE", "77SE",…
$ Gear                    <chr> "GOV", "GOV", "GOV", "GOV", "GOV", "GOV", "GOV", "GOV", "GOV", "GOV", "…
$ SweepLength             <int> 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60, 60,…
$ GearExceptions          <chr> "S", "S", "S", "S", "S", "S", "S", "S", "S", "S", "S", "S", "S", "S", "…
$ DoorType                <chr> "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "…
$ StationName             <chr> "140", "141", "142", "143", "144", "145", "146", "147", "148", "149", "…
$ HaulNumber              <int> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, …
$ Year                    <int> 2025, 2025, 2025, 2025, 2025, 2025, 2025, 2025, 2025, 2025, 2025, 2025,…
$ Month                   <int> 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 8,…
$ Day                     <int> 18, 18, 18, 19, 19, 19, 19, 20, 20, 20, 20, 20, 21, 21, 21, 21, 22, 22,…
$ StartTime               <chr> "0607", "0945", "1239", "0556", "0856", "1214", "1414", "0518", "0756",…
$ DepthStratum            <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ HaulDuration            <int> 15, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 30, 15, 30, 30, 30,…
$ DayNight                <chr> "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", "…
$ ShootLatitude           <dbl> 58.5348, 58.8454, 58.8040, 58.0472, 57.8045, 57.6475, 57.4658, 57.1847,…
$ ShootLongitude          <dbl> 10.9335, 10.4956, 10.2735, 11.1504, 11.2550, 10.7695, 10.8735, 11.4434,…
$ HaulLatitude            <dbl> 58.5497, 58.8147, 58.7866, 58.0171, 57.7746, 57.6215, 57.4433, 57.1547,…
$ HaulLongitude           <dbl> 10.9390, 10.4935, 10.3223, 11.1636, 11.2622, 10.7397, 10.8351, 11.4509,…
$ StatisticalRectangle    <chr> "46G0", "46G0", "46G0", "45G1", "44G1", "44G0", "43G0", "43G1", "42G1",…
$ BottomDepth             <int> 73, 134, 160, 108, 59, 26, 41, 69, 57, 45, 34, 40, 22, 19, 32, 33, 24, …
$ HaulValidity            <chr> "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "V", "…
$ HydrographicStationID   <chr> "623", "624", "625", "628", "629", "630", "631", "632", "633", "634", "…
$ StandardSpeciesCode     <chr> "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "…
$ BycatchSpeciesCode      <chr> "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "…
$ DataType                <chr> "C", "C", "C", "C", "C", "C", "C", "C", "C", "C", "C", "C", "C", "C", "…
$ NetOpening              <dbl> 5.1, 5.0, 4.9, 4.6, 4.8, 4.9, 5.3, 4.6, 4.8, 4.9, 4.8, 4.9, 4.5, 4.4, 4…
$ Rigging                 <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Tickler                 <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Distance                <dbl> 1685, 3426, 3426, 3445, 3408, 3389, 3389, 3389, 3408, 3426, 3426, 3389,…
$ WarpLength              <int> 230, 420, 500, 340, 220, 150, 160, 240, 210, 180, 160, 170, 150, 150, 1…
$ WarpDiameter            <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ WarpDensity             <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ DoorSurface             <dbl> 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6.5, 6…
$ DoorWeight              <int> 1050, 1050, 1050, 1050, 1050, 1050, 1050, 1050, 1050, 1050, 1050, 1050,…
$ DoorSpread              <dbl> 60.5, 73.5, 74.1, 66.0, 61.1, 60.3, 56.7, 68.5, 59.3, 61.0, 63.5, 62.8,…
$ WingSpread              <dbl> 19.3, 21.4, 21.5, 20.0, 19.1, 18.9, 18.3, 20.5, 19.0, 19.3, 19.5, 19.3,…
$ Buoyancy                <int> 168, 218, 218, 168, 168, 168, 168, 168, 168, 168, 168, 168, 168, 168, 1…
$ KiteArea                <dbl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ GroundRopeWeight        <dbl> 811, 811, 811, 811, 811, 811, 811, 811, 811, 811, 811, 811, 811, 811, 8…
$ TowDirection            <int> 11, 182, 124, 167, 173, 212, 223, 172, 150, 119, 190, 34, 6, 7, 99, 354…
$ SpeedGround             <dbl> 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3.7, 3…
$ SpeedWater              <dbl> 4.2, 4.2, 4.9, 4.2, 4.3, 3.8, 3.7, 3.8, 4.4, 4.8, 4.2, 4.0, 4.0, 3.7, 4…
$ SurfaceCurrentDirection <int> 315, 360, 360, 360, 315, 180, 180, 315, 315, 315, 45, 180, NA, NA, 225,…
$ SurfaceCurrentSpeed     <dbl> 0.3, 0.3, 0.8, 0.4, 0.5, 0.3, 0.3, 0.3, 0.3, 0.5, 0.4, 0.3, NA, NA, 0.4…
$ BottomCurrentDirection  <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ BottomCurrentSpeed      <dbl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ WindDirection           <int> 248, 298, 300, 193, 150, 77, 56, 85, 49, 28, 12, 24, 319, 332, 10, 31, …
$ WindSpeed               <int> 2, 10, 5, 3, 5, 10, 7, 7, 5, 5, 7, 7, 7, 7, 8, 7, 7, 9, 8, 8, 7, 6, 9, …
$ SwellDirection          <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ SwellHeight             <dbl> 0.4, 0.8, 1.0, 1.0, 0.8, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 0.8, 0.8, 1…
$ SurfaceTemperature      <dbl> 18.5, 18.9, 17.9, 17.4, 18.6, 18.2, 18.3, 18.3, 18.9, 18.8, 19.2, 18.8,…
$ BottomTemperature       <dbl> 9.3, 8.8, 8.4, 8.8, 11.2, 17.4, 15.0, 10.1, 11.5, 12.4, 10.4, 9.8, 13.2…
$ SurfaceSalinity         <dbl> 26.2, 26.2, 30.4, 32.3, 24.8, 31.1, 24.2, 24.2, 20.0, 20.8, 19.1, 20.8,…
$ BottomSalinity          <dbl> 35.1, 35.1, 35.1, 35.1, 34.6, 32.7, 33.3, 33.3, 33.5, 33.1, 32.9, 33.3,…
$ ThermoCline             <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ ThermoClineDepth        <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ CodendMesh              <int> 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20, 20,…
$ SecchiDepth             <dbl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ Turbidity               <dbl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ TidePhase               <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ TideSpeed               <dbl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ PelagicSamplingType     <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ MinTrawlDepth           <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ MaxTrawlDepth           <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ SurveyIndexArea         <chr> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ EDOM                    <int> 20251106, 20251106, 20251106, 20251106, 20251106, 20251106, 20251106, 2…
$ ReasonHaulDisruption    <lgl> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA,…
$ DateofCalculation       <chr> "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "", "",…
```

And then this:

```{r, eval = FALSE}
icesDatras::get_datras_unaggregated_data("HH", "NS-IBTS", 2025, 3) |>
  filter(!is.na(ReasonHaulDisruption)) |> count()
´´´

´´´
  n
1 0
```

```{r, eval = FALSE}
icesDatras::get_datras_unaggregated_data("HH", "NS-IBTS", 2025, 3) |> 
  filter(DateofCalculation != "" | is.na(DateofCalculation)) |> 
  count()
```

´´´
  n
1 0
```

The EDOM variable values look suspiciously like what is DateofCalculation

Action: Check for variable-value mixup


