---
title: "variables"
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE.
  warnings = FALSE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(obus)
```

```{r}
hl <-
  dr_con("HH") |> 
  select(.id, Survey, Quarter, Year, DataType, HaulDuration) |> 
  left_join(dr_con("HL", trim = FALSE))
hl2 <-
  hl |> 
  select(.id, SpeciesValidity:ValidAphiaID)
dictionary <- dr_con("dictionary")
vocab <- dr_con("vocabulary")
```

## HL row type

One of the key variable in the HL table is "SpeciesValidity". In hindsight the variable name itself is very cryptic and is e.g. easily confused by the novice with "ValidAphiaID". I like like to think of this variable as ...

```{r, echo = FALSE}
hl |> 
  count(SpeciesValidity, name = "rows") |> 
  full_join(vocab |> filter(new == "SpeciesValidity") |> 
              #mutate(key = ifelse(key == "-9", NA, key)) |> 
              select(SpeciesValidity = key, description)) |> 
  mutate(SpeciesValidity = as.integer(SpeciesValidity),
         "%" = round(rows / sum(rows, na.rm = FALSE) * 100, 2),
         .after = rows) |> 
  arrange(SpeciesValidity) |>
  knitr::kable(caption = "SpeciesValidity (SpecVal), number of rows in HL table and description.")
```

When thinking about the HL construct one can classify the variables into two groups (those affected by HH-DataType and HH-HaulDuration are marked with ^*):

* Control variables
  * SpeciesValidity
  * SpeciesCategory
  * LengthCode
  * LengthType
* Value variables
  * ValidAphiaId
  * SpeciesSex
  * DevelopmentStage
  * LengthClass
  * NumberAtLength^*
  * SpeciesCategoryWeight^*
  * SubsamplingFactor
  * SubsampledNumber^*
  * SubsampleWeight^*
  * TotalNumber
  * SpeciesCategoryWeight
  

```{r}
d <- 
  hl2 |> 
  #filter(SpeciesValidity == "4") |> 
  mutate(TotalN = ifelse(is.na(TotalNumber), 0L, 1L),
         SubsampledN = ifelse(is.na(SubsampledNumber), 0L, 1L),
         SubsampleW = ifelse(is.na(SubsampleWeight), 0L, 1L),
         SpeciesCategoryW = ifelse(is.na(SpeciesCategoryWeight), 0L, 1L),
         LengthClass = ifelse(is.na(LengthClass), 0L, 1L),
         NumberAtLength = ifelse(is.na(NumberAtLength), 0L, 1L)) |> 
  count(SpeciesValidity, TotalN, SubsampledN, SubsampleW, 
        SpeciesCategoryW, LengthClass, NumberAtLength) |> 
  mutate(SpeciesValidity = as.integer(SpeciesValidity)) |> 
  group_by(SpeciesValidity) |> 
  mutate("%" = n / sum(n, na.rm = FALSE) * 100) |> 
  ungroup() |> 
  collect() |> 
  arrange(SpeciesValidity, desc(n)) 
d |> 
  knitr::kable(digits = 2, caption = "Overview: Binary classification of value variables indication presence (1) or missing values (0).")
```

```{r}
d |> 
  filter(SpeciesValidity == 0) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 0, Invalid information. Use in data products: no.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 1) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 1, Representative length information (length measurements, numbers and category catch weights). Use in data products: yes.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 2) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 2, Representative length information only for commercial/assessed species. Full set of species to be completed next year after a two-years ban to allow national data use. Use in data products: yes, only for commercial/assessed species.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 4) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 4, Total number only. No length measurements or category catch weight. Use in data products: yes.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 5) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 5, Presence/absence only. No length measurements, numbers or category catch weights. Use in DATRAS products: yes.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 6) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 6, Category catch weight only. No length measurements or numbers. Use in DATRAS products: yes.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 7) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 7, Numbers and category catch weights only. No length measurements. Use in DATRAS products: yes.",
               digits = 2)
```

```{r}
d |> 
  filter(SpeciesValidity == 10) |> 
  select(-SpeciesValidity) |> 
  knitr::kable(caption = "SpeciesValidity == 10, Length measurements and numbers only. No category catch weights. Use in data products: yes.",
               digits = 2)
```

```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```





```{r}
hl |> count(SpeciesValidity) |> mutate(SpeciesValidity = as.integer(SpeciesValidity)) |> 
  arrange(SpeciesValidity)
```


# Variables values that depend on DataType and HaulDuration

* NumberAtLength
* 


# Standardize everthing to the haul

```{r}
hl |> 
  select(.id, Survey, Quarter, Year, latin, SpeciesSex, DevelopmentStage, SpeciesCategory, TotalNumber, SubsampledNumber, SubsampleWeight, SubsamplingFactor, DataType, HaulDuration) |> 
  distinct() |> 
  filter(SubsamplingFactor > 1)

hl |> 
  select(.id, Survey, Quarter, Year, latin, SpeciesSex, DevelopmentStage, SpeciesCategory, TotalNumber, SubsampledNumber, SubsampleWeight, SubsamplingFactor, DataType, HaulDuration, NumberAtLength) |> 
  group_by(.id, Survey, Quarter, Year, latin, SpeciesSex, DevelopmentStage, SpeciesCategory) |>
  # Why do I have to use na.rm, ie what are NA values here
  mutate(total_n = sum(NumberAtLength, na.rm = TRUE)) |> 
  ungroup() |> 
  select(-NumberAtLength) |> 
  distinct() |> 
  filter(SubsamplingFactor > 1)




```

### Lingo

In order for me to make some sense of things, my vocabulary:

For each subfactor?, species, sex and developmental stage we have:

* n_meas: This is the number **measured** in a haul per species, sex and developmental stage
* n_haul: This is the number in a haul
* n_hour: This is the standardized number fish per 1 hour towing

Subfactor bookkeeping is of interest in:

* QC
* Some statistics?



### TotalNumber (TotalNo)

* From vocabulary: "Total number of specimens per category (unique combination of cruise haul, species, sex, devstage and subsampling category identifier). For data type C, values should be standardised to 60 minutes."


```{r}
hl |> 
  mutate(n_meas_cat = case_when(DataType == "C" ~ TotalNumber / 60 * HaulDuration,
                                  .default = TotalNumber))
```



### SpeciesCategory (CatIdentifier)

From [FAQ](https://www.ices.dk/data/Documents/DATRAS/DATRAS_FAQs.pdf):

* "The CatIdentifier is the ID number given to a particular length- or weight-based subsampling of species with the same sex code in the same haul."
* "There can be reported up to **5** categories per species in one haul."

```{r}
hl |> collect() |> count(SpeciesCategory) |> collect() |> 
  ggplot(aes(factor(SpeciesCategory), n)) + geom_col() + scale_y_log10() +
  labs(x = "Species category", y = "Number of HL records")
```

```{r}
hl |> 
  select(Survey, SpeciesCategory) |> 
  collect() |> 
  group_by(Survey) |> 
  summarise(max = max(SpeciesCategory, na.rm = TRUE)) |> 
  ggplot(aes(max, reorder(Survey, max))) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 30, by = 2)) +
  labs(x = "Maximum SpeciesCategory", y = "Survey")
```

```{r}
dr_con("HH") |> 
  select(.id, Year) |> 
  left_join(hl) |> 
  select(Survey, Year, SpeciesCategory) |> 
  group_by(Survey, Year) |> 
  summarise(max = max(SpeciesCategory, na.rm = TRUE)) |> 
  ggplot(aes(Year, Survey, fill = factor(max))) +
  geom_tile() +
  scale_fill_viridis_d(direction = -1) +
  #scale_x_continuous(breaks = seq(0, 30, by = 2)) +
  labs(x = "Year", y = "Survey", fill = "Max/nSpeciesCategory")
```

```{r}

```

```{r}

```

```{r}

```

```{r}
library(tidyverse)
devtools::load_all()
vc <- dr_con("vocabulary")
vc |> filter(new == "SpeciesCategory")
hl <- dr_con("HL", trim = FALSE)
hl |> glimpse()



hl |> glimpse()
hl |> filter(is.na(LengthClass)) |> glimpse()

hl |> 
  filter(.id == "BITS:1991:1:LV:90MX:LBT:3:20",
         latin == "Enchelyopus cimbrius") |> 
  glimpse()
# If length class is missing, expect no count
hl |> 
  filter(is.na(LengthClass)) |> 
  filter(!is.na(NumberAtLength))
# If no length class, expect only one species record per station
hl |> 
  filter(is.na(LengthClass)) |> 
  count(.id, latin) |> 
  arrange(desc(n)) |> 
  collect() |> 
  slice(1:50) |> 
  knitr::kable()
# May have to take into account Sex
hl |> 
  filter(is.na(LengthClass)) |> 
  count(.id, latin, SpeciesSex) |> 
  arrange(desc(n)) |> 
  collect() |> 
  slice(1:50) |> 
  knitr::kable()
hl |> 
  filter(.id == "BTS:2023:3:NL:64T2:BT8:41F334:34",
         latin == "Amblyraja radiata") |> 
  #filter(is.na(LengthClass)) |> 
  collect() |> 
  janitor::remove_constant() |> 
  view()

hl |> 
  filter(.id == "BTS:2017:3:NL:64T2:BT8:40F043:43",
         latin == "Echinus") |> 
  #filter(is.na(LengthClass)) |> 
  collect() |> 
  janitor::remove_constant() |> 
  view()

# classify measurement type:
hl |> 
  mutate(sample = case_when(!is.na(LengthClass) ~ "length measurment",
                            .default = "rest")) |> 
  count(sample) |> 
  mutate(p = round(n / sum(n) * 100, 1))
lh <- function(x) {
  ifelse(!is.na(x), "yes", "no")
}
hl |> glimpse()
d <- 
  hl |> 
  mutate(.TotalNumber = ifelse(!is.na(TotalNumber), "yes", "no"),
         .SubsampledNumber = ifelse(!is.na(SubsampledNumber), "yes", "no"),
         .SubsampleWeight = ifelse(!is.na(SubsampleWeight), "yes", "no"),
         .SpeciesCategoryWeight = ifelse(!is.na(SpeciesCategoryWeight), "yes", "no"),
         .LengthClass = ifelse(!is.na(LengthClass), "yes", "no")) |> 
  collect() |> 
  unite(.TotalNumber:.LengthClass, col = "TN_SN_SW_SCW_LC", sep = "_")



d |> 
  count(TN_SN_SW_SCW_LC) |> 
  mutate(p = round( n / sum(n) * 100, 1)) |> 
  arrange(-n) |> 
  filter(str_ends(TN_SN_SW_SCW_LC, "_yes")) |> 
  knitr::kable()
d |> 
  filter(TN_SN_SW_SCW_LC == "no_no_no_no_yes") |> 
  glimpse()
d |> 
  filter(TN_SN_SW_SCW_LC == "no_yes_no_no_yes") |> 
  glimpse()

hl |> 
  filter(is.na(TotalNumber), is.na(SubsampledNumber), is.na(SubsampleWeight),
         is.na(SpeciesCategoryWeight), is.na(LengthClass)) |> 
  glimpse()
         
         


```
