---
title: "Accessing DATRAS data"
format:
  html:
    footnotes-hover: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(obus)
library(dplyr)
```

## Background

Working with DATRAS^[**Background to expand:** DATRAS is the pan-European trawl survey database maintained by ICES, covering [N surveys] from [year range], with three main record types: haul header (HH), catch-at-length (HL), and catch-at-age (CA). Consider linking to the [DATRAS website](https://datras.ices.dk).] data presents a practical challenge: the full dataset is
large (the HL table alone runs to ~14 million rows across all surveys), and
the traditional access route — the ICES web API — forces you to download one
survey at a time in a format that is slow to parse. The {obus} package offers
two complementary approaches that improve on this, described below.

---

## 1. API-based access

Both API methods return data as an ordinary R data frame and apply filters
(survey, year, quarter) before downloading. They are appropriate when you need
a specific slice of the data and the full dataset is not required.

### 1.1 The legacy ICES Web Services

The [DATRAS Web Services](https://datras.ices.dk/WebServices/DATRASWebService.asmx)
have been available for many years and underpin the {icesDatras} package:

```{r}
#| eval: false
icesDatras::getDATRAS(record = "HH", survey = "NS-IBTS",
                      years = 2025, quarters = 1)
```

The process is slow^[**To expand:** The main drawbacks are XML parsing overhead, one survey per call meaning looping is needed for multi-survey work, and column types being returned inconsistently (often all character).] — in particular due to parsing XML responses — and retrieves one survey at a time.

### 1.2 The DATRAS Download API

ICES also provides a
[Download API](https://datras.ices.dk/Data_products/Download/DATRASDownloadAPI.aspx)
that delivers a zipped CSV instead of XML. {obus} wraps this endpoint:

```{r}
#| eval: false
# One survey, one year range, one quarter range
get_datras_unaggregated_data(recordtype = "HH",
                              survey     = "NS-IBTS",
                              year       = "2000:2025",
                              quarter    = "1")
```

Compared to the legacy Web Services, this is faster and returns correct column
types out of the box. The limitation is the same: one survey per call, so
fetching all surveys still requires looping.^[**To confirm:** Whether the Download API has been stabilised or is still experimental, and whether rate limits or authentication requirements apply.]

---

## 2. Parquet-based access (recommended for full-dataset work)

All surveys across all years are compiled into a small set of parquet files^[**To add:** 1–2 sentences on what Parquet is for unfamiliar readers — a columnar binary format that is compact, typed, and readable by Arrow and DuckDB without loading the whole file into memory.]
and hosted on a web server.^[**To confirm:** How often `get_datras.R` is run to refresh these files and how far behind the live DATRAS database they typically are.] The base URL is:

```{r}
base_url <- "https://heima.hafro.is/~einarhj/data/datras/"
```

Four tables are available:

| File | Content |
|---|---|
| `HH.parquet` | Haul header — station metadata, gear, position, timing |
| `HL.parquet` | Catch at length — length-frequency records by species and haul |
| `CA.parquet` | Catch at age — individual biological samples |
| `species_worms.parquet` | Species list — AphiaID, latin name, common name (from WoRMS) |

There are two ways to work with these files.

### 2.1 Reading fully into R (`arrow::read_parquet`)

The simplest approach: download the entire table into an R data frame.

```{r}
#| eval: false
hh <- arrow::read_parquet(paste0(base_url, "HH.parquet"))
hl <- arrow::read_parquet(paste0(base_url, "HL.parquet"))
```

This is convenient for HH (haul header), which is small enough to hold
comfortably in memory. For HL and CA the full tables are larger; if you only
need a subset it is more efficient to use the connection approach below.

### 2.2 Connecting via DuckDB (`dr_con`) — the preferred approach

`dr_con()` opens a lazy connection to the parquet file through an in-process
DuckDB instance. No data is downloaded until you explicitly request it:

```{r}
hh <- dr_con("HH")
hl <- dr_con("HL")
ca <- dr_con("CA")
```

The object returned is a `tbl_duckdb_connection` — it behaves like a data
frame in `dplyr` pipelines but the computation happens inside DuckDB, not R.
Rows are only transferred to R when you call `collect()`.

```{r}
# Peek at structure — no data transferred yet
hl |> glimpse()
```

Note `Rows: ??` in the output: DuckDB has not scanned the file to count rows.
This is intentional — the point is to push filtering down to DuckDB so that
only the rows you actually need cross the network.

---

## 3. The lazy evaluation workflow

The typical pattern with `dr_con()` is:

```
connect → filter/mutate/join → collect()
```

Everything before `collect()` is translated to SQL and executed by DuckDB
against the parquet file. Only the result lands in R. For example, to get
North Sea IBTS haul positions for 2020–2025:

```{r}
#| eval: false
hh_ns <- hh |>
  filter(Survey == "NS-IBTS", Year >= 2020) |>
  select(Survey, Year, Quarter, .id, ShootLongitude, ShootLatitude) |>
  collect()
```

Joins across tables also work before collection:

```{r}
#| eval: false
# Cod catches in NS-IBTS, with haul metadata
cod <- hh |>
  filter(Survey == "NS-IBTS") |>
  select(.id, Year, Quarter, HaulDuration) |>
  left_join(
    hl |> filter(latin == "Gadus morhua"),
    by = ".id"
  ) |>
  collect()
```

The join is executed inside DuckDB: only the matching rows from the HL
file are read, not the full 14-million-row table.

::: {.callout-note}
DuckDB can read parquet files directly over HTTPS. The connection in `dr_con()`
points at the remote URL; there is no local copy of the data.
:::

---

## 4. Pre-computed variables in the parquet files

The production pipeline (`get_datras.R`) applies several transformations
before writing the parquet files, so these derived variables are already
present when you connect:

| Variable | Table | Description |
|---|---|---|
| `.id` | HH, HL, CA | Haul identifier: `Survey:Year:Quarter:Country:Platform:Gear:StationName:HaulNumber` |
| `date` | HH | Shoot date (`lubridate::make_date(Year, Month, Day)`) |
| `time` | HH | Shoot timestamp (from `StartTime` in HHMM format) |
| `latin` | HL, CA | Scientific name from WoRMS, joined on `ValidAphiaID` |
| `length_cm` | HL, CA | Standardised length in cm, accounting for `LengthCode` |
| `n_haul` | HL | Numbers caught per haul, corrected for `DataType` and `SubsamplingFactor` |
| `n_hour` | HL | `n_haul` raised to a 60-minute standard effort |

The `.id` key is the linking field between HH, HL, and CA — all joins across
tables should be made on `.id`.

The `n_haul` and `n_hour` columns account for the haul-level `DataType`:

| `DataType` | Meaning | `n_haul` formula |
|---|---|---|
| `C` | Data recorded as CPUE (per hour) | `NumberAtLength × SubsamplingFactor × HaulDuration / 60` |
| `R` | Data recorded by haul | `NumberAtLength × SubsamplingFactor` |
| `P` | Pseudocategory sampling | `NumberAtLength × SubsamplingFactor` |
| `S` | Subsampled data | `NumberAtLength × SubsamplingFactor` |
| `-9` / `NA` | Invalid haul | `NA` |

### HL trimming

When `trim = TRUE` (the default), `dr_con("HL")` returns a reduced column set
focused on the variables needed for length-frequency analysis:

```{r}
hl |> glimpse()
```

Pass `trim = FALSE` to retrieve all columns, including `TotalNumber`,
`SpeciesCategoryWeight`, `SpeciesCategory`, and the full set of HL exchange
variables.

---

## 5. The ICES datacenter parquet store

The ICES datacenter is separately exploring hosting DATRAS data as parquet
files on a cloud (Azure Blob) server. At the time of writing this is
experimental^[**To confirm:** Whether this store is still limited to NS-IBTS or has expanded to other surveys, its current stability status, and who to contact at ICES for access details.] and covers NS-IBTS only:

```{r}
url_ices <- "https://stdatrastest001.blob.core.windows.net/datras/ns-ibts/HH.parquet"
arrow::read_parquet(url_ices) |> glimpse()
```

The column names and types in this store follow the raw DATRAS exchange format
and differ from the {obus} parquet files — the derived variables (`.id`,
`latin`, `n_haul`, etc.) are not present. It is anticipated that this store
will eventually cover all surveys and become the authoritative source,
at which point `https://heima.hafro.is/~einarhj/data/datras/` will be
retired.
