---
title: "Accessing DATRAS data"
format:
  html:
    footnotes-hover: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(duckdb)
library(duckdbfs)
library(obus)
library(tidyverse)
```

## Background

Working with DATRAS data presents a practical challenge: the full
dataset is large (the HL table alone runs to \~14 million rows across
all surveys), and the traditional access route — ICES web API —
is limited to accessing and downloading data with filtering on **predefined**
variables (Survey, Year, Quarter, Aphia) in a format that is large and hence 
slow to download, requires good downstream datafile bookkeeping and coding to build
up a complete dataset. Complementary approach that improve on this
are described below.


## 1. API-based access

Both ICES DATRRAS API methods return data as an ordinary R data frame working on one
survey at a time.

### The legacy DATRAS Web Services

The [DATRAS Web
Services](https://datras.ices.dk/WebServices/DATRASWebService.asmx) have
been available for many years and underpins the {icesDatras} package:

```{r eval = FALSE}
icesDatras::getDATRAS(record = "HL", survey = "NS-IBTS",
                      years = 2020:2025, quarters = 1:4)
```

The process is reasonably fast if one is only interested one Survey,
but very slow if one wants the full DATRAS dataset, even for a single survey.
The main reason is because xml-tags are wrapped around each 
field value resulting in large uncompressed file size being transferred over the 
web. This gets compounded by the API request actually being limited to one survey x year x quarter at a time.

### The DATRAS Download API

ICES has recently provided a [Download
API](https://datras.ices.dk/Data_products/Download/DATRASDownloadAPI.aspx)
that delivers a zipped CSV instead of XML. {icesDatras} wraps this
endpoint:

```{r}
#| eval: false
# One survey, one year range, one quarter range
icesDatras::get_datras_unaggregated_data(recordtype = "HL",
                              survey     = "NS-IBTS",
                              year       = "2000:2025",
                              quarter    = "1:4")
```

Compared to the legacy Web Services, this is much faster: the response is a **zipped CSV** rather than raw xml, which means less data travels over the network and no XML parsing overhead on the downstream side. The limitation remains the same: filtering is restricted to the predefined variables (Survey, Year, Quarter).

------------------------------------------------------------------------

## 2. Parquet-based access

Parquet is a binary, columnar file format. It has built-in compression per column. Because a column holds values of the same type — often with low cardinality (e.g. Survey is the same string repeated thousands of times) — compression ratios are high. The file size on disk and over the network is normally a fraction of the equivalent CSV^[The full HL table is ~100 MB as csv, while only ~20 MB as parquet].

All surveys across all years have been compiled into a set of parquet files and hosted temporarily on a web server. The base URL is:

```{r}
base_url <- "https://heima.hafro.is/~einarhj/datras"
```

### Reading fully into R (`arrow::read_parquet`)

The simplest approach is to read the entire dataset directly into R by:

```{r}
#| eval: false
hh <- arrow::read_parquet(file.path(base_url, "HH.parquet"))
hl <- arrow::read_parquet(file.path(base_url, "HL.parquet"))
ca <- arrow::read_parquet(file.path(base_url, "CA.parquet"))
```

This is convenient for haul data (HH), which is relatively small. For HL and CA the full tables are larger; if you
only need a subset it is more efficient to use the connection approach below.

The ICES datacenter is separately experimenting with hosting DATRAS data
as parquet files on an S3-compatible cloud store (Azure Blob). At the
time of writing this is limited to NS-IBTS only, and the files follow
the raw DATRAS exchange format to the letter (including -9 for NA-values).

```{r}
url_ices <- "https://stdatrastest001.blob.core.windows.net/datras/ns-ibts/HL.parquet"
hl <- arrow::read_parquet(url_ices)
```

### Connecting via DuckDB (`dr_con`)

In addition to the compressed columnar format parquet stores each column contiguously, so reading only the columns you need means only those bytes are touched. For a table with 35 columns where you need 5, you skip roughly 85% of the file.

Within each column, parquet further divides the data into row groups — horizontal sections of the data.^[Typically 100,000–1,000,000 rows each. The file metadata stores the minimum value, maximum value, and null count for every column in every row group.] A query engine reads the metadata statistics first: row groups that do not contain matching values are skipped entirely, without reading a single data byte. This means targeted byte-range requests for only the relevant sections of the file.

The combination matters especially for remote access: a query engine like DuckDB can read a parquet file over HTTPS and apply "predicate pushdown" — filtering rows and selecting columns before transferring data to R. For a table like HL with ~14 million rows and 35 columns, a well-filtered query may only pull a few thousand values across the network, regardless of the upstream file size.

`dr_con()` opens a lazy connection to the parquet file through an
in-process DuckDB instance. No data is downloaded until you explicitly
request it:

```{r}
hh <- dr_con("HH")
hl <- dr_con("HL")
ca <- dr_con("CA")
```

The object returned is a `tbl_duckdb_connection` — it behaves like a
data frame in `dplyr` pipelines but the computation happens inside
DuckDB, not R. Rows are only transferred to R when you call `collect()`.

```{r}
# Peek at structure — no data transferred yet
hl |> glimpse()
```

Note `Rows: ??` in the output: DuckDB has not scanned the file to count
rows, we just get a wiff of the actual dataset.


#### Lazy evaluation workflow

The typical pattern with `dr_con()` is:

```         
connect → filter/select/mutate/join/summarise → collect()
```

Everything before `collect()` is translated to SQL and executed by
DuckDB against the parquet file. Only the result lands in R. For
example, to get North Sea IBTS haul positions for 2020–2025:

```{r}
#| eval: false
hh_ns <- hh |>
  filter(Survey == "NS-IBTS", Year >= 2020) |>
  select(Survey, Year, Quarter, .id, ShootLongitude, ShootLatitude) |>
  collect()
```

For those familiar with using dplyr-verbs to process data most of those function as well as many base-R functions have an equalent sql-backend. E.g. one can get all survey stations for the third quarter in 2025 and add to that the number of cod observed using the following script:

```{r}
data <-
  # Process the data in DuckDB
  hh |> 
  filter(Year == 2025,
         Quarter == 3) |> 
  left_join(hl |> 
              filter(latin == "Gadus morhua") |> 
              group_by(.id) |> 
              summarise(n_haul = sum(n_haul, na.rm = TRUE)),
            by = join_by(.id)) |> 
  # Import the data into R
  collect() |> 
  mutate(n_haul = replace_na(n_haul, 0))
```

Here all the code steps prior to the collect command are automatically translated to SQL and passed to the in-process DuckDB. It is only at collect step that the data is actually downloaded and imported into R. More importantly, only the variables Year, Quarter, latin, .id and n are ever passed over the web. In addition only certain chunks of the parquet files (read: rows), those that "fall within the range" of the values in the variables are equivalently passed, the rest being ignored.

### Working with a local copy

One can easily download the full DATRAS paruet data and work from a local copy. Here the HH, HL and CA paquet files are downloaded and stored in the default "data"-directory.

```{r eval = FALSE}
obus::dr_download()
```

Now user can choose to either read in the whole datasets into R via read_parquet or create a duckdb-connection via:

```{r eval = FALSE}
hh <- open_dataset("data/HH.parquet")
hl <- open_dataset("data/HL.parquet")
ca <- open_dataset("data/CA.parquet")
```

