---
title: "Two catch tables for DATRAS"
subtitle: "`HL_length` and `HL_summary` — a proposal, what they reproduce, and the conventions that shape them."
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    citations-hover: true
    reference-location: document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE,
                      message = FALSE, warning = FALSE)
options(width = 120)
# The argument in this article is arithmetic, so the default three significant
# figures actively hide it: a raising factor of 1.4466 prints as 1.45, and
# 1.45 x 21 does not give the reported 30. Five digits put the numbers the
# prose quotes on the screen the reader is looking at. Seven, not five, because
# HL_summary's two routes differ at the fourth decimal (467.82 vs 467.824)
# and that difference is the point of the comparison.
options(pillar.sigfig = 7)
```

This article is about two tables — `HL_length` and `HL_summary` — that
`{obus}` derives from the DATRAS HL record, and about why the record makes
them necessary. It describes what the package does now; it does not propose
changes to it.

**What it argues for is the tables, not the package.** They are the
calculation nearly every DATRAS analysis performs before it does anything else,
and the right home for them is ICES's own product suite, next to CPUEL —
computed once, centrally, and served to everyone, rather than re-implemented
once per institute, once per language and once per working group.

This is not a proposal in the abstract. Built from these two tables and nothing
else, `{obus}` computes ICES's own CPUEL product for every survey in the
archive, where ICES computes it for a few, and agrees with it on most counts
where the two overlap, the exceptions named below; and the `{DATRAS}` and `{DATRASextra}` packages
run on them **unmodified**, reproducing numbers at length, totals per haul,
swept area and a stratified abundance index with no difference at all — the
index a working group would report comes out the same to every digit. Both
demonstrations are below, with the places they fall short named rather than
glossed. If ICES served these two tables, [`{obus}`](https://github.com/einarhjorleifsson/obus)
would not need to exist, and that would be the better outcome.

```{r}
library(obus)
library(dplyr)
library(ggplot2)
```

```{r, include = FALSE}
# Every number the prose quotes is measured on render and referenced inline,
# so the text cannot drift from the archive (Working Principle 8). Numbers a
# displayed chunk prints are taken from that chunk's result; the ones the
# prose needs before any chunk has shown them are measured here. The article
# used to type its analytical figures in as "dated evidence" without dates,
# and after the archive rebuild of 2026-10-04 a dozen of them disagreed with
# the chunk printed directly above them. The whole article knits in about two
# minutes, so nothing is too expensive to measure.
M <- list(
  len_rows   = dr_con("HL_length")  |> count() |> pull(n),
  sum_rows   = dr_con("HL_summary") |> count() |> pull(n),
  sum_groups = dr_con("HL_summary") |> distinct(.id, Valid_Aphia) |>
                 count() |> pull(n),
  # HL_summary rows with no HL_length rows behind them. Stated in the prose
  # AND computed by the displayed chunk just below it, so a drift here would
  # show up as the article contradicting itself on the same screen.
  sum_nolen  = dr_con("HL_summary") |>
                 anti_join(dr_con("HL_length") |>
                             distinct(.id, Valid_Aphia, SpeciesValidity),
                           by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
                 count() |> pull(n)
)
# Hauls touched by that, so the prose can say whether it is concentrated or
# spread. The claim it supports is "a third of hauls", which will drift.
M$sum_hauls <- dr_con("HL_summary") |> distinct(.id) |> count() |> pull(n)
M$sum_nolen_hauls <- dr_con("HL_summary") |>
  anti_join(dr_con("HL_length") |>
              distinct(.id, Valid_Aphia, SpeciesValidity),
            by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  distinct(.id) |> count() |> pull(n)
M$sum_nolen_pct <- 100 * M$sum_nolen / M$sum_rows
M$sum_nolen_haul_pct <- 100 * M$sum_nolen_hauls / M$sum_hauls

# Part I: how much of the archive is DataType C, and how short its hauls are.
hl_dt <- dr_con("HL") |> inner_join(dr_con("HH") |> select(.id, DataType), by = ".id")
x <- dr_con("HH") |>
  summarise(hauls = n(), c = sum(as.integer(DataType == "C"), na.rm = TRUE),
            c_short = sum(as.integer(DataType == "C" & HaulDuration < 60), na.rm = TRUE)) |>
  collect()
M$c_haul_pct  <- 100 * x$c / x$hauls
M$c_short_pct <- 100 * x$c_short / x$c
x <- hl_dt |> summarise(rows = n(), c = sum(as.integer(DataType == "C"), na.rm = TRUE)) |> collect()
M$c_row_pct <- 100 * x$c / x$rows

# How much of the format is in use. -9 and NA are not codes.
M$n_datatype <- dr_con("HH") |> filter(DataType != "-9") |> distinct(DataType) |> count() |> pull(n)
M$n_lengthcode <- dr_con("HL") |> filter(!is.na(LengthCode), LengthCode != "-9") |>
  distinct(LengthCode) |> count() |> pull(n)
M$n_category <- dr_con("HL") |> filter(!is.na(SpeciesCategory)) |>
  distinct(SpeciesCategory) |> count() |> pull(n)
M$platforms <- dr_con("HH") |> distinct(Country, Platform) |> count(Country) |> collect() |>
  (\(d) setNames(d$n, d$Country))()
x <- dr_con("HL") |> filter(!is.na(LengthCode), LengthCode != "-9") |>
  group_by(Country, Platform, Year, Valid_Aphia) |>
  summarise(codes = n_distinct(LengthCode), .groups = "drop") |>
  summarise(n = n(), one = sum(as.integer(codes == 1))) |> collect()
M$code_constant_pct <- 100 * x$one / x$n

# The units that submit: countries, and platform-years of the platforms the
# displayed chunk counts (5,000 length rows or more).
x <- hl_dt |> group_by(Country) |>
  summarise(rows = n(), dt = n_distinct(DataType), .groups = "drop") |>
  filter(rows >= 5000) |> summarise(n = n(), one = sum(as.integer(dt == 1))) |> collect()
M$countries <- x$n; M$countries_one <- x$one
big <- hl_dt |> group_by(Country, Platform) |> summarise(rows = n(), .groups = "drop") |>
  filter(rows >= 5000) |> select(Country, Platform)
x <- hl_dt |> semi_join(big, by = c("Country", "Platform")) |>
  group_by(Country, Platform, Year) |>
  summarise(dt = n_distinct(DataType), .groups = "drop") |>
  summarise(n = n(), one = sum(as.integer(dt == 1))) |> collect()
M$platform_years <- x$n; M$platform_years_one <- x$one
x <- dr_con("HH") |> filter(Survey %in% c("NS-IBTS", "BITS"), DataType != "-9") |>
  group_by(Survey, Country) |> summarise(dt = n_distinct(DataType), .groups = "drop") |>
  summarise(n = n(), one = sum(as.integer(dt == 1))) |> collect()
M$programmes <- x$n; M$programmes_one <- x$one
x <- dr_con("HL") |> filter(!is.na(LengthCode), LengthCode != "-9") |>
  group_by(Survey) |> summarise(k = n_distinct(LengthCode), .groups = "drop") |>
  summarise(n = n(), multi = sum(as.integer(k > 1))) |> collect()
M$surveys <- x$n; M$surveys_lc_multi <- x$multi

# Part II: CPUEL's coverage and zeros, and how heavily surveys sex their catch.
cpuel <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CPUEL.parquet")
M$cpuel_surveys <- cpuel |> distinct(Survey) |> count() |> pull(n)
x <- cpuel |> summarise(rows = n(), zero = sum(as.integer(n_hour == 0), na.rm = TRUE)) |> collect()
M$cpuel_rows <- x$rows; M$cpuel_zero <- x$zero
M$sexed <- dr_con("HL_length") |> group_by(Survey) |>
  summarise(rows = n(), sexed = sum(as.integer(SpeciesSex %in% c("F", "M", "B")), na.rm = TRUE),
            .groups = "drop") |>
  filter(rows >= 100000) |> collect() |>
  mutate(pct = 100 * sexed / rows) |> arrange(desc(pct))

# What makes a row unique: the coarser key, and how often each field splits it.
x <- dr_con("HL_length") |>
  mutate(ds = coalesce(DevelopmentStage, "<NA>"), lt = coalesce(LengthType, "<NA>"),
         sv = coalesce(SpeciesValidity, "<NA>")) |>
  group_by(.id, Valid_Aphia, length_mm, SpeciesSex) |>
  summarise(ds = n_distinct(ds), lt = n_distinct(lt), sv = n_distinct(sv),
            acc = n_distinct(accuracy), .groups = "drop") |>
  summarise(n = n(), ds = sum(as.integer(ds > 1)), lt = sum(as.integer(lt > 1)),
            sv = sum(as.integer(sv > 1)), acc = sum(as.integer(acc > 1))) |>
  collect()
M$coarse <- x
M$ds_na_pct <- dr_con("HL_length") |>
  summarise(p = 100 * sum(as.integer(is.na(DevelopmentStage))) / n()) |> pull(p)
# HL_length rows that combine more than one submitted row.
x <- dr_con("HL") |> filter(!is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia, LengthClass, LengthCode, LengthType, SpeciesSex,
           DevelopmentStage, SpeciesValidity) |>
  summarise(n = n(), .groups = "drop") |>
  summarise(rows = n(), combined = sum(as.integer(n > 1))) |> collect()
M$combined_one_in <- x$rows / x$combined
# TotalNumber over several sex rows: one value repeated, or one per sex.
x <- dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage,
           SpeciesSex, TotalNumber) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage) |>
  summarise(n_sex = n_distinct(SpeciesSex), n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_sex > 1) |>
  summarise(n = n(), repeated = sum(as.integer(n_val == 1)),
            per_sex = sum(as.integer(n_val == n_sex))) |> collect()
M$sex_repeated_pct <- 100 * x$repeated / x$n
M$sex_per_sex_pct  <- 100 * x$per_sex / x$n
M$sex_na_pct <- dr_con("HL") |>
  summarise(p = 100 * sum(as.integer(is.na(SpeciesSex))) / n()) |> pull(p)

# Part III: the share of haul-species records with a suspect or unexplained flag.
M$flagged_pct <- 100 * (dr_con("hl_flag") |>
  left_join(dr_con("hl_flag_code") |> select(code, kind), by = "code") |>
  filter(kind %in% c("suspect", "unexplained")) |>
  distinct(.id, Valid_Aphia) |> count() |> pull(n)) /
  (dr_con("HL_summary") |> distinct(.id, Valid_Aphia) |> count() |> pull(n))

# Thousands separators, so the prose reads the way it always has, and percent.
N <- function(x) formatC(as.numeric(x), format = "d", big.mark = ",")
P <- function(x, digits = 0) sprintf(paste0("%.", digits, "f%%"), x)
```

Every claim below is either a quotation from an ICES document or a measurement
of the archive, given as code you can run. [How this article was
made](#how-this-article-was-made), in the appendices, says what counts as
evidence and how the figures are kept from drifting as the archive grows.

# Part I. What is wrong with the HL table

## A good record, and a poor table

HL does the job it was designed for — capturing, faithfully and completely,
what a country submitted for a haul. That is not the same job as being the
table you analyse from, and this part is about the difference. Nothing
below is an argument that the format is broken; it is an argument that the
record and the analysis want different shapes, and that the conversion between
them is being done separately by everyone who needs it.

What a country sends ICES for one haul is a set of length rows: a length
class, a number, a raising factor relating a measured sub-sample to the whole
catch, and — repeated on every row — the totals for that species category in
number and weight.

**What that number means depends on how the country chose to submit.** Under
most conventions it is fish *measured*, and the raising factor is what relates
it to fish caught. Under `DataType` C it is neither: the submitter has already
raised it *and* already scaled it to one hour of towing, so the row carries a
catch rate and the factor is 1. Note which way that scaling goes:
`r P(M$c_short_pct)` of `C` hauls are shorter than an hour, most of them
half-hour tows, so the number written on the row is usually **larger** than
the number of fish actually caught, not smaller. This is not a corner case.
`C` covers `r P(M$c_haul_pct)` of all the hauls in the archive, and
`r P(M$c_row_pct)` of its length rows.

Given the `DataType`, any one row is unambiguous. **It is the column that is
ambiguous**, across submissions — and the `DataType` is not in HL at all. It
is in the haul record, which is why no amount of reading an HL row tells you
what its numbers are.

Getting from a submission to a catch is the same short list of steps for every
survey in the archive:

- raise the measured numbers by the sub-sampling factor — except where the
  convention has already done it;
- turn the length class into a length, in known units;
- decide what to do when the raising factor is missing, or the haul lasted no
  time at all;
- avoid adding up a total the submitter repeated across sex rows, sampling
  categories or record types.

Nearly every DATRAS analysis begins there, and they are the same steps every
time.

## Three tallies in one table {#what-is-wrong-with-hl}

The steps above are bookkeeping. The reason they are *easy to get wrong* is
structural, and it is worth naming before the tables that fix it.

**HL is one table doing three jobs.** Its columns are three different tallies
of three different things:

| Tally | Columns | What it counts |
|---|---|---|
| **by length** | `LengthClass`, `NumberAtLength` | fish in each length class |
| **of the sub-sample** | `SubsampledNumber`, `SubsamplingFactor`, `SubsampleWeight` | what was examined, and how it scales to the catch |
| **of the catch** | `TotalNumber`, `SpeciesCategoryWeight` | what came aboard |

Only the first has a row of its own. The other two are one value each, and
because the file offers only one kind of row they are **written onto
every length row of that category**. [One haul, from submission to catch at
length](#one-haul-from-submission-to-catch-at-length) below shows exactly this
in a sprat record.

That is not a DATRAS eccentricity. It is the textbook case of what Wickham
[@wickham2014tidy] describes as storing *multiple observational units* — several
different things being counted — in one table, and the consequence is exactly
the one the format's users keep meeting: **a column you must not add up.** Add `TotalNumber` down a
species' length rows and you count the same catch once for every length class
it happened to be measured in. Every repeated-value trap in [DATRAS data issues](data-issues.html) comes
back to this one thing.

**The three tallies are not merely stacked — they can come apart.** Under
pseudo-category sampling the sub-sample really is finer than the catch: the
raising factor belongs to the sub-category, while the total and the weight
belong to the main category above it. Take every `P` main category that was
split into more than one sub-category, and ask which of the three fields move
together:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "P", !is.na(SpeciesCategory)) |>
  mutate(main_category = substr(SpeciesCategory, 1, 1)) |>
  distinct(.id, Valid_Aphia, SpeciesValidity, main_category, SpeciesCategory,
           SubsamplingFactor, TotalNumber, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, main_category) |>
  summarise(sub_categories = n_distinct(SpeciesCategory),
            factors = n_distinct(SubsamplingFactor),
            totals  = n_distinct(TotalNumber),
            weights = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(sub_categories > 1) |>
  summarise(main_categories = n(),
            factor_varies   = sum(as.integer(factors > 1)),
            total_constant  = sum(as.integer(totals == 1)),
            weight_constant = sum(as.integer(weights == 1))) |>
  collect()
```

In almost every one of them the raising factor changes from sub-category to
sub-category while the total and the weight stay put. Three levels of
description, and one kind of row to write them all on.

That is why ICES's own sum `TotalNumber = SubsampledNumber ×
SubsamplingFactor` cannot hold under `P`: its three terms do not all describe
the same thing. [Pseudo category
sampling](data-issues.html#totalnumber-across-categories) therefore needs treating on its own
rather than in a sentence.

Two further properties compound it:

- **`LengthClass` is a number whose unit *and resolution* are in another
  column.** `LengthCode` says whether the value is millimetres or centimetres,
  so the number alone does not say what it measures — but it also says how wide
  the length class is, and that is the harder half.
- **`NumberAtLength` is a quantity whose identity is in another *table*.** As
  above, it is a measured count or an hourly rate depending on `DataType`,
  which lives in HH. You cannot tell from the HL row in front of you.

**The resolution is the harder half because the unit does not imply it.** Two
codes report millimetres and two of the three centimetre codes are not 1 cm:
`.` is 1 mm classes and `0` is 5 mm; `1`, `2` and `5` are 1, 2 and 5 cm
(though the archive holds no `2`). So a
`LengthClass` of 150 is the same 15 cm fish under `.` and under `0`, but in one
case it stands for a class 1 mm wide and in the other 5 mm. Across the archive
the classes run from 1 mm to 5 cm — a fiftyfold range — and `LengthClass` is
the class's **lower edge**, not the fish, so without the code you know neither
where the class starts nor how far it extends.

**Most of that variation is good practice, not sloppiness.** Resolution is
chosen to suit the fish: within one country, platform and year, the code is
constant for a given species in `r P(M$code_constant_pct)` of cases — and the choices are the ones a
biologist would make, herring and sprat measured in 5 mm classes, cod, plaice,
haddock and whiting in 1 cm.

```{r}
dr_con("HL") |>
  filter(!is.na(LengthCode), LengthCode != "-9") |>
  group_by(Country, Platform, Year, Valid_Aphia) |>
  summarise(codes = n_distinct(LengthCode), .groups = "drop") |>
  summarise(platform_year_species = n(),
            one_code_throughout   = sum(as.integer(codes == 1)),
            more_than_one         = sum(as.integer(codes > 1))) |>
  collect()
```

**What bites is that the choice is the submitter's, not the species'.** The
same fish is measured at different resolutions by different programmes:

```{r}
widths <- dr_con("HL") |>
  filter(Valid_Aphia %in% c(126436, 126417), !is.na(LengthCode), LengthCode != "-9") |>
  left_join(dr_con("species") |> select(Valid_Aphia, species), by = "Valid_Aphia") |>
  count(species, LengthCode) |>
  mutate(class_width = case_when(LengthCode == "." ~ "1 mm", LengthCode == "0" ~ "5 mm",
                                 LengthCode == "1" ~ "1 cm", LengthCode == "2" ~ "2 cm",
                                 LengthCode == "5" ~ "5 cm")) |>
  arrange(species, desc(n)) |>
  collect()
widths
```

Cod arrives in all four widths — mostly 1 cm, but
`r N(widths$n[widths$species == "Atlantic cod" & widths$LengthCode == "."])`
length rows at 1 mm and a few thousand at 5 cm, a fiftyfold spread for one
species. Herring is
mostly 5 mm and substantially 1 mm. So the moment you pool a species across
surveys — which is what a stratified index does — you are pooling classes of
different widths.

Why that matters is arithmetic, not bookkeeping. A count in a 5 cm class and a
count in a 1 mm class are not the same measurement, and no unit conversion makes
them so: at equal abundance the wider class simply collects more fish. Add them
together and the result is not a length distribution that is wrong by some
correctable factor — it is not a length distribution. `{obus}` therefore
resolves `LengthCode` once, on reading, into `length_mm`, `length_cm` and
`accuracy`, so the class width travels with every row instead of being
recoverable only by going back to HL.

**Why this is easy to miss.** The conventions vary far more between surveys
than they do inside any one of them. Most surveys have submitted under a
single convention for their whole history, and it is always either `R`
throughout or `C` throughout:

```{r}
dr_con("HH") |>
  filter(DataType != "-9") |>
  distinct(Survey, DataType) |>
  group_by(Survey) |>
  summarise(conventions = n(),
            submits = ifelse(n() == 1, min(DataType, na.rm = TRUE), "mixed"),
            .groups = "drop") |>
  count(conventions, submits) |>
  collect()
```

Neither `S` nor `P` ever defines a survey on its own: every `S` haul in the
archive sits inside a survey that also submits `R`, and `P` appears only
in surveys that also use `R` and `C`. So the two conventions that stand alone
are exactly the two that disagree about what `NumberAtLength` counts.

The practical effect is that within one survey `NumberAtLength` usually means
one stable thing, and nothing in the record warns you that it means something
else in the next survey along. The column is clear locally and ambiguous only
archive-wide, which is the kind of problem that survives a long time, because
every individual dataset is self-consistent. The mixed surveys are where it
becomes visible without leaving home — NS-IBTS carries `R`, `C` and `P`
together, BITS `R` and `C`. Even there it tends to nest:
`r M$programmes_one` of the `r M$programmes` national programmes inside those
two surveys have only ever used one convention, so a single country's series
can still look perfectly uniform.

**And no one who submits ever uses the whole format.** The archive as a whole
draws on `r M$n_datatype` `DataType` conventions, `r M$n_lengthcode`
`LengthCode` values and `r M$n_category` `SpeciesCategory` codes. The unit
that actually samples and uploads is nearer a platform than a country —
`Platform` stands in for the institutes within a country, and Germany runs
`r M$platforms[["DE"]]` of them, the Netherlands `r M$platforms[["NL"]]`,
Norway `r M$platforms[["NO"]]` — so count what each country and platform
between them ever uses:

```{r}
pl <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  group_by(Country, Platform) |>
  summarise(rows           = n(),
            data_types     = n_distinct(DataType),
            length_codes   = n_distinct(LengthCode),
            category_codes = n_distinct(SpeciesCategory), .groups = "drop") |>
  filter(rows >= 5000) |>
  summarise(platforms                  = n(),
            recording_one_way_only     = sum(as.integer(data_types == 1)),
            most_data_types_by_any_one = max(data_types),
            most_category_codes_by_any_one = max(category_codes)) |>
  collect()
pl
```

`r pl$recording_one_way_only` of the `r pl$platforms` record their catch one
way only, and no single platform ever uses more than
`r pl$most_data_types_by_any_one` of the `r M$n_datatype` conventions. The
closer you look at the unit that does the work, the smaller its corner of the
format gets: at country level `r M$countries_one` of `r M$countries` never
vary, at platform level `r pl$recording_one_way_only` of `r pl$platforms`.
Adding time tightens it again — in `r N(M$platform_years_one)` of
`r N(M$platform_years)` platform-years the catch is reported one way
throughout.

That is the position someone is in when they upload. They sample in a single,
settled way, and must then express it in a format built wide enough to hold
everybody else's way of working too — most of which is irrelevant to them, and
therefore most of which they have little reason to understand well. The parts
they never use are the parts they are most likely to fill in wrongly.

This is worth stating plainly, because the register in [DATRAS data issues](data-issues.html) could otherwise
read as a list of other people's mistakes. It is better read as the cost of one
row shape serving every sampling practice in the community at once.

Length units run the other way and are worse for it: **`r M$surveys_lc_multi`
of the `r M$surveys` surveys use more than one `LengthCode`**, so there the
variation is inside the series rather than between them.

The remedy for the first problem — several different things being counted in
one table — is to give each of them a table of its own, and that is what
`HL_length` and `HL_summary` are: not a packaging decision, but the splitting
apart that the record has been asking for. Each holds one row per observation
of one kind, and each column means one thing in one unit: `length_mm` is
always millimetres, `n_haul` is always a whole-haul count, `n_hour` is always
a rate per hour. The branching on `DataType` and `LengthCode` is worked out
once, when the tables are built, instead of in every analysis that reads
them.

**Two tables, three tallies.** `HL_length` is the length tally and
`HL_summary` the catch tally. The sub-sample tally gets no table of its own:
`SubsamplingFactor`, `SubsampledNumber` and `SubsampleWeight` appear in
neither file. What survives of it is `n_measured`, the count of fish actually
put through the calipers, carried alongside the lengths in both tables; the
raising factor is applied and then discarded.

**Is a third table warranted? Probably not, and the reason is what it would be
used for.** The sub-sample tally answers questions about *method* — how much
of the catch was examined, whether a raising factor is plausible, how a
provider's sampling changed over time. Those are quality-control and
methodological questions, and they are asked at the record level by people who
want the records: HH and HL themselves, which ICES already serves and which
`{obus}` exposes unchanged through `dr_con_raw()`. A derived table would be a
second copy of something nobody was struggling to obtain. The catch tables
exist because the catch questions need arithmetic that is easy to get wrong;
the sampling questions need the raw fields undisturbed, which is a different
service.

One piece of the sub-sample tally does earn a place in both tables, and it is
carried: `n_measured`. How much of a length distribution was measured rather
than raised is a question asked *about a catch*, at the level of the catch, which
is why it belongs here and the factor does not. What genuinely disappears is
`SubsampleWeight`, and with it the only candidate cross-check the weight path
has — that is a real loss rather than a tidy-up, and it is [an open
question](data-issues.html#what-remains-open) rather than a settled one.

# Part II. The two tables

## The two tables, formally {#the-two-tables-formally}

So much for why they have the shape they do. This is what they actually
contain. Both are one flat parquet file, read on demand.

```{r, include = FALSE}
# The column list is READ FROM THE PUBLISHED FILES; only the descriptions are
# authored here. A column added to obus and not described below renders as a
# visible gap rather than quietly vanishing from this article (Principle 8).
# n_haul and n_measured mean different things in the two tables, so the
# descriptions are per-table rather than per-name.
KEY <- list(
  HL_length  = c(".id", "Valid_Aphia", "length_mm", "accuracy", "LengthType",
                 "SpeciesSex", "DevelopmentStage", "SpeciesValidity"),
  HL_summary = c(".id", "Valid_Aphia", "SpeciesValidity")
)
COMMON <- c(
  .id         = "Haul identifier, built by obus from the eight fields ICES uses to identify a haul",
  Survey      = "Survey code, e.g. `NS-IBTS`",
  Year        = "Year of the haul",
  Quarter     = "Quarter of the haul",
  Valid_Aphia = "WoRMS AphiaID of the accepted taxon",
  latin       = "Accepted scientific name",
  species     = "Common name",
  rank        = "Taxonomic rank of the identification (`Species`, `Genus`, `Family`, ...)",
  SpeciesValidity = "ICES `SpecVal` -- the *record type*, i.e. what data is present; not a quality flag"
)
DESC <- list(
  HL_length = c(COMMON,
    length_mm  = "Lower bound of the length bin, in millimetres",
    length_cm  = "`length_mm / 10`, carried because the two user communities count in different units",
    accuracy   = "Bin width in centimetres, from ICES `LngtCode`",
    LengthType = "ICES `LenMeasType` -- which landmark was measured (total, standard, pre-anal-fin, ...)",
    SpeciesSex = "ICES `Sex`, as reported: `F`, `M`, `U`, `B` (berried) or `NA` (never assessed)",
    DevelopmentStage = "ICES `DevStage`, as reported",
    n_haul     = "Number at this length in the whole haul: `NumberAtLength` x `SubsamplingFactor`",
    n_hour     = "`n_haul` standardised to one hour fished",
    n_measured = "`NumberAtLength` as submitted, before raising -- how many fish went through the calipers. `NA` where `DataType == \"C\"`, which reports a rate"),
  HL_summary = c(COMMON,
    n_totalnumber = "The submitted `TotalNo` -- the one per-species haul total present whether or not anything was measured",
    n_totalnumber_hour = "`n_totalnumber` standardised to one hour fished",
    n_haul     = "The same quantity reached the other way, by raising the length frequencies. The two do not always agree, and that gap is a finding",
    w_haul     = "Submitted `CatCatchWgt` for the species in the haul",
    w_hour     = "`w_haul` standardised to one hour fished",
    n_measured = "Fish actually measured, before raising. `0` for a species with no length data; `NA` where `DataType == \"C\"`",
    p_females  = "Proportion female, derived from the per-sex counts")
)
spec_table <- function(tbl) {
  x  <- dr_con(tbl) |> head(1) |> collect()
  d  <- DESC[[tbl]]
  nm <- names(x)
  data.frame(
    Column = sprintf("`%s`", nm),
    Type   = vapply(x, function(v) class(v)[1], character(1)),
    Key    = ifelse(nm %in% KEY[[tbl]], "*", ""),
    Meaning = ifelse(nm %in% names(d), d[nm],
                     "**undocumented — add it to `DESC` in this article**"),
    row.names = NULL, check.names = FALSE)
}
```

**`HL_length`** — one row per species at one length class in one haul, for
species that were actually measured. `*` marks the fields that together
identify a row.

```{r, echo = FALSE}
knitr::kable(spec_table("HL_length"))
```

**`HL_summary`** — one row per species in one haul, for *every* species
recorded, measured or not.

```{r, echo = FALSE}
knitr::kable(spec_table("HL_summary"))
```

Three things are worth saying about those keys before Part III takes them
apart. `HL_length` is identified by eight fields because ICES's own record
identity puts them there, and dropping one silently merges counts that were
separate observations — [what makes a row
unique](#what-makes-a-row-unique) measures how often. If you want the numbers
added up more coarsely, `dr_hl_collapse()` is the route to use; it carries
guards that a
hand-rolled `group_by()` loses; it is described under [what makes a row
unique](#what-makes-a-row-unique). And `SpeciesValidity` is in
both keys because it is a record type rather than a quality flag — filtering
it away discards real records. Adding across it needs one rule per column:
where a species has several record types in a haul, `HL_summary` almost always
carries the same submitted total and weight on each, so those are taken once,
while the length-derived `n_haul` is summed. `dr_add_catch()` does exactly
that.

**You never need an AphiaID.** Both tables carry `latin`, `species` and
`rank` alongside `Valid_Aphia`, so a species is selected by name:

```{r, eval = FALSE}
dr_con("HL_length") |> filter(species == "Atlantic herring")
dr_con("HL_length") |> filter(latin == "Sprattus sprattus")
```

That is not cosmetic. `Valid_Aphia == 126425` is unreadable, unverifiable on
sight, and a typo in it returns an empty table rather than an error — which is
the same failure mode as a wrong survey code, and just as quiet.

## The same calculation as CPUEL, and where the numbers differ

ICES already publishes this table. It is called **CPUEL** — catch per unit
effort at length: numbers per hour by haul, species and length class,
[@ices2021units] — and it is the natural starting point for length-based work,
computed centrally so that nobody has to raise a sub-sample again.

It covers **`r M$cpuel_surveys` of the archive's `r M$surveys` surveys**:

```{r}
cpuel <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CPUEL.parquet")
cpuel |> distinct(Survey) |> count() |> collect()

dr_con("HH") |> distinct(Survey) |> count() |> collect()
```

**`HL_length` computes the same quantity for all `r M$surveys`.** That is a
claim about definition and coverage: raised numbers at length, per haul,
species and length class, by the same arithmetic, for every survey in the
archive rather than for `r M$cpuel_surveys` of them.

Whether the two also agree *number for number* where they overlap is a
separate question, and it deserves a straight answer rather than a match rate.
Take every CPUEL record with a real catch and ask whether `HL_length` has the
same haul, species and length, and if so whether the count agrees. CPUEL
identifies a haul by six fields, not eight, so hauls whose six fields are
ambiguous are left out, as are hauls HH marks non-valid and years after the
last one CPUEL computed for each survey:

```{r}
#| code-fold: true
#| code-summary: "The comparison, all surveys at once"
cutoff <- cpuel |> filter(n_hour > 0) |> group_by(Survey) |>
  summarise(cutoff = max(Year, na.rm = TRUE), .groups = "drop")
comparable <- dr_con("HH") |>
  filter(HaulValidity == "V") |>
  inner_join(cutoff, by = "Survey") |> filter(Year <= cutoff) |>
  mutate(short_key = paste(Survey, Year, Quarter, Platform, Gear, HaulNumber, sep = ":")) |>
  group_by(short_key) |> filter(n() == 1) |> ungroup() |>
  select(Survey, .id, short_key)
ours <- dr_con("HL_length") |>
  select(.id, Valid_Aphia, length_mm, n_hour) |>
  inner_join(comparable, by = ".id") |>
  group_by(Survey, short_key, Valid_Aphia, length_mm) |>
  summarise(n_obus = sum(n_hour, na.rm = TRUE), .groups = "drop")
agreement <- cpuel |>
  filter(n_hour > 0) |>
  mutate(short_key = paste(Survey, Year, Quarter, Platform, Gear, HaulNumber, sep = ":")) |>
  semi_join(comparable, by = "short_key") |>
  group_by(Survey, short_key, Valid_Aphia = aphia, length_mm) |>
  summarise(n_cpuel = sum(n_hour, na.rm = TRUE), .groups = "drop") |>
  left_join(ours, by = c("Survey", "short_key", "Valid_Aphia", "length_mm")) |>
  group_by(Survey) |>
  summarise(cpuel_rows = n(),
            matched    = sum(as.integer(!is.na(n_obus)), na.rm = TRUE),
            agree      = sum(as.integer(!is.na(n_obus) & abs(n_obus - n_cpuel) < 0.01),
                             na.rm = TRUE),
            .groups = "drop") |>
  collect() |>
  mutate(rows_matched_pct = round(100 * matched / cpuel_rows, 1),
         counts_agree_pct = round(100 * agree / matched, 1)) |>
  arrange(desc(cpuel_rows))
agreement
```

```{r, include = FALSE}
A <- colSums(agreement[, c("cpuel_rows", "matched", "agree")])
six <- agreement |> filter(!Survey %in% c("BITS", "SWC-IBTS"))
swc <- agreement |> filter(Survey == "SWC-IBTS")
```

Across `r N(A[["cpuel_rows"]])` comparable records,
**`r P(100 * (1 - A[["matched"]] / A[["cpuel_rows"]]), 1)` of CPUEL's rows
have no `HL_length` partner and
`r P(100 * (1 - A[["agree"]] / A[["matched"]]), 1)` of its counts differ** —
`r N(A[["cpuel_rows"]] - A[["matched"]])` rows and
`r N(A[["matched"]] - A[["agree"]])` counts. For two tables that should be the
same calculation, that is not a rounding residual.

Some of it is understood. The two files are independent downloads with their
own calculation vintages, so a haul revised on one side and not the other will
differ, and the comparison cannot separate that from a real disagreement. On
BITS, CPUEL gear-standardises cod and nothing else and does not
always preserve the submitted length grid — both documented, and
both choices a generalised product would have to make deliberately rather than
inherit. Set BITS and SWC-IBTS aside and counts agree on
`r P(100 * sum(six$agree) / sum(six$matched), 1)` for the remaining
`r nrow(six)` surveys. **SWC-IBTS is unexplained, and remains so**: it agrees
on `r N(swc$agree)` of `r N(swc$matched)` matched counts.

So: the same calculation, demonstrably; the same numbers, not yet. Closing
that gap is work neither this article nor the package has finished, and a
reader deciding whether to use one table in place of the other should know
which of the two claims they are relying on. [Building CPUEL from
`HL_length`](#building-cpuel-from-hl_length) in the appendices shows what the
comparison looks like on a single survey-quarter.

**Two further differences are by design, not residual.**

- **Sex.** CPUEL carries `sex` as `NA` on every survey. `HL_length` keeps
  `SpeciesSex` wherever it was recorded, and some surveys sex much of what
  they measure — `r P(M$sexed$pct[1])` of `r M$sexed$Survey[1]`'s length
  records are female, male or berried, `r P(M$sexed$pct[2])` of
  `r M$sexed$Survey[2]`'s and `r P(M$sexed$pct[3])` of `r M$sexed$Survey[3]`'s.
- **Zero rows.** `r P(100 * M$cpuel_zero / M$cpuel_rows)` of CPUEL's rows are
  explicit zeros; `HL_length` adds none. That is not an oversight but the
  reason it generalises. A zero only means *none caught* if you know the
  species was in scope for that haul, and settling that universe is a
  survey-by-survey judgement. Leaving it out is what lets the same function
  work identically on all `r M$surveys` surveys instead of stopping at
  `r M$cpuel_surveys` — and it leaves the choice of universe with you.

None of that is a complaint about CPUEL. ICES's Data Centre builds and
maintains real, working products — CPUE by length, by age, catch weight,
abundance indices — computed centrally so that the tools that use them can
read them straight off. That is the right shape, and it is the shape this article is asking
for more of. The argument is narrower than it may sound: the length-frequency
half of that suite does not need the zero-fill question answered first, so it
can be built for every survey.

## What this saves you later

`HL_length` and `HL_summary` exist so that those four steps — raise the
measured numbers, resolve the length units, decide what to do when the factor
or the duration is missing, and avoid adding up a total the submitter repeated
— are done once, the same way for the whole archive,
following ICES's own written procedure. And so that the hauls where they
cannot be done cleanly are visible rather than quietly absorbed.

Done by hand, none of the four is hard, exactly. They are a lot of
bookkeeping to get right every time, and easy to get subtly wrong once — and a
species silently missing from a haul reads exactly like a species nobody ever
looked for. Everything that follows starts from the same table: length spectra,
abundance and biomass per haul, CPUE, stratified indices, species composition,
age–length keys. Getting the first step wrong once is getting it wrong
everywhere.

## One haul, from submission to catch at length

A single sprat record shows the whole shape of it — five length rows, exactly
as the country submitted them:

```{r}
dr_con("HL") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(LengthCode, LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber, SpeciesCategoryWeight) |>
  arrange(LengthClass) |>
  collect()
```

Four things have to happen before that is a catch at length:

- **The units have to be resolved.** `LengthCode` `.` says the `LengthClass`
  values are millimetres, so 115 means 11.5 cm and each class is 1 mm wide.
- **The numbers have to be raised.** Only part of the catch was measured. The
  11 sprat measured at 11.5 cm stand for 11 × 4.177 = 45.9 fish in the catch.
- **The sub-sample is not the catch.** The 112 in `SubsampledNumber` is how
  many fish went through the calipers, not how many came aboard.
- **The total is written once but repeated everywhere.** `TotalNumber` 467.82
  is a single figure for the whole species in this haul, and it stands on all
  five rows. Add that column up and you get the catch five times over. This is
  the problem named [above](#what-is-wrong-with-hl), in one record: a figure
  that belongs to the species has nowhere of its own to live, so it is copied
  onto every length row.

`HL_length` carries the raised numbers at length, in resolved units:

```{r}
dr_con("HL_length") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(length_cm, accuracy, n_haul, n_hour, n_measured) |>
  arrange(length_cm) |>
  collect()
```

and `HL_summary` carries one row for the species in the haul, with the
reported total beside the one rebuilt from the lengths:

```{r}
dr_con("HL_summary") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(species, n_totalnumber, n_haul, n_measured, w_haul) |>
  collect()
```

467.82 against 467.824. The reported total and the raised length frequencies
agree here to four thousandths of a fish, which is simply what happens when a
raising factor that is not a whole number is applied to whole fish. [Two
routes to a catch total](data-issues.html#two-routes-to-a-catch-total) is about the hauls
where they do not agree.


## Which table for which job

The columns are listed in [The two tables,
formally](#the-two-tables-formally). What a column list cannot show is which
table to reach for, and the one difference between them that catches people
out:

| | `HL_length` | `HL_summary` |
|---|---|---|
| Which species | only those actually measured | every species recorded, measured or not |
| Rows | `r N(M$len_rows)` | `r N(M$sum_rows)` |

| What you want to know | Where to get it |
|---|---|
| length spectra, size-based indices, inputs to age–length keys, length–weight work | `HL_length` |
| numbers or weighed biomass per haul, CPUE, species composition, stratified indices | `HL_summary`, through `dr_add_catch()`, which puts the zeros back and takes a total repeated across record types once |
| where a species occurs — distribution, occupancy, range | `HL_summary`, because it lists every species recorded and not only the measured ones |
| species richness or diversity per haul | `HL_summary`. Counting species in `HL_length` instead loses at least one species in `r P(M$sum_nolen_haul_pct)` of hauls |
| biomass of a length-measured species, **estimated** rather than weighed | `HL_length`, via `dr_add_length_mid()` then `dr_add_predicted_weight()` |
| how much of a catch was sub-sampled, anything per `SpeciesCategory`, per-category weights, or rebuilding an exchange file | `dr_con("HL")` — the submitted records, unchanged apart from `.id` |
| age, individual weight, maturity, age–length keys | `dr_con("CA")` — `{obus}` derives no CA table |

**Biomass comes two ways, and they are different quantities.** `HL_summary`'s
`w_haul` is what the catch actually weighed, as reported. For any species that
was length-measured you can also *estimate* biomass from the lengths, applying
a length–weight relationship class by class — which also gives you biomass *at
length*, which no weighed figure does:

```{r}
dr_con("HL_length") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  collect() |>
  dr_add_length_mid() |>
  dr_add_predicted_weight() |>
  summarise(fish = sum(n_haul), estimated_grams = sum(w_haul_pred))
```

It is an estimate, and it rests on coefficients from outside the submission, so
it will not match the weighed figure exactly — here it comes to about four
fifths of it. `dr_compare_length_weight()` sets the two against each other
across a whole survey.

**Which species is the difference that catches people out.** A haul's record
routinely includes species that were only counted in bulk, only weighed, or
only noted as present. Those have no `HL_length` row at all — not a zero, no
row. A species list taken from `HL_length` is therefore a list of what was
*measured*, not of what was *caught*. [Why two tables and not
one](#why-two-tables-and-not-one) measures how much that matters.

Both tables are published as parquet files and read on demand: ask for one
survey, year or species and only those rows are fetched, so you can look at a
single haul without downloading the archive.


## Why two tables and not one

The structural answer is [above](#what-is-wrong-with-hl): HL holds two
different things at once, so separating them gives two tables and not one. But that
argument would be academic if the split did not also fall along a seam the
data actually has. It does, in two places, and both come from the data rather
than from convenience.

**Part of a haul's species roster is never length-measured.** A haul's record
routinely includes species that were only counted in bulk, only weighed, or
only noted as present. `r N(M$sum_nolen)` of `HL_summary`'s
`r N(M$sum_rows)` rows — **`r sprintf("%.1f%%", M$sum_nolen_pct)`** — have no
length rows behind them at all. That is a minority of records, but it is not a
rare corner of the archive: those records fall in
`r N(M$sum_nolen_hauls)` of `r N(M$sum_hauls)` hauls,
**`r sprintf("%.0f%%", M$sum_nolen_haul_pct)`** of them, so a length table
alone would quietly drop at least one species from a third of the hauls it
covers:

```{r}
dr_con("HL_summary") |>
  anti_join(dr_con("HL_length") |> distinct(.id, Valid_Aphia, SpeciesValidity),
            by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  count() |>
  collect()
```

Those species cannot appear in a length table, and giving them a placeholder
length row would invent a measurement that was never made. `HL_length` is
therefore, by definition, only the species that were measured, while
`HL_summary` is the full list of what was caught.

**Numbers and weights survive being split by sex differently.** `HL_length`
carries `SpeciesSex` as reported, because `NumberAtLength` is a genuine
per-sex count. `SpeciesCategoryWeight` is not reliably per-sex — DATRAS
repeats one weight across a species' sex rows — so a table split by sex would
state a weight that was never weighed. Keeping the two quantities in tables of
different shape is what lets each be right. This is worked through in [what
makes a row unique](#what-makes-a-row-unique) below.

## Why these columns, and which of them could go

These two tables are a **proposal**, not a settled product, and the column set
is deliberately generous. Several columns can be worked out from others, and that
is a choice rather than an oversight — while the shape of the tables is still
open, adding a column later is disruptive and removing one is not.

Three of the redundancies are worth naming, because a reader will spot them:

- **`length_cm` is exactly `length_mm / 10`** — every row, no exceptions. It
  is carried because the two communities that would use these tables count in
  different units. The exchange format and CPUEL are
  millimetre; `{DATRAS}`, `{DATRASextra}` and the length–weight relationship
  *W = aL^b* are centimetre, as is `accuracy`. Carrying both means neither side
  converts.
- **The per-hour columns are unit conversions.** No row in either table carries
  an hourly value whose per-haul counterpart is missing — measured as zero for
  `n_hour`, `n_totalnumber_hour` and `w_hour` alike — and the reconstruction
  `n_haul / HaulDuration × 60` is exact to machine precision. They are carried
  because per hour is what CPUEL publishes and what most later work asks
  for. The alternative is to publish `HaulDuration` instead and let the user
  standardise to an hour, a half hour or a swept area — which is arguably the
  better table, and the one that stops presuming what the next analysis
  needs.
- **`latin` and `species` are one join away** from `Valid_Aphia` and
  `dr_con("species")`. They are carried so that someone can filter on a common
  name without first knowing an AphiaID.

`n_measured` is the one you could *not* work back to, because some rows add up
raw rows that were raised by different factors, so no single divisor recovers
it. It is also the one the analysis tools never read: `{DATRAS}` uses
`NumberAtLength` once, when it reads the file, to produce `Count`, and never
looks at it again. It is kept for two reasons. It is the only number left on
the rows where the raising factor is missing, and how much of a length
distribution was actually measured rather than raised is a question worth
being able to ask.

`.id` is the largest single column in the table, and it is not redundant:
five of the eight haul fields it concatenates — `Country`, `Platform`,
`Gear`, `StationName`, `HaulNumber` — appear nowhere else.

**Which of these survive is not a package author's decision** — and not any
one group's either. The survey working groups know what their data can
support; the Data Centre knows what it can build, serve and keep correct; and
the assessment and ecosystem groups are who the columns are
actually for. If ICES adopts tables of this shape, what they carry is a
decision for those three together, and the sensible direction of travel is to
trim. The cost that matters is not storage — the whole of `HL_length` is under
200 MB and no column in it is expensive — but **interface**: every column is
one more thing a user has to understand and a maintainer has to keep true. This section says what each one
duplicates and why it is there, so that conversation starts from the design
rather than from taste.

## What makes a row unique {#what-makes-a-row-unique}

ICES states the record's identity in the field descriptions, four times over —
once each under `TotalNumber`, `SubsampledNumber`, `NumberAtLength` and
`SpeciesCategoryWeight` — and all four name the same five components, the

> unique combination of cruise haul, species, sex, devstage and subsampling
> category identifier

(the phrasing varies slightly between the four cells; the components do not).
So an HL record is identified by **haul × species × sex × `DevelopmentStage` ×
`SpeciesCategory`**, and the length rows add `LengthClass` and `LengthType`.
The two `{obus}` tables follow that, with the sub-sampling category summed
away — which is safe because a category is a sampling stratum and not a
biological one: the North Sea IBTS manual [@ices2020sisp10] has one created
when a catch holds two or more distinct length groups so that each can be
sub-sampled at its own rate, so the categories partition a single catch and
adding them back returns it.

| Table | One row per |
|---|---|
| `HL_length` | `.id` × `Valid_Aphia` × `length_mm` × `length_cm` × `accuracy` × `LengthType` × `SpeciesSex` × `DevelopmentStage` × `SpeciesValidity` |
| `HL_summary` | `.id` × `Valid_Aphia` × `SpeciesValidity` |

Each is exactly unique there — no repeated combinations at all, over
`r N(M$len_rows)` rows — and at anything coarser it is not, so **add `n_haul` up rather than
assuming one row**:

```{r}
dr_con("HL_length") |>
  count(.id, Valid_Aphia, length_mm, length_cm, accuracy,
        LengthType, SpeciesSex, DevelopmentStage, SpeciesValidity) |>
  filter(n > 1) |>
  count() |>
  collect()
```

`LengthType` and `DevelopmentStage` are part of what makes a row unique, not
labels carried along, because ICES's record identity puts them there: one haul
can measure the same species at the same length under two measurement
conventions, or as both egg-bearing and unstaged, and each is its own count.
Over the `r N(M$coarse$n)` combinations at the coarser `.id` × `Valid_Aphia` ×
`length_mm` × `SpeciesSex` level, `r N(M$coarse$ds)` split on
`DevelopmentStage`, `r N(M$coarse$lt)` on `LengthType`, `r N(M$coarse$sv)` on
`SpeciesValidity` and `r N(M$coarse$acc)` on `accuracy`. Rare — but each split
is a separate count, and collapsing one is silent. `DevelopmentStage` is the
easiest to overlook, being `NA` for `r P(M$ds_na_pct, 1)` of rows, and
dropping it would add a berried record to an unstaged one.

**If you want the numbers added up more coarsely, ask for that — do not do it by hand.**
Everyone else's table is one of these with something added away, so wanting
that is entirely normal. `dr_hl_collapse()` does it, whether the table is
still on disk or already pulled into memory:

```{r, eval = FALSE}
dr_con("HL_length") |>
  dr_hl_collapse(c("SpeciesSex", "DevelopmentStage", "SpeciesValidity"))
```

It exists for what a `group_by()` silently loses rather than for the grouping
itself: the all-`NA` case, where `sum(x, na.rm = TRUE)` over a group with
nothing in it is `0` in R and `NULL` in SQL, so an unknown raising comes back
as a confident zero; and the `DataType == "C"` rule, which keeps a rate from
being reported as a count. It also refuses what cannot be summed —
`accuracy` and `LengthType` where a group actually mixes them, since two
records at one `length_mm` measured to 1 cm and to 5 cm are different bins.
`{DATRAS}`'s `addSpectrum()` collapses a mixed `LengthCode` to the coarsest with
only a warning; this errors and names the group. It does not reproduce
`HL_summary`: that table's totals come from the submitted `TotalNumber`, which
is a different quantity, as [two routes to a catch
total](data-issues.html#two-routes-to-a-catch-total) sets out.

That choice has one measurable cost, in [duplicated length
series](data-issues.html#duplicated-length-series).

## Why `HL_summary` is not split by sex

`HL_length` carries `SpeciesSex`; `HL_summary` does not. That difference is
deliberate, and it comes from what the submissions contain rather than from any
limitation of the package.

`NumberAtLength` is a genuine per-sex, per-length count, and — checked over the
whole archive — never repeats one placeholder value across a species' sex rows.
`TotalNumber` does. Where a haul's record splits a species by sex,
`r P(M$sex_per_sex_pct)` of the time the total genuinely varies from sex to
sex, and `r P(M$sex_repeated_pct)` of the time one species-level figure has
simply been written onto every sex row. Per-sex arithmetic can tell those two apart, and `{obus}`
does exactly that — the measurement, and what the package does about it, are in
[one `TotalNumber` written across several sex
rows](data-issues.html#totalnumber-across-sex-rows).

`SpeciesCategoryWeight` has no comparable check, and its
repeats-versus-varies pattern tracks `TotalNumber`'s only about half the time.
Splitting `HL_summary` by sex would therefore misstate `w_haul` for an unknown
share of species with no way to flag which. Sex-specific *numbers* stay
available by adding up `HL_length`; sex-specific *weight* is not recoverable in
general. That is a property of the source data, not a gap in `{obus}`.

Sex codes are kept as reported — `"F"`, `"M"`, `"U"`, `"B"` or `NA`. `"B"`
(berried, egg-bearing) is not folded into `"F"`, and `"U"` (examined,
undetermined) stays distinct from `NA` (never examined). The female proportion
`HL_summary` publishes as `p_females` is one step away from `HL_length`:

```{r, eval = FALSE}
dr_con("HL_length") |>
  group_by(.id, Valid_Aphia) |>
  summarise(
    p_females = sum(n_haul[SpeciesSex %in% c("F", "B")]) /
                sum(n_haul[SpeciesSex %in% c("F", "M", "B")]),
    .groups = "drop"
  )
```

## Do the derived tables still say what the submissions said?

Yes, and this is checked rather than argued. `HL_length`
has been compared cell by cell against the `{DATRAS}` package's own `Count`
[@kristensen2023datraspkg], over every species and survey in the archive.
**No cell differed by as much as half a fish**, the largest difference anywhere
being floating-point dust, and no row appeared in `HL_length` that was not in
the submitted record. The field-by-field harness,
[`data-raw/CHECK_datras_interop.R`](https://github.com/einarhjorleifsson/obus/blob/main/data-raw/CHECK_datras_interop.R),
repeats the comparison on `{DATRASextra}`'s NS-IBTS dab fixture; it needs the
`{DATRAS}` package, so it is not re-run when this page is built (last run
2026-10-06: 25 checks, no differences).

The differences run one way only. A few rows are in the submitted record and
not in `HL_length`, and they are all the same case: the raising factor is
missing, so there is no honest number to publish and `{obus}` returns `NA`
rather than assuming the catch was fully measured.

```{r}
dr_con("HL_length") |>
  summarise(rows = n(), no_raised_number = sum(as.integer(is.na(n_haul)))) |>
  collect()
```

`{DATRAS}` assumes 1 in exactly these rows and so reports a number; `{obus}`
does not, because ICES makes the field mandatory. That is a choice, argued in
[a mandatory `SubsamplingFactor` that is
absent](data-issues.html#subsamplingfactor-absent), not an error.

The tables are faithful, but they are **not lossless**. Nearly every row in
`HL_length` comes from exactly one submitted HL row. About one row in
`r round(M$combined_one_in, -1)` combines several, and where that happens the raising factor and the
sub-sampling category are added away and cannot be recovered from the
published table. Also gone at this level: `SubsampledNumber`,
`SubsampleWeight`, per-category `SpeciesCategoryWeight` (`w_haul` is added up
to the species) and any weight split by sex.

## The haul identifier, `.id`

`.id` is an `{obus}` construct; ICES has no such field. It is the eight fields
ICES's own exchange tables use to key a haul, concatenated with `:` —

`Survey : Year : Quarter : Country : Platform : Gear : StationName : HaulNumber`

— for example `"BITS:1991:1:DE:06S1:H20:33:28"`. It always has eight fields:
a missing one (most often `StationName`) is written as the literal string
`"NA"`, so `.id` splits back into its eight columns, two hauls cannot collide
by having different fields missing, and the result is byte-identical to the
`{DATRAS}` package's own `haul.id`. The token is `"NA"` and not `-9`, because
the archive stores the field as missing and writing `-9` back would invent a
value it does not contain. No `StationName` in the archive is itself `"NA"`,
so the token is unambiguous. When every field is `NA` the result is `NA`.

`dr_add_id()` accepts only the current names. A legacy-named table must be
renamed first (`opus::op_rename()`); `Ship`, `StNo` and `HaulNo` are not
handled.

# Part III. Where the data is ambiguous

Part II is arithmetic the format specifies. What remains is a small set of
constructs a single record cannot settle on its own: places where two
submissions write the same shape and mean different things, or write a value
the format does not allow. They touch `r P(M$flagged_pct, 1)` of
`HL_summary`'s records, and each has a rule in the code that builds the two
tables. They are written up separately, one entry each, with a real record, its
reach, what the tables do with it, and the question that would settle it, in
[DATRAS data issues](data-issues.html), because the people who can act on them
are the institutes that submit and the ICES Data Centre rather than the
analysts who use the tables.

What belongs here are the conventions that are not ambiguous at all, but that a
consumer reading HL flat will get wrong.

## Conventions the code resolves once

These are not ambiguous. They are settled in the format, and are listed only
because a consumer reading HL flat will get them wrong.

| Convention | What a flat read does | What `{obus}` does |
|---|---|---|
| `LengthClass` holds millimetres and centimetres in one column, per `LengthCode` | a tenfold error on a length | resolved on reading into `length_mm`, `length_cm` and `accuracy` |
| `LengthClass` is a bin **lower bound**, not a midpoint | biases any weight predicted from it, since \eqn{W = aL^b} is convex | carried as the bound it is; `dr_add_length_mid()` where a midpoint is wanted |
| `-9` marks an absent value | a sentinel enters the arithmetic | nulled once, upstream, per field; never written back |
| One species can appear under several `SpeciesValidity` record types, usually with the same total and weight repeated on each | the same fish counted twice | `SpeciesValidity` is part of what makes an `HL_summary` row unique; `dr_add_catch()` takes a repeated total once and sums `n_haul` |
| `DataType` `C` reports `NumberAtLength` as an hourly rate | a rate summed as a count | `n_measured` is `NA` under `C`, never `0` |
| Invalid hauls are present unless excluded | silent inclusion | `haulval` has no default — the survey's call, not the package's |

# Appendices

Reference and evidence material, moved out of the reading path. Nothing
here is required to follow the argument; all of it is cited from the text
where it matters.


## Building CPUEL from `HL_length` {#building-cpuel-from-hl_length}

The clearest way to see that these are the same table is to build one from the
other. CPUEL's payload is numbers per hour by haul, species and length class —
`.id`, `aphia`, `length_mm`, `n_hour` — with `sex` empty throughout.
`HL_length` is finer than that in three ways it keeps on purpose (sex,
development stage and length measurement type), so CPUEL is one aggregation
away:

```{r}
cpuel_like <- dr_con("HL_length") |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1) |>
  group_by(.id, Valid_Aphia, length_mm) |>
  summarise(n_hour = sum(n_hour, na.rm = TRUE), .groups = "drop")
```

Set that against ICES's own file for the same hauls:

```{r}
cpuel_q <- cpuel |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1, n_hour > 0) |>
  select(.id, Valid_Aphia = aphia, length_mm, cpuel = n_hour)
cpuel_cells <- cpuel_q |> count() |> pull(n)

cmp <- cpuel_like |>
  inner_join(cpuel_q, by = c(".id", "Valid_Aphia", "length_mm")) |>
  summarise(hauls = n_distinct(.id),
            cells = n(),
            agree = sum(ifelse(abs(n_hour - cpuel) < 0.001, 1, 0))) |>
  collect()
cmp
```

Over the `r cmp$hauls` hauls both sides cover, the rebuild matches
**`r P(100 * cmp$cells / cpuel_cells, 1)` of CPUEL's non-zero cells** and
agrees with **`r P(100 * cmp$agree / cmp$cells, 1)`** of those to a
thousandth of a fish per hour. A worked haul, three cod:

```{r}
cpuel_like |>
  inner_join(cpuel_q, by = c(".id", "Valid_Aphia", "length_mm")) |>
  filter(.id == "NS-IBTS:2023:1:DE:06NI:GOV:104:22", Valid_Aphia == 126436) |>
  arrange(length_mm) |>
  collect()
```

The causes of the residual are named, though not quantified: nothing in it has
been traced to the arithmetic. The two files are independent downloads with
their own calculation vintages, so a haul revised on one side and not the
other differs, and this comparison cannot tell that apart from a genuine
disagreement. On BITS two further, documented things apply: CPUEL gear-standardises cod and nothing else and does not
always preserve the submitted length grid — both of which a
generalised product would have to decide about explicitly rather than inherit.

**The zero rows are the part that is not one aggregation away**, and
deliberately so. `r P(100 * M$cpuel_zero / M$cpuel_rows)` of CPUEL is
explicit zeros — `r N(M$cpuel_zero)` of `r N(M$cpuel_rows)` rows — and they
encode a judgement about which species were in scope for a haul. Add them
yourself and the judgement stays yours:

```{r}
hauls   <- cpuel_like |> distinct(.id) |> collect()
species <- cpuel_like |> distinct(Valid_Aphia) |> collect()    # your universe, not ours

filled <- tidyr::crossing(hauls, species) |>
  left_join(collect(cpuel_like), by = c(".id", "Valid_Aphia")) |>
  tidyr::replace_na(list(n_hour = 0, length_mm = 0))
filled
```

On this one survey-quarter that turns `r nrow(hauls)` hauls and
`r nrow(species)` species into `r N(nrow(filled))` rows,
`r N(sum(filled$n_hour == 0))` of them explicit zeros — a proportion like
CPUEL's, arrived at from a universe you chose rather than one you inherited.

That is the whole distance between `HL_length` and a CPUEL product covering
`r M$surveys` surveys rather than `r M$cpuel_surveys`: one `group_by()`, plus
a species universe that only a survey group can properly set.

## Existing analyses run on them unchanged

A product also has to leave standing what is already built on the exchange
tables. So: starting from these two tables instead of the raw HL record, can
the usual chain still be run — numbers at length raised to the haul, summed to
total numbers and weight per haul, divided by swept area and carried through to
a stratified abundance index?

[`data-raw/CHECK_datras_adapter.R`](https://github.com/einarhjorleifsson/obus/blob/main/data-raw/CHECK_datras_adapter.R)
answers that against real numbers. It
assembles an object whose HL comes **only** from `HL_length` and `HL_summary`,
never from raw HL, runs [`{DATRAS}`](https://github.com/DTUAqua/DATRAS) and
`{DATRASextra}` on it unmodified, and compares with the same steps run on their
own bundled fixture — NS-IBTS 2020–2023, quarters 1 and 3, dab (*Limanda
limanda*), 2,651 valid hauls. The adapter is about 60 lines: rename to the
legacy field names, add the five HH quantities the packages derive for
themselves, and reshape the two catch tables into the single HL frame they
expect. **No arithmetic on any number** — the raised number at length is
`n_haul` exactly as published.

| Step | Quantity | Difference |
|---|---|---|
| numbers at length, by length class | `N`, 41 classes | same classes, **max gap 0** over 108,691 cells |
| total numbers per haul | `HaulN` | **max gap 0** over 2,651 hauls |
| total weight per haul, from the length–weight lookup | `HaulWgt` | **max gap 0** |
| swept area | `SweptArea` | **max gap 0** |
| stratified abundance index, per hour, by statistical rectangle | index and CV | **identical** |

Thirteen checks, no differences (last run 2026-10-06). The stratified index is the
figure a survey working group would report, and it comes out the same to every
digit:

```
  Year    index         cv
  2020 771.7314 0.07114170
  2021 712.8609 0.09926699
  2022 823.4249 0.06559173
  2023 739.3506 0.05788805
```

Two caveats. Species recorded in a haul but never measured enter through
`HL_summary`, carrying no length — which is how the exchange format carries
them, and what a species-richness count needs; a length table alone would lose
them. And **CA is out of scope**: `{obus}` derives no age table, so nothing
age-, maturity- or individual-weight-based is exercised.

Underneath sits the field-by-field comparison,
[`data-raw/CHECK_datras_interop.R`](https://github.com/einarhjorleifsson/obus/blob/main/data-raw/CHECK_datras_interop.R)
— 25 checks, no differences (last run 2026-10-06), including the
raised number at length (max gap 2.3 × 10⁻¹³ over 24,960 cells) and the
measured number at length, `NumberAtLength`, against `n_measured` (max gap 0 over
16,196 cells).


## Field names: two layers, both ICES's {#two-naming-layers}

Field names in this codebase come from two places — **both of them ICES's** —
and there is deliberately no third.

| Layer | What it is | Example |
|---|---|---|
| **1. ICES legacy exchange names** | The names the exchange file still carries, and the ones the older ICES documents quoted here use. | `HLNoAtLngt`, `SubFactor`, `CatIdentifier` |
| **2. ICES current names** | ICES's replacement names, published in the current field descriptions. What the raw parquet archive carries, and what every `{obus}` table carries too. | `NumberAtLength`, `SubsamplingFactor`, `SpeciesCategory` |

**The translation between the two layers is ICES's own, not this package's.**
ICES is renaming DATRAS's fields table by table, and publishes the
old-to-new pairs itself in two places: the `getDatrasFieldList` operation on
the DATRAS web service, which returns each current `FieldName` alongside the
`FieldNameOld` it was renamed from, and the *DATRAS New Headers Lookup*
spreadsheet of December 2024 [@ices2024newheaders], which carries the same
pairs in one sheet per record type. Layer 2 is therefore not a scheme `{opus}`
invented and `{obus}` adopted — it is ICES's own target naming, taken up ahead of
the migration's completion. The migration's progress is visible in the service:
HL has flipped, and `getHLdataNewHeaders` returns all 29 of its fields under
the current names, while HH, CA and LT are still served under legacy names
only.

The documentation has already moved ahead of the service: the current field
descriptions name every field in layer 2, so a reader working from ICES's
workbook and a reader working from the `{obus}` tables see the same column names.
Legacy names survive in four places:

- the exchange files themselves;
- the web-service operations that have not yet flipped;
- the older ICES documents quoted later in this article;
- and, in a handful of cells, the current descriptions' own prose — where
  formulas still read `TotalNo=SUM(HLNoAtLngt)` and *"If DataType is C,
  CatIdentifier is always 1"* beneath fields now called `TotalNumber` and
  `SpeciesCategory`.

Quotations below keep whatever names the quoted document uses.

**`{obus}` invents no field names.** `dr_con_raw("HL")` and `dr_con("HL_length")`
both carry `Valid_Aphia` and `SpeciesSex`, so a join between the raw archive and
a derived table needs no name mapping, and the crosswalk below describes every
column in every table.

A shorter `sex` column is a tempting simplification, and it is the wrong one.
ICES's legacy `Sex` maps to `SpeciesSex` in HL but to `IndividualSex` in CA —
two genuinely different measurements, one a property of a length-frequency
record and the other of an individual fish. A single `sex` column flattens
them.

### The legacy-to-current crosswalk

The mapping is neither hand-authored here nor taken from ICES's published
lists on trust. `{opus}` checks every pair ICES publishes against the field names
the corresponding operation actually returns, keeps what survives
(`opus::op_datras_field_list()`), reduces that to a collision-checked 1:1 map
(`opus::op_datras_rename_crosswalk()`), and embeds the result in the raw
parquet footer. `{obus}` reads it back from the same file as the data, so it
cannot describe a different vintage than the data it accompanies:

```{r}
xw <- opus::op_crosswalk("HL")
xw
```

`r nrow(xw)` rows — two more than either ICES list. Neither `getDatrasFieldList` nor the
December 2024 lookup documents HL's `DateofCalculation` or `Valid_Aphia` under
any name, though both are real fields that `getHLdata` and
`getHLdataNewHeaders` alike return; they appear here unrenamed, which is what
those two operations show them to be.

Check names against this output rather than from memory. Most pairs
are guessable; at least one is not — **`LenMeasType` maps to `LengthType`**,
which is easy to mistake for a length *unit* field when it is a measurement
*convention* field. `Sex` → `SpeciesSex` is the pair worth memorising, because ICES's one legacy
`Sex` splits across two current names — `SpeciesSex` in HL and
`IndividualSex` in CA — so it is the clearest case of why the crosswalk is
per table rather than global.


## An engineering note: R and SQL disagree about `NA`

Not a DATRAS matter, but the single largest source of silent defects in this
code. `{obus}` runs the same functions eagerly (data frames, in the tests) and
lazily (DuckDB, in the build), and the two backends differ in four ways.
Anyone extending these functions needs all four.

**1. Joining.** SQL treats `NULL = NULL` as unknown, so a lazy join drops
every row with an `NA` key — and `SpeciesSex` is `NA` in `r P(M$sex_na_pct)` of HL, with
`SpeciesCategory`, `SpeciesValidity` and `TotalNumber` able to be `NA` too.
Every join in `dr_hl_summary()` whose key can hold `NA` carries
`na_matches = "na"`, which dbplyr renders as `IS NOT DISTINCT FROM` and DuckDB
supports natively. Without it, every `NA`-sex row fails to find its own
length-derived expectation, never reconciles, falls to the fallback, and
undercounts. Confirmed on a real case
(`BITS:1999:1:DK:26HI:TVS:017492:22`, aphia 127143): true total 4, and 3
without the flag.

**2. Aggregating over an all-`NA` group.** `sum(x, na.rm = TRUE)` is `0` in R
and `NULL` in SQL. Every aggregation here carries an explicit non-`NA` counter
(`.n_ok`, `.h_ok`, `.w_ok`, `.wh_ok`) and forces `NA` when the counter is
zero, so both backends agree. Without it the eager path silently converts a
deliberate `NA` — a missing `SubsamplingFactor`, a non-positive
`HaulDuration`, a species that was never weighed — straight back into the
confident `0` it was meant to prevent.

**3. Self-reference inside `summarise()`.** dplyr evaluates expressions in
order, so `n_ok = sum(!is.na(n_haul))` written *after* `n_haul = sum(n_haul)`
sees the new scalar; SQL resolves every aggregate against the source column.
`dr_hl_length()` renames to `.n_raw`/`.h_raw` before aggregating, so the two
agree by construction rather than by luck.

**4. String concatenation with `NA`.** On a plain data frame R's `paste(sep =)`
renders an `NA` field as the literal two-character string `"NA"`; on a
`tbl_lazy`, dbplyr translates the same call to DuckDB's `CONCAT_WS()`, which
drops `NA` fields instead. `.id` is the join key between HH, HL and CA, so it
must come out identical whichever backend built it. `.dr_concat_ws()`
substitutes the token `"NA"` for a missing field before concatenating, using
`is.na()`, `case_when()` and `paste0()`, which translate the same way on both,
so every `.id` keeps its eight fields on either backend. When every field is
`NA` it returns `NA` rather than a string of separators.

Each of these produced a real, silent, wrong number at some point. Keeping
both an eager test suite and a lazy build is what catches them; neither does
alone.


## How this article was made {#how-this-article-was-made}

**What counts as evidence.** Every claim is either a quotation from an ICES
document or a measurement of the archive, and the measurements are given as
code that reads the published tables directly, so they need nothing but the
package.

**A note on the numbers, because DATRAS grows.** Every number in the text is
**measured when this article is rendered**, not typed in, so it cannot drift
from the code printed beside it or from the archive. The one exception is the
pair of `{DATRAS}` harnesses, which need packages this page does not load; their
results carry the date of their last run. Claims that should survive the
archive growing are written in that form — *only*, *exactly one*, *never* — so
that if one ever stops being true, that is a finding rather than a stale
number.

**Use of AI.** The text and code of this article were written with an AI
assistant (Claude, by Anthropic) under the direction of the author, who is
responsible for its content. Every number in it is computed by the code shown,
from the published archive.

**On the package.** [`{obus}`](https://github.com/einarhjorleifsson/obus) is a
reference implementation of the proposal and a demonstration that it holds for
the whole archive rather than part of it. Every rule it applies is quoted from
an ICES document; every figure is reproducible from the published archive with
the code shown. The nearest thing to a proof of the proposal is that [CPUEL
itself can be rebuilt from `HL_length`](#building-cpuel-from-hl_length),
worked through above.

## References {.unnumbered}

::: {#refs}
:::
