---
title: "Two catch tables for DATRAS"
subtitle: "`HL_length` and `HL_summary` — a proposal, what they reproduce, and the conventions that shape them."
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
    citations-hover: true
    reference-location: document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE,
                      message = FALSE, warning = FALSE)
options(width = 120)
# The argument in this article is arithmetic, so the default three significant
# figures actively hide it: a raising factor of 1.4466 prints as 1.45, and
# 1.45 x 21 does not give the reported 30. Five digits put the numbers the
# prose quotes on the screen the reader is looking at. Seven, not five, because
# HL_summary's two routes differ at the fourth decimal (467.82 vs 467.824)
# and that difference is the point of the comparison.
options(pillar.sigfig = 7)
```

This article is about two tables — `HL_length` and `HL_summary` — that
`{obus}` derives from the DATRAS HL record, and about why the record makes
them necessary. It describes what the package does now; it does not propose
changes to it.

**What it argues for is the tables, not the package.** They are the
calculation nearly every DATRAS analysis performs before it does anything else,
and the right home for them is ICES's own product suite, next to CPUEL —
computed once, centrally, and served to everyone, rather than re-implemented
once per institute, once per language and once per working group.

This is not a proposal in the abstract. Built from these two tables and nothing
else, `{obus}` rebuilds ICES's own CPUEL product and agrees with it on the
overwhelming majority of cells; and the `{DATRAS}` and `{DATRASextra}` packages
run on them **unmodified**, reproducing numbers at length, totals per haul,
swept area and a stratified abundance index with no difference at all — the
index a working group would report comes out the same to every digit. Both
demonstrations are below, with the places they fall short named rather than
glossed. If ICES served these two tables, [`{obus}`](https://github.com/einarhjorleifsson/obus)
would not need to exist, and that would be the better outcome.

```{r}
library(obus)
library(dplyr)
library(ggplot2)
```

```{r, include = FALSE}
# Structural counts, measured once here and referenced inline throughout, so
# they cannot drift apart as the archive grows (Working Principle 8). Before
# this block they were typed into the prose, and HL_summary's group count had
# already been written down at two different vintages in two different files.
# Analytical findings below -- reconciliation buckets, confidence intervals,
# rounding regimes -- are NOT computed here: they are dated evidence, and
# re-running them on every render would embed whole analyses in the article.
M <- list(
  len_rows   = dr_con("HL_length")  |> count() |> pull(n),
  sum_rows   = dr_con("HL_summary") |> count() |> pull(n),
  sum_groups = dr_con("HL_summary") |> distinct(.id, Valid_Aphia) |>
                 count() |> pull(n),
  # HL_summary rows with no HL_length rows behind them. Stated in the prose
  # AND computed by the displayed chunk just below it, so a drift here would
  # show up as the article contradicting itself on the same screen.
  sum_nolen  = dr_con("HL_summary") |>
                 anti_join(dr_con("HL_length") |>
                             distinct(.id, Valid_Aphia, SpeciesValidity),
                           by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
                 count() |> pull(n)
)
# Hauls touched by that, so the prose can say whether it is concentrated or
# spread. The claim it supports is "a third of hauls", which will drift.
M$sum_hauls <- dr_con("HL_summary") |> distinct(.id) |> count() |> pull(n)
M$sum_nolen_hauls <- dr_con("HL_summary") |>
  anti_join(dr_con("HL_length") |>
              distinct(.id, Valid_Aphia, SpeciesValidity),
            by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  distinct(.id) |> count() |> pull(n)
M$sum_nolen_pct <- 100 * M$sum_nolen / M$sum_rows
M$sum_nolen_haul_pct <- 100 * M$sum_nolen_hauls / M$sum_hauls
# Thousands separators, so the prose reads the way it always has.
N <- function(x) formatC(as.numeric(x), format = "d", big.mark = ",")
```

Every claim below is either a quotation from an ICES document or a measurement
of the archive, given as code you can run. [How this article was
made](#how-this-article-was-made), in the appendices, says what counts as
evidence and how the figures are kept from drifting as the archive grows.

# Part I. What is wrong with the HL table

## A good record, and a poor table

HL does the job it was designed for — capturing, faithfully and completely,
what a country submitted for a haul. That is not the same job as being the
table you analyse from, and this part is about the difference. Nothing
below is an argument that the format is broken; it is an argument that the
record and the analysis want different shapes, and that the conversion between
them is being done separately by everyone who needs it.

What a country sends ICES for one haul is a set of length rows: a length
class, a number, a raising factor relating a measured sub-sample to the whole
catch, and — repeated on every row — the totals for that species category in
number and weight.

**What that number means depends on how the country chose to submit.** Under
most conventions it is fish *measured*, and the raising factor is what relates
it to fish caught. Under `DataType` C it is neither: the submitter has already
raised it *and* already scaled it to one hour of towing, so the row carries a
catch rate and the factor is 1. Note which way that scaling goes: nearly nine
`C` hauls in ten are shorter than an hour, most of them half-hour tows, so the
number written on the row is usually **larger** than the number of fish
actually caught, not smaller. This is not a corner case. `C` covers about a
third of all the hauls in the archive, and about a third of every length row
in it.

Given the `DataType`, any one row is unambiguous. **It is the column that is
ambiguous**, across submissions — and the `DataType` is not in HL at all. It
is in the haul record, which is why no amount of reading an HL row tells you
what its numbers are.

Getting from a submission to a catch is the same short list of steps for every
survey in the archive:

- raise the measured numbers by the sub-sampling factor — except where the
  convention has already done it;
- turn the length class into a length, in known units;
- decide what to do when the raising factor is missing, or the haul lasted no
  time at all;
- avoid adding up a total the submitter repeated across sex or category rows.

Nearly every DATRAS analysis begins there, and they are the same steps every
time.

## Three tallies in one table {#what-is-wrong-with-hl}

The steps above are bookkeeping. The reason they are *easy to get wrong* is
structural, and it is worth naming before the tables that fix it.

**HL is one table doing three jobs.** Its columns are three different tallies
of three different things:

| Tally | Columns | What it counts |
|---|---|---|
| **by length** | `LengthClass`, `NumberAtLength` | fish in each length class |
| **of the sub-sample** | `SubsampledNumber`, `SubsamplingFactor`, `SubsampleWeight` | what was examined, and how it scales to the catch |
| **of the catch** | `TotalNumber`, `SpeciesCategoryWeight` | what came aboard |

Only the first has a row of its own. The other two are one value each, and
because the file offers only one kind of row they are **written onto
every length row of that category**. [One haul, from submission to catch at
length](#one-haul-from-submission-to-catch-at-length) below shows exactly this
in a sprat record.

That is not a DATRAS eccentricity. It is the textbook case of what Wickham
[@wickham2014tidy] describes as storing *multiple observational units* — several
different things being counted — in one table, and the consequence is exactly
the one the format's users keep meeting: **a column you must not add up.** Add `TotalNumber` down a
species' length rows and you count the same catch once for every length class
it happened to be measured in. Every repeated-value trap in Part III comes
back to this one thing.

**The three tallies are not merely stacked — they can come apart.** Under
pseudo-category sampling the sub-sample really is finer than the catch: the
raising factor belongs to the sub-category, while the total and the weight
belong to the main category above it. Take every `P` main category that was
split into more than one sub-category, and ask which of the three fields move
together:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "P", !is.na(SpeciesCategory)) |>
  mutate(main_category = substr(SpeciesCategory, 1, 1)) |>
  distinct(.id, Valid_Aphia, SpeciesValidity, main_category, SpeciesCategory,
           SubsamplingFactor, TotalNumber, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, main_category) |>
  summarise(sub_categories = n_distinct(SpeciesCategory),
            factors = n_distinct(SubsamplingFactor),
            totals  = n_distinct(TotalNumber),
            weights = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(sub_categories > 1) |>
  summarise(main_categories = n(),
            factor_varies   = sum(as.integer(factors > 1)),
            total_constant  = sum(as.integer(totals == 1)),
            weight_constant = sum(as.integer(weights == 1))) |>
  collect()
```

In almost every one of them the raising factor changes from sub-category to
sub-category while the total and the weight stay put. Three levels of
description, and one kind of row to write them all on.

That is why ICES's own sum `TotalNumber = SubsampledNumber ×
SubsamplingFactor` cannot hold under `P`: its three terms do not all describe
the same thing. [Pseudo category
sampling](#totalnumber-across-categories) therefore needs treating on its own
rather than in a sentence.

Two further properties compound it:

- **`LengthClass` is a number whose unit *and resolution* are in another
  column.** `LengthCode` says whether the value is millimetres or centimetres,
  so the number alone does not say what it measures — but it also says how wide
  the length class is, and that is the harder half.
- **`NumberAtLength` is a quantity whose identity is in another *table*.** As
  above, it is a measured count or an hourly rate depending on `DataType`,
  which lives in HH. You cannot tell from the HL row in front of you.

**The resolution is the harder half because the unit does not imply it.** Two
codes report millimetres and two of the three centimetre codes are not 1 cm:
`.` is 1 mm classes and `0` is 5 mm; `1`, `2` and `5` are 1, 2 and 5 cm. So a
`LengthClass` of 150 is the same 15 cm fish under `.` and under `0`, but in one
case it stands for a class 1 mm wide and in the other 5 mm. Across the archive
the classes run from 1 mm to 5 cm — a fiftyfold range — and `LengthClass` is
the class's **lower edge**, not the fish, so without the code you know neither
where the class starts nor how far it extends.

**Most of that variation is good practice, not sloppiness.** Resolution is
chosen to suit the fish: within one country, platform and year, the code is
constant for a given species in 96% of cases — and the choices are the ones a
biologist would make, herring and sprat measured in 5 mm classes, cod, plaice,
haddock and whiting in 1 cm.

```{r}
dr_con("HL") |>
  filter(!is.na(LengthCode), LengthCode != "-9") |>
  group_by(Country, Platform, Year, Valid_Aphia) |>
  summarise(codes = n_distinct(LengthCode), .groups = "drop") |>
  summarise(platform_year_species = n(),
            one_code_throughout   = sum(as.integer(codes == 1)),
            more_than_one         = sum(as.integer(codes > 1))) |>
  collect()
```

**What bites is that the choice is the submitter's, not the species'.** The
same fish is measured at different resolutions by different programmes:

```{r}
dr_con("HL") |>
  filter(Valid_Aphia %in% c(126436, 126417), !is.na(LengthCode), LengthCode != "-9") |>
  left_join(dr_con("species") |> select(Valid_Aphia, species), by = "Valid_Aphia") |>
  count(species, LengthCode) |>
  mutate(class_width = case_when(LengthCode == "." ~ "1 mm", LengthCode == "0" ~ "5 mm",
                                 LengthCode == "1" ~ "1 cm", LengthCode == "2" ~ "2 cm",
                                 LengthCode == "5" ~ "5 cm")) |>
  arrange(species, desc(n)) |>
  collect()
```

Cod arrives in all four widths — mostly 1 cm, but 122,188 length rows at 1 mm
and a few thousand at 5 cm, a fiftyfold spread for one species. Herring is
mostly 5 mm and substantially 1 mm. So the moment you pool a species across
surveys — which is what a stratified index does — you are pooling classes of
different widths.

Why that matters is arithmetic, not bookkeeping. A count in a 5 cm class and a
count in a 1 mm class are not the same measurement, and no unit conversion makes
them so: at equal abundance the wider class simply collects more fish. Add them
together and the result is not a length distribution that is wrong by some
correctable factor — it is not a length distribution. `{obus}` therefore
resolves `LengthCode` once, on reading, into `length_mm`, `length_cm` and
`accuracy`, so the class width travels with every row instead of being
recoverable only by going back to HL.

**Why this is easy to miss.** The conventions vary far more between surveys
than they do inside any one of them. Most surveys have submitted under a
single convention for their whole history, and it is always either `R`
throughout or `C` throughout:

```{r}
dr_con("HH") |>
  filter(DataType != "-9") |>
  distinct(Survey, DataType) |>
  group_by(Survey) |>
  summarise(conventions = n(),
            submits = ifelse(n() == 1, min(DataType, na.rm = TRUE), "mixed"),
            .groups = "drop") |>
  count(conventions, submits) |>
  collect()
```

Neither `S` nor `P` ever defines a survey on its own: every `S` haul in the
archive sits inside a survey that otherwise submits `R`, and `P` appears only
in surveys that also use `R` and `C`. So the two conventions that stand alone
are exactly the two that disagree about what `NumberAtLength` counts.

The practical effect is that within one survey `NumberAtLength` usually means
one stable thing, and nothing in the record warns you that it means something
else in the next survey along. The column is clear locally and ambiguous only
archive-wide, which is the kind of problem that survives a long time, because
every individual dataset is self-consistent. The mixed surveys are where it
becomes visible without leaving home — NS-IBTS carries `R`, `C` and `P`
together, BITS `R` and `C`. Even there it tends to nest: about half the
national programmes inside those two surveys have only ever used one
convention, so a single country's series can still look perfectly uniform.

**And no one who submits ever uses the whole format.** The archive as a whole
draws on five `DataType` conventions, five `LengthCode` values and fourteen
`SpeciesCategory` codes. The unit that actually samples and uploads is nearer a
platform than a country — `Platform` stands in for the institutes within a
country, and Germany runs nineteen of them, the Netherlands thirteen, Norway
twelve — so count what each country and platform between them ever uses:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  group_by(Country, Platform) |>
  summarise(rows           = n(),
            data_types     = n_distinct(DataType),
            length_codes   = n_distinct(LengthCode),
            category_codes = n_distinct(SpeciesCategory), .groups = "drop") |>
  filter(rows >= 5000) |>
  summarise(platforms                  = n(),
            recording_one_way_only     = sum(as.integer(data_types == 1)),
            most_data_types_by_any_one = max(data_types),
            most_category_codes_by_any_one = max(category_codes)) |>
  collect()
```

Eighty of the ninety-seven record their catch one way only, and no single
platform ever uses more than three of the five conventions. The closer you look
at the unit that does the work, the smaller its corner of the format gets: at
country level thirteen of twenty-two never vary, at platform level eighty of
ninety-seven. Adding time tightens it again — in 1,229 of 1,347 platform-years
the catch is reported one way throughout.

That is the position someone is in when they upload. They sample in a single,
settled way, and must then express it in a format built wide enough to hold
everybody else's way of working too — most of which is irrelevant to them, and
therefore most of which they have little reason to understand well. The parts
they never use are the parts they are most likely to fill in wrongly.

This is worth stating plainly, because the register in Part III could otherwise
read as a list of other people's mistakes. It is better read as the cost of one
row shape serving every sampling practice in the community at once. It is better read as the cost of one
row shape serving twenty different sampling practices.


Length units run the other way and are worse for it: **20 of the 29 surveys
use more than one `LengthCode`**, so there the variation is inside the series
rather than between them.

The remedy for the first problem — several different things being counted in
one table — is to give each of them a table of its own, and that is what
`HL_length` and `HL_summary` are: not a packaging decision, but the splitting
apart that the record has been asking for. Each holds one row per observation
of one kind, and each column means one thing in one unit: `length_mm` is
always millimetres, `n_haul` is always a whole-haul count, `n_hour` is always
a rate per hour. The branching on `DataType` and `LengthCode` is worked out
once, when the tables are built, instead of in every analysis that reads
them.

**Two tables, three tallies.** `HL_length` is the length tally and
`HL_summary` the catch tally. The sub-sample tally gets no table of its own:
`SubsamplingFactor`, `SubsampledNumber` and `SubsampleWeight` appear in
neither file. What survives of it is `n_measured`, the count of fish actually
put through the calipers, carried alongside the lengths in both tables; the
raising factor is applied and then discarded.

**Is a third table warranted? Probably not, and the reason is what it would be
used for.** The sub-sample tally answers questions about *method* — how much
of the catch was examined, whether a raising factor is plausible, how a
provider's sampling changed over time. Those are quality-control and
methodological questions, and they are asked at the record level by people who
want the records: HH and HL themselves, which ICES already serves and which
`{obus}` exposes unchanged through `dr_con_raw()`. A derived table would be a
second copy of something nobody was struggling to obtain. The catch tables
exist because the catch questions need arithmetic that is easy to get wrong;
the sampling questions need the raw fields undisturbed, which is a different
service.

One piece of the sub-sample tally does earn a place in both tables, and it is
carried: `n_measured`. How much of a length distribution was measured rather
than raised is a question asked *about a catch*, at the level of the catch, which
is why it belongs here and the factor does not. What genuinely disappears is
`SubsampleWeight`, and with it the only candidate cross-check the weight path
has — that is a real loss rather than a tidy-up, and it is [an open
question](#what-remains-open) rather than a settled one.

# Part II. The two tables

## The two tables, formally {#the-two-tables-formally}

So much for why they have the shape they do. This is what they actually
contain. Both are one flat parquet file, read on demand.

```{r, include = FALSE}
# The column list is READ FROM THE PUBLISHED FILES; only the descriptions are
# authored here. A column added to obus and not described below renders as a
# visible gap rather than quietly vanishing from this article (Principle 8).
# n_haul and n_measured mean different things in the two tables, so the
# descriptions are per-table rather than per-name.
KEY <- list(
  HL_length  = c(".id", "Valid_Aphia", "length_mm", "accuracy", "LengthType",
                 "SpeciesSex", "DevelopmentStage", "SpeciesValidity"),
  HL_summary = c(".id", "Valid_Aphia", "SpeciesValidity")
)
COMMON <- c(
  .id         = "Haul identifier, built by obus from the eight fields ICES uses to identify a haul",
  Survey      = "Survey code, e.g. `NS-IBTS`",
  Year        = "Year of the haul",
  Quarter     = "Quarter of the haul",
  Valid_Aphia = "WoRMS AphiaID of the accepted taxon",
  latin       = "Accepted scientific name",
  species     = "Common name",
  rank        = "Taxonomic rank of the identification (`Species`, `Genus`, `Family`, ...)",
  SpeciesValidity = "ICES `SpecVal` -- the *record type*, i.e. what data is present; not a quality flag"
)
DESC <- list(
  HL_length = c(COMMON,
    length_mm  = "Lower bound of the length bin, in millimetres",
    length_cm  = "`length_mm / 10`, carried because the two user communities count in different units",
    accuracy   = "Bin width in centimetres, from ICES `LngtCode`",
    LengthType = "ICES `LenMeasType` -- which landmark was measured (total, standard, pre-anal-fin, ...)",
    SpeciesSex = "ICES `Sex`, as reported: `F`, `M`, `U`, `B` (berried) or `NA` (never assessed)",
    DevelopmentStage = "ICES `DevStage`, as reported",
    n_haul     = "Number at this length in the whole haul: `NumberAtLength` x `SubsamplingFactor`",
    n_hour     = "`n_haul` standardised to one hour fished",
    n_measured = "`NumberAtLength` as submitted, before raising -- how many fish went through the calipers. `NA` where `DataType == \"C\"`, which reports a rate"),
  HL_summary = c(COMMON,
    n_totalnumber = "The submitted `TotalNo` -- the one per-species haul total present whether or not anything was measured",
    n_totalnumber_hour = "`n_totalnumber` standardised to one hour fished",
    n_haul     = "The same quantity reached the other way, by raising the length frequencies. The two do not always agree, and that gap is a finding",
    w_haul     = "Submitted `CatCatchWgt` for the species in the haul",
    w_hour     = "`w_haul` standardised to one hour fished",
    n_measured = "Fish actually measured, before raising. `0` for a species with no length data; `NA` where `DataType == \"C\"`",
    p_females  = "Proportion female, derived from the per-sex counts")
)
spec_table <- function(tbl) {
  x  <- dr_con(tbl) |> head(1) |> collect()
  d  <- DESC[[tbl]]
  nm <- names(x)
  data.frame(
    Column = sprintf("`%s`", nm),
    Type   = vapply(x, function(v) class(v)[1], character(1)),
    Key    = ifelse(nm %in% KEY[[tbl]], "*", ""),
    Meaning = ifelse(nm %in% names(d), d[nm],
                     "**undocumented — add it to `DESC` in this article**"),
    row.names = NULL, check.names = FALSE)
}
```

**`HL_length`** — one row per species at one length class in one haul, for
species that were actually measured. `*` marks the fields that together
identify a row.

```{r, echo = FALSE}
knitr::kable(spec_table("HL_length"))
```

**`HL_summary`** — one row per species in one haul, for *every* species
recorded, measured or not.

```{r, echo = FALSE}
knitr::kable(spec_table("HL_summary"))
```

Three things are worth saying about those keys before Part III takes them
apart. `HL_length` is identified by eight fields because ICES's own record
identity puts them there, and dropping one silently merges counts that were
separate observations — [what makes a row
unique](#what-makes-a-row-unique) measures how often. If you want the numbers
added up more coarsely, `dr_HL_collapse()` is the route to use; it carries
guards that a
hand-rolled `group_by()` loses; it is described under [what makes a row
unique](#what-makes-a-row-unique). And `SpeciesValidity` is in
both keys because it is a record type rather than a quality flag — filtering
it away discards real records.

**You never need an AphiaID.** Both tables carry `latin`, `species` and
`rank` alongside `Valid_Aphia`, so a species is selected by name:

```{r, eval = FALSE}
dr_con("HL_length") |> filter(species == "Atlantic herring")
dr_con("HL_length") |> filter(latin == "Sprattus sprattus")
```

That is not cosmetic. `Valid_Aphia == 126425` is unreadable, unverifiable on
sight, and a typo in it returns an empty table rather than an error — which is
the same failure mode as a wrong survey code, and just as quiet.

## The same calculation as CPUEL, and where the numbers differ

ICES already publishes this table. It is called **CPUEL** — catch per unit
effort at length: numbers per hour by haul, species and length class,
[@ices2021units] — and it is the natural starting point for length-based work,
computed centrally so that nobody has to raise a sub-sample again.

It covers **8 of the archive's 29 surveys**:

```{r}
duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CPUEL.parquet") |>
  distinct(Survey) |> count() |> collect()

dr_con("HH") |> distinct(Survey) |> count() |> collect()
```

**`HL_length` computes the same quantity for all 29.** That is a claim about
definition and coverage: raised numbers at length, per haul, species and
length class, by the same arithmetic, for every survey in the archive rather
than for eight of them.

Whether the two also agree *number for number* where they overlap is a
separate question, and it deserves a straight answer rather than a match rate.
Across 5.5 million comparable records, **1.5% of CPUEL's rows have no
`HL_length` partner and 5.9% of its counts differ** — roughly eighty thousand
rows and three hundred thousand counts. For two tables that should be the same
calculation, that is not a rounding residual.

Some of it is understood. The two files are independent downloads with their
own calculation vintages, so a haul revised on one side and not the other will
differ, and the comparison cannot separate that from a real disagreement. On
BITS, CPUEL gear-standardises cod and nothing else and does not
always preserve the submitted length grid — both documented, and
both choices a generalised product would have to make deliberately rather than
inherit. Set BITS and SWC-IBTS aside and counts agree on 97.4% for the
remaining six surveys. **SWC-IBTS is unexplained, and remains so.**

So: the same calculation, demonstrably; the same numbers, not yet. Closing
that gap is work neither this article nor the package has finished, and a
reader deciding whether to use one table in place of the other should know
which of the two claims they are relying on. [Building CPUEL from
`HL_length`](#building-cpuel-from-hl_length) in the appendices shows what the
comparison looks like on a single survey-quarter.

**Two further differences are by design, not residual.**

- **Sex.** CPUEL carries `sex` as `NA` on every survey. `HL_length` keeps
  `SpeciesSex` wherever it was recorded, and some surveys record it heavily —
  everything on ROCKALL, 64% of SWC-IBTS, 76% of EVHOE.
- **Zero rows.** Roughly half of CPUEL's rows are explicit zeros; `HL_length`
  adds none. That is not an oversight but the reason it generalises. A zero
  only means *none caught* if you know the species was in scope for that haul,
  and settling that universe is a survey-by-survey judgement. Leaving it out
  is what lets the same function work identically on all 29 surveys instead of
  stopping at eight — and it leaves the choice of universe with you.

None of that is a complaint about CPUEL. ICES's Data Centre builds and
maintains real, working products — CPUE by length, by age, catch weight,
abundance indices — computed centrally so that the tools that use them can
read them straight off. That is the right shape, and it is the shape this article is asking
for more of. The argument is narrower than it may sound: the length-frequency
half of that suite does not need the zero-fill question answered first, so it
can be built for every survey.

## What this saves you later

`HL_length` and `HL_summary` exist so that those four steps — raise the
measured numbers, resolve the length units, decide what to do when the factor
or the duration is missing, and avoid adding up a total the submitter repeated
— are done once, the same way for the whole archive,
following ICES's own written procedure. And so that the hauls where they
cannot be done cleanly are visible rather than quietly absorbed.

Done by hand, none of the four is hard, exactly. They are a lot of
bookkeeping to get right every time, and easy to get subtly wrong once — and a
species silently missing from a haul reads exactly like a species nobody ever
looked for. Everything that follows starts from the same table: length spectra,
abundance and biomass per haul, CPUE, stratified indices, species composition,
age–length keys. Getting the first step wrong once is getting it wrong
everywhere.

## One haul, from submission to catch at length

A single sprat record shows the whole shape of it — five length rows, exactly
as the country submitted them:

```{r}
dr_con("HL") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(LengthCode, LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber, SpeciesCategoryWeight) |>
  arrange(LengthClass) |>
  collect()
```

Four things have to happen before that is a catch at length:

- **The units have to be resolved.** `LengthCode` `.` says the `LengthClass`
  values are millimetres, so 115 means 11.5 cm and each class is 1 mm wide.
- **The numbers have to be raised.** Only part of the catch was measured. The
  11 sprat measured at 11.5 cm stand for 11 × 4.177 = 45.9 fish in the catch.
- **The sub-sample is not the catch.** The 112 in `SubsampledNumber` is how
  many fish went through the calipers, not how many came aboard.
- **The total is written once but repeated everywhere.** `TotalNumber` 467.82
  is a single figure for the whole species in this haul, and it stands on all
  five rows. Add that column up and you get the catch five times over. This is
  the problem named [above](#what-is-wrong-with-hl), in one record: a figure
  that belongs to the species has nowhere of its own to live, so it is copied
  onto every length row.

`HL_length` carries the raised numbers at length, in resolved units:

```{r}
dr_con("HL_length") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(length_cm, accuracy, n_haul, n_hour, n_measured) |>
  arrange(length_cm) |>
  collect()
```

and `HL_summary` carries one row for the species in the haul, with the
reported total beside the one rebuilt from the lengths:

```{r}
dr_con("HL_summary") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  select(species, n_totalnumber, n_haul, n_measured, w_haul) |>
  collect()
```

467.82 against 467.824. The reported total and the raised length frequencies
agree here to four thousandths of a fish, which is simply what happens when a
raising factor that is not a whole number is applied to whole fish. [Two
routes to a catch total](#two-routes-to-a-catch-total) is about the hauls
where they do not agree.


## Which table for which job

The columns are listed in [The two tables,
formally](#the-two-tables-formally). What a column list cannot show is which
table to reach for, and the one difference between them that catches people
out:

| | `HL_length` | `HL_summary` |
|---|---|---|
| Which species | only those actually measured | every species recorded, measured or not |
| Rows | `r N(M$len_rows)` | `r N(M$sum_rows)` |

| What you want to know | Where to get it |
|---|---|
| length spectra, size-based indices, inputs to age–length keys, length–weight work | `HL_length` |
| numbers or weighed biomass per haul, CPUE, species composition, stratified indices | `HL_summary` |
| where a species occurs — distribution, occupancy, range | `HL_summary`, because it lists every species recorded and not only the measured ones |
| species richness or diversity per haul | `HL_summary`. Counting species in `HL_length` instead loses at least one in a third of hauls |
| biomass of a length-measured species, **estimated** rather than weighed | `HL_length`, via `dr_add_length_mid()` then `dr_add_predicted_weight()` |
| how much of a catch was sub-sampled, anything per `SpeciesCategory`, per-category weights, or rebuilding an exchange file | `dr_con("HL")` — the submitted records, unchanged apart from `.id` |
| age, individual weight, maturity, age–length keys | `dr_con("CA")` — `{obus}` derives no CA table |

**Biomass comes two ways, and they are different quantities.** `HL_summary`'s
`w_haul` is what the catch actually weighed, as reported. For any species that
was length-measured you can also *estimate* biomass from the lengths, applying
a length–weight relationship class by class — which also gives you biomass *at
length*, which no weighed figure does:

```{r}
dr_con("HL_length") |>
  filter(.id == "NS-IBTS:2017:3:DK:26D4:GOV:68:33", Valid_Aphia == 126425) |>
  collect() |>
  dr_add_length_mid() |>
  dr_add_predicted_weight() |>
  summarise(fish = sum(n_haul), estimated_grams = sum(w_haul_pred))
```

It is an estimate, and it rests on coefficients from outside the submission, so
it will not match the weighed figure exactly — here it comes to about four
fifths of it. `dr_compare_length_weight()` sets the two against each other
across a whole survey.

**Which species is the difference that catches people out.** A haul's record
routinely includes species that were only counted in bulk, only weighed, or
only noted as present. Those have no `HL_length` row at all — not a zero, no
row. A species list taken from `HL_length` is therefore a list of what was
*measured*, not of what was *caught*. [Why two tables and not
one](#why-two-tables-and-not-one) measures how much that matters.

Both tables are published as parquet files and read on demand: ask for one
survey, year or species and only those rows are fetched, so you can look at a
single haul without downloading the archive.


## Why two tables and not one

The structural answer is [above](#what-is-wrong-with-hl): HL holds two
different things at once, so separating them gives two tables and not one. But that
argument would be academic if the split did not also fall along a seam the
data actually has. It does, in two places, and both come from the data rather
than from convenience.

**Part of a haul's species roster is never length-measured.** A haul's record
routinely includes species that were only counted in bulk, only weighed, or
only noted as present. `r N(M$sum_nolen)` of `HL_summary`'s
`r N(M$sum_rows)` rows — **`r sprintf("%.1f%%", M$sum_nolen_pct)`** — have no
length rows behind them at all. That is a minority of records, but it is not a
rare corner of the archive: those records fall in
`r N(M$sum_nolen_hauls)` of `r N(M$sum_hauls)` hauls,
**`r sprintf("%.0f%%", M$sum_nolen_haul_pct)`** of them, so a length table
alone would quietly drop at least one species from a third of the hauls it
covers:

```{r}
dr_con("HL_summary") |>
  anti_join(dr_con("HL_length") |> distinct(.id, Valid_Aphia, SpeciesValidity),
            by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  count() |>
  collect()
```

Those species cannot appear in a length table, and giving them a placeholder
length row would invent a measurement that was never made. `HL_length` is
therefore, by definition, only the species that were measured, while
`HL_summary` is the full list of what was caught.

**Numbers and weights survive being split by sex differently.** `HL_length`
carries `SpeciesSex` as reported, because `NumberAtLength` is a genuine
per-sex count. `SpeciesCategoryWeight` is not reliably per-sex — DATRAS
repeats one weight across a species' sex rows — so a table split by sex would
state a weight that was never weighed. Keeping the two quantities in tables of
different shape is what lets each be right. This is worked through in [what
makes a row unique](#what-makes-a-row-unique) below.

## Why these columns, and which of them could go

These two tables are a **proposal**, not a settled product, and the column set
is deliberately generous. Several columns can be worked out from others, and that
is a choice rather than an oversight — while the shape of the tables is still
open, adding a column later is disruptive and removing one is not.

Three of the redundancies are worth naming, because a reader will spot them:

- **`length_cm` is exactly `length_mm / 10`** — every row, no exceptions. It
  is carried because the two communities that would use these tables count in
  different units. The exchange format and CPUEL are
  millimetre; `{DATRAS}`, `{DATRASextra}` and the length–weight relationship
  *W = aL^b* are centimetre, as is `accuracy`. Carrying both means neither side
  converts.
- **The per-hour columns are unit conversions.** No row in either table carries
  an hourly value whose per-haul counterpart is missing — measured as zero for
  `n_hour`, `n_totalnumber_hour` and `w_hour` alike — and the reconstruction
  `n_haul / HaulDuration × 60` is exact to machine precision. They are carried
  because per hour is what CPUEL publishes and what most later work asks
  for. The alternative is to publish `HaulDuration` instead and let the user
  standardise to an hour, a half hour or a swept area — which is arguably the
  better table, and the one that stops presuming what the next analysis
  needs.
- **`latin` and `species` are one join away** from `Valid_Aphia` and
  `dr_con("species")`. They are carried so that someone can filter on a common
  name without first knowing an AphiaID.

`n_measured` is the one you could *not* work back to, because some rows add up
raw rows that were raised by different factors, so no single divisor recovers
it. It is also the one the analysis tools never read: `{DATRAS}` uses
`NumberAtLength` once, when it reads the file, to produce `Count`, and never
looks at it again. It is kept for two reasons. It is the only number left on
the rows where the raising factor is missing, and how much of a length
distribution was actually measured rather than raised is a question worth
being able to ask.

`.id` is the largest single column in the table, and it is not redundant:
five of the eight haul fields it concatenates — `Country`, `Platform`,
`Gear`, `StationName`, `HaulNumber` — appear nowhere else.

**Which of these survive is not a package author's decision** — and not any
one group's either. The survey working groups know what their data can
support; the Data Centre knows what it can build, serve and keep correct; and
the assessment and ecosystem groups are who the columns are
actually for. If ICES adopts tables of this shape, what they carry is a
decision for those three together, and the sensible direction of travel is to
trim. The cost that matters is not storage — the whole of `HL_length` is under
200 MB and no column in it is expensive — but **interface**: every column is
one more thing a user has to understand and a maintainer has to keep true. This section says what each one
duplicates and why it is there, so that conversation starts from the design
rather than from taste.

## What makes a row unique {#what-makes-a-row-unique}

ICES states the record's identity in the field descriptions, four times over —
once each under `TotalNumber`, `SubsampledNumber`, `NumberAtLength` and
`SpeciesCategoryWeight` — and all four name the same five components, the

> unique combination of cruise haul, species, sex, devstage and subsampling
> category identifier

(the phrasing varies slightly between the four cells; the components do not).
So an HL record is identified by **haul × species × sex × `DevelopmentStage` ×
`SpeciesCategory`**, and the length rows add `LengthClass` and `LengthType`.
The two `{obus}` tables follow that, with the sub-sampling category summed
away — which is safe because a category is a sampling stratum and not a
biological one: the North Sea IBTS manual [@ices2020sisp10] has one created
when a catch holds two or more distinct length groups so that each can be
sub-sampled at its own rate, so the categories partition a single catch and
adding them back returns it.

| Table | One row per |
|---|---|
| `HL_length` | `.id` × `Valid_Aphia` × `length_mm` × `length_cm` × `accuracy` × `LengthType` × `SpeciesSex` × `DevelopmentStage` × `SpeciesValidity` |
| `HL_summary` | `.id` × `Valid_Aphia` × `SpeciesValidity` |

Each is exactly unique there — no repeated combinations at all, over 14 million
rows — and at anything coarser it is not, so **add `n_haul` up rather than
assuming one row**:

```{r}
dr_con("HL_length") |>
  count(.id, Valid_Aphia, length_mm, length_cm, accuracy,
        LengthType, SpeciesSex, DevelopmentStage, SpeciesValidity) |>
  filter(n > 1) |>
  count() |>
  collect()
```

`LengthType` and `DevelopmentStage` are part of what makes a row unique, not
labels carried along, because ICES's record identity puts them there: one haul
can measure the same species at the same length under two measurement
conventions, or as both egg-bearing and unstaged, and each is its own count.
Over the 13,992,024 combinations at the coarser `.id` × `Valid_Aphia` ×
`length_mm` × `SpeciesSex` level, 5,611 split on `DevelopmentStage` (0.040%),
4,044 on `LengthType` (0.029%), and 15 each on `SpeciesValidity` and
`accuracy` (0.0001%). Rare — but each split is a separate count, and collapsing
one is silent. `DevelopmentStage` is the easiest to overlook, being `NA` for
99.8% of rows, and dropping it would add a berried record to an unstaged one.

**If you want the numbers added up more coarsely, ask for that — do not do it by hand.**
Everyone else's table is one of these with something added away, so wanting
that is entirely normal. `dr_HL_collapse()` does it, whether the table is
still on disk or already pulled into memory:

```{r, eval = FALSE}
dr_con("HL_length") |>
  dr_HL_collapse(c("SpeciesSex", "DevelopmentStage", "SpeciesValidity"))
```

It exists for what a `group_by()` silently loses rather than for the grouping
itself: the all-`NA` case, where `sum(x, na.rm = TRUE)` over a group with
nothing in it is `0` in R and `NULL` in SQL, so an unknown raising comes back
as a confident zero; and the `DataType == "C"` rule, which keeps a rate from
being reported as a count. It also refuses what cannot be summed —
`accuracy` and `LengthType` where a group actually mixes them, since two
records at one `length_mm` measured to 1 cm and to 5 cm are different bins.
`{DATRAS}`'s `addSpectrum()` collapses a mixed `LengthCode` to the coarsest with
only a warning; this errors and names the group. It does not reproduce
`HL_summary`: that table's totals come from the submitted `TotalNumber`, which
is a different quantity, as [two routes to a catch
total](#two-routes-to-a-catch-total) sets out.

That choice has one measurable cost, in [duplicated length
series](#duplicated-length-series) below.

## Why `HL_summary` is not split by sex

`HL_length` carries `SpeciesSex`; `HL_summary` does not. That difference is
deliberate, and it comes from what the submissions contain rather than from any
limitation of the package.

`NumberAtLength` is a genuine per-sex, per-length count, and — checked over the
whole archive — never repeats one placeholder value across a species' sex rows.
`TotalNumber` does. Where a haul's record splits a species by sex, roughly two
thirds of the time the total genuinely varies from sex to sex, and the
remaining third of the time one species-level figure has simply been written
onto every sex row. Per-sex arithmetic can tell those two apart, and `{obus}`
does exactly that — the measurement, and what the package does about it, are in
[one `TotalNumber` written across several sex
rows](#totalnumber-across-sex-rows).

`SpeciesCategoryWeight` has no comparable check, and its
repeats-versus-varies pattern tracks `TotalNumber`'s only about half the time.
Splitting `HL_summary` by sex would therefore misstate `w_haul` for an unknown
share of species with no way to flag which. Sex-specific *numbers* stay
available by adding up `HL_length`; sex-specific *weight* is not recoverable in
general. That is a property of the source data, not a gap in `{obus}`.

Sex codes are kept as reported — `"F"`, `"M"`, `"U"`, `"B"` or `NA`. `"B"`
(berried, egg-bearing) is not folded into `"F"`, and `"U"` (examined,
undetermined) stays distinct from `NA` (never examined). The female proportion
`HL_summary` publishes as `p_females` is one step away from `HL_length`:

```{r, eval = FALSE}
dr_con("HL_length") |>
  group_by(.id, Valid_Aphia) |>
  summarise(
    p_females = sum(n_haul[SpeciesSex %in% c("F", "B")]) /
                sum(n_haul[SpeciesSex %in% c("F", "M", "B")]),
    .groups = "drop"
  )
```

## Do the derived tables still say what the submissions said?

Yes, and this is checked rather than argued. `HL_length`
has been compared cell by cell against the `{DATRAS}` package's own `Count`
[@kristensen2023datraspkg], over every species and survey in the archive — more
than thirteen million haul × species × length cells. **No cell differed by as
much as half a fish**, the largest difference anywhere being floating-point
dust, and no row appeared in `HL_length` that was not in the submitted record.
The comparison is reproduced by `data-raw/CHECK_datras_interop.R`; it needs the
`{DATRAS}` package, so it is not re-run when this page is built.

The differences run one way only. A few rows are in the submitted record and
not in `HL_length`, and they are all the same case: the raising factor is
missing, so there is no honest number to publish and `{obus}` returns `NA`
rather than assuming the catch was fully measured.

```{r}
dr_con("HL_length") |>
  summarise(rows = n(), no_raised_number = sum(as.integer(is.na(n_haul)))) |>
  collect()
```

`{DATRAS}` assumes 1 in exactly these rows and so reports a number; `{obus}`
does not, because ICES makes the field mandatory. That is a choice, argued in
[a mandatory `SubsamplingFactor` that is
absent](#subsamplingfactor-absent), not an error.

The tables are faithful, but they are **not lossless**. Nearly every row in
`HL_length` comes from exactly one submitted HL row. About one row in three
hundred combines several, and where that happens the raising factor and the
sub-sampling category are added away and cannot be recovered from the
published table. Also gone at this level: `SubsampledNumber`,
`SubsampleWeight`, per-category `SpeciesCategoryWeight` (`w_haul` is added up
to the species) and any weight split by sex.

## The haul identifier, `.id`

`.id` is an `{obus}` construct; ICES has no such field. It is the eight fields
ICES's own exchange tables use to key a haul, concatenated with `:` —

`Survey : Year : Quarter : Country : Platform : Gear : StationName : HaulNumber`

— for example `"BITS:1991:1:DE:06S1:H20:33:28"`. `NA` fields (most often
`StationName`) are skipped rather than rendered as the literal string `"NA"`,
matching the convention the published archive already uses. When every field
is `NA` the result is `NA`, not the empty string, because an empty string
would collide every such row onto one identifier.

`dr_add_id()` accepts only the current names. A legacy-named table must be
renamed first (`opus::op_rename()`); `Ship`, `StNo` and `HaulNo` are not
handled.

# Part III. Where the data is ambiguous

Part II is arithmetic the format specifies. This part is about the places it
does not — a small set of constructs the archive contains that a single record
cannot settle on its own, where two submissions write the same shape and mean
different things, or write a value the format does not allow. For most of
DATRAS there is nothing to say here, and that is not said often enough: the
great majority of fields mean what the field descriptions say they mean, and
are used that way.

**First, how much of the archive this is about.** Every record `{obus}` touches
that is affected by anything in this part is marked in `dr_con("hl_flag")`, and
each marking says whether it is a genuine defect or simply the format behaving
as documented. Counting only the genuine ones:

```{r}
dr_con("hl_flag") |>
  left_join(dr_con("hl_flag_code") |> select(code, kind), by = "code") |>
  filter(kind %in% c("suspect", "unexplained")) |>
  summarise(records_affected = n_distinct(paste0(.id, "|", Valid_Aphia)),
            hauls_affected   = n_distinct(.id)) |>
  collect()
```

Against `HL_summary`'s `r N(M$sum_rows)` records, that is under three in every
hundred, sitting in about one haul in eleven. **For the other ninety-seven
records in a hundred the two tables are right and the arithmetic in Part II is
all there is to it.**

**So why spend a whole part on the remainder?** Because of the shape of the
error. The two tables are *vaguely right* everywhere — right to the fish for
almost all of the archive, and right in substance even where a submission is
odd. A small number of records are *precisely wrong*: a particular country, in
a particular year, writing a particular number in a way that cannot mean what
it appears to mean. Being vaguely right is a good position to argue from. Being
able to point at exactly which records are precisely wrong is more useful
still, because it can be acted on.

And it can only be acted on in one place. Where a submission writes one number
in a way that could mean two things, no derivation — ours, `{DATRAS}`'s, or
ICES's own — can recover which was meant. The fix is in the data, at source.

**And that source is the national institute that uploaded the record.** Only
the programme that made a submission still holds the sampling sheets behind it,
so only that programme can say what a number was meant to mean and correct it.
The Data Centre is a middleman here: DATRAS stores what it is sent.

**Which is not to say the submitters got it wrong.** The format gives them one
kind of row for three different tallies, so a total belonging to the species
has nowhere of its own to go and must be written onto every length row — [the
problem Part I describes](#what-is-wrong-with-hl), along with the fact that no
institute ever exercises more than a corner of the format it has to submit to.
Writing one total across several sex rows is not carelessness; it is close to
what the record shape asks for. And the same structure that makes the ambiguity
easy to create makes it impossible to resolve later, because the field that
would settle it was never given a column.

Nor is there anyone outside to refer this to. The format is set through ICES
working groups staffed by scientists employed by these same institutes, and the
Data Centre works for their consortium. The people who wrote the format, the
people who submit to it and the people who can repair it are one community. So
the fix has two halves, and both are theirs: correcting the particular records
below, and giving the three tallies tables of their own so that the next decade
of submissions cannot express the ambiguity at all. The second is the argument
of this whole article.

Each construct is shown the same way: **a real record**, so its shape is
visible; the archive-wide count, so its scale is; what `{obus}` does with it;
and the question it leaves. They are ordered by reach: how many surveys write
the construct, then how many stations, then how many records. The first four
are written by fourteen surveys or more and touch thousands of stations; the
last five are narrow enough to name the surveys and countries involved.

## 1. One `TotalNumber` written across several sex rows {#totalnumber-across-sex-rows}

A haul, a species and a sampling category can carry several rows because the
catch was sexed. `TotalNumber` is written on each. Here is a plaice record
where each sex reports its own:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:1999:1:DK:26HI:TVS:017492:22", Valid_Aphia == 127143) |>
  group_by(SpeciesSex) |>
  summarise(TotalNumber        = max(TotalNumber, na.rm = TRUE),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsamplingFactor  = max(SubsamplingFactor, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Three rows; each `TotalNumber` matches that sex's own measured lengths at a
factor of 1. The haul total is 4. But two of the three values are the same
number, so a consumer that removes duplicates gets 3 — the unsexed row and the
male row agree by coincidence, not because one is a copy of the other.

Other submissions do the opposite, and write one whole-category total on every
sex row. The archive is split roughly a third to two thirds:

```{r}
dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage,
           SpeciesSex, TotalNumber) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_sex > 1) |>
  summarise(groups             = n(),
            one_value_repeated = sum(as.integer(n_val == 1)),
            one_value_per_sex  = sum(as.integer(n_val == n_sex)),
            neither            = sum(as.integer(n_val > 1 & n_val < n_sex))) |>
  collect()
```

Nothing in the record says which a group is. Add them up and the repeated kind
is inflated; collapse them and the per-sex kind is understated.

**Is it ad hoc, or a house style?** If repeating is a submitter's habit it
should be near-total for that submitter; if it is a slip it should be a small
minority everywhere. So ask it of each country that sends data, showing both
how often it happens and how much data each country contributes:

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-alt: "Canada repeats one total on every sex row in 100% of its records and
#|   is also one of the largest contributors. The other fourteen countries all lie
#|   between 8 and 23 per cent."
dr_con("HL") |>
  distinct(Country, .id, Valid_Aphia, SpeciesCategory, SpeciesValidity,
           DevelopmentStage, SpeciesSex, TotalNumber) |>
  group_by(Country, .id, Valid_Aphia, SpeciesCategory, SpeciesValidity,
           DevelopmentStage) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_sex > 1) |>
  group_by(Country) |>
  summarise(records = n(), repeated = sum(as.integer(n_val == 1)), .groups = "drop") |>
  filter(records >= 500) |>
  mutate(pct = 100 * repeated / records) |>
  collect() |>
  mutate(Country = reorder(Country, pct)) |>
  ggplot(aes(pct, Country)) +
  geom_segment(aes(x = 0, xend = pct, yend = Country), colour = "grey80") +
  geom_point(aes(size = records, colour = pct > 90)) +
  scale_size_area(max_size = 9, breaks = c(1000, 10000, 50000),
                  labels = c("1,000", "10,000", "50,000"),
                  name = "multi-sex\nrecords") +
  scale_colour_manual(values = c(`FALSE` = "grey30", `TRUE` = "firebrick"),
                      guide = "none") +
  scale_x_continuous(limits = c(0, 104), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "records where one total is written on every sex row (%)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

Both, as it turns out, and one country is on its own. Canada repeats the total
in **every single** one of its multi-sex records, without exception, and it is
one of the largest contributors of them, which is why it accounts for about
seventy per cent of all the repeats in the archive. For Can-Mar this is plainly
a house style and can be handled by rule.

Every other country lies between eight and twenty-three per cent. That is not a
convention but case-by-case variation, and it is where the ambiguity actually
bites, because there it cannot be settled by knowing who sent the file. Note
also that a high rate and a large contribution are different things: Spain sits
near the bottom of the scale but sends a great many records, so it contributes
more repeats than the Netherlands at the top of it.

**What `{obus}` does.** `dr_HL_summary()` tests each sex's own `TotalNumber`
against its own length rows raised by its own factor — the check the record
above passes three times. A sex that reconciles is trusted and summed on its
own. Sexes that do not reconcile, or that have no length data to check
against, fall back to collapsing duplicate values, but only among values no
reconciling sex has already claimed.

**For ICES.** Is a `TotalNumber` repeated across sex rows meant as the category
total, or as each sex's own? If the former, should the sexed rows carry it at
all?

## 2. The same shape, for `SpeciesCategoryWeight`

A *Lepidorhombus boscii* record, three sexes, counts plainly per-sex:

```{r}
dr_con("HL") |>
  filter(.id == "SP-NORTH:1990:4:ES:29CS:BAK:022:22", Valid_Aphia == 127145) |>
  group_by(SpeciesSex) |>
  summarise(TotalNumber           = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Counts of 44, 73 and 1 — genuinely different, genuinely per-sex. The weight is
4250 on all three rows, which cannot be per-sex: three catches that differ
seventy-fold in number do not weigh the same. It is one category weight written
three times. Sum it and the species gains 12,750 g where it caught 4,250.

Archive-wide the weight splits much as the count does:

```{r}
dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesValidity, SpeciesCategory,
           SpeciesSex, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, SpeciesCategory) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_w   = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(n_sex > 1) |>
  summarise(groups              = n(),
            one_weight_repeated = sum(as.integer(n_w == 1)),
            one_weight_per_sex  = sum(as.integer(n_w == n_sex))) |>
  collect()
```

Weight is worse off than number, because it has no arithmetic cross-check. The
count above could be tested against the length rows that produced it. A weight
can be tested against nothing in the record — the reasoning that unmasked the
4250 was biological, not arithmetic.

**What `{obus}` does.** Collapses duplicate weights without sex in the key —
the opposite of the count treatment in 1. The asymmetry is forced by the data,
not chosen: the reconciliation that arbitrates counts has no weight equivalent.

**For ICES.** The same question as 1, for weight. Because there is no
cross-check: is there any field a consumer can use to tell a repeated weight
from a genuine per-sex one? And is `DevelopmentStage` part of the
`SpeciesCategoryWeight` key, as the field description's wording implies, given
that sex demonstrably is not?

## 3. Two routes to a catch total, and where they part {#two-routes-to-a-catch-total}

`HL_summary` carries two count columns because DATRAS states the catch total
twice: once directly, as the submitted `TotalNumber`, and once implicitly, as
the length rows raised by their own factor. Where a submission is well formed
the two agree, and most do:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(reported = max(TotalNumber, na.rm = TRUE),
            raised   = sum(NumberAtLength * SubsamplingFactor, na.rm = TRUE),
            .groups  = "drop") |>
  filter(!is.na(reported), !is.na(raised)) |>
  summarise(groups          = n(),
            agree_within_1  = sum(as.integer(abs(reported - raised) < 1)),
            differ_1_plus   = sum(as.integer(abs(reported - raised) >= 1)),
            reported_higher = sum(as.integer(reported - raised >= 1)),
            raised_higher   = sum(as.integer(raised - reported >= 1))) |>
  collect()
```

91.6% agree to within one fish, which is what a non-integer factor applied to
whole fish does. 133,874 groups differ by a fish or more, and it is
`TotalNumber` that is higher in all but 3,437 of them.

What a gap looks like, at its simplest — an anchovy record from NIGFS, one
category, no sub-sampling to complicate it:

```{r}
dr_con("HL") |>
  filter(.id == "NIGFS:2019:4:GB-NIR:74RY:ROT:90:10", Valid_Aphia == 126426) |>
  select(LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber) |>
  collect()
```

Three length rows, one fish each. The factor is 1, so nothing is raised. Yet
the record states that 6 were measured and 6 were caught. Three anchovies are
in the totals and absent from the length distribution, and no arithmetic in
the format says which count is right.

At the other extreme the gaps are enormous — `TotalNumber` in the hundred
thousands against raised lengths in the hundreds — but those turn out not to
be disagreements at all. They are one total shared across sampling categories,
which is the construct in [4](#totalnumber-across-categories).

**What `{obus}` does.** Carries both routes in separate columns —
`n_totalnumber` from the submission, `n_haul` from the raised lengths — never
averaged and never silently preferred. They are named differently on purpose:
calling both `n_haul` invited exactly the confusion this section is about.

**For ICES.** When the two routes disagree by more than rounding, which is
authoritative? And should a consumer be able to tell, from the record alone,
that a `TotalNumber` spans categories rather than describing its own row?

## 4. One `TotalNumber` written across several sampling categories {#totalnumber-across-categories}

The same shape one dimension over — and here the record answers the question
itself. A whiting split into two sampling categories:

```{r}
dr_con("HL") |>
  filter(.id == "NS-IBTS:2022:3:GB-SCT:748S:GOV:141:141", Valid_Aphia == 126438) |>
  group_by(SpeciesCategory) |>
  summarise(length_rows           = n(),
            sum_NumberAtLength    = sum(NumberAtLength, na.rm = TRUE),
            SubsamplingFactor     = max(SubsamplingFactor, na.rm = TRUE),
            TotalNumber           = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Category 11 raises to 187 × 1.893 = 353.99; category 12 contributes 7 more.
Together that is 360.99, and both rows report `TotalNumber` 361 — the residue
is the rounding of a four-decimal factor. So the shared 361 is the two
categories *together*, not each one's own, and the shared 58100 is one weight
written twice. Read per-category, this species doubles.

```{r}
dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesValidity, SpeciesCategory, TotalNumber) |>
  group_by(.id, Valid_Aphia, SpeciesValidity) |>
  summarise(n_cat = n_distinct(SpeciesCategory),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_cat > 1) |>
  summarise(groups            = n(),
            one_value_shared  = sum(as.integer(n_val == 1)),
            one_value_per_cat = sum(as.integer(n_val == n_cat))) |>
  collect()
```

3,362 of 57,285 multi-category groups share one value. They concentrate on the
category pairs `1`+`2` and `11`+`12`, and on NS-IBTS, BTS and SCOWCGFS.

**Why would a total be shared at all?** Pseudo-category sampling is the obvious
candidate: under `DataType` `P` a catch is split into sub-categories of one
weighing, so one total covering them is what the design implies. If that is the
mechanism, `P` should be far more prone to it than `R`:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  distinct(.id, DataType, Valid_Aphia, SpeciesValidity, SpeciesCategory, TotalNumber) |>
  group_by(.id, DataType, Valid_Aphia, SpeciesValidity) |>
  summarise(n_cat = n_distinct(SpeciesCategory),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_cat > 1) |>
  group_by(DataType) |>
  summarise(multi_category = n(),
            shared_total = sum(as.integer(n_val == 1)), .groups = "drop") |>
  mutate(pct = round(100 * shared_total / multi_category, 1)) |>
  collect()
```

It is: half of all multi-category `P` groups share a total, against 4% under
`R`, and none at all under `C` or `S`. So pseudo-category sampling is one real
mechanism — but it accounts for 1,150 of the 3,362. The other 2,212 sit under
`R`, where ICES documents no sub-category convention, and those remain
unexplained. The weight splits the same way, for the same two `DataType`s.

**What `{obus}` does.** The fallback collapse leaves `SpeciesCategory` out of
its key, so a shared value is counted once rather than once per category. A
category whose own data reconciles is still credited separately.

**For ICES.** When two categories carry one identical `TotalNumber`, is that
the species total for both strata — as the record above demonstrates — or a
submission error? The weight beside it behaves the same way, and there the
answer is partly known: under pseudo-category sampling it is documented
convention, which is why `{obus}` collapses it. See [the same weight written on
two category codes](#the-same-weight-written-on-two-category-codes).

## 5. A mandatory `SubsamplingFactor` that is absent {#subsamplingfactor-absent}

`SubsamplingFactor` is mandatory, and `1` is a specific claim — *not
sub-sampled*. Some rows carry neither the claim nor a factor:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), is.na(SubsamplingFactor),
         NumberAtLength != 0, !is.na(LengthClass)) |>
  summarise(rows = n(), hauls = n_distinct(.id)) |>
  collect()
```

24,298 length rows over 5,013 hauls, 97.9% of them Can-Mar.

The records are not silent, though. Nearly all of them still report both
`TotalNumber` and `SubsampledNumber`, and the ratio of those two is what the
factor would have been:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), is.na(SubsamplingFactor),
         NumberAtLength != 0, !is.na(LengthClass)) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesSex) |>
  summarise(sum_nal = sum(NumberAtLength, na.rm = TRUE),
            tot     = max(TotalNumber, na.rm = TRUE),
            sub     = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(sub), sub > 0) |>
  summarise(groups          = n(),
            sub_eq_lengths  = sum(as.integer(abs(sub - sum_nal) < 0.5)),
            implied_1       = sum(as.integer(abs(tot - sub) < 0.5)),
            implied_over_1  = sum(as.integer(tot - sub >= 0.5)),
            implied_under_1 = sum(as.integer(sub - tot >= 0.5))) |>
  collect()
```

14,419 of 14,433 groups carry both, and in **every one of them**
`SubsampledNumber` equals the sum of the length rows exactly — so the ratio is
well founded rather than a guess. In 10,416 it comes to 1, so reading the
absence as *not sub-sampled* would have been right. In 3,968 it is above 1, so
that reading understates the catch; 3,955 of those are Can-Mar and 13 DYFS, so
this is one submitter's practice rather than a format-wide pattern.

The remaining 35 are the reason not to simply act on the ratio: they imply a
factor **below** 1, a total smaller than the sub-sample it was raised from,
which is not a sub-sampling factor at all. In those records `TotalNumber` and
`SubsampledNumber` cannot both be right, and nothing says which to keep.

**What `{obus}` does.** Propagates `NA`, and reconstructs nothing. An absent
mandatory field means *no information was supplied*, which is not the same
statement as `1`; substituting one for the other asserts something the
submission does not contain. This is a deliberate difference from the
`{DATRAS}` package, which reads `R` plus a missing factor as 1.

**For ICES.** Should a missing `SubsamplingFactor` be read as 1, as unknown, or
reconstructed as `TotalNumber` / `SubsampledNumber`? The 3,968 groups show the
first is not safe. Whether the third is safe depends on something only ICES can
say — whether `TotalNumber` is authoritative when the factor is absent — and
the 35 incoherent groups show it is not universally so. `{obus}` will not
reconstruct the factor until that is settled.

## 6. A `SubsamplingFactor` below 1

The raising factor is what you multiply a measured sub-sample by to reach the
whole catch, so it cannot be less than 1 without saying that the catch was
smaller than the sample drawn from it. Some submissions say exactly that. Two
John dory, measured and reported by EVHOE:

```{r}
dr_con("HL") |>
  filter(.id == "EVHOE:2011:4:FR:35HT:GOV:P1111:67", Valid_Aphia == 127427) |>
  select(LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber, SpeciesCategoryWeight) |>
  collect()
```

The record's own arithmetic is impeccable: two fish measured, a factor of
0.615, and a reported total of 2 \eqn{\times} 0.615 = 1.23. Every identity in
the format holds. It is the factor that cannot be true — two dory were in the
hand and 320 g of them were weighed, so the catch was at least two fish, not
1.23.

```{r}
dr_con("HL") |>
  filter(!is.na(SubsamplingFactor), SubsamplingFactor < 1) |>
  summarise(length_rows   = n(),
            hauls         = n_distinct(.id),
            surveys       = n_distinct(Survey),
            lowest_factor = min(SubsamplingFactor, na.rm = TRUE)) |>
  collect()
```

372 length rows over 56 hauls and four surveys, spread across 2000 to 2022 and
reaching as low as 0.24 — SP-NORTH and SP-ARSA (ES), BITS (DK and DE) and
EVHOE (FR). It is rare, and it is not confined to one programme or one year.

**What `{obus}` does.** Applies the factor as submitted. A reported value is not
second-guessed, and inventing a replacement would be asserting something the
record does not contain. The consequence is visible rather than hidden, and it
has a signature anyone can test for: the published catch comes out smaller than
the number of fish measured.

```{r}
dr_con("HL_summary") |>
  filter(!is.na(n_haul), !is.na(n_measured), n_measured > 0) |>
  summarise(records = n(),
            catch_below_measured = sum(as.integer(n_haul < n_measured - 0.001))) |>
  collect()
```

**For ICES.** A factor below 1 is not a value the format allows. Which half of
the record is wrong — the factor, or the counts it was applied to? Until that
is settled these records cannot be raised, only reported as they stand.

## 7. `DataType` `C` with a `SubsamplingFactor` above 1

`DataType` `C` means the catch is reported as a rate per hour, and the format
requires `SubsamplingFactor` to be 1. These submissions report both:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType, HaulDuration), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  distinct(.id, Valid_Aphia, DataType, HaulDuration, SubsamplingFactor) |>
  arrange(desc(SubsamplingFactor)) |>
  head(3) |>
  collect()
```

A sprat record labelled as an hourly rate, on a 30-minute haul, sub-sampled at
one part in 168. The two statements cannot both hold.

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  summarise(rows = n(), hauls = n_distinct(.id),
            max_factor = max(SubsamplingFactor, na.rm = TRUE)) |>
  collect()
```

1,832 length rows over 36 hauls — BITS, Lithuania 2008–2016 and Denmark 2018.

**Which of the two statements is wrong?** The label and the factor imply
different arithmetic, so ask the record. Under `C` the total should be the
length rows scaled by duration; under `R` it should be the length rows raised
by the factor. Only one of those can reproduce the reported `TotalNumber`:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType, HaulDuration), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesSex) |>
  summarise(s   = sum(NumberAtLength, na.rm = TRUE),
            f   = max(SubsamplingFactor, na.rm = TRUE),
            d   = max(HaulDuration, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  summarise(groups         = n(),
            matches_R_rule = sum(as.integer(abs(tot - s * f) < 0.5)),
            matches_C_rule = sum(as.integer(abs(tot - s * d / 60) < 0.5)),
            sub_eq_lengths = sum(as.integer(abs(sub - s) < 0.5))) |>
  collect()
```

102 of 113 groups reproduce under the `R` rule and **none at all** under the
`C` rule, and `SubsampledNumber` matches the length rows in 108. So the
`NumberAtLength` here is a measured sub-sample, not an hourly rate: **the
`DataType` label is the wrong one, not the factor.**

**What `{obus}` does.** Applies both multipliers, `SubsamplingFactor` and
`HaulDuration/60`, and flags the records `CNT_C_SUBFACTOR_CONFLICT`. Applying
the duration alone understates them by up to 168 times.

**For ICES.** Is the `DataType` wrong on these hauls, or the factor? They pass
upload validation, and 35 of the 36 carry `HaulValidity` `V`.

## 8. A length row that reports no fish

One Atlantic cod record, and almost every count identity in the format fails on
it at once:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:1992:1:SE:77AR:GOV:50:1", Valid_Aphia == 126436) |>
  summarise(length_rows       = n(),
            rows_reporting_0  = sum(as.integer(NumberAtLength == 0)),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsampledNumber  = max(SubsampledNumber, na.rm = TRUE),
            TotalNumber       = max(TotalNumber, na.rm = TRUE),
            SubsamplingFactor = max(SubsamplingFactor, na.rm = TRUE)) |>
  collect()
```

22 length rows, 15 of which report no fish at that length. The remaining rows
sum to 8, while `SubsampledNumber` claims 30 were measured and `TotalNumber`
claims a catch of 28 — at a factor of 1, so no raising can reconcile them.

Zero-count length rows are rare:

```{r}
dr_con("HL") |>
  filter(NumberAtLength == 0) |>
  summarise(rows = n(), hauls = n_distinct(.id),
            species = n_distinct(Valid_Aphia)) |>
  collect()
```

37 rows over 8 hauls and 14 species, and 32 of them sit on `SpeciesValidity`
`1`, which states that length data is present.

**They do not share one explanation.** The natural guess is a fixed length grid
written out in full, which would put every zero between the smallest and
largest occupied class. Inspecting all 14 haul × species groups that contain
one, that is not what they are: six groups report **nothing but** zeros, so no
fish were measured at any length; in the rest a single zero sits among real
counts, sometimes inside the occupied range and sometimes outside it. Whatever
produces them is not one mechanism, and 37 rows is too few to find it from the
data alone.

**What `{obus}` does.** Drops them, so a length class that contains no fish
cannot enter a length distribution.

**For ICES.** Is a zero `NumberAtLength` a presence marker, a placeholder, or
an error? And on the record above, which of the three counts is right?

## 9. `HaulDuration` of zero or less

Every hourly rate divides by duration, and every `C` per-haul count multiplies
by it:

```{r}
dr_con("HH") |>
  filter(HaulDuration <= 0) |>
  count(HaulDuration, DataType) |>
  collect()
```

217 hauls at zero, 200 of them `C`, and two negative. The exposure to the two
catch tables is far smaller than that, because only 7 of those 219 hauls carry
any HL record at all — a duration cannot spoil a catch that was never reported:

```{r}
dr_con("HH") |>
  filter(HaulDuration <= 0) |>
  inner_join(dr_con("HL") |> distinct(.id), by = ".id") |>
  count(DataType, HaulValidity) |>
  collect()
```

Seven hauls, and every one of them already carries `HaulValidity` `I` — so a
survey excluding invalid hauls never meets them, and the other 212 have no
length record to spoil. This is a question about the format rather than a live
source of error in these two tables.

**What `{obus}` does.** `NA`, never `0` and never `Inf`. Left alone the
arithmetic returns a confident `0` for a `C` haul and `Inf` for every other
kind, and both read as observations rather than as gaps.

**For ICES.** Is a zero `HaulDuration` a missing value that should have been
`-9`? A negative one cannot be anything else — and both pass upload validation.

## Conventions the code resolves once

These are not ambiguous. They are settled in the format, and are listed only
because a consumer reading HL flat will get them wrong.

| Convention | What a flat read does | What `{obus}` does |
|---|---|---|
| `LengthClass` holds millimetres and centimetres in one column, per `LengthCode` | a tenfold error on a length | resolved on reading into `length_mm`, `length_cm` and `accuracy` |
| `LengthClass` is a bin **lower bound**, not a midpoint | biases any weight predicted from it, since \eqn{W = aL^b} is convex | carried as the bound it is; `dr_add_length_mid()` where a midpoint is wanted |
| `-9` marks an absent value | a sentinel enters the arithmetic | nulled once, upstream, per field; never written back |
| One species can appear under several `SpeciesValidity` record types | the same fish counted twice | `SpeciesValidity` is part of what makes an `HL_summary` row unique |
| `DataType` `C` reports `NumberAtLength` as an hourly rate | a rate summed as a count | `n_measured` is `NA` under `C`, never `0` |
| Invalid hauls are present unless excluded | silent inclusion | `haulval` has no default — the survey's call, not the package's |

## The nine questions, together

A list to take to the Data Centre, not a conclusion. Every figure below is
produced by the code above it, and the rows are ordered by reach — how many
surveys write the construct, then how many stations, then how many records.
None of them is a reason to compute the two tables differently; all of them are
reasons to ask what the submission meant.

| | Construct | What the archive does | Surveys | Stations | Records |
|---|---|---|---|---|---|
| 1 | `TotalNumber` over several sex rows | one value repeated in 73,013 groups, one per sex in 131,792 | 28 | 29,118 | 73,013 groups |
| 2 | `SpeciesCategoryWeight` over several sex rows | repeated in 85,210, one per sex in 121,330 | 24 | 26,802 | 85,210 groups |
| 3 | the two routes to a catch total | differ by \eqn{\ge} 1 fish in 133,874 of 1,592,596 groups | 15 | 16,036 | 133,874 groups |
| 4 | `TotalNumber` over several sampling categories | one value shared in 3,362 of 57,285 groups | 14 | 2,378 | 3,362 groups |
| 5 | mandatory `SubsamplingFactor` absent | recoverable in 14,419 groups; implied factor above 1 in 3,968 | 3 | 5,013 | 24,298 rows |
| 6 | `SubsamplingFactor` below 1 | a catch smaller than the sample it came from | 4 | 56 | 372 rows |
| 7 | `DataType` `C` with a factor above 1 | both stated, the factor reaching 167.9 | 1 | 36 | 1,832 rows |
| 8 | `NumberAtLength` of 0 | a length class with no fish in it | 5 | 8 | 37 rows |
| 9 | `HaulDuration` of 0 or less | 217 zero, 2 negative; the 7 carrying records are all invalid hauls | 2 | 7 | 65 rows |

## What remains open

Stated plainly, because a reader is entitled to know the edges of this.

| Open item | Size |
|---|---|
| Can-Mar's whole-fish disagreement — two candidate mechanisms, with a test that would separate them (`PLAN-qc-checks.md` s8) | 31,047 groups |
| The >50 tail after both the duplication and factor-rounding accounts | 165 `TotalNumber`-higher + 43 length-higher, of 443 |
| The 6–50 bucket where neither identity reconciles | ~301 of 3,885 |
| Whether `SubsamplingFactor = SpeciesCategoryWeight / SubsampleWeight` generalises, which would give the weight path its first arithmetic cross-check | untested |
| `DataType` C weight standardisation — potentially a third of the archive, against 0.08% for the weight repeats that remain | untested |
| Whether excluding `SpeciesSex` from the weight key loses anything | untested |
| The intersection of the duplicated length series with the records whose weight is still summed twice — predicted empty | not computed |


# Appendices

Reference and evidence material, moved out of the reading path. Nothing
here is required to follow the argument; all of it is cited from the text
where it matters.


The weight questions that remain share one instrument that has not been built:
**mean weight per fish, `w_haul / n_haul`, stratified by species**. That
quantity is biologically bounded within roughly an order of magnitude, so any
mechanism that scales weight without scaling number leaves a self-normalising
signature — which is what makes the whole family testable at once. The
individual cases, with their tests, are in `PLAN-qc-checks.md` section 8.

## Building CPUEL from `HL_length` {#building-cpuel-from-hl_length}

The clearest way to see that these are the same table is to build one from the
other. CPUEL's payload is numbers per hour by haul, species and length class —
`.id`, `aphia`, `length_mm`, `n_hour` — with `sex` empty throughout.
`HL_length` is finer than that in three ways it keeps on purpose (sex,
development stage and length measurement type), so CPUEL is one aggregation
away:

```{r}
cpuel_like <- dr_con("HL_length") |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1) |>
  group_by(.id, Valid_Aphia, length_mm) |>
  summarise(n_hour = sum(n_hour, na.rm = TRUE), .groups = "drop")
```

Set that against ICES's own file for the same hauls:

```{r}
cpuel <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CPUEL.parquet") |>
  filter(Survey == "NS-IBTS", Year == 2023, Quarter == 1, n_hour > 0) |>
  select(.id, Valid_Aphia = aphia, length_mm, cpuel = n_hour)

cpuel_like |>
  inner_join(cpuel, by = c(".id", "Valid_Aphia", "length_mm")) |>
  summarise(cells = n(),
            agree = sum(ifelse(abs(n_hour - cpuel) < 0.001, 1, 0))) |>
  collect()
```

Over the 317 hauls both sides cover, the rebuild matches **96.6% of CPUEL's
non-zero cells** and agrees with **96.45%** of those to a thousandth of a fish
per hour. A worked haul, three cod:

```{r}
cpuel_like |>
  inner_join(cpuel, by = c(".id", "Valid_Aphia", "length_mm")) |>
  filter(.id == "NS-IBTS:2023:1:DE:06NI:GOV:104:22", Valid_Aphia == 126436) |>
  arrange(length_mm) |>
  collect()
```

The causes of the residual are named, though not quantified: nothing in it has
been traced to the arithmetic. The two files are independent downloads with
their own calculation vintages, so a haul revised on one side and not the
other differs, and this comparison cannot tell that apart from a genuine
disagreement. On BITS two further, documented things apply: CPUEL gear-standardises cod and nothing else and does not
always preserve the submitted length grid — both of which a
generalised product would have to decide about explicitly rather than inherit.

**The zero rows are the part that is not one aggregation away**, and
deliberately so. Half of CPUEL is explicit zeros — 5,874,543 of 11,774,468
rows, each carrying `length_mm = 0` — and they encode a judgement about which
species were in scope for a haul. Add them yourself and the judgement stays
yours:

```{r}
hauls   <- cpuel_like |> distinct(.id) |> collect()
species <- cpuel_like |> distinct(Valid_Aphia) |> collect()    # your universe, not ours

tidyr::crossing(hauls, species) |>
  left_join(collect(cpuel_like), by = c(".id", "Valid_Aphia")) |>
  tidyr::replace_na(list(n_hour = 0, length_mm = 0))
```

On this one survey-quarter that turns 318 hauls and 141 species into 81,557
rows, 38,513 of them explicit zeros — the same roughly-half proportion CPUEL
carries, arrived at from a universe you chose rather than one you inherited.

That is the whole distance between `HL_length` and a CPUEL product covering
29 surveys rather than 8: one `group_by()`, plus a species universe that only a
survey group can properly set.

## Existing analyses run on them unchanged

A product also has to leave standing what is already built on the exchange
tables. So: starting from these two tables instead of the raw HL record, can
the usual chain still be run — numbers at length raised to the haul, summed to
total numbers and weight per haul, divided by swept area and carried through to
a stratified abundance index?

`data-raw/CHECK_datras_adapter.R` answers that against real numbers. It
assembles an object whose HL comes **only** from `HL_length` and `HL_summary`,
never from raw HL, runs [`{DATRAS}`](https://github.com/DTUAqua/DATRAS) and
`{DATRASextra}` on it unmodified, and compares with the same steps run on their
own bundled fixture — NS-IBTS 2020–2023, quarters 1 and 3, dab (*Limanda
limanda*), 2,651 valid hauls. The adapter is about 60 lines: rename to the
legacy field names, add the five HH quantities the packages derive for
themselves, and reshape the two catch tables into the single HL frame they
expect. **No arithmetic on any number** — the raised number at length is
`n_haul` exactly as published.

| Step | Quantity | Difference |
|---|---|---|
| numbers at length, by length class | `N`, 41 classes | same classes, **max gap 0** over 108,691 cells |
| total numbers per haul | `HaulN` | **max gap 0** over 2,651 hauls |
| total weight per haul, from the length–weight lookup | `HaulWgt` | **max gap 0** |
| swept area | `SweptArea` | **max gap 0** |
| stratified abundance index, per hour, by statistical rectangle | index and CV | **identical** |

Eleven checks, no differences (re-run 2026-09-08). The last of them is the
figure a survey working group would report, and it comes out the same to every
digit:

```
  Year    index         cv
  2020 771.7314 0.07114170
  2021 712.8609 0.09926699
  2022 823.4249 0.06559173
  2023 739.3506 0.05788805
```

Two caveats. Species recorded in a haul but never measured enter through
`HL_summary`, carrying no length — which is how the exchange format carries
them, and what a species-richness count needs; a length table alone would lose
them. And **CA is out of scope**: `{obus}` derives no age table, so nothing
age-, maturity- or individual-weight-based is exercised.

Underneath sits the field-by-field comparison,
`data-raw/CHECK_datras_interop.R` — 25 checks, no differences, including the
raised number at length (max gap 2.3 × 10⁻¹³ over 24,960 cells) and the
measured number at length, `NumberAtLength`, against `n_measured` (max gap 0 over
16,196 cells).


## Field names: two layers, both ICES's {#two-naming-layers}

Field names in this codebase come from two places — **both of them ICES's** —
and there is deliberately no third.

| Layer | What it is | Example |
|---|---|---|
| **1. ICES legacy exchange names** | The names the exchange file still carries, and the ones the older ICES documents quoted here use. | `HLNoAtLngt`, `SubFactor`, `CatIdentifier` |
| **2. ICES current names** | ICES's replacement names, published in the current field descriptions. What the raw parquet archive carries, and what every `{obus}` table carries too. | `NumberAtLength`, `SubsamplingFactor`, `SpeciesCategory` |

**The translation between the two layers is ICES's own, not this package's.**
ICES is renaming DATRAS's fields table by table, and publishes the
old-to-new pairs itself in two places: the `getDatrasFieldList` operation on
the DATRAS web service, which returns each current `FieldName` alongside the
`FieldNameOld` it was renamed from, and the *DATRAS New Headers Lookup*
spreadsheet of December 2024 [@ices2024newheaders], which carries the same
pairs in one sheet per record type. Layer 2 is therefore not a scheme `{opus}`
invented and `{obus}` adopted — it is ICES's own target naming, taken up ahead of
the migration's completion. The migration's progress is visible in the service:
HL has flipped, and `getHLdataNewHeaders` returns all 29 of its fields under
the current names, while HH, CA and LT are still served under legacy names
only.

The documentation has already moved ahead of the service: the current field
descriptions name every field in layer 2, so a reader working from ICES's
workbook and a reader working from the `{obus}` tables see the same column names.
Legacy names survive in four places:

- the exchange files themselves;
- the web-service operations that have not yet flipped;
- the older ICES documents quoted later in this article;
- and, in a handful of cells, the current descriptions' own prose — where
  formulas still read `TotalNo=SUM(HLNoAtLngt)` and *"If DataType is C,
  CatIdentifier is always 1"* beneath fields now called `TotalNumber` and
  `SpeciesCategory`.

Quotations below keep whatever names the quoted document uses.

**`{obus}` invents no field names.** `dr_con_raw("HL")` and `dr_con("HL_length")`
both carry `Valid_Aphia` and `SpeciesSex`, so a join between the raw archive and
a derived table needs no name mapping, and the crosswalk below describes every
column in every table.

A shorter `sex` column is a tempting simplification, and it is the wrong one.
ICES's legacy `Sex` maps to `SpeciesSex` in HL but to `IndividualSex` in CA —
two genuinely different measurements, one a property of a length-frequency
record and the other of an individual fish. A single `sex` column flattens
them.

### The legacy-to-current crosswalk

The mapping is neither hand-authored here nor taken from ICES's published
lists on trust. `{opus}` checks every pair ICES publishes against the field names
the corresponding operation actually returns, keeps what survives
(`opus::op_datras_field_list()`), reduces that to a collision-checked 1:1 map
(`opus::op_datras_rename_crosswalk()`), and embeds the result in the raw
parquet footer. `{obus}` reads it back from the same file as the data, so it
cannot describe a different vintage than the data it accompanies:

```{r}
opus::op_crosswalk("HL")
```

29 rows — two more than either ICES list. Neither `getDatrasFieldList` nor the
December 2024 lookup documents HL's `DateofCalculation` or `Valid_Aphia` under
any name, though both are real fields that `getHLdata` and
`getHLdataNewHeaders` alike return; they appear here unrenamed, which is what
those two operations show them to be.

Check names against this output rather than from memory. Most pairs
are guessable; at least one is not — **`LenMeasType` maps to `LengthType`**,
which is easy to mistake for a length *unit* field when it is a measurement
*convention* field. `Sex` → `SpeciesSex` is the pair worth memorising, because ICES's one legacy
`Sex` splits across two current names — `SpeciesSex` in HL and
`IndividualSex` in CA — so it is the clearest case of why the crosswalk is
per table rather than global.


## An engineering note: R and SQL disagree about `NA`

Not a DATRAS matter, but the single largest source of silent defects in this
code. `{obus}` runs the same functions eagerly (data frames, in the tests) and
lazily (DuckDB, in the build), and the two backends differ in four ways.
Anyone extending these functions needs all four.

**1. Joining.** SQL treats `NULL = NULL` as unknown, so a lazy join drops
every row with an `NA` key — and `SpeciesSex` is `NA` in 54% of HL, with
`SpeciesCategory`, `SpeciesValidity` and `TotalNumber` able to be `NA` too.
Every join in `dr_HL_summary()` whose key can hold `NA` carries
`na_matches = "na"`, which dbplyr renders as `IS NOT DISTINCT FROM` and DuckDB
supports natively. Without it, every `NA`-sex row fails to find its own
length-derived expectation, never reconciles, falls to the fallback, and
undercounts. Confirmed on a real case
(`BITS:1999:1:DK:26HI:TVS:017492:22`, aphia 127143): true total 4, and 3
without the flag.

**2. Aggregating over an all-`NA` group.** `sum(x, na.rm = TRUE)` is `0` in R
and `NULL` in SQL. Every aggregation here carries an explicit non-`NA` counter
(`.n_ok`, `.h_ok`, `.w_ok`, `.wh_ok`) and forces `NA` when the counter is
zero, so both backends agree. Without it the eager path silently converts a
deliberate `NA` — a missing `SubsamplingFactor`, a non-positive
`HaulDuration`, a species that was never weighed — straight back into the
confident `0` it was meant to prevent.

**3. Self-reference inside `summarise()`.** dplyr evaluates expressions in
order, so `n_ok = sum(!is.na(n_haul))` written *after* `n_haul = sum(n_haul)`
sees the new scalar; SQL resolves every aggregate against the source column.
`dr_HL_length()` renames to `.n_raw`/`.h_raw` before aggregating, so the two
agree by construction rather than by luck.

**4. String concatenation with `NA`.** On a plain data frame R's `paste(sep =)`
renders an `NA` field as the literal two-character string `"NA"`; on a
`tbl_lazy`, dbplyr translates the same call to DuckDB's `CONCAT_WS()`, which
drops `NA` fields instead. `.id` is the join key between HH, HL and CA, so it
must come out identical whichever backend built it. `.dr_concat_ws()` builds
the `NA`-skipping behaviour explicitly from `is.na()`, `case_when()` and
`paste0()`, which translate the same way on both — with one deliberate
departure from raw `CONCAT_WS()`: when every field is `NA` it returns `NA`,
not `CONCAT_WS()`'s `""`, because an empty string would collide every such row
onto one `.id`.

Each of these produced a real, silent, wrong number at some point. Keeping
both an eager test suite and a lazy build is what catches them; neither does
alone.


## Why the two routes disagree

[Part III](#two-routes-to-a-catch-total) shows that the submitted `TotalNumber`
and the raised length rows differ by a fish or more in 133,874 groups. Most of
that has one mechanism, and it is arithmetic done before the file was
submitted.

Define `F = TotalNumber / SubsampledNumber` — the whole multiplier the
submitter actually used, read straight off the record with nothing
reconstructed. Split `DataType` `C` groups by whether that multiplier is a
whole number:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "C", Survey != "Can-Mar", !is.na(LengthClass),
         NumberAtLength != 0, !is.na(TotalNumber),
         !is.na(SubsampledNumber), SubsampledNumber > 0) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(s   = sum(NumberAtLength, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  mutate(whole_number_factor = abs(tot / sub - round(tot / sub)) < 1e-9) |>
  group_by(whole_number_factor) |>
  summarise(groups = n(),
            whole_fish_gap = sum(as.integer(abs(tot - s) >= 1)), .groups = "drop") |>
  mutate(rate_pct = round(100 * whole_fish_gap / groups, 2)) |>
  collect()
```

0.19% against 29.49% — a relative risk of about 150, using no reconstruction
whatsoever. A non-integer multiplier is the single strongest predictor of a
whole-fish disagreement in the archive.

The mechanism follows. If the submitter raised each length row separately,
`round(NumberAtLength / F)` should recover the raw integer counts, and those
should sum to the `SubsampledNumber` the record reports independently:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "C", Survey != "Can-Mar", !is.na(LengthClass),
         NumberAtLength != 0, !is.na(TotalNumber),
         !is.na(SubsampledNumber), SubsampledNumber > 0) |>
  mutate(F = TotalNumber / SubsampledNumber) |>
  filter(F > 0) |>
  mutate(m = round(NumberAtLength / F)) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(F = max(F, na.rm = TRUE), sum_m = sum(m, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  group_by(whole_number_factor = abs(F - round(F)) < 1e-9) |>
  summarise(groups = n(),
            reconstruction_matches = sum(as.integer(abs(sum_m - sub) < 0.5)),
            .groups = "drop") |>
  collect()
```

It does, in 86% of the non-integer groups. So the submitter computed
`NumberAtLength = round(m * F)` row by row while computing `TotalNumber` once
over the whole sample. The per-row rounding residuals do not cancel, and their
sum is the gap. **The fish are missing from the submitted file before `{obus}`
reads it**, which is why no choice in the derivation can recover them.

## Can-Mar

Can-Mar behaves differently enough to be worth separating. The article used to
record two candidate mechanisms and no way to choose between them: a raising
factor lost in a provider-side conversion, or fish counted into `TotalNumber`
that never entered a length row. They make different predictions, so they can
be told apart.

**A lost conversion would apply to a whole haul.** If a factor went missing in
processing, every species in that haul should show the same ratio of reported
total to measured lengths. Take Can-Mar hauls where three or more species show
a gap at `SubsamplingFactor` 1, and ask how many carry one shared ratio:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(Survey == "Can-Mar", DataType %in% c("R", "P", "S"),
         !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia) |>
  summarise(s = sum(NumberAtLength, na.rm = TRUE),
            f = max(SubsamplingFactor, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(f), f == 1, s > 0, tot - s >= 1) |>
  group_by(.id) |>
  summarise(species = n(), distinct_ratios = n_distinct(round(tot / s, 3)),
            .groups = "drop") |>
  filter(species >= 3) |>
  summarise(hauls                 = n(),
            all_species_one_ratio = sum(as.integer(distinct_ratios == 1)),
            any_two_agree         = sum(as.integer(distinct_ratios < species)),
            every_ratio_different = sum(as.integer(distinct_ratios == species))) |>
  collect()
```

Not one haul in 4,523 has a single ratio across its species. In 4,225 of them
every species has a different one, and in only 298 do even two species agree.
**The lost-factor reading is refuted**, at least as anything applied at haul
level.

**The other mechanism predicts whole fish.** Fish counted but never measured
leave an integer gap, and they leave the record's own sub-sample field siding
with the length rows rather than with the total:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(Survey == "Can-Mar", DataType %in% c("R", "P", "S"),
         !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia) |>
  summarise(s = sum(NumberAtLength, na.rm = TRUE),
            f = max(SubsamplingFactor, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(f), f == 1, s > 0, tot - s >= 1) |>
  summarise(groups         = n(),
            gap_whole_fish = sum(as.integer(abs((tot - s) - round(tot - s)) < 1e-9)),
            sub_eq_lengths = sum(as.integer(abs(sub - s) < 0.5)),
            sub_eq_total   = sum(as.integer(abs(sub - tot) < 0.5)),
            gap_of_1       = sum(as.integer(tot - s == 1)),
            gap_over_10    = sum(as.integer(tot - s > 10))) |>
  collect()
```

Every one of the 25,953 gaps is a whole number of fish — unlike the rounding
mechanism above, which leaves fractional residues. Where `SubsampledNumber` is
present it matches the length rows in 10,183 groups and the total in **none**.
So the record's own account of what was measured corroborates the lengths, and
`TotalNumber` alone carries the extra fish: exactly one fish in 4,239 groups,
more than ten in 12,647.

**What is left open.** This refutes a haul-level conversion and fits fish
counted but not measured. It does not exclude a factor lost per species, though
the ratios tell against that too: they concentrate on 4/3, 3/2, 5/4, 6/5 and
7/6, which are what a handful of unmeasured fish produce and implausible as
sub-sampling factors — nobody sub-samples a catch at eight sevenths. Treat
Can-Mar separately and prefer the reported total.

**For ICES.** Does Can-Mar record fish in `TotalNumber` that are deliberately
not length-measured, and if so, is that intended under the format?

## Duplicated length series {#duplicated-length-series}

This is the one place where `{obus}`'s choice of what makes a row unique
measurably moves the number. One BITS record, one species, and the whole length
series written twice under two `LengthType` codes:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:2017:4:DK:26HF:TVS:552:42", Valid_Aphia == 127082) |>
  group_by(LengthType) |>
  summarise(length_rows       = n(),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsampledNumber  = max(SubsampledNumber, na.rm = TRUE),
            TotalNumber       = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Two identical series. Every one of the 19 length classes appears under both
codes carrying the same count, and both copies report the same
`SubsampledNumber` 57, the same `TotalNumber` 57 and the same 4500 g.
`dr_HL_length()` treats `LengthType` as part of what makes a row unique, so it
adds both copies and returns 114 where `TotalNumber` reports 57 — a ratio of
exactly 2.

How far this reaches:

```{r}
dr_con("HL_length") |>
  group_by(.id, Valid_Aphia, SpeciesValidity, SpeciesSex, DevelopmentStage,
           length_mm, accuracy) |>
  summarise(length_types = n_distinct(LengthType),
            distinct_counts = n_distinct(n_measured), .groups = "drop") |>
  filter(length_types > 1, distinct_counts == 1) |>
  summarise(duplicated_rows = n(),
            haul_species_groups = n_distinct(paste0(.id, "|", Valid_Aphia))) |>
  collect()
```

**These look like coding artefacts rather than two real measurement
conventions.** `LengthType` codes are supposed to be different length
definitions, so the same fish measured both ways would give *different*
`LengthClass` values. Series identical in length class, count, sub-sample,
total and weight cannot be two genuine measurements of one catch. *What would
settle it:* confirming with ICES that the codes are non-convertible
measurement definitions.

`{obus}` does not merge them, because `LengthType` genuinely splits other
records, and merging on a hypothesis would lose those.

## The same weight written on two category codes

Keeping the category is right when two categories are two separately weighed
parts of the catch. It is wrong when the same weight has simply been written on
two codes — and under **pseudo-category sampling (`DataType` P) that is the
documented convention**, not an error. The pseudo-category memo states
*"CatCatchweight value is same in each subcategory"*, and the archive agrees:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "P", !is.na(SpeciesCategory)) |>
  mutate(main_category = substr(SpeciesCategory, 1, 1)) |>
  distinct(.id, Valid_Aphia, SpeciesValidity, main_category,
           SpeciesCategory, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, main_category) |>
  summarise(sub_categories = n_distinct(SpeciesCategory),
            distinct_weights = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(sub_categories > 1) |>
  summarise(main_categories = n(),
            one_shared_weight = sum(as.integer(distinct_weights == 1))) |>
  collect()
```

1,456 of 1,464 — 99.5% carry a single identical weight across their
sub-categories, which is the convention the memo describes. The structural reason is in the sampling
itself — under P the raising factor *is* category weight divided by sample
weight, so the category weight must appear on every sub-category row of that
category.

**`{obus}` therefore collapses P weights on the main category**, not on the
reported code. Under P the `SpeciesCategory` is two digits, the first being the
main category, so `11`, `12` and `13` are one weighing and not three. Take the
worked case that used to be wrong:

```{r}
dr_con_raw("HL") |>
  dr_add_id() |>
  filter(.id == "NS-IBTS:2016:1:GB-SCT:748S:GOV:12:12", Valid_Aphia == 126437) |>
  distinct(SpeciesCategory, TotalNumber, SpeciesCategoryWeight) |>
  collect()
```

Both sub-categories carry 105,300 g, which is one catch weighed once.
`HL_summary` returns it once:

```{r}
dr_con("HL_summary") |>
  filter(.id == "NS-IBTS:2016:1:GB-SCT:748S:GOV:12:12", Valid_Aphia == 126437) |>
  select(n_totalnumber, w_haul) |> collect()
```

Left uncollapsed this one haul would report 210,600 g for a 105,300 g catch,
and archive-wide the P repeats were worth 410,122,351 g — close to 29% of all
`DataType` P weight mass, because sub-category sampling is used precisely on
the big catches. The records are still marked, as `WGT_CAT_REPEAT_P` in
`dr_con("hl_flag")`, with `kind` `property`: the flag records that the
submission carries the repetition, not that the published weight is wrong.

**What is deliberately not collapsed.** The same shape occurs under `DataType`
R, where ICES documents no sub-category convention at all, and there the cause
is not established: of 209 such records, 135 are a single platform and year
(NS-IBTS 2025 Q3, GB, platform 74E9), and among those where `TotalNumber` also
differs some ties look implausible (two categories both at exactly 9,992 g) and
some entirely plausible (two tiny catches both at 12 g). Collapsing them would
assert a convention no document states, so `{obus}` sums them and flags them
`WGT_CAT_REPEAT_R`, `kind` `unexplained`. A further 36 P records share one
weight *across* main categories, which the memo does not describe either.

Together those are the whole of the remaining excess. Set the published
`w_haul` against the alternative of collapsing every repeated weight whatever
the `DataType`:

```{r}
published <- dr_con("HL_summary") |>
  group_by(.id, Valid_Aphia, SpeciesValidity) |>
  summarise(w_pub = sum(w_haul, na.rm = TRUE), .groups = "drop")

collapsed <- dr_con_raw("HL") |>
  dr_add_id() |>
  inner_join(dr_con_raw("HH") |> dr_add_id() |> select(.id, DataType), by = ".id") |>
  filter(DataType != "C") |>
  distinct(.id, Valid_Aphia, SpeciesValidity, SpeciesCategoryWeight, DataType) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, DataType) |>
  summarise(w_alt = sum(SpeciesCategoryWeight, na.rm = TRUE), .groups = "drop")

published |>
  inner_join(collapsed, by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  filter(w_pub > w_alt + 0.5) |>
  group_by(DataType) |>
  summarise(records = n(), excess_g = sum(w_pub - w_alt)) |>
  collect()
```

**18,513,763 g in all — 0.080% of the archive's `w_haul` mass** of
23,263,243,838 g. Before the P collapse the same comparison returned
428,460,512 g, or 1.84%. The flag table marks 1,491 P and 209 R records.

**The numbers are unaffected either way.** For all 1,491 P repeated-weight
records the underlying length frequencies **differ**, so the mechanism that
repeats a weight never touches `n_haul`. Conversely the weight path ignores
`DevelopmentStage` and `LengthType`, so the [duplicated length
series](#duplicated-length-series) that inflate `n_haul` cannot
inflate `w_haul` there. The two affected sets are expected to be disjoint; that
has **not been computed**, and a non-empty result would be a new finding.


## How this article was made {#how-this-article-was-made}

**What counts as evidence.** Every claim is either a quotation from an ICES
document or a measurement of the archive, and the measurements are given as
code that reads the published tables directly, so they need nothing but the
package.

**A note on the numbers, because DATRAS grows.** The structural counts — how
many rows each table has, how many haul x species groups, how many of those
carry no length data — are **measured when this article is rendered**, not
typed in, so they cannot drift apart from each other or from the archive. The
analytical figures are different: the CPUEL reconciliation, the rounding
regimes, the confidence intervals took real work to establish, and re-running
them on every render would embed whole analyses in a reference article. Those
are stated once, as dated evidence, and the claims they support are written in
the form that survives the archive growing — *only*, *exactly one*, *never* —
so that if one ever stops being true, that is a finding rather than a stale
number.

**On the package.** [`{obus}`](https://github.com/einarhjorleifsson/obus) is a
reference implementation of the proposal and a demonstration that it holds for
the whole archive rather than part of it. Every rule it applies is quoted from
an ICES document; every figure is reproducible from the published archive with
the code shown. The nearest thing to a proof of the proposal is that [CPUEL
itself can be rebuilt from `HL_length`](#building-cpuel-from-hl_length),
worked through above.

## References {.unnumbered}

::: {#refs}
:::
