---
title: "Catch products: from raw exchange data to survey indices"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4
)
```

Three functions cover most survey-index workflows:

| Function | Input | Output | Zero-fill |
|---|---|---|---|
| `dr_catch_by_length()` | raw HH + HL | one row per haul × species × length bin | no |
| `dr_catch_by_haul()` | output of above | one row per haul × species | yes |
| `dr_expand_length()` | output of above | one row per haul × species × length bin | yes |

The typical pipeline is:

```
dr_catch_by_length(hh, hl)  →  filter  →  dr_catch_by_haul(hh)
                                        ↘  dr_expand_length(hh)
```

`dr_catch_by_length()` reads the `NumberAtLength` field and applies
DataType-aware subsampling corrections. It returns only observed rows — no
zeros. The downstream functions add the zero hauls needed for unbiased CPUE
estimation.

```{r load}
library(obus)
library(dplyr)
library(ggplot2)

hh  <- dr_con("HH")
hl  <- dr_con("HL")

# Build the base catch table once (lazy); filter downstream
cbl <- dr_catch_by_length(hh, hl, haulval = "V")
```


---

## NS-IBTS Q1 — North Sea demersal survey

### Why zero hauls matter

Species are absent from many hauls. Dropping those zeros inflates the mean CPUE
and makes trends unreliable. `dr_catch_by_haul()` adds an explicit zero row for
every haul where a species was not caught (within the same Survey/Year/Quarter
stratum).

```{r ns_ibts_zero_demo}
species_ns <- c("Atlantic herring", "Atlantic cod")

cpue_ns <- cbl |>
  filter(Survey == "NS-IBTS", Quarter == 1, species %in% species_ns) |>
  dr_catch_by_haul(hh) |>
  collect()

cpue_ns_nozero <- cpue_ns |> filter(n_hour > 0)

bind_rows(
  cpue_ns        |> mutate(zeros = "with zero hauls"),
  cpue_ns_nozero |> mutate(zeros = "without zero hauls")
) |>
  ggplot(aes(Year, n_hour, colour = zeros)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", size = 0.3) +
  facet_wrap(~species, scales = "free_y") +
  scale_colour_manual(values = c("with zero hauls" = "#2166ac",
                                 "without zero hauls" = "#d73027")) +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = NULL, colour = NULL,
       title = "NS-IBTS Q1 — effect of zero-haul inclusion on CPUE") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Including zero hauls consistently lowers the mean and widens the confidence
interval — the difference is most pronounced in years with low prevalence.

### Length distributions for selected years

`dr_expand_length()` propagates zero observations across the full length × haul
grid, enabling length-structured indices. Here we show cod length distributions
for three decades.

```{r ns_ibts_length}
years_sel <- c(1985, 1995, 2005, 2015)

cbl |>
  filter(Survey == "NS-IBTS", Quarter == 1, species == "Atlantic cod",
         Year %in% years_sel) |>
  dr_expand_length(hh) |>
  collect() |>
  ggplot(aes(length_cm, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon",
               aes(ymin = after_stat(ymin), ymax = after_stat(ymax)),
               alpha = 0.25, fill = "#2166ac") +
  stat_summary(fun = mean, geom = "line", colour = "#2166ac") +
  facet_wrap(~Year, scales = "free_y") +
  labs(x = "Length (cm)", y = "Numbers per hour (mean ± 95 % boot CI)",
       title = "NS-IBTS Q1 — Atlantic cod length distribution") +
  theme_bw()
```

---

## BTS Q3 — Southern North Sea flatfish survey

BTS covers the Dutch, Belgian and English coastal waters in summer. Common dab
and plaice dominate; this survey also records seahorses consistently.

```{r bts_cpue}
species_bts <- c("common dab", "European plaice", "common sole")

cbl |>
  filter(Survey == "BTS", Quarter == 3, species %in% species_bts) |>
  dr_catch_by_haul(hh) |>
  collect() |>
  ggplot(aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               colour = "#1b7837", size = 0.3) +
  facet_wrap(~species, scales = "free_y") +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = NULL,
       title = "BTS Q3 — flatfish CPUE indices") +
  theme_bw()
```

### Seahorse — a rare bycatch indicator

Seahorses (*Hippocampus* spp.) are rarely caught but their occurrence in trawl
surveys is used as an indicator of seagrass habitat condition. BTS has the most
consistent records in DATRAS.

```{r bts_seahorse}
seahorse_aphia <- cbl |>
  filter(Survey == "BTS", grepl("Hippocampus", latin)) |>
  distinct(aphia, latin, species) |>
  collect()

cbl |>
  filter(Survey == "BTS", aphia %in% !!seahorse_aphia$aphia) |>
  dr_catch_by_haul(hh) |>
  collect() |>
  mutate(present = n_hour > 0) |>
  group_by(Year, species) |>
  summarise(prevalence = mean(present),
            mean_cpue  = mean(n_hour),
            .groups = "drop") |>
  ggplot(aes(Year)) +
  geom_col(aes(y = prevalence), fill = "#c2a5cf", alpha = 0.7) +
  geom_line(aes(y = mean_cpue / max(mean_cpue) * max(prevalence)),
            colour = "#762a83", linewidth = 0.8) +
  facet_wrap(~species) +
  scale_y_continuous(
    name = "Proportion of hauls with catch (bars)",
    sec.axis = sec_axis(~ . * 1, name = "Scaled mean CPUE (line)")
  ) +
  labs(x = NULL,
       title = "BTS Q3 — seahorse prevalence and relative CPUE") +
  theme_bw()
```

---

## SP-ARSA Q1 — Southern Iberian Atlantic survey

SP-ARSA covers the Gulf of Cádiz (southern Spain / northern Morocco). The small
pelagic and demersal communities here differ markedly from the North Sea.

```{r sparsa_cpue}
species_sp <- c("anchovy", "European hake")

cbl |>
  filter(Survey == "SP-ARSA", Quarter == 1, species %in% species_sp) |>
  dr_catch_by_haul(hh) |>
  collect() |>
  ggplot(aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               colour = "#b35806", size = 0.3) +
  facet_wrap(~species, scales = "free_y") +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = NULL,
       title = "SP-ARSA Q1 — anchovy and hake CPUE") +
  theme_bw()
```

### Hake length distribution

European hake is a key commercial species in Iberian waters. Length
distributions illustrate how recruitment varies between years.

```{r sparsa_length}
years_sp <- c(2005, 2010, 2015, 2020)

cbl |>
  filter(Survey == "SP-ARSA", Quarter == 1, species == "European hake",
         Year %in% years_sp) |>
  dr_expand_length(hh) |>
  collect() |>
  ggplot(aes(length_cm, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon",
               aes(ymin = after_stat(ymin), ymax = after_stat(ymax)),
               alpha = 0.25, fill = "#b35806") +
  stat_summary(fun = mean, geom = "line", colour = "#b35806") +
  facet_wrap(~Year, scales = "free_y") +
  labs(x = "Length (cm)", y = "Numbers per hour (mean ± 95 % boot CI)",
       title = "SP-ARSA Q1 — European hake length distribution") +
  theme_bw()
```

---

## How fast is it?

Ten of the most common NS-IBTS Q1 species — full CPUE pipeline from raw
exchange data through to a collected data frame, timed with `system.time()`.

```{r timing}
species_10 <- c(
  "Atlantic herring", "European sprat", "Norway pout",
  "whiting", "haddock", "common dab",
  "Atlantic cod", "grey gurnard", "Atlantic mackerel",
  "European plaice"
)

t <- system.time(
  cpue_10 <- cbl |>
    filter(Survey == "NS-IBTS", Quarter == 1, species %in% species_10) |>
    dr_catch_by_haul(hh) |>
    collect()
)

cat(sprintf("10 species, all NS-IBTS Q1 years: %.1f s\n", t["elapsed"]))
```

```{r timing_plot, fig.height=6}
cpue_10 |>
  mutate(species = factor(species, levels = species_10)) |>
  ggplot(aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               colour = "#2166ac", size = 0.2) +
  facet_wrap(~species, scales = "free_y", ncol = 2) +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)", x = NULL,
       title = "NS-IBTS Q1 — ten species CPUE indices") +
  theme_bw()
```

---

## A note on `dr_expand_length()` and large selections

::: {.callout-warning}
## Filter before expanding

`dr_expand_length()` builds a **complete haul × species × length-bin grid** for
every combination present in the Survey/Year/Quarter stratum. The output size
grows as:

$$N_\text{hauls} \times N_\text{species} \times N_\text{length bins}$$

Filtering to one species before calling `dr_expand_length()` is fast. Passing
many species — or worse, an unfiltered `cbl` across multiple surveys — will
produce a very large table and may exhaust memory or take minutes to collect.

**Rules of thumb:**

- Always filter to a single survey, quarter, and a small set of species before
  calling `dr_expand_length()`.
- If you only need haul-level totals (no length structure), use
  `dr_catch_by_haul()` instead — it never materialises the length grid and is
  substantially faster.
- When working with many species, iterate species-by-species rather than passing
  them all at once.

```r
# Avoid — expands the full grid for all species across all surveys
cbl |> dr_expand_length(hh) |> collect()

# Correct — filter to one species first
cbl |>
  filter(Survey == "NS-IBTS", Quarter == 1, species == "Atlantic cod") |>
  dr_expand_length(hh) |>
  collect()
```
:::

