---
title: "DATRAS data issues"
subtitle: "What the archive's records cannot settle on their own — a register for data providers, survey groups and the ICES Data Centre."
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    citations-hover: true
    reference-location: document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE,
                      message = FALSE, warning = FALSE)
options(width = 120, pillar.sigfig = 7)
# Every number the prose quotes is computed on render and referenced inline,
# so the text cannot drift from the archive (Working Principle 8). When a
# provider resubmits a record, the next render shows it; that is the point.
N <- function(x) formatC(as.numeric(x), format = "d", big.mark = ",")
P <- function(x, digits = 0) sprintf(paste0("%.", digits, "f%%"), x)
```

```{r}
library(obus)
library(dplyr)
library(ggplot2)
```

This article lists the places where a DATRAS record says something it cannot
mean, or could mean two things, and what `{obus}` does with each. It is
written for the people who can act on them: the national institutes that
submit, the survey groups that set the conventions, and the ICES Data Centre
that hosts the archive. An analyst who wants catch tables rather than their
history should start from [Two catch tables for DATRAS](catch-tables.html):
the tables there already handle everything below that can be handled.

**Most of DATRAS needs no entry here.** The great majority of fields mean what
the field descriptions say they mean, and are used that way. What follows is
the minority — and much of it is less a list of mistakes than the cost of one
record shape. HL gives a submitter one kind of row for three different tallies,
so a total that belongs to a species has nowhere of its own to go and is
written onto every length row; and no institute ever exercises more than a
corner of the format it submits to ([Part I of the catch-tables
article](catch-tables.html#what-is-wrong-with-hl) measures both). Writing one
total across several sex rows is close to what the record shape asks for.

**`{obus}` flags; it does not correct.** Where a submission writes one number
in a way that could mean two things, no derivation — ours, the `{DATRAS}`
package's, or ICES's own — can recover which was meant. Only the programme
that made the submission still holds the sampling sheets behind it, so the fix
is at source, and every entry below ends with the question that would settle
it and who can answer it.

**Every number here is measured when the page is rendered.** When a record is
corrected and resubmitted, the next render shows the difference, so the page
doubles as a status report on the archive it describes.

## How much of the archive this concerns

`{obus}` marks every catch record touched by anything below in
`dr_con("hl_flag")`, one row per record and issue, and says what kind of
thing each flag records. The kind is not a severity: `property` is the format
behaving as documented, `intrinsic` is arithmetic that cannot come out exact,
`suspect` is a recognisable defect, and `unexplained` is a disagreement with no
established cause. Counting only the last two:

```{r}
reach <- dr_con("hl_flag") |>
  left_join(dr_con("hl_flag_code") |> select(code, kind), by = "code") |>
  filter(kind %in% c("suspect", "unexplained")) |>
  summarise(records_affected = n_distinct(paste0(.id, "|", Valid_Aphia)),
            hauls_affected   = n_distinct(.id)) |>
  collect()
reach

totals <- dr_con("HL_summary") |>
  summarise(records = n_distinct(paste0(.id, "|", Valid_Aphia)),
            hauls   = n_distinct(.id)) |>
  collect()
```

`r N(reach$records_affected)` of `r N(totals$records)` haul-species records
(`r P(100 * reach$records_affected / totals$records, 1)`), in
`r P(100 * reach$hauls_affected / totals$hauls)` of hauls. For the rest, the
records are what they appear to be.

## Finding your own records

The flag table joins to the haul table on `.id`, so any submitter can list
their own flagged records — by survey, country and year — and see which issue
each carries:

```{r, eval = FALSE}
dr_con("hl_flag") |>
  left_join(dr_con("hl_flag_code") |> select(code, kind, label), by = "code") |>
  inner_join(dr_con("HH") |> select(.id, Survey, Year, Country), by = ".id") |>
  filter(Country == "DK", Survey == "BITS") |>       # your country and survey
  count(Year, code, label) |>
  collect()
```

Across countries, the share of each country's records that carries a suspect
or unexplained flag, and how much of that share is one single issue:

```{r}
hh_country <- dr_con("HH") |> select(.id, Country)
flagged <- dr_con("hl_flag") |>
  left_join(dr_con("hl_flag_code") |> select(code, kind), by = "code") |>
  filter(kind %in% c("suspect", "unexplained")) |>
  inner_join(hh_country, by = ".id")

by_country <- dr_con("HL_summary") |>
  distinct(.id, Valid_Aphia) |>
  inner_join(hh_country, by = ".id") |>
  count(Country, name = "records") |>
  left_join(flagged |> distinct(Country, .id, Valid_Aphia) |> count(Country, name = "flagged"),
            by = "Country") |>
  left_join(flagged |> count(Country, code) |>
              group_by(Country) |>
              mutate(share = n / sum(n)) |>
              filter(n == max(n)) |>
              ungroup() |>
              select(Country, commonest_issue = code, share),
            by = "Country") |>
  collect() |>
  mutate(flagged = coalesce(flagged, 0),
         per_1000_records = round(1000 * flagged / records, 1),
         commonest_issue_share = P(100 * share)) |>
  select(Country, records, per_1000_records, commonest_issue, commonest_issue_share) |>
  arrange(desc(per_1000_records))
by_country
```

```{r, include = FALSE}
rate <- function(cc) by_country$per_1000_records[by_country$Country == cc]
```

**Read the rate together with the last column.** A high rate made of one issue
is one consistent practice — a convention applied the same way to every
record, which is handled by one rule and settled by one answer. A low rate
spread over many issues is case-by-case variation. The table is a map of where
the questions below land, not a ranking. Canada's rate
(`r rate("CA")` per thousand records) is one raising-factor practice
([B5](#can-mar)), and Lithuania's (`r rate("LT")`) one `DataType` label
([B2](#c-with-a-factor-above-1)).

## How each entry is laid out

Each entry shows **a real record**, so its shape is visible; **how far it
reaches**, measured; **whether it moves a product** — numbers or biomass per
haul, a length distribution, an index — and could produce a trend or contrast
that is not there; **what `{obus}` does**, and the `hl_flag` code that marks
it; and **the question**, addressed to whoever can answer it. The entries are
grouped by who that is. [The register at a glance](#at-a-glance) collects
them, with their status.

# A. Questions about the specification, for ICES

These are places where the format, or ICES's own documents about it, leave a
reading open. No single submitter can settle them.

## A1. What `TotalNumber` means {#totalnumber-definition}

**ICES's documents disagree.** The field description defines the total as the
sum of the measured numbers, `TotalNo=SUM(HLNoAtLngt)` [@ices2026fielddesc].
The DATRAS FAQ instructs that the total equal the measured numbers raised by
the sub-sampling factor [@ices2014datrasfaqs], and the IBTSWG traced recent
submission errors to specifications it called ambiguous, asking for them to be
made unambiguous [@ices2023ibtswg]. The field description's own example file
takes the raised reading. The archive does too:

```{r}
tn_def <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), !is.na(LengthClass),
         NumberAtLength != 0, !is.na(TotalNumber)) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex,
           DevelopmentStage) |>
  summarise(measured    = sum(NumberAtLength, na.rm = TRUE),
            factor_lo   = min(SubsamplingFactor, na.rm = TRUE),
            factor_hi   = max(SubsamplingFactor, na.rm = TRUE),
            TotalNumber = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  filter(factor_lo == factor_hi, factor_hi >= 2) |>     # one factor, and the two readings far apart
  summarise(groups              = n(),
            measured_x_factor   = sum(as.integer(abs(TotalNumber - measured * factor_hi) < 0.5)),
            measured_only       = sum(as.integer(abs(TotalNumber - measured) < 0.5))) |>
  collect()
tn_def
```

Where a single factor of 2 or more separates the two readings, the total is the
raised number in `r P(100 * tn_def$measured_x_factor / tn_def$groups)` of
groups and the plain measured count in `r N(tn_def$measured_only)` of
`r N(tn_def$groups)`. So the field description's formula is the one that does
not describe the data.

**The definition also changed, and the data carry no marker of when.** The
March 2024 field descriptions [@ices2024marfielddesc] turned the total from one
figure per haul and species — repeated across its categories and sexes, and so
not to be added up — into one figure per category, which is. Survey years from
2024 can only have been submitted under the new definition, so if submitters
had followed it, a total shared across a species' categories would become rarer
there. It does not:

```{r}
shared_by_period <- dr_con("HL") |>
  distinct(.id, Year, Valid_Aphia, SpeciesValidity, SpeciesCategory, TotalNumber) |>
  group_by(.id, Year, Valid_Aphia, SpeciesValidity) |>
  summarise(categories = n_distinct(SpeciesCategory),
            totals     = n_distinct(TotalNumber), .groups = "drop") |>
  filter(categories > 1) |>
  mutate(period = if_else(Year >= 2024, "2024 onwards", "before 2024")) |>
  group_by(period) |>
  summarise(multi_category = n(),
            one_total_shared = sum(as.integer(totals == 1)), .groups = "drop") |>
  collect() |>
  mutate(pct_shared = round(100 * one_total_shared / multi_category, 1))
shared_by_period
```

**Does it move a product?** It decides whether the two routes to a haul total
should agree ([A6](#two-routes-to-a-catch-total)), and whether a total may be
added across categories ([A4](#totalnumber-across-categories)) — for any
species caught in more than one category, the difference between counting it
once and counting it twice.

**What `{obus}` does.** Publishes the submitted total as `n_totalnumber` and
the raised length frequencies as `n_haul`, side by side, and decides between
category readings record by record by checking each category's own arithmetic
([A4](#totalnumber-across-categories)).

**For ICES.** Which definition is authoritative, and from what date? And can
the archive say, per record, which definition it was submitted under?

## A2. One `TotalNumber` written across several sex rows {#totalnumber-across-sex-rows}

A haul, a species and a sampling category can carry several rows because the
catch was sexed. `TotalNumber` is written on each. Here is a plaice record
where each sex reports its own:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:1999:1:DK:26HI:TVS:017492:22", Valid_Aphia == 127143) |>
  group_by(SpeciesSex) |>
  summarise(TotalNumber        = max(TotalNumber, na.rm = TRUE),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsamplingFactor  = max(SubsamplingFactor, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Three rows; each `TotalNumber` matches that sex's own measured lengths at a
factor of 1. The haul total is 4. But two of the three values are the same
number, so a consumer that removes duplicates gets 3 — the unsexed row and the
male row agree by coincidence, not because one is a copy of the other.

Other submissions do the opposite, and write one whole-category total on every
sex row:

```{r}
sex_rows <- dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage,
           SpeciesSex, TotalNumber) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, DevelopmentStage) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_sex > 1) |>
  summarise(groups             = n(),
            one_value_repeated = sum(as.integer(n_val == 1)),
            one_value_per_sex  = sum(as.integer(n_val == n_sex)),
            neither            = sum(as.integer(n_val > 1 & n_val < n_sex))) |>
  collect()
sex_rows
```

`r P(100 * sex_rows$one_value_repeated / sex_rows$groups)` of multi-sex groups
repeat one value and `r P(100 * sex_rows$one_value_per_sex / sex_rows$groups)`
carry one per sex. Nothing in the record says which a group is. Add them up and
the repeated kind is inflated; collapse them and the per-sex kind is
understated.

**Is it ad hoc, or a house style?** If repeating is a submitter's habit it
should be near-total for that submitter; if it is a slip it should be a small
minority everywhere. So ask it of each country, showing both how often it
happens and how much data each country contributes:

```{r}
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-alt: "Share of multi-sex records where one total is written on every sex row, by country, with point size showing how many multi-sex records the country contributes. One country repeats the total in all of its records; the others are spread across a much lower range."
sex_by_country <- dr_con("HL") |>
  distinct(Country, .id, Valid_Aphia, SpeciesCategory, SpeciesValidity,
           DevelopmentStage, SpeciesSex, TotalNumber) |>
  group_by(Country, .id, Valid_Aphia, SpeciesCategory, SpeciesValidity,
           DevelopmentStage) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_sex > 1) |>
  group_by(Country) |>
  summarise(records = n(), repeated = sum(as.integer(n_val == 1)), .groups = "drop") |>
  filter(records >= 500) |>
  mutate(pct = 100 * repeated / records) |>
  collect()

sex_by_country |>
  mutate(Country = reorder(Country, pct)) |>
  ggplot(aes(pct, Country)) +
  geom_segment(aes(x = 0, xend = pct, yend = Country), colour = "grey80") +
  geom_point(aes(size = records, colour = pct > 90)) +
  scale_size_area(max_size = 9, breaks = c(1000, 10000, 50000),
                  labels = c("1,000", "10,000", "50,000"),
                  name = "multi-sex\nrecords") +
  scale_colour_manual(values = c(`FALSE` = "grey30", `TRUE` = "firebrick"),
                      guide = "none") +
  scale_x_continuous(limits = c(0, 104), expand = expansion(mult = c(0, 0.02))) +
  labs(x = "records where one total is written on every sex row (%)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

```{r, include = FALSE}
ca_sex <- sex_by_country |> filter(Country == "CA")
others <- sex_by_country |> filter(Country != "CA")
```

Both, as it turns out, and one country is on its own. Canada repeats the total
in `r P(ca_sex$pct)` of its `r N(ca_sex$records)` multi-sex records, which
makes it plainly a house style that can be handled by rule. Every other
country lies between `r P(min(others$pct))` and `r P(max(others$pct))`. That
is not a convention but case-by-case variation, and it is where the ambiguity
actually bites, because there it cannot be settled by knowing who sent the
file.

**Does it move a product?** A total summed across sex rows inflates the haul
total of every repeated-kind group; one collapsed across them understates every
per-sex group. Either way the error lands per haul, so a mean over hauls
inherits it in proportion to how often the species is sexed.

**What `{obus}` does.** `dr_hl_summary()` tests each sex's own `TotalNumber`
against its own length rows raised by its own factor — the check the record
above passes three times. A sex that reconciles is trusted and summed on its
own. Sexes that do not reconcile, or that have no length data to check
against, fall back to collapsing duplicate values, but only among values no
reconciling sex has already claimed.

**For ICES.** Is a `TotalNumber` repeated across sex rows meant as the category
total, or as each sex's own? If the former, should the sexed rows carry it at
all?

## A3. The same shape, for `SpeciesCategoryWeight` {#weight-across-sex-rows}

A *Lepidorhombus boscii* record, three sexes, counts plainly per-sex:

```{r}
dr_con("HL") |>
  filter(.id == "SP-NORTH:1990:4:ES:29CS:BAK:022:22", Valid_Aphia == 127145) |>
  group_by(SpeciesSex) |>
  summarise(TotalNumber           = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Counts of 44, 73 and 1 — genuinely different, genuinely per-sex. The weight is
4250 on all three rows, which cannot be per-sex: three catches that differ
seventy-fold in number do not weigh the same. It is one category weight written
three times. Sum it and the species gains 12,750 g where it caught 4,250.

Archive-wide the weight splits much as the count does:

```{r}
weight_sex <- dr_con("HL") |>
  distinct(.id, Valid_Aphia, SpeciesValidity, SpeciesCategory,
           SpeciesSex, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, SpeciesCategory) |>
  summarise(n_sex = n_distinct(SpeciesSex),
            n_w   = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(n_sex > 1) |>
  summarise(groups              = n(),
            one_weight_repeated = sum(as.integer(n_w == 1)),
            one_weight_per_sex  = sum(as.integer(n_w == n_sex))) |>
  collect()
weight_sex
```

Weight is worse off than number, because it has no arithmetic cross-check. The
count above could be tested against the length rows that produced it. A weight
can be tested against nothing in the record — the reasoning that unmasked the
4250 was biological, not arithmetic.

**Does it move a product?** Biomass per haul, for every repeated-kind group, by
the number of sex rows it was written on.

**What `{obus}` does.** Collapses duplicate weights without sex in the key —
the opposite of the count treatment in [A2](#totalnumber-across-sex-rows). The
asymmetry is forced by the data, not chosen: the reconciliation that
arbitrates counts has no weight equivalent.

**For ICES.** The same question as A2, for weight. Because there is no
cross-check: is there any field a consumer can use to tell a repeated weight
from a genuine per-sex one? And is `DevelopmentStage` part of the
`SpeciesCategoryWeight` key, as the field description's wording implies, given
that sex demonstrably is not?

## A4. One `TotalNumber` written across several sampling categories {#totalnumber-across-categories}

The same shape one dimension over — and here the record answers the question
itself. A whiting split into two sampling categories:

```{r}
dr_con("HL") |>
  filter(.id == "NS-IBTS:2022:3:GB-SCT:748S:GOV:141:141", Valid_Aphia == 126438) |>
  group_by(SpeciesCategory) |>
  summarise(length_rows           = n(),
            sum_NumberAtLength    = sum(NumberAtLength, na.rm = TRUE),
            SubsamplingFactor     = max(SubsamplingFactor, na.rm = TRUE),
            TotalNumber           = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Category 11 raises to 187 × 1.893 = 353.99; category 12 contributes 7 more.
Together that is 360.99, and both rows report `TotalNumber` 361 — the residue
is the rounding of a four-decimal factor. So the shared 361 is the two
categories *together*, not each one's own, and the shared 58100 is one weight
written twice. Read per-category, this species doubles.

```{r}
cat_shared <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  distinct(.id, DataType, Valid_Aphia, SpeciesValidity, SpeciesCategory, TotalNumber) |>
  group_by(.id, DataType, Valid_Aphia, SpeciesValidity) |>
  summarise(n_cat = n_distinct(SpeciesCategory),
            n_val = n_distinct(TotalNumber), .groups = "drop") |>
  filter(n_cat > 1) |>
  group_by(DataType) |>
  summarise(multi_category = n(),
            shared_total = sum(as.integer(n_val == 1)), .groups = "drop") |>
  mutate(pct = round(100 * shared_total / multi_category, 1)) |>
  collect()
cat_shared
```

```{r, include = FALSE}
cs <- function(dt, col) cat_shared[[col]][cat_shared$DataType == dt]
cat_total <- sum(cat_shared$shared_total)
```

`r N(cat_total)` of `r N(sum(cat_shared$multi_category))` multi-category
groups share one value. **Why would a total be shared at all?** Pseudo-category
sampling is the obvious candidate: under `DataType` `P` a catch is split into
sub-categories of one weighing, so one total covering them is what the design
implies [@vaishav2021pseudocat]. If that is the mechanism, `P` should be far
more prone to it than `R` — and it is: `r P(cs("P", "pct"))` of multi-category
`P` groups share a total, against `r P(cs("R", "pct"))` under `R`. But `P`
accounts for only `r N(cs("P", "shared_total"))` of the `r N(cat_total)`. The
other `r N(cs("R", "shared_total"))` sit under `R`, where ICES documents no
sub-category convention, and those remain unexplained.

**Does it move a product?** Read per category, every one of these species
doubles in the haul (or more, with more categories) — in numbers and in weight.

**What `{obus}` does.** The fallback collapse leaves `SpeciesCategory` out of
its key, so a shared value is counted once rather than once per category. A
category whose own data reconciles is still credited separately.

**For ICES.** When two categories carry one identical `TotalNumber`, is that
the species total for both strata — as the record above demonstrates — or a
submission error? And under `R`, where no sub-category convention is
documented, what are these? The weight beside them behaves the same way; see
[A8](#the-same-weight-written-on-two-category-codes).

## A5. One total repeated across record types {#totalnumber-across-validity}

`SpeciesValidity` is a record type — what kind of data a record holds (lengths,
a total number only, presence only, and so on) — and HL allows a species more
than one in a haul. Where it has two, the submitted total and weight are almost
always written on both. A Can-Mar witch flounder, lengths split between a
validity-`1` and a validity-`5` record:

```{r}
dr_con("HL_summary") |>
  filter(.id == "Can-Mar:1979:4:CA:18LH:W2A:5:73", Valid_Aphia == 127136) |>
  select(SpeciesValidity, n_totalnumber, n_haul, n_measured, w_haul) |>
  collect()
```

Five fish measured on each record, so ten in all; one total of 9 and one
weight of 1,975 g, written twice. Add the records up and the haul catches 18
fish weighing twice what it did.

```{r}
validity <- dr_con("HL_summary") |>
  group_by(Survey, .id, Valid_Aphia) |>
  filter(n() > 1) |>
  summarise(spread = sd(n_totalnumber, na.rm = TRUE),
            level  = mean(n_totalnumber, na.rm = TRUE), .groups = "drop") |>
  mutate(same_total = coalesce(spread, 0) <= 1e-9 * abs(level)) |>
  group_by(Survey) |>
  summarise(pairs = n(),
            same_total = sum(as.integer(same_total), na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(different_totals = pairs - same_total) |>
  arrange(desc(pairs))
validity
```

`r N(sum(validity$same_total))` of `r N(sum(validity$pairs))` haul-species
pairs with more than one record type carry the same total on each; "the same"
allows for float noise, because the build adds category totals in no fixed
order. The `r N(sum(validity$different_totals))` that differ are unexplained.
Four of them, all NL-BSAS 2023, share a pattern the rest do not — the second
record's total is the first's halved and rounded up:

```{r}
dr_con("HL_summary") |>
  filter(Survey == "NL-BSAS", Year == 2023) |>
  group_by(.id, Valid_Aphia) |>
  filter(n() > 1, n_distinct(n_totalnumber) > 1) |>
  ungroup() |>
  select(.id, latin, SpeciesValidity, n_totalnumber, n_haul) |>
  arrange(.id, SpeciesValidity) |>
  collect()
```

**Does it move a product?** Archive-wide very little; per haul, a doubling.
Until 2026-10-06 `dr_add_catch()` summed across record types, which doubled
every one of these hauls and moved NL-BSAS numbers per hour by about 3% and
DYFS weight per hour by about 2%.

**What `{obus}` does.** `dr_add_catch()` takes a submitted total or weight that
is the same on every record once, sums totals that genuinely differ, and always
sums `n_haul`, because the length rows behind it are never repeated across
record types.

**For ICES.** Is the total on a second record type meant to restate the
species total, or to add to it? And for the NL-BSAS pairs: is the second total
the count of a separately handled part of the catch?

## A6. Two routes to a catch total, and where they part {#two-routes-to-a-catch-total}

DATRAS states a catch total twice: once directly, as the submitted
`TotalNumber`, and once implicitly, as the length rows raised by their own
factor. Where a submission is well formed the two agree, and most do:

```{r}
routes <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(reported = max(TotalNumber, na.rm = TRUE),
            raised   = sum(NumberAtLength * SubsamplingFactor, na.rm = TRUE),
            .groups  = "drop") |>
  filter(!is.na(reported), !is.na(raised)) |>
  summarise(groups          = n(),
            agree_within_1  = sum(as.integer(abs(reported - raised) < 1)),
            differ_1_plus   = sum(as.integer(abs(reported - raised) >= 1)),
            reported_higher = sum(as.integer(reported - raised >= 1)),
            raised_higher   = sum(as.integer(raised - reported >= 1))) |>
  collect()
routes
```

`r P(100 * routes$agree_within_1 / routes$groups, 1)` agree to within one
fish, which is what a non-integer factor applied to whole fish does.
`r N(routes$differ_1_plus)` groups differ by a fish or more, and it is
`TotalNumber` that is higher in all but `r N(routes$raised_higher)` of them.

What a gap looks like, at its simplest — an anchovy record from NIGFS, one
category, no sub-sampling to complicate it:

```{r}
dr_con("HL") |>
  filter(.id == "NIGFS:2019:4:GB-NIR:74RY:ROT:90:10", Valid_Aphia == 126426) |>
  select(LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber) |>
  collect()
```

Three length rows, one fish each. The factor is 1, so nothing is raised. Yet
the record states that 6 were measured and 6 were caught. Three anchovies are
in the totals and absent from the length distribution, and no arithmetic in
the format says which count is right.

**Most of the gap has one mechanism, and it happens before submission.**
Define `F = TotalNumber / SubsampledNumber` — the whole multiplier the
submitter actually used, read straight off the record. Split `DataType` `C`
groups by whether that multiplier is a whole number:

```{r}
rounding <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "C", Survey != "Can-Mar", !is.na(LengthClass),
         NumberAtLength != 0, !is.na(TotalNumber),
         !is.na(SubsampledNumber), SubsampledNumber > 0) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(s   = sum(NumberAtLength, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  mutate(whole_number_factor = abs(tot / sub - round(tot / sub)) < 1e-9) |>
  group_by(whole_number_factor) |>
  summarise(groups = n(),
            whole_fish_gap = sum(as.integer(abs(tot - s) >= 1)), .groups = "drop") |>
  mutate(rate_pct = round(100 * whole_fish_gap / groups, 2)) |>
  collect()
rounding
```

```{r, include = FALSE}
r_int  <- rounding$rate_pct[rounding$whole_number_factor]
r_frac <- rounding$rate_pct[!rounding$whole_number_factor]
```

`r P(r_int, 2)` against `r P(r_frac, 2)` — a relative risk of about
`r round(r_frac / r_int, -1)`, with no reconstruction at all. If the submitter
raised each length row separately, `round(NumberAtLength / F)` should recover
the raw integer counts, and those should sum to the `SubsampledNumber` the
record reports independently:

```{r}
reconstruct <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "C", Survey != "Can-Mar", !is.na(LengthClass),
         NumberAtLength != 0, !is.na(TotalNumber),
         !is.na(SubsampledNumber), SubsampledNumber > 0) |>
  mutate(F = TotalNumber / SubsampledNumber) |>
  filter(F > 0) |>
  mutate(m = round(NumberAtLength / F)) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesValidity, SpeciesSex) |>
  summarise(F = max(F, na.rm = TRUE), sum_m = sum(m, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  group_by(whole_number_factor = abs(F - round(F)) < 1e-9) |>
  summarise(groups = n(),
            reconstruction_matches = sum(as.integer(abs(sum_m - sub) < 0.5)),
            .groups = "drop") |>
  collect()
reconstruct
```

```{r, include = FALSE}
rec_frac <- reconstruct |> filter(!whole_number_factor)
```

It does, in `r P(100 * rec_frac$reconstruction_matches / rec_frac$groups)` of
the non-integer groups. So the submitter computed `NumberAtLength = round(m ×
F)` row by row while computing `TotalNumber` once over the whole sample. The
per-row rounding residuals do not cancel, and their sum is the gap. **The fish
are missing from the submitted file before anyone reads it.**

At the other extreme the gaps are enormous — `TotalNumber` in the hundred
thousands against raised lengths in the hundreds — but those turn out not to be
disagreements at all: they are one total shared across sampling categories
([A4](#totalnumber-across-categories)). Can-Mar behaves differently enough to
have its own entry ([B5](#can-mar)), and a length distribution submitted twice
breaks the agreement the other way ([B4](#distribution-submitted-twice)).

**Does it move a product? For fish, barely; for invertebrates, a great deal.**
Many invertebrates are counted and never measured, so the length route cannot
see them at all:

```{r}
FISH <- c("Teleostei", "Elasmobranchii", "Holocephali", "Actinopteri",
          "Petromyzonti", "Myxini", "Cladistia")
length_route <- dr_con("HL_length") |>
  group_by(.id, Valid_Aphia, SpeciesValidity) |>
  summarise(length_hour = sum(n_hour, na.rm = TRUE), .groups = "drop")

by_group <- dr_con("HL_summary") |>
  filter(!is.na(n_totalnumber_hour)) |>
  left_join(length_route, by = c(".id", "Valid_Aphia", "SpeciesValidity"),
            na_matches = "na") |>
  left_join(dr_con("species") |> select(Valid_Aphia, class), by = "Valid_Aphia") |>
  mutate(group = if_else(class %in% FISH, "fish", "everything else")) |>
  group_by(group) |>
  summarise(records       = n(),
            total_only    = sum(as.integer(is.na(length_hour))),
            totals_route  = sum(n_totalnumber_hour, na.rm = TRUE),
            length_route  = sum(length_hour, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  mutate(total_only_pct     = round(100 * total_only / records, 1),
         length_over_totals = round(length_route / totals_route, 3))
by_group
```

```{r, include = FALSE}
fish  <- by_group |> filter(group == "fish")
inv   <- by_group |> filter(group == "everything else")
```

For fish the length route carries `r round(100 * fish$length_over_totals)`
fish in every hundred the totals route does; for everything else,
`r round(100 * inv$length_over_totals)`, because
`r P(inv$total_only_pct)` of those records have no lengths at all. Which route
a series is built on decides an invertebrate trend, and choosing between them
is a design decision rather than a footnote.

**What `{obus}` does.** Carries both routes in separate columns —
`n_totalnumber` from the submission, `n_haul` from the raised lengths — never
averaged and never silently preferred.

**For ICES.** When the two routes disagree by more than rounding, which is
authoritative? And should the format ask submitters to raise the sample once
rather than row by row, so that the two can agree?

## A7. A mandatory `SubsamplingFactor` that is absent {#subsamplingfactor-absent}

`SubsamplingFactor` is mandatory, and `1` is a specific claim — *not
sub-sampled*. Some rows carry neither the claim nor a factor:

```{r}
sf_missing <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), is.na(SubsamplingFactor),
         NumberAtLength != 0, !is.na(LengthClass)) |>
  summarise(rows = n(), hauls = n_distinct(.id),
            can_mar = sum(as.integer(Survey == "Can-Mar"))) |>
  collect()
sf_missing
```

`r N(sf_missing$rows)` length rows over `r N(sf_missing$hauls)` hauls,
`r P(100 * sf_missing$can_mar / sf_missing$rows, 1)` of them Can-Mar. The
records are not silent, though. Nearly all of them still report both
`TotalNumber` and `SubsampledNumber`, and the ratio of those two is what the
factor would have been:

```{r}
implied <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType %in% c("R", "P", "S"), is.na(SubsamplingFactor),
         NumberAtLength != 0, !is.na(LengthClass)) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesSex) |>
  summarise(sum_nal = sum(NumberAtLength, na.rm = TRUE),
            tot     = max(TotalNumber, na.rm = TRUE),
            sub     = max(SubsampledNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(sub), sub > 0) |>
  summarise(groups          = n(),
            sub_eq_lengths  = sum(as.integer(abs(sub - sum_nal) < 0.5)),
            implied_1       = sum(as.integer(abs(tot - sub) < 0.5)),
            implied_over_1  = sum(as.integer(tot - sub >= 0.5)),
            implied_under_1 = sum(as.integer(sub - tot >= 0.5))) |>
  collect()
implied
```

In `r N(implied$sub_eq_lengths)` of `r N(implied$groups)` groups
`SubsampledNumber` equals the sum of the length rows exactly, so the ratio is
well founded rather than a guess. In `r N(implied$implied_1)` it comes to 1,
so reading the absence as *not sub-sampled* would have been right. In
`r N(implied$implied_over_1)` it is above 1, so that reading understates the
catch. The remaining `r N(implied$implied_under_1)` are the reason not to
simply act on the ratio: they imply a factor **below** 1, a total smaller than
the sub-sample it was raised from. In those records `TotalNumber` and
`SubsampledNumber` cannot both be right, and nothing says which to keep.

**Does it move a product?** Reading a missing factor as 1 — what the
`{DATRAS}` package does with `DataType` `R` — understates the catch in every
group where the implied factor is above 1, which is concentrated in Can-Mar.

**What `{obus}` does.** Propagates `NA`, and reconstructs nothing; the records
carry `CNT_SF_MISSING`. An absent mandatory field means *no information was
supplied*, which is not the same statement as `1`.

**For ICES.** Should a missing `SubsamplingFactor` be read as 1, as unknown, or
reconstructed as `TotalNumber` / `SubsampledNumber`? The groups with an
implied factor above 1 show the first is not safe; whether the third is safe
depends on whether `TotalNumber` is authoritative when the factor is absent,
and the groups below 1 show it is not universally so.

## A8. The same weight written on two category codes {#the-same-weight-written-on-two-category-codes}

Keeping the category is right when two categories are two separately weighed
parts of the catch. It is wrong when the same weight has simply been written on
two codes — and under **pseudo-category sampling (`DataType` P) that is the
documented convention**, not an error. The pseudo-category memo states *"CatCatchweight
value is same in each subcategory"* [@vaishav2021pseudocat], and the archive
agrees:

```{r}
p_weight <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType == "P", !is.na(SpeciesCategory)) |>
  mutate(main_category = substr(SpeciesCategory, 1, 1)) |>
  distinct(.id, Valid_Aphia, SpeciesValidity, main_category,
           SpeciesCategory, SpeciesCategoryWeight) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, main_category) |>
  summarise(sub_categories = n_distinct(SpeciesCategory),
            distinct_weights = n_distinct(SpeciesCategoryWeight), .groups = "drop") |>
  filter(sub_categories > 1) |>
  summarise(main_categories = n(),
            one_shared_weight = sum(as.integer(distinct_weights == 1))) |>
  collect()
p_weight
```

`r N(p_weight$one_shared_weight)` of `r N(p_weight$main_categories)` carry a
single identical weight across their sub-categories. The structural reason is
in the sampling itself: under P the raising factor *is* category weight divided
by sample weight, so the category weight must appear on every sub-category row.
`{obus}` therefore collapses P weights on the main category — the first digit
of a two-digit `SpeciesCategory` — and marks the records `WGT_CAT_REPEAT_P`,
`kind` `property`.

The same shape occurs under `DataType` R, where ICES documents no sub-category
convention, and there the cause is not established. Set the published `w_haul`
against the alternative of collapsing every repeated weight whatever the
`DataType`:

```{r}
published <- dr_con("HL_summary") |>
  group_by(.id, Valid_Aphia, SpeciesValidity) |>
  summarise(w_pub = sum(w_haul, na.rm = TRUE), .groups = "drop")

collapsed <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(DataType != "C") |>
  distinct(.id, Valid_Aphia, SpeciesValidity, SpeciesCategoryWeight, DataType) |>
  group_by(.id, Valid_Aphia, SpeciesValidity, DataType) |>
  summarise(w_alt = sum(SpeciesCategoryWeight, na.rm = TRUE), .groups = "drop")

excess <- published |>
  inner_join(collapsed, by = c(".id", "Valid_Aphia", "SpeciesValidity")) |>
  filter(w_pub > w_alt + 0.5) |>
  group_by(DataType) |>
  summarise(records = n(), excess_g = sum(w_pub - w_alt)) |>
  collect()
excess

all_weight <- dr_con("HL_summary") |> summarise(g = sum(w_haul, na.rm = TRUE)) |> pull(g)
r_rep <- dr_con("hl_flag") |> filter(code == "WGT_CAT_REPEAT_R") |>
  inner_join(dr_con("HH") |> select(.id, Survey, Year, Country, Platform), by = ".id") |>
  count(Survey, Year, Country, Platform, sort = TRUE) |>
  collect()
```

`{obus}` sums the R repeats and flags them `WGT_CAT_REPEAT_R`, `kind`
`unexplained`: collapsing them would assert a convention no document states.
They are `r N(sum(r_rep$n))` records, `r N(r_rep$n[1])` of them from one
platform-year (`r r_rep$Survey[1]` `r r_rep$Year[1]`, `r r_rep$Country[1]`,
platform `r r_rep$Platform[1]`), and together with the P records that share a
weight *across* main categories they are the whole of the remaining excess:
`r N(sum(excess$excess_g))` g, or `r P(100 * sum(excess$excess_g) / all_weight, 2)`
of the archive's `w_haul`.

**Does it move a product?** Collapsed, the P repeats are the difference between
a catch weighed once and the same catch counted per sub-category — large,
because sub-category sampling is used precisely on the big catches. The R
repeats are small archive-wide and concentrated in one submission.

**For ICES.** Do the R repeats follow a convention that should be documented,
or are they submission errors? And should P records sharing a weight across
*main* categories, which the memo does not describe, be collapsed too?

## A9. One length series written under two `LengthType` codes {#duplicated-length-series}

One BITS record, one species, and the whole length series written twice under
two `LengthType` codes:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:2017:4:DK:26HF:TVS:552:42", Valid_Aphia == 127082) |>
  group_by(LengthType) |>
  summarise(length_rows        = n(),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsampledNumber   = max(SubsampledNumber, na.rm = TRUE),
            TotalNumber        = max(TotalNumber, na.rm = TRUE),
            SpeciesCategoryWeight = max(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop") |>
  collect()
```

Two identical series: every length class appears under both codes carrying the
same count, and both copies report the same `SubsampledNumber`, `TotalNumber`
and weight. `LengthType` codes are different length definitions, so the same
fish measured both ways would give *different* classes. Series identical in
class, count, sub-sample, total and weight cannot be two genuine measurements
of one catch. How far this reaches:

```{r}
lt_dup <- dr_con("HL_length") |>
  group_by(.id, Valid_Aphia, SpeciesValidity, SpeciesSex, DevelopmentStage,
           length_mm, accuracy) |>
  summarise(length_types = n_distinct(LengthType),
            distinct_counts = n_distinct(n_measured), .groups = "drop") |>
  filter(length_types > 1, distinct_counts == 1) |>
  summarise(duplicated_rows = n(),
            haul_species_groups = n_distinct(paste0(.id, "|", Valid_Aphia))) |>
  collect()
lt_dup
```

**Does it move a product?** It doubles `n_haul` for these
`r N(lt_dup$haul_species_groups)` haul-species groups — this is the one place
where `{obus}`'s choice of what makes a row unique measurably moves a number.
The weight path ignores `LengthType`, so `w_haul` is not inflated.

**What `{obus}` does.** Keeps both copies, because `LengthType` genuinely
splits other records and merging on a hypothesis would lose those.

**For ICES and the submitting institute.** Are the `LengthType` codes
non-convertible measurement definitions? If so, are these series one
measurement submitted twice?

## A10. `HaulDuration` outside its range {#haulduration}

Every hourly rate divides by duration, and every `C` per-haul count multiplies
by it. The field is documented as 1–120 minutes:

```{r}
duration <- dr_con("HH") |>
  filter(HaulDuration < 1 | HaulDuration > 120) |>
  mutate(range = case_when(HaulDuration < 0  ~ "negative",
                           HaulDuration == 0 ~ "zero",
                           TRUE              ~ "over 120 min")) |>
  count(range, Survey, HaulValidity) |>
  arrange(range, desc(n)) |>
  collect()
duration

with_catch <- dr_con("HH") |>
  filter(HaulDuration <= 0) |>
  semi_join(dr_con("HL") |> distinct(.id), by = ".id") |>
  count(HaulValidity) |>
  collect()
with_catch
```

The zero and negative durations matter less than their count suggests: only
`r sum(with_catch$n)` of those hauls carry any catch record, all of them
`HaulValidity` `r paste(unique(with_catch$HaulValidity), collapse = ", ")`, so
a survey excluding invalid hauls never meets them. The long tows are a
different matter: most are valid hauls in one survey, so either that survey
tows longer than the format anticipates or the field is wrong.

**Does it move a product?** A per-hour rate is undefined at zero and
understated by a wrong long duration; neither is visible downstream.

**What `{obus}` does.** `NA`, never `0` and never `Inf`, for a non-positive
duration (`REC_DUR_NONPOS`). Long durations are used as submitted.

**For ICES and the survey group.** Is a zero `HaulDuration` a missing value
that should have been `-9`? A negative one cannot be anything else, and both
pass upload validation. Are the long tows real?

## A11. Maturity: several scales in one column {#maturity-scales}

`IndividualMaturity` holds codes from at least five scales at once — numeric
1–6, ICES 61–66, roman numerals, R-codes and letters:

```{r}
maturity <- dr_con("CA") |>
  filter(!is.na(IndividualMaturity)) |>
  mutate(family = case_when(
    IndividualMaturity %in% c("1", "2", "3", "4", "5", "6")          ~ "numeric 1-6",
    IndividualMaturity %in% c("61", "62", "63", "64", "65", "66")    ~ "ICES 61-66",
    IndividualMaturity %in% c("I", "II", "III", "IV", "V", "VI",
                              "VII", "VIII", "IX", "X")               ~ "roman numerals",
    IndividualMaturity %like% "R%"                                    ~ "R-codes",
    TRUE                                                              ~ "letters")) |>
  count(MaturityScale, family) |>
  arrange(desc(n)) |>
  collect()
maturity

scale_coverage <- dr_con("CA") |>
  filter(!is.na(IndividualMaturity)) |>
  mutate(decade = (Year %/% 10) * 10) |>
  group_by(decade) |>
  summarise(rows = n(), with_scale = sum(as.integer(!is.na(MaturityScale))),
            .groups = "drop") |>
  collect() |>
  mutate(pct_with_scale = round(100 * with_scale / rows, 1)) |>
  arrange(decade)
scale_coverage
```

`MaturityScale`, the field that says which scale a code belongs to, came late:
it is on `r P(scale_coverage$pct_with_scale[nrow(scale_coverage)])` of rows in
the latest decade and on under a tenth before the 2010s. For most historical
rows the scale can only be inferred from the shape of the code, which works
because the families do not overlap — but it is inference, not a lookup. And
where `MaturityScale` is present it is not always consistent with the codes
beside it.

**Does it move a product?** A maturity ogive pooled across submitters or
decades mixes scales unless it is harmonised first.

**What `{obus}` does.** Nothing; it publishes no maturity product, for this
reason.

**For ICES and the survey groups.** Can `MaturityScale` be made mandatory and
back-filled where the scale is known, and the codes that contradict it
resolved?

# B. Records a submitting institute could correct

Here the record contradicts itself or the format, and the institute that
submitted it holds the information that would correct it.

## B1. A `SubsamplingFactor` below 1 {#factor-below-1}

The raising factor is what you multiply a measured sub-sample by to reach the
whole catch, so it cannot be less than 1 without saying that the catch was
smaller than the sample drawn from it. Two John dory, measured and reported by
EVHOE:

```{r}
dr_con("HL") |>
  filter(.id == "EVHOE:2011:4:FR:35HT:GOV:P1111:67", Valid_Aphia == 127427) |>
  select(LengthClass, NumberAtLength, SubsamplingFactor,
         SubsampledNumber, TotalNumber, SpeciesCategoryWeight) |>
  collect()
```

The record's own arithmetic is impeccable: two fish measured, a factor of
0.615, and a reported total of 2 × 0.615 = 1.23. Every identity in the format
holds. It is the factor that cannot be true — two dory were in the hand and 320
g of them were weighed.

```{r}
below_1 <- dr_con("HL") |>
  filter(!is.na(SubsamplingFactor), SubsamplingFactor < 1) |>
  group_by(Survey, Country) |>
  summarise(length_rows = n(), hauls = n_distinct(.id),
            years = paste(min(Year, na.rm = TRUE), max(Year, na.rm = TRUE), sep = "-"),
            lowest_factor = min(SubsamplingFactor, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(length_rows)) |>
  collect()
below_1
```

**Does it move a product?** The published catch comes out smaller than the
number of fish measured, a signature anyone can test for:

```{r}
dr_con("HL_summary") |>
  filter(!is.na(n_haul), !is.na(n_measured), n_measured > 0) |>
  summarise(records = n(),
            catch_below_measured = sum(as.integer(n_haul < n_measured - 0.001))) |>
  collect()
```

**What `{obus}` does.** Applies the factor as submitted; a reported value is
not second-guessed.

**For the submitting institutes.** Which half of the record is wrong — the
factor, or the counts it was applied to?

## B2. `DataType` `C` with a `SubsamplingFactor` above 1 {#c-with-a-factor-above-1}

`DataType` `C` means the catch is reported as a rate per hour, and the format
requires `SubsamplingFactor` to be 1 [@ices2014datrasfaqs]. These submissions
report both:

```{r}
dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType, HaulDuration), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  distinct(.id, Valid_Aphia, DataType, HaulDuration, SubsamplingFactor) |>
  arrange(desc(SubsamplingFactor)) |>
  head(3) |>
  collect()
```

A record labelled as an hourly rate, on a 30-minute haul, sub-sampled at one
part in more than a hundred. The two statements cannot both hold.

```{r}
c_factor <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType, HaulValidity), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  group_by(Survey, Country) |>
  summarise(rows = n(), hauls = n_distinct(.id),
            years = paste(min(Year, na.rm = TRUE), max(Year, na.rm = TRUE), sep = "-"),
            valid_hauls = n_distinct(if_else(HaulValidity == "V", .id, NA_character_)),
            max_factor = max(SubsamplingFactor, na.rm = TRUE), .groups = "drop") |>
  collect()
c_factor
```

**Which of the two statements is wrong?** Under `C` the total should be the
length rows scaled by duration; under `R` it should be the length rows raised by
the factor. Only one can reproduce the reported `TotalNumber`:

```{r}
which_rule <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType, HaulDuration), by = ".id") |>
  filter(DataType == "C", SubsamplingFactor > 1) |>
  group_by(.id, Valid_Aphia, SpeciesCategory, SpeciesSex) |>
  summarise(s   = sum(NumberAtLength, na.rm = TRUE),
            f   = max(SubsamplingFactor, na.rm = TRUE),
            d   = max(HaulDuration, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  summarise(groups         = n(),
            matches_R_rule = sum(as.integer(abs(tot - s * f) < 0.5)),
            matches_C_rule = sum(as.integer(abs(tot - s * d / 60) < 0.5)),
            sub_eq_lengths = sum(as.integer(abs(sub - s) < 0.5))) |>
  collect()
which_rule
```

`r which_rule$matches_R_rule` of `r which_rule$groups` groups reproduce under
the `R` rule and `r which_rule$matches_C_rule` under the `C` rule. So the
`NumberAtLength` here is a measured sub-sample, not an hourly rate: **the
`DataType` label is the wrong one, not the factor.**

**Does it move a product?** Applying the duration alone, as the label
instructs, understates these catches by up to the factor — over a hundredfold.

**What `{obus}` does.** Applies both multipliers and flags the records
`CNT_C_SUBFACTOR_CONFLICT`.

**For the submitting institutes.** Is the `DataType` wrong on these hauls, or
the factor? They pass upload validation, and nearly all carry `HaulValidity`
`V`.

## B3. A length row that reports no fish {#zero-numberatlength}

One Atlantic cod record, and almost every count identity in the format fails on
it at once:

```{r}
dr_con("HL") |>
  filter(.id == "BITS:1992:1:SE:77AR:GOV:50:1", Valid_Aphia == 126436) |>
  summarise(length_rows        = n(),
            rows_reporting_0   = sum(as.integer(NumberAtLength == 0)),
            sum_NumberAtLength = sum(NumberAtLength, na.rm = TRUE),
            SubsampledNumber   = max(SubsampledNumber, na.rm = TRUE),
            TotalNumber        = max(TotalNumber, na.rm = TRUE),
            SubsamplingFactor  = max(SubsamplingFactor, na.rm = TRUE)) |>
  collect()
```

Most of its length rows report no fish at that length; the rest sum to 8, while
`SubsampledNumber` claims 30 were measured and `TotalNumber` a catch of 28 — at
a factor of 1, so no raising can reconcile them. Zero-count length rows are
rare:

```{r}
zero_nal <- dr_con("HL") |>
  filter(NumberAtLength == 0) |>
  summarise(rows = n(), hauls = n_distinct(.id),
            species = n_distinct(Valid_Aphia),
            surveys = n_distinct(Survey)) |>
  collect()
zero_nal
```

They do not share one explanation: some groups report nothing but zeros, so no
fish were measured at any length; in others a single zero sits among real
counts. `r zero_nal$rows` rows is too few to find a mechanism from the data
alone.

**Does it move a product?** No: a length class with no fish cannot enter a
length distribution.

**What `{obus}` does.** Drops them.

**For the submitting institutes.** Is a zero `NumberAtLength` a presence
marker, a placeholder, or an error? And on the record above, which of the three
counts is right?

## B4. A length distribution submitted twice {#distribution-submitted-twice}

ICES's field descriptions identify a length record by its haul, species, sex,
development stage and sampling category [@ices2026fielddesc], so one length
class can appear once in a category. Some records carry the same length
distribution twice, which doubles the length route while the submitted total
stays right. A BITS plaice haul:

```{r}
dup_id <- "BITS:2021:1:DK:26HF:TVS:77:35"
dr_con("HL") |>
  filter(.id == dup_id, Valid_Aphia == 127143) |>
  count(SpeciesCategory, SpeciesSex, LengthClass, NumberAtLength,
        name = "times_submitted") |>
  count(times_submitted, name = "length_classes") |>
  collect()

dr_con("HL_summary") |>
  filter(.id == dup_id, Valid_Aphia == 127143) |>
  select(n_totalnumber, n_haul, n_totalnumber_hour) |>
  collect()
```

Every length class is there twice, and the raised lengths come to double the
reported total. `{obus}` marks such records `CNT_LEN_DUP`:

```{r}
dr_con("hl_flag") |>
  filter(code == "CNT_LEN_DUP") |>
  inner_join(dr_con("HH") |> select(.id, Survey, Year, Country), by = ".id") |>
  group_by(Survey, Country) |>
  summarise(records = n(),
            years = paste(min(Year, na.rm = TRUE), max(Year, na.rm = TRUE), sep = "-"),
            .groups = "drop") |>
  arrange(desc(records)) |>
  collect()
```

The flag compares rows within one category. A distribution resubmitted under a
second category code would not be caught by it; whether any are is an open
check ([E](#what-remains-open)).

**Does it move a product?** Anything built on the length route — numbers at
length, and any total or rate summed from them — doubles for these records.

**What `{obus}` does.** Flags; keeps both copies; prefer the reported total for
these records.

**For the submitting institutes.** Are these distributions submitted twice, and
can they be resubmitted once?

## B5. Can-Mar: fish in the total that never entered a length row {#can-mar}

Can-Mar's two routes disagree in a pattern of their own. A lost raising factor
would apply to a whole haul — every species with the same ratio of reported
total to measured lengths. Take Can-Mar hauls where three or more species show
a gap at a factor of 1, and ask how many carry one shared ratio:

```{r}
canmar_hauls <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(Survey == "Can-Mar", DataType %in% c("R", "P", "S"),
         !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia) |>
  summarise(s = sum(NumberAtLength, na.rm = TRUE),
            f = max(SubsamplingFactor, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(f), f == 1, s > 0, tot - s >= 1) |>
  group_by(.id) |>
  summarise(species = n(), distinct_ratios = n_distinct(round(tot / s, 3)),
            .groups = "drop") |>
  filter(species >= 3) |>
  summarise(hauls                 = n(),
            all_species_one_ratio = sum(as.integer(distinct_ratios == 1)),
            any_two_agree         = sum(as.integer(distinct_ratios < species)),
            every_ratio_different = sum(as.integer(distinct_ratios == species))) |>
  collect()
canmar_hauls
```

`r N(canmar_hauls$all_species_one_ratio)` of `r N(canmar_hauls$hauls)` hauls
have a single ratio across their species, so **a haul-level lost factor is
refuted**. The other candidate — fish counted but never measured — predicts
whole-fish gaps, and a sub-sample field that sides with the length rows rather
than with the total:

```{r}
canmar_gaps <- dr_con("HL") |>
  inner_join(dr_con("HH") |> select(.id, DataType), by = ".id") |>
  filter(Survey == "Can-Mar", DataType %in% c("R", "P", "S"),
         !is.na(LengthClass), NumberAtLength != 0) |>
  group_by(.id, Valid_Aphia) |>
  summarise(s = sum(NumberAtLength, na.rm = TRUE),
            f = max(SubsamplingFactor, na.rm = TRUE),
            sub = max(SubsampledNumber, na.rm = TRUE),
            tot = max(TotalNumber, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(tot), !is.na(f), f == 1, s > 0, tot - s >= 1) |>
  summarise(groups         = n(),
            gap_whole_fish = sum(as.integer(abs((tot - s) - round(tot - s)) < 1e-9)),
            sub_eq_lengths = sum(as.integer(abs(sub - s) < 0.5)),
            sub_eq_total   = sum(as.integer(abs(sub - tot) < 0.5)),
            gap_of_1       = sum(as.integer(tot - s == 1)),
            gap_over_10    = sum(as.integer(tot - s > 10))) |>
  collect()
canmar_gaps
```

`r N(canmar_gaps$gap_whole_fish)` of `r N(canmar_gaps$groups)` gaps are a
whole number of fish. Where `SubsampledNumber` is present it matches the length
rows in `r N(canmar_gaps$sub_eq_lengths)` groups and the total in
`r N(canmar_gaps$sub_eq_total)`. So the record's own account of what was
measured corroborates the lengths, and `TotalNumber` alone carries the extra
fish. The ratios concentrate on 4/3, 3/2, 5/4 and the like — what a handful of
unmeasured fish produce, and implausible as sub-sampling factors.

**Does it move a product?** Built on the length route, Can-Mar catches are
understated wherever this happens; built on the totals route, they are not.

**What `{obus}` does.** Carries both routes; the records carry
`CNT_SF_RESET` or `CNT_RAISE_LOST`. Treat Can-Mar separately and prefer the
reported total.

**For the submitting institute.** Does Can-Mar record fish in `TotalNumber`
that are deliberately not length-measured, and if so, is that intended under
the format?

## B6. BITS catch weights a hundredfold too low, 1991–2002 {#bits-weight-unit}

What a single fish weighs is the one diagnostic a weight field cannot escape.
In the early BITS first-quarter survey, for three common species:

```{r}
#| fig-width: 7.5
#| fig-height: 3.6
#| fig-alt: "Median grams per fish by year in BITS quarter 1, for Denmark, Germany and Sweden. Denmark and Germany sit around two orders of magnitude below Sweden until 1994 and 2002 respectively, then rise to the Swedish level."
g_per_fish <- dr_con("HL") |>
  filter(Survey == "BITS", Quarter == 1, Year <= 2006,
         Valid_Aphia %in% c(126436, 127143, 126417),      # cod, plaice, herring
         Country %in% c("DE", "DK", "SE")) |>
  distinct(.id, Year, Country, Valid_Aphia, TotalNumber, SpeciesCategoryWeight) |>
  filter(TotalNumber > 0, SpeciesCategoryWeight > 0) |>
  mutate(g = SpeciesCategoryWeight / TotalNumber) |>
  collect() |>
  group_by(Year, Country) |>
  summarise(median_g = median(g), .groups = "drop")

ggplot(g_per_fish, aes(Year, median_g, colour = Country)) +
  geom_line() + geom_point(size = 1) +
  scale_y_log10() +
  labs(x = NULL, y = "median grams per fish (log scale)", colour = NULL) +
  theme_minimal(base_size = 11)
```

German and Danish fish weigh a few grams; Swedish fish from the same survey,
quarter and years weigh what a fish of their length weighs. The lengths in the
same records are normal. The error corrects on a date per country — Denmark
after 1994, Germany after 2002 — and its size, about two orders of magnitude,
points at a unit: weights recorded in hundreds of grams in a field defined as
grams. The IBTSWG has recorded exactly this kind of correction before, against
another submission [@ices2026ibtswg].

**Does it move a product? Yes, and it can manufacture one.** A BITS biomass
series from the reported weights rises by more than a hundredfold across the
correction dates, for every species. Numbers and lengths are unaffected.

**What `{obus}` does.** Nothing: no flag catches it. The records are
self-consistent and merely impossible, and every `hl_flag` test is a test of a
record against itself.

**For the submitting institutes.** What unit were BITS catch weights recorded
in before the correction dates, and can the records be resubmitted in grams?

## B7. Haul positions: missing, or copied {#haul-positions}

`ShootLatitude`/`ShootLongitude` are where the gear went in;
`HaulLatitude`/`HaulLongitude` where it came out. The second pair is often
absent, and where both exist they are sometimes identical:

```{r}
positions <- dr_con("HH") |>
  group_by(Country) |>
  summarise(hauls = n(),
            no_end_position = sum(as.integer(is.na(HaulLatitude) | is.na(HaulLongitude))),
            end_equals_shoot = sum(as.integer(ShootLatitude == HaulLatitude &
                                              ShootLongitude == HaulLongitude), na.rm = TRUE),
            coarse = sum(as.integer(ShootLatitude == round(ShootLatitude, 1)), na.rm = TRUE),
            .groups = "drop") |>
  filter(hauls > 250) |>
  collect() |>
  mutate(no_end_pct    = round(100 * no_end_position / hauls, 1),
         identical_pct = round(100 * end_equals_shoot / (hauls - no_end_position), 1),
         coarse_pct    = round(100 * coarse / hauls, 1)) |>
  select(Country, hauls, no_end_pct, identical_pct, coarse_pct) |>
  arrange(desc(identical_pct))
positions

position_totals <- dr_con("HH") |>
  summarise(hauls = n(),
            no_end = sum(as.integer(is.na(HaulLatitude) | is.na(HaulLongitude))),
            copied = sum(as.integer(ShootLatitude == HaulLatitude &
                                    ShootLongitude == HaulLongitude), na.rm = TRUE)) |>
  collect()
```

`r N(position_totals$no_end)` of `r N(position_totals$hauls)` hauls
(`r P(100 * position_totals$no_end / position_totals$hauls)`) have no end
position, and `r N(position_totals$copied)` have one identical to the shoot
position.

An identical pair is not rounding: the `coarse_pct` column shows how many
positions are recorded to one decimal, and the countries with most identical
pairs record far fewer than that coarsely. These are fine-grained positions
copied from one slot to the other.

**Does it move a product?** Towed distance from positions, and anything built
on it: a missing end position needs a fallback, and a copied one returns a
plausible zero-distance tow instead of an honest gap.

**What `{obus}` does.** Nothing yet; `{obus}` computes no distance. Treat an
identical pair as a missing end position.

**For the submitting institutes.** Were the end positions recorded? If so, can
they be resubmitted; if not, can the field be left empty rather than copied?

## B8. Placeholders and out-of-range values in HH {#hh-ranges}

The opus dictionary carries each HH field's declared range, so one query tests
every declared range at once:

```{r}
declared <- opus::op_dict("HH") |>
  filter(!is.na(range_min), !is.na(range_max), r_type %in% c("integer", "numeric")) |>
  transmute(name, declared_min = as.numeric(range_min), declared_max = as.numeric(range_max))

observed <- dr_con("HH") |>
  summarise(across(all_of(declared$name),
                   list(lo = ~ min(.x, na.rm = TRUE), hi = ~ max(.x, na.rm = TRUE)))) |>
  collect() |>
  tidyr::pivot_longer(everything(), names_to = c("name", ".value"),
                      names_sep = "_(?=lo$|hi$)")

breaches <- declared |>
  left_join(observed, by = "name") |>
  filter(lo < declared_min | hi > declared_max)
breaches
```

`r nrow(breaches)` of `r nrow(declared)` fields with a declared range breach it,
and the breaches sort into three kinds: **zero standing in for "not recorded"**
(a haul latitude, a door spread, a wing spread, a codend mesh of 0); **a
suspiciously round maximum** (999 for a direction, 99.9 knots for a trawling
speed); and **values that are simply wrong** (a door spread of several hundred
metres, a wind speed of hundreds of metres per second). None appears in any
sentinel list, so they arrive as ordinary numbers and are averaged in as
ordinary numbers:

```{r}
dr_con("HH") |>
  summarise(hauls      = n(),
            lat_zero   = sum(as.integer(HaulLatitude == 0), na.rm = TRUE),
            mesh_zero  = sum(as.integer(CodendMesh == 0), na.rm = TRUE),
            dir_999    = sum(as.integer(TowDirection == 999 | WindDirection == 999), na.rm = TRUE),
            speed_99.9 = sum(as.integer(SpeedGround == 99.9), na.rm = TRUE)) |>
  collect()
```

**Does it move a product?** Only through the fields an analysis uses — gear
geometry and speed for swept area — and only for a few hundred hauls at most.

**What `{obus}` does.** Nothing; these are not flagged. Whether they should be,
as a family alongside the HL codes, is open.

**For the submitting institutes and ICES.** Should these be `-9`? And should
upload validation enforce the declared ranges it publishes?

## B9. Age data: records that name no haul, and impossible values {#ca-values}

```{r}
ca <- dr_con("CA")
ca_issues <- tibble(
  age_records        = ca |> count() |> pull(n),
  naming_no_haul     = ca |> anti_join(dr_con("HH") |> select(.id), by = ".id") |> count() |> pull(n),
  no_station_or_haul = ca |> filter(is.na(StationName), is.na(HaulNumber)) |> count() |> pull(n),
  age_below_0        = ca |> filter(Age < 0) |> count() |> pull(n),
  age_99             = ca |> filter(Age == 99) |> count() |> pull(n),
  weight_zero        = ca |> filter(IndividualWeight == 0) |> count() |> pull(n)
)
ca_issues
```

- **Records that name no haul.** `r N(ca_issues$naming_no_haul)` age records
  (`r P(100 * ca_issues$naming_no_haul / ca_issues$age_records, 1)`) match no
  haul in HH, nearly all because they carry neither a station name nor a haul
  number. `{obus}` publishes them; an `inner_join()` to HH drops them without
  a word.
- **Ages below zero.** The range is 0–99; `r N(ca_issues$age_below_0)` rows
  fall below it, almost all at `-1` — absence written as a negative number,
  which drags down any mean age it reaches.
- **Ages of exactly 99.** `r ca_issues$age_99` fish, after a real tail that
  thins out smoothly and stops in the fifties: a sentinel the declared ceiling
  has accidentally made legal.
- **Individual weights of zero.** `r ca_issues$weight_zero` rows; not a fish.

**Does it move a product?** Mean age and age–length keys, through the negative
and 99 ages; anything joining age data to hauls, through the records that name
none.

**For the submitting institutes.** Can the haul of each unlinked record be
supplied, and the negative and 99 ages be resubmitted as missing?

## B10. Identifications {#identifications}

Two kinds of record a plausibility check would stop. First, birds in a trawl
survey — plausible as single drowned individuals, not in the numbers recorded
here:

```{r}
dr_con("HL_summary") |>
  filter(Valid_Aphia %in% c(232039, 159048)) |>       # smew and Baird's sandpiper
  select(.id, latin, n_totalnumber, w_haul) |>
  collect()
```

Second, a species group one country left unresolved for a period. Gurnards in
the North Sea and Channel surveys, the share of each country's catch recorded
only to genus or family:

```{r}
unresolved <- c(125598, 126178, 150636)                 # Eutrigla, Chelidonichthys, Triglidae
gurnards   <- c(unresolved, 127259, 127261, 127262, 127263, 150637)
gurnard <- dr_con("HL_summary") |>
  filter(Survey %in% c("NS-IBTS", "FR-CGFS", "BTS"), Valid_Aphia %in% gurnards) |>
  inner_join(dr_con("HH") |> select(.id, Country), by = ".id") |>
  mutate(unresolved = Valid_Aphia %in% unresolved) |>
  group_by(Country, Year) |>
  summarise(total = sum(n_haul, na.rm = TRUE),
            unresolved = sum(if_else(unresolved, n_haul, 0), na.rm = TRUE),
            .groups = "drop") |>
  collect()

gurnard |>
  group_by(Country) |>
  summarise(total = sum(total), pct_unresolved = round(100 * sum(unresolved) / sum(total), 1),
            years_mostly_unresolved = sum(unresolved > total / 2), .groups = "drop") |>
  filter(total > 1000) |>
  arrange(desc(pct_unresolved))
```

One country stands out, in a block of years with a clean beginning and end;
its neighbours fishing the same grounds resolve their gurnards to species
throughout.

**Does it move a product?** Any per-species gurnard series over those years
loses that country's catch.

**For the submitting institutes.** Can the smew and sandpiper records be
checked against the sampling sheets? And is there any finer identification
behind the unresolved gurnards?

# C. ICES's derived products

ICES computes products from the archive centrally. Two of them can be checked
against the records they are built from.

## C1. CPUEL {#cpuel}

CPUEL is ICES's catch per unit effort at length [@ices2021units].
[The catch-tables article](catch-tables.html#the-same-calculation-as-cpuel-and-where-the-numbers-differ)
rebuilds it and compares the two number for number; three properties of the
product itself are worth knowing:

```{r}
cpuel <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/CPUEL.parquet")

short_key_collisions <- dr_con("HH") |>
  filter(HaulValidity == "V") |>
  mutate(short_key = paste(Survey, Year, Quarter, Platform, Gear, HaulNumber, sep = ":")) |>
  count(Survey, short_key) |>
  filter(n > 1) |>
  count(Survey, name = "ambiguous_keys", sort = TRUE) |>
  collect()
short_key_collisions

cpuel_validity <- cpuel |>
  filter(n_hour > 0) |>
  distinct(.id) |>
  inner_join(dr_con("HH") |> select(.id, HaulValidity), by = ".id") |>
  count(HaulValidity) |>
  collect()
cpuel_validity

cpuel |> summarise(rows = n(), rows_with_a_sex = sum(as.integer(!is.na(sex)))) |> collect()
```

- **Its haul identifier is not unique.** CPUEL keys a haul by six fields, and
  `r N(sum(short_key_collisions$ambiguous_keys))` six-field keys name more than
  one valid haul. CPUEL's own values are computed on the full identifier, but
  joining a CPUEL row back to HH for position or gear can land on the wrong
  haul.
- **It keeps `r sum(cpuel_validity$n[cpuel_validity$HaulValidity == "I"])`
  hauls HH marks invalid (`I`)**, among
  `r N(sum(cpuel_validity$n))` with a catch.
- **Its sex column is empty on every survey**, including surveys that sex much
  of their catch: the sexes are summed and the column they were summed over is
  left behind. A `group_by(sex)` on it runs without complaint and says nothing.

**For ICES.** Should CPUEL carry the eight-field haul key, drop non-valid hauls,
and either fill or drop its sex column? And what explains the SWC-IBTS
disagreement?

## C2. The FlexFile's calculated gear spreads {#flexfile}

The FlexFile (`FL.parquet`) carries swept areas computed from door and wing
spreads, with a flag saying whether each spread was observed or calculated by
ICES from published rules [@ices2022wksae]. Some calculated spreads are
negative, and carry straight through into a negative swept area:

```{r}
fl <- duckdbfs::open_dataset("https://heima.hafro.is/~einarhj/datras/FL.parquet")
fl |>
  summarise(negative_wing  = sum(as.integer(Cal_WingSpread < 0), na.rm = TRUE),
            negative_door  = sum(as.integer(Cal_DoorSpread < 0), na.rm = TRUE),
            negative_swept = sum(as.integer(SweptAreaWSKM2 < 0 | SweptAreaDSKM2 < 0),
                                 na.rm = TRUE)) |>
  collect()
```

The mechanism is in ICES's published calculation code. The rule table writes
its formulas with `log` and no base, and the production script rewrites every
`log` to `log10` before evaluating
([utilities.R](https://github.com/ices-tools-prod/DATRAS/blob/master/FlexFile_calculation/utilities.R)).
Refitting the published rules on the hauls they apply to recovers the published
coefficients in natural log, not base 10, for nearly every stratum — so the
rules are evaluated in a base they were not fitted in, which shrinks the
calculated spreads and pushes intercept-dominated rules through zero. One
further rule, for a single Norwegian vessel and year, puts the wings outside the
doors on every haul it applies to. The refit and its arithmetic check are in
datrasdoodle2's appendix on spread methods.

**Does it move a product?** Every swept area built on a calculated spread. In
that appendix's comparison, published calculated door spreads sit at about 57%
of the same rules evaluated in natural log.

**For ICES.** In which base were the WKSAE rules fitted, per stratum? Can the
rule table write the base explicitly, and the pipeline refuse a negative
spread?

# D. What is not an error

These shape every analysis but are the format working as documented. They are
listed so that a submitter can see what `{obus}` does *not* flag.

- **`DataType` `C` reports a rate per hour**, already raised, with the
  sub-sampling factor set to 1; `{obus}` converts it back to a per-haul count
  and leaves `n_measured` missing (`CNT_C_SUBSAMPLE_HIDDEN`).
- **Under pseudo-category sampling the category weight is written on every
  sub-category** ([A8](#the-same-weight-written-on-two-category-codes)).
- **Species counted, weighed or noted present but never measured** have no
  length rows (`CNT_NO_LENGTH`); they are `HL_summary` records with no
  `HL_length` partner.
- **What a survey records widens over time.** An absence is only a zero if the
  taxon was being recorded:

```{r}
FISH <- c("Teleostei", "Elasmobranchii", "Holocephali", "Actinopteri",
          "Petromyzonti", "Myxini", "Cladistia")
dr_con("HL_summary") |>
  filter(Survey == "NS-IBTS", Quarter == 1) |>
  distinct(Year, Valid_Aphia) |>
  left_join(dr_con("species") |> select(Valid_Aphia, phylum, class), by = "Valid_Aphia") |>
  collect() |>
  mutate(group = case_when(class %in% FISH            ~ "fish",
                           phylum == "Echinodermata"  ~ "echinoderms",
                           phylum == "Mollusca"       ~ "molluscs",
                           phylum == "Arthropoda"     ~ "arthropods",
                           TRUE                       ~ "other"),
         decade = (Year %/% 10) * 10) |>
  group_by(decade, group) |>
  summarise(taxa = n_distinct(Valid_Aphia), .groups = "drop") |>
  tidyr::pivot_wider(names_from = group, values_from = taxa, values_fill = 0)
```

- **`Quarter` is the survey's label, not the calendar quarter** of the haul,
  where a survey group has agreed so:

```{r}
dr_con("HH") |>
  filter(!is.na(Month)) |>
  mutate(calendar_quarter = (Month - 1) %/% 3 + 1) |>
  filter(calendar_quarter != Quarter) |>
  count(Survey, Quarter, calendar_quarter, sort = TRUE) |>
  head(5) |>
  collect()
```

- **A gear specification can stand in a measurement field.** A constant
  nominal spread looks the same as a sensor reading:

```{r}
dr_con("HH") |>
  filter(Survey == "Can-Mar") |>
  count(Gear, WingSpread, DoorSpread, sort = TRUE) |>
  head(3) |>
  collect()
```

- **`HaulValidity` is more than valid or invalid**: `I` is invalid, `C` is
  calibration and not used in products, `A` is excluded from indices, and BITS
  products use `V` and `N` [@ices2014datrasfaqs]. `{obus}` keeps every code and
  leaves the choice to the analysis.

# E. What remains open {#what-remains-open}

Stated plainly, because a reader is entitled to know the edges of this
register.

| Open item | Where it is tracked |
|---|---|
| The fish haul-species pairs whose two routes differ by more than rounding and are neither total-only nor duplicated — flagged `unexplained` (`CNT_DIRECTIONAL`, `CNT_LEN_HIGH`, `CNT_SF_UNRECONCILED`, `CNT_OTHER`) | [`TODO.md`](https://github.com/einarhjorleifsson/obus/blob/main/TODO.md) |
| Whether every distribution submitted twice carries `CNT_LEN_DUP`, including resubmissions under a second category code ([B4](#distribution-submitted-twice)) | `TODO.md` |
| The record-type pairs whose totals genuinely differ ([A5](#totalnumber-across-validity)) | — |
| Whether `SubsamplingFactor = SpeciesCategoryWeight / SubsampleWeight` generalises, which would give the weight path its first arithmetic cross-check | [`PLAN-qc-checks.md`](https://github.com/einarhjorleifsson/obus/blob/main/PLAN-qc-checks.md) §8 |
| `DataType` `C` weight standardisation | `PLAN-qc-checks.md` §8 |
| Whether excluding `SpeciesSex` from the weight key loses anything ([A3](#weight-across-sex-rows)) | `PLAN-qc-checks.md` §8 |
| Whether the length series duplicated under two `LengthType` codes ([A9](#duplicated-length-series)) intersect the records whose weight is still summed twice — predicted empty | not computed |
| A unit check on catch weights, which would have caught [B6](#bits-weight-unit): mean weight per fish, `w_haul / n_haul`, is biologically bounded within about an order of magnitude per species | `PLAN-qc-checks.md` §8 |

# F. Kept in `{opus}`

Issues about what ICES publishes *about* the data, rather than what the records
say — the vocabularies, the field list and its renames, field types, the two
meanings of `-9`, and the download endpoints — belong to `{opus}`, which stages
the archive, and are written up in its
[issues article](https://github.com/einarhjorleifsson/opus/blob/main/articles/issues.qmd).

# The register at a glance {#at-a-glance}

| | Issue | Ask | Scale | Status |
|---|---|---|---|---|
| A1 | What `TotalNumber` means | ICES | the archive follows the raised reading in `r P(100 * tn_def$measured_x_factor / tn_def$groups)` of testable groups | open |
| A2 | `TotalNumber` across sex rows | ICES | repeated in `r N(sex_rows$one_value_repeated)` groups, one per sex in `r N(sex_rows$one_value_per_sex)` | open |
| A3 | Weight across sex rows | ICES | repeated in `r N(weight_sex$one_weight_repeated)` groups | open |
| A4 | `TotalNumber` across categories | ICES | shared in `r N(cat_total)` groups | open |
| A5 | Totals across record types | ICES | `r N(sum(validity$pairs))` pairs, `r N(sum(validity$different_totals))` with different totals | open |
| A6 | The two routes to a total | ICES | differ by a fish or more in `r N(routes$differ_1_plus)` groups | open |
| A7 | `SubsamplingFactor` absent | ICES | `r N(sf_missing$rows)` length rows | open |
| A8 | One weight on two category codes | ICES | `r N(sum(r_rep$n))` R records unexplained | open |
| A9 | Series under two `LengthType` codes | ICES, institutes | `r N(lt_dup$haul_species_groups)` haul-species groups | open |
| A10 | `HaulDuration` out of range | ICES, survey groups | `r N(sum(duration$n))` hauls | open |
| A11 | Maturity scales | ICES, survey groups | `r nrow(distinct(maturity, family))` code families in one column | open |
| B1 | Factor below 1 | institutes | `r N(sum(below_1$length_rows))` length rows | open |
| B2 | `C` with a factor above 1 | institutes | `r N(sum(c_factor$rows))` length rows | open |
| B3 | Zero `NumberAtLength` | institutes | `r zero_nal$rows` rows | open |
| B4 | Distribution submitted twice | institutes | flagged `CNT_LEN_DUP` | open |
| B5 | Can-Mar unmeasured fish | institute | `r N(canmar_gaps$groups)` groups | open |
| B6 | BITS weights too low | institutes | DE 1991–2002, DK 1991–1994 | open |
| B7 | Haul positions | institutes | `r N(position_totals$no_end)` hauls with no end position, `r N(position_totals$copied)` copied | open |
| B8 | HH placeholders and ranges | institutes, ICES | `r nrow(breaches)` fields | open |
| B9 | Age data | institutes | `r N(ca_issues$naming_no_haul)` records naming no haul | open |
| B10 | Identifications | institutes | a handful of records; one country's gurnards | open |
| C1 | CPUEL | ICES | `r N(sum(short_key_collisions$ambiguous_keys))` ambiguous haul keys | open |
| C2 | FlexFile spreads | ICES | negative spreads; rules evaluated in the wrong base | open |

## Answers received

When a question above is answered, the answer is recorded here with its date
and source, and the entry's status changes. None has been recorded yet.

## How this article is made

Every number in it is computed by the code shown, from the archive as
`{obus}` publishes it, when the page is rendered. The one reference to work not
re-run here is the refit behind [C2](#flexfile), which needs the `{DATRAS}`
packages.

**Use of AI.** The text and code of this article were written with an AI
assistant (Claude, by Anthropic) under the direction of the author, who is
responsible for its content.

## References {.unnumbered}

::: {#refs}
:::
