---
title: "DATRAS conventions, and what obus does about them"
subtitle: "What the ICES exchange format specifies, where its documents disagree with each other, and how obus reads them."
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This is a reference article, organised by subject rather than by function. It
states, for each part of the DATRAS exchange format that shapes
`dr_HL_length()` and `dr_HL_summary()`: what ICES's documents say, where those
documents contradict one another, what the archive actually contains, and what
obus does. It describes the current behaviour of the package. It does not
propose changes to it.

Use the table of contents. The sections that most often send a reader here are
[Sub-sampling, the raising factor and `DataType`](#sub-sampling-the-raising-factor-and-datatype),
[Two routes to a catch total](#two-routes-to-a-catch-total), and
[Catch weight](#catch-weight-catcatchwgt).

## What counts as evidence

**Only ICES's own published documents.** For this package that means the
primary sources kept in `~/R/Pakkar/imbus/DATRAS/external/`, principally:

| Source | Short name used below |
|---|---|
| `DATRAS_Field_descriptions_and_example_file_*.xlsx` (vintages May 2022, March 2023, March 2024, September 2025, December 2025) | the field descriptions, or "the spreadsheet" |
| `DATRAS_FAQs` v3.0, August 2014 | the FAQ |
| `Indices_Calculation_Steps_IBTS`, `Indices_Calculation_Steps_BITS` | the index procedures |
| `DATRAS_dataproducts_units` v3.0 and v3.1 | the units specification |
| `WKDATR13` (ICES CM 2013/SSGESST:05) | the 2013 data-review workshop |
| `WKABSENS` 2021 | the absence-record workshop |
| IBTSWG reports, 2024 and 2025 meetings | IBTSWG 2024, IBTSWG 2025 |
| `Reporting_Pseudo_Category_To_DATRAS` (Vaishav, 19/01/2021, marked *Draft*) | the pseudo-category memo |

Authored commentary elsewhere in the `imbus` tree is a pointer to where to
look, not evidence. Dates in this article identify a document vintage or an
archive snapshot and nothing else; archive figures were measured against the
full archive on 2026-08-31.

Two of these sources are marked as drafts on their own covers — the pseudo-
category memo, and the 2006 variance-estimation workshop report cited in the
[discrepancy register](#register-of-cross-document-discrepancies). Claims that
rest on them are flagged as such each time.

## Three naming layers

Field names in this codebase come from three places, and confusing them is the
commonest source of a broken join.

| Layer | What it is | Example |
|---|---|---|
| **1. ICES legacy exchange names** | The names in the ICES exchange file and in every ICES document quoted here. | `HLNoAtLngt`, `SubFactor`, `CatIdentifier` |
| **2. opus current names** | What the raw parquet archive actually carries. `opus` stages the exchange tables under these. | `NumberAtLength`, `SubsamplingFactor`, `SpeciesCategory` |
| **3. obus renames** | Two renames obus applies on top, **for the derived tables only**. | `Valid_Aphia` → `aphia`, `SpeciesSex` → `sex` |

Layer 3 is exactly two renames and nothing else. They are applied in
`data-raw/DATASET_products.R` when `HL_length` and `HL_summary` are built, so
`dr_con_raw("HL")` carries `Valid_Aphia` and `SpeciesSex`, while
`dr_con("HL_length")` carries `aphia` and `sex`. Neither `aphia` nor `sex`
appears as a `new_name` in the crosswalk below, because neither is an ICES or
an opus name.

### The legacy-to-current crosswalk

The mapping is not hand-authored here. It is embedded in the raw parquet
footer and read back from the same file as the data, so it cannot describe a
different vintage than the data it accompanies:

```{r}
opus::op_crosswalk("HL")
```

```
          old_name              new_name
        RecordType          RecordHeader
            Survey                Survey
           Quarter               Quarter
           Country               Country
              Ship              Platform
              Gear                  Gear
         SweepLngt           SweepLength
            GearEx        GearExceptions
          DoorType              DoorType
              StNo           StationName
            HaulNo            HaulNumber
              Year                  Year
      SpecCodeType       SpeciesCodeType
          SpecCode           SpeciesCode
           SpecVal       SpeciesValidity
               Sex            SpeciesSex
           TotalNo           TotalNumber
     CatIdentifier       SpeciesCategory
            NoMeas      SubsampledNumber
         SubFactor     SubsamplingFactor
            SubWgt       SubsampleWeight
       CatCatchWgt SpeciesCategoryWeight
          LngtCode            LengthCode
         LngtClass           LengthClass
        HLNoAtLngt        NumberAtLength
          DevStage      DevelopmentStage
       LenMeasType            LengthType
 DateofCalculation     DateofCalculation
       Valid_Aphia           Valid_Aphia
```

29 rows. Check names against this output rather than from memory. Most pairs
are guessable; at least one is not — **`LenMeasType` maps to `LengthType`**,
which is easy to mistake for a length *unit* field when it is a measurement
*convention* field. `Sex` → `SpeciesSex` is the pair that then takes obus's
own rename to `sex`, and it is the clearest illustration of all three layers
sitting on one column.

## The record key

### The haul key, `.id`

`.id` is an obus construct; ICES has no such field. It is the eight fields
ICES's own exchange tables use to key a haul, concatenated with `:` —

`Survey : Year : Quarter : Country : Platform : Gear : StationName : HaulNumber`

— for example `"BITS:1991:1:DE:06S1:H20:33:28"`. `NA` fields (most often
`StationName`) are skipped rather than rendered as the literal string `"NA"`,
matching the convention the published archive already uses. When every field
is `NA` the result is `NA`, not the empty string, because an empty string
would collide every such row onto one identifier.

`dr_add_id()` accepts only opus's current names. A legacy-named table must be
renamed first (`opus::op_rename()`); `Ship`, `StNo` and `HaulNo` are not
handled.

### The HL record key

From March 2024 the field descriptions define both `TotalNo` (obus:
`TotalNumber`) and `CatCatchWgt` (obus: `SpeciesCategoryWeight`) as totals

> per category (unique combination of cruise haul, species, sex, devstage and
> subsampling category identifier)

so an HL record is keyed by **haul × species × sex × `DevStage` ×
`CatIdentifier`**, and the length rows add `LngtClass` and `LenMeasType` to
that.

The May 2022 and March 2023 vintages read *"for the unique combination of
cruise haul, species, sex, and category identifier"* — sex present, devstage
absent. `DevStage` entered the key in March 2024. The quoted five-part
wording above must not be attributed to the earlier vintages.

### The grain of the two obus tables

| Table | Grain |
|---|---|
| `HL_length` | `.id` × `aphia` × `length_mm` × `length_cm` × `accuracy` × `LengthType` × `sex` × `DevelopmentStage` × `SpeciesValidity` |
| `HL_summary` | `.id` × `aphia` × `SpeciesValidity` |

Both are exactly unique at their stated grain (verified: zero duplicated
groups, over 14,001,605 rows in `HL_length`). At any coarser key neither is,
so sum `n_haul` rather than assuming one row.

`LengthType` and `DevelopmentStage` are grain dimensions in `HL_length`, not
carried attributes, because the ICES record key puts them there: one haul can
measure the same species and length under two measurement conventions, or as
both egg-bearing and unstaged, and each is its own count. Measured
archive-wide: 5,476 groups split on `DevelopmentStage`, 885 on `LengthType`,
15 on `SpeciesValidity`, 14 on `accuracy`. `DevelopmentStage` is easy to
overlook — it is `NA` for 99.8% of rows — but dropping it would sum a berried
record together with an unstaged one.

That grain choice has one measurable cost, in the
[duplicated length series](#duplicated-length-series) section below.

### Why `HL_summary` is not keyed on sex

`HL_length` carries `sex`; `HL_summary` does not. The asymmetry is deliberate.

`HLNoAtLngt` is a genuine per-sex, per-length count, and — verified
full-archive — never repeats a placeholder value across sex sub-rows. `TotalNo`
does: across 212,497 haul × species groups reporting more than one sex, it
varies genuinely by sex for 66% of them and is a species-level placeholder
repeated identically across every sex sub-row for the other 34%. Per-sex
arithmetic reconciliation can tell those apart, and obus does exactly that
(see [the deduplication](#how-obus-deduplicates-totalno)).

`CatCatchWgt` has no analogous check, and its repeats-versus-varies pattern
tracks `TotalNo`'s only about half the time. Keying `HL_summary` on sex would
therefore silently misrepresent `w_haul` for an unknown share of species with
no per-row way to flag which. Sex-specific *counts* stay recoverable by
aggregating `HL_length`; sex-specific *weight* is not recoverable in general.
That is a property of the source data, not a gap in obus.

`sex` codes are kept as reported: `"F"`, `"M"`, `"U"`, `"B"`, or `NA`. `"B"`
(berried, egg-bearing) is not pre-merged into `"F"`, and `"U"` (assessed,
undetermined) stays distinct from `NA` (never assessed). A caller wanting the
old female-proportion can derive it after grouping `sex` away:

```{r}
dr_con("HL_length") |>
  dplyr::group_by(.id, aphia) |>
  dplyr::summarise(
    p_females = sum(n_haul[sex %in% c("F", "B")]) /
                sum(n_haul[sex %in% c("F", "M", "B")])
  )
```

## `SpecVal` is a record type, not a quality grade

The field descriptions call `SpecVal` (obus: `SpeciesValidity`) the

> Validity of species records, e.g. to specify if species has only been
> counted, measured, weighed, etc.

The codes describe *what data is present*, not how good it is:

| Code | What is present |
|---|---|
| `1` | Lengths, numbers and weights |
| `4` | Total number only |
| `5` | Presence / absence |
| `6` | Weight only |
| `7` | Numbers and weights, no lengths |

There is no ordering, so there is no "most valid" code to collapse to, and the
format deliberately allows more than one record type per species per haul.

**obus:** `SpeciesValidity` is part of `HL_summary`'s grain. 1,219 of
2,290,203 groups (0.05%) carry more than one record type; the commonest pairs
are `{1,5}` and `{4,7}`. Filtering to a single code makes `.id` × `aphia`
exactly unique (verified at codes `"1"`, `"4"`, `"5"` and `"7"`). Aggregating
over `.id` × `aphia` without also grouping on `SpeciesValidity` mixes record
types and double-counts.

```{r}
dr_con("HL_summary") |>
  dplyr::filter(SpeciesValidity == "1")   # the usual starting point
```

**The documents disagree on which codes are valid, and this is
survey-specific.** `Indices_Calculation_Steps_BITS` states a rule —

> CPUE values are only estimated for hauls with validity codes "V" and "N" and
> for species with species validity code "SpecValCode" 1 or NULL

— and its Annex 2 repeats `SpecValCode = 1 OR NULL`, so BITS treats *missing*
as valid. `Indices_Calculation_Steps_IBTS` contains no occurrence of `SpecVal`
at all. A rule stated for one survey and absent for another is not a rule of
the format.

**Can-Mar is the exception that breaks the usual filter.** It puts real
`LngtClass`/`HLNoAtLngt` data on `"5"` rows, where every other survey uses
`"5"` for a length-free presence marker. Filtering to `"1"` there discards
genuine length data. Treat Can-Mar separately.

## `-9`, and what a missing value means

The `General Notes` sheet instructs submitters:

> For the fields with no information but header, please submit -9

(The December 2025 wording added *"but header"* to the sentence that stood
before it; the change carries no version note.) So the format has no concept
of "blank" — an absent value is submitted as `-9`. Where a field has a
vocabulary, `-9` carries a per-field meaning:

| Field | Meaning of `-9` |
|---|---|
| `Sex` (obus: `SpeciesSex`) | "not sampled" — distinct from `U`, "Unidentified" |
| `DevStage` | "No information" |
| `DataType` | "Invalid hauls" — a real classification, not an absence |

**The archive:** zero `-9` values anywhere in HL. `opus` resolves them when
building the archive, nulling any whose vocabulary label means absence and
keeping the three that assert something (`Tickler`, `AgePlusGroup`,
`DataType`). That line — keep a sentinel that states a fact, null one that
states nothing — is why an `NA` `sex` in obus reads as "not sampled".

**obus:** never re-introduces a sentinel. It cannot tell which `NA`s were
`-9`, and a sentinel placed back into a numeric column would corrupt any sum
taken over it.

**Dummy records.** The FAQ prescribes a placeholder HL record for hauls with
no catch. Its example is

```
HL,1,SCO,SCO3,GOV,60,-9,P,S01,1,2010,W,126417,0,-9,-9,-9,-9,-9,-9,-9,-9,-9,-9
```

24 fields, matching the HL layout, with `SpecVal` = `0` at position 14 — while
the FAQ's own adjacent instruction is to *"report -9 for all other fields"*.
These records are legitimate artefacts and must be excluded before any
reconciliation of counts. (An earlier reading of this record placed the `0` in
the `CatIdentifier` position; it does not fall there, and the FAQ's
`CatIdentifier` rule is untouched by it.)

## Length codes and length classes

`LngtCode` (obus: `LengthCode`) declares the resolution, and `LngtClass`
(obus: `LengthClass`) **carries mixed units within one column**. Both the units
specification and the format sheet say so:

> LngtClass — Length class in mm for mm and 1/2-cm classes, in cm for cm
> classes

**obus** resolves this once, at read time, into three columns:

| `LengthCode` | Unit of `LengthClass` | `length_mm` | `length_cm` | `accuracy` (cm) |
|---|---|---|---|---|
| `.` | mm | `LengthClass` | `LengthClass / 10` | 0.1 |
| `0` | mm (½-cm classes) | `LengthClass` | `LengthClass / 10` | 0.5 |
| `1` | cm | `LengthClass * 10` | `LengthClass` | 1 |
| `2` | cm (2-cm classes) | `LengthClass * 10` | `LengthClass` | 2 |
| `5` | cm (5-cm classes) | `LengthClass * 10` | `LengthClass` | 5 |
| anything else, or `-9` | — | `NA` | `NA` | `NA` |

`accuracy` is a grain dimension, not a note: a 1-cm and a 5-cm series for the
same species in the same haul are different observations.

Note that the format standard's `Vocab` column — which should point at the
ICES vocabulary that enumerates these codes — is populated in exactly **one
cell** across all three unaggregated sheets of the December 2025 workbook (CA
`PreservationMethod`). It is empty for `DataType`, `SpecVal`, `Sex`,
`LngtCode`, `DevStage` and `LenMeasType`: 0 of 70 HH rows and 0 of 26 HL rows.
The self-described format standard never enumerates its own code values.

## Sub-sampling, the raising factor and `DataType`

This is the core of the format, and the part obus depends on most directly.

### What each field means

| Legacy (obus) | Meaning | Authority |
|---|---|---|
| `HLNoAtLngt` (`NumberAtLength`) | **Raw measured count** in the sub-sample, per length class per category. Per-hour for `DataType` C only. | Dec 2025 HL row 25: *"Number of measured fish for the length class per category…"*; FAQ Annex I: *"HLNoAtLenght: number of fish measured in the (sub) sample"* |
| `NoMeas` (`SubsampledNumber`) | `Sum(HLNoAtLngt)` over length classes within the category. | Dec 2025 HL row 19: *"NoMeas=Sum(HLNoAtLngt)."* |
| `SubFactor` (`SubsamplingFactor`) | The raising factor. ≥ 1, four decimals, per category. For `DataType` S it must be the **compound** of split and species factors. | Dec 2025 HL row 20; FAQ Annex I Ex. 1: *"Subfactor = 8\*2 = 16"* |
| `TotalNo` (`TotalNumber`) | **Raised** count = `NoMeas × SubFactor`. Non-mandatory. | FAQ p. 18 (R): *"Make sure that TotalNo = Sum (HLNoAtLngt) x SubFactor"*; FAQ Annex I: *"Total number (TotalNo): subfact\*nomeas"* |

`NumberAtLength`, `SubsampledNumber` and `SubsamplingFactor` are
`Mandatory = Yes` in every vintage. `TotalNumber` is not.

### The `SubFactor` rule by `DataType`

The `DataType` vocabulary:

| Code | Label |
|---|---|
| `-9` | Invalid hauls |
| `C` | Data calculated as CPUE (number per hour) |
| `P` | Pseudocategory sampling |
| `R` | Data by haul |
| `S` | Sub sampled data |

and the `SubFactor` rule:

> If DataType is C, it should be reported as 1. If DataType is S, it is always
> >1. If DataType is R, the SubFactor is = 1 or >1 for different species
> depending on whether they were subsampled.

So `1` is a specific claim — *this species was not sub-sampled* — and it is
not the same statement as "no factor was supplied", which the format expects
as `-9`.

**obus:** a missing `SubsamplingFactor` propagates to `NA` for every
`DataType`, including `"R"`. This is a deliberate departure from the DATRAS R
package, which treats `"R"` plus a missing factor as 1. That convention
asserts a fact the submission does not contain: it silently converts "the
raising factor is unknown" into "the catch was fully measured", understating
the raised count whenever the true factor was greater than 1. Following the
format instead costs 24,298 length rows (0.17%), worth 0.003% of the archive's
total `n_haul`, and 98% of them Can-Mar.

### Pseudo category sampling (`P`) and the two-part category identifier

`R` is one sub-sampling rate per species. `P` is size-stratified: several
`CatIdentifier` grades, each sub-sampled separately, under one main category.

Under `P` the `CatIdentifier` is **two digits**: the first is the main
category, the second the sub-category, so main category 1 has sub-categories
`11`, `12`, `13`, and main category 2 has `21`, `22`, `23`. The only document
that describes the convention is the pseudo-category memo, which is **marked
Draft** and unpublished:

> If datatype is "P" and there is subcategory PRESENT then Catidentifier value
> will be 11, 12,13… 21,21,23…31,32,33..

and, in the same memo:

> Subfactors value are not the same within subcategory
>
> CatCatchweight value is same in each subcategory
>
> Length class are unique in each subcategory

Two-digit `CatIdentifier` codes occur **only** under `DataType` `P` in the
archive: `S` and `C` show zero affected blocks.

**Published ICES documentation for `P` does not exist.** From March 2024, HH
row 29 enumerates the code — *"P (pseudocategory sampling, only used by
Scotland)"* — and that is all. HL row 20 (`SubFactor`) enumerates C, S and R
and is silent on P. HL row 22 (`CatCatchWgt`) says *"For datatype R or S …
For datatype C …"* and is silent on P. The FAQ, which carries every raising
formula in the corpus, contains no occurrence of the word "pseudo" and its
`DataType` paragraph reads *"Data Type can be reported as R …, S …, C …"*. The
FAQ does not know `P` exists.

**obus:** treats `P` exactly as `R` and `S` — raised count is
`NumberAtLength × SubsamplingFactor`, summed over the rows — because it uses
the row-level formula for every `DataType` and never assumes one category per
species. This is a defensible choice in a documented silence, not an
application of a documented rule.

Note also that the shape `P` describes is submitted under `R` in practice:
34,531 `R` groups carry more than one `SpeciesCategory` each with its own
rate, across many surveys and back to at least 2009. obus is unaffected by
that, for the same reason.

### What obus computes

```
DataType == "C"                 n_haul = NumberAtLength * SubsamplingFactor * HaulDuration / 60
                                n_hour = NumberAtLength * SubsamplingFactor
DataType in ("R", "P", "S")     n_haul = NumberAtLength * SubsamplingFactor
                                n_hour = n_haul / HaulDuration * 60
anything else, including "-9"   n_haul = NA, n_hour = NA
```

This is the ICES index procedure, step for step.
`Indices_Calculation_Steps_IBTS`, section II:

> d. Raise data from sub sampling to total number per haul by multiplying the
> number at length with the sub sampling factor
>
> e. Sum number at haul by length class over category and sex
>
> f. If data type is not CPUE (C) then NoAthaul=number\*(60/HaulDuration)

`Indices_Calculation_Steps_BITS` states the same three steps independently,
adding *"select valid data … for each category and sex"*. The pseudo-category
memo describes the same two steps again. `dr_HL_length()` is those steps.

The `DataType` `C` back-multiplication by `HaulDuration/60` — recovering an
absolute per-haul count from a reported rate — is **also** documented by ICES,
in `WKABSENS` 2021, section 3.3 step 17:

> Estimate numbers (Count) / a. If HLNoAgeLngt == NA, use TotalNo and set
> SubFactor = 1 / b. Multiplier = HaulDur/60 if DataType in ("C") / c.
> Multiplier = SubFactor if DataType in ("S","R") / d. Count = HLNoAtLngt \*
> Multiplier

IBTSWG 2024 (ICES Sci. Rep. 04:65) performs the inverse transform on the
weight field for the same reason: *"CatchWgtraw = (CatchWgtCPUE \* HaulDur) /
60 min"*.

There is one divergence worth naming: WKABSENS applies `HaulDur/60` **only**
for `C`, with no `SubFactor` term, whereas obus multiplies by
`SubsamplingFactor` as well. The two agree wherever the format is followed,
because the format requires `SubFactor = 1` for `C`.

## Two routes to a catch total

There are two independent routes to the same number, and obus keeps both, in
different tables, named apart on purpose:

| Route | obus | Source |
|---|---|---|
| Reported | `HL_summary$n_totalnumber` | the recorded `TotalNo` |
| Reconstructed | `sum(HL_length$n_haul)` | raised length frequencies |

`HL_summary` needs the reported route because 16% of its rows are species with
no length data at all, where no length-derived figure exists. `HL_length`
needs the reconstructed route because it is the only one that resolves to a
length class.

Summing `HL_length` reproduces `Σ(NumberAtLength × SubsamplingFactor)` exactly
(1,925,444 groups, zero differing), so nothing is lost by not duplicating it
into `HL_summary`:

```{r}
dr_con("HL_summary") |>
  dplyr::left_join(
    dr_con("HL_length") |>
      dplyr::group_by(.id, aphia, SpeciesValidity) |>
      dplyr::summarise(n_length = sum(n_haul), .groups = "drop"),
    by = c(".id", "aphia", "SpeciesValidity")
  )
```

### The documentation error: `TotalNo=SUM(HLNoAtLngt)`

The current field descriptions contain a sentence that is **false**, and it is
provably false using nothing but ICES's own files.

December 2025, HL row 17 (`TotalNo`) reads `TotalNo=SUM(HLNoAtLngt).` HL row
19 (`NoMeas`), in the same sheet, reads `NoMeas=Sum(HLNoAtLngt).` They cannot
both be true when `SubFactor > 1`, because the same sheet also requires
`TotalNo = NoMeas × SubFactor`.

**The Example file inside the same workbook settles it arithmetically.** HL
rows 4–22, `SpecCode` 126417, HH `DataType` `R`:

| Quantity | Value |
|---|---|
| The 19 `HLNoAtLngt` values | 2, 2, 5, 6, 16, 28, 22, 32, 21, 28, 31, 14, 12, 6, 4, 3, 3, 1, 1 |
| Their sum | **237** |
| Reported `NoMeas` | **237** — so `NoMeas = Sum(HLNoAtLngt)` is TRUE |
| Reported `TotalNo` | **9766.059** — so `TotalNo = SUM(HLNoAtLngt)` is FALSE, by a factor of 41 |
| Reported `SubFactor` | **41.207**, and 237 × 41.207 = **9766.059** exactly |

So `TotalNo = Sum(HLNoAtLngt) × SubFactor` is the live rule. This example is
byte-identical in all five vintages from May 2022 to December 2025; it has
contradicted the sentence continuously for four years. Three further ICES
statements agree with the example and against the sentence: FAQ p. 18's
`DataType` R block, FAQ Annex I's bullet list, and both FAQ worked examples
(Ex. 1, a `DataType` S case: *"Nomeas = ∑(Hlnoat length) = 50 / Subfactor =
8\*2 = 16 / TotalNo = nomeas \* Subfactor = 50\*16 = 800"*).

**The defect is a stale edit, and it can be dated.** Comparing vintages
directly:

| Vintage | HL row 25 (`HLNoAtLngt`) | HL row 19 (`NoMeas`) |
|---|---|---|
| May 2022, March 2023 | *"Number of fish at this category in this haul. For CPUE data should be adjusted with time. **HLNoAtLngt=NoMeas\*Subfactor**"* — i.e. already **raised** | no formula at all |
| March 2024 onwards | *"Number of **measured** fish for the length class per category…"* — the `=NoMeas*Subfactor` identity deleted | gains *"NoMeas=Sum(HLNoAtLngt)."* |

Row 17 was rewritten in the same pass — its scope clause was replaced and a
`DataType` C sentence added — but the formula fragment `TotalNo=SUM(HLNoAtLngt).`
survived the edit. The March 2024 `General Notes` confirm all three cells were
touched together: *"Improved descriptions for : RecordType , SweepLngt,
StatRec, DayNight, DataType, CatCatchWgt, TotalNo, NoMeas, HLNoAtLngt."*

The stale fragment is additionally **true for `DataType` C only**, where
`SubFactor` is 1 and both sides are per-hour — FAQ p. 19 states it for C
explicitly — which is very likely why it went unnoticed for R, S and P.

**The March 2024 change was not silent in the ICES record, only in the
spreadsheet.** IBTSWG 2024 §3.3 minutes an intersessional group that met on
15 September 2023 to *"review the field specifications of the DATRAS format,
of which the ambiguity had likely led to some of the documented issues"*, with
*"particular attention … to the clarification and harmonisation of
specifications for the ambiguous fields … CatCatchWgt, TotalNo, NoMeas and
HlNoAtLngt, in particular the description of their aggregation levels"*. The
recommendations were reviewed by WGDG and the format descriptions updated. The
`General Notes` line "Improved descriptions" understates that; the change
itself is minuted.

### Is there a reading under which obus is wrong?

Two, and both are dead.

**Reading 1 — "don't raise at all."** Take `TotalNo=SUM(HLNoAtLngt)`
literally: the length sum needs no `SubFactor`, and obus over-raises. Refuted
by the Example file, by the FAQ's R rule, by Annex I, and by both worked
examples — four independent ICES sources against one sentence.

**Reading 2 — "the length rows are already raised, so obus double-raises."**
This is the pre-March-2024 spreadsheet definition, in ICES's own words:
`HLNoAtLngt=NoMeas*Subfactor`. Anyone building against the May 2022 or March
2023 standard would conclude obus is wrong. Refuted by the Example file, which
was already present, unchanged, in those very same workbooks, and shows
`Sum(HLNoAtLngt) = NoMeas = 237`, not 9766. The pre-2024 formula was also
incoherent on its face: it is a per-row identity that can only hold when there
is exactly one length class.

### The scope of `TotalNo` is documented three incompatible ways

| Source | Scope of `TotalNo` |
|---|---|
| FAQ, 2014 to present, p. 18 | *"the total number of fish of one species, sex, and category in the given haul"* — per species × sex × category |
| Spreadsheet, ≤ March 2023, row 17 | *"Total number of specimens in the haul for the species."* — a haul × species total, correctly repeated on every sex and category row |
| Spreadsheet, March 2024 → | per species × sex × devstage × category, matching the FAQ |

Both submitter conventions are present in the archive. **Nothing in the record
marks which convention a given row follows.** This is not a case of obus
deviating from a documented rule; it is two mutually exclusive documented
rules that the data itself does not distinguish.

### How obus deduplicates `TotalNo` {#how-obus-deduplicates-totalno}

`dr_HL_summary()` resolves the ambiguity per group rather than by picking one
convention globally:

1. **Reconcile.** For each `(.id, aphia, SpeciesCategory, SpeciesValidity,
   sex, DevelopmentStage)` subgroup, compare its **own** recorded `TotalNo`
   against its **own** length-derived expectation. A subgroup that reconciles
   (within 0.5) is *trusted* and summed individually.
2. **Fall back.** Subgroups that do not reconcile — or have no length data to
   check against, e.g. bulk-only categories — fall back to a
   `distinct()`-collapse whose key **excludes both `sex` and
   `SpeciesCategory`**, and excludes values already claimed by a reconciling
   subgroup so nothing is counted twice.

Both branches are needed. A plain `distinct()` on the value alone gets the
common cases right but collapses two genuinely different sexes that happen to
report the same small number: a real BITS haul with F (own total 2), M (own
total 1) and unsexed (own total 1, coincidentally equal to M's) has a true
total of 4, and `distinct()`-on-value-alone gives 3.

`SpeciesCategory` is excluded from the *fallback* key because some surveys
report two categories sharing one identical, non-reconciling `TotalNo` — in
practice entirely NS-IBTS, SCOROC and SCOWCGFS, always the category pair
`"11"`+`"12"`, 1,430 haul × species groups archive-wide. That value is the
species-level total for both strata together, so keeping `SpeciesCategory` in
the collapse key sums it once per category and inflates the total by close to
2×. The *trusted* branch still credits any category whose own data genuinely
reconciles.

`DevelopmentStage` is in the key because the ICES record key puts it there.
Archive-wide it moves almost nothing: 289 groups span more than one
`DevStage`, 21 of them with colliding `TotalNo`, worth +29 fish out of
1,443,326,029. Correct by the book and invisible in practice.

**How well the hybrid does, measured.** On the closest reproducible
population — 1,150 groups sharing one `TotalNo` across two-digit categories,
all `DataType` P — obus returns 3,252,517 against an independent length sum of
3,252,507. A naive sum that keeps `SpeciesCategory` in the collapse key gives
6,584,935, i.e. **2.02× too high**; that half of the comparison is
established. A blanket collapse that always drops `sex` and `SpeciesCategory`
gives 3,252,503 on the same population — within 0.0004% of obus. So the
hybrid's advantage over the naive route is demonstrated; its advantage over a
blanket collapse is **not established** on the archive as it stands, and
should not be claimed.

### Where the two routes disagree

Comparison must be made **in submission units**. For `DataType == "C"` both
sides are scaled by `HaulDuration/60`, which turns whole fish into fractions:
a genuine one-fish disagreement in a 30-minute haul appears as 0.5 and slips
under any sub-unit tolerance. Divide the scaling out first; then a gap is in
fish.

Of 1,912,556 comparable groups:

| Gap, in submission units | Groups | Share | |
|---|---|---|---|
| Exactly zero | 1,702,315 | 89.01% | reconciles |
| Less than one fish | 136,645 | 7.14% | **arithmetic — not a data issue** |
| One or more whole fish | 73,596 | **3.85%** | real |

The sub-fish population is arithmetic and nothing else: a non-integer raising
factor applied to integer counts cannot give an integer, and the reported
`TotalNo` is an integer. It carries no information about the data and is
excluded from everything below. Reporting it as part of a "disagreement rate"
only inflates the number. Independent reproductions of the whole comparison
land at 1,912,200–1,912,600 comparable groups, 88.3–89.0% exact, and
73,400–74,200 whole-fish, so bucket counts below move by a few dozen groups
depending on the exact comparability filter.

**Direction, by gap size** (non-Can-Mar; Can-Mar is separated because it is a
single suspected event):

| Gap | Groups | `TotalNo` higher |
|---|---|---|
| 1 fish | 22,559 | 48.5% |
| 2 fish | 9,475 | 53.8% |
| 3–5 | 6,995 | 62.5% |
| 6–10 | 2,442 | **86.3%** |
| 11–50 | 1,449 | **81.5%** |
| >50 | 462 | 38.5% |

Small gaps are near-symmetric; middling gaps point one way more than four
times in five; and above 50 fish the direction **flips**, so the length sum is
the higher side in 61.5% of that tail. The obvious summary statistic is
actively misleading here: the overall direction split is 55/45, which looks
like noise and is not — it is a symmetric population, a strongly directional
one, and a reversed tail added together.

### What explains the disagreement

The whole-fish residual breaks down as follows. Established causes and
labelled hypotheses are marked.

| Population | Groups | Cause | Status |
|---|---|---|---|
| Can-Mar | 30,214 | — | **hypothesis only** (see below) |
| 1–2 fish, five directional pockets | 9,452 of ~38,700 | Can-Mar, non-unit `SubFactor`, PT-IBTS / FR-CGFS, BTS, EVHOE | established |
| 1–2 fish, remainder | 29,145 | genuine symmetric noise — 50.27% one way, 95% CI [49.69–50.85], p = 0.36 | established |
| 6–50 fish | 2,738 of 3,885 | `DataType` C with a non-integer raising factor, rounded by the submitter **per length row** | established |
| 6–50 fish | 618 | `DataType` C, `NoMeas` missing — same family by `DataType` and direction | probable |
| 6–50 fish | 112 | exact 2× / 3× integer-ratio scaling error | established |
| 6–50 fish | 116 | non-C, `NoMeas` arbitrates one side as the outlier | established |
| 6–50 fish | 301 | non-C where neither identity reconciles; C with integer factor | **unexplained** |
| >50 fish | 214 of 443 | [duplicated length series](#duplicated-length-series) | established |
| >50 fish | 21+ of 64 | `SubFactor` reported as a rounded integer, over-raising `TotalNo` | established |
| >50 fish | 165 `TotalNo`-higher + 43 length-higher | — | **unexplained** |

(The 6–50 and >50 sub-counts were computed on an independent reproduction of
the same buckets, with denominators 3,885 and 443, which is why they do not
sum to the 3,891 and 462 in the direction table above.)

**The strongest single result.** Define `F = TotalNo / NoMeas` in submission
units — the whole multiplier the submitter actually used. Over 404,833
non-Can-Mar `DataType` C groups:

| `F` | Groups | Whole-fish disagreement rate |
|---|---|---|
| Integer | 298,653 | **0.16%** |
| Non-integer | 106,180 | **29.35%** |

A relative risk of **187×**, using no reconstruction whatsoever. Haul-duration
divisibility alone does not produce it: even where `60 %% HaulDuration == 0`,
the rate is 32.83% with a non-integer factor against 0.14% with an integer
one.

The mechanism is reconstructible. `m_i = round(NumberAtLength_i / F)` recovers
the submitter's raw integer counts, and those `m_i` sum **exactly** to the
independently reported `NoMeas` in 87–96% of groups. So the submitter computed
`NumberAtLength = round(m_i × F)` row by row while computing
`TotalNo = F × Σm_i` once. The per-row residuals do not cancel; their sum is
the gap. **The fish are missing from the submitted file before obus ever reads
it.**

**The 1–2 fish and 6–50 fish populations are one mechanism, not two.** Inside
the 6–50 bucket, classified by the provider's rounding regime (n = 2,393
classifiable): symmetric rounders (n = 326) are 26.1% `TotalNo`-higher; mixed
(n = 363) 89.5%; **truncators — providers flooring more than 90% of rows
(n = 1,704) — 99.9%**. Truncating providers, as a share of their own groups:
PT-IBTS 46.1%, FR-WCGFS 30.4%, FR-CGFS 15.1%, BITS 10.8%, NS-IBTS 9.6%,
SE-SOUND 4.9%. Magnitude scales with the number of occupied length classes;
direction is set by floor-versus-round-half. Catch size is not a driver
(median |gap| = 0.02 in every `TotalNo` size class from under 50 to over
5,000).

### Can-Mar: a hypothesis, not a finding

Can-Mar contributes 30,214 whole-fish groups — 41% of the residual. Median gap
**13 fish**, **98%** with `TotalNo` higher, and **88.2%** carrying
`SubsamplingFactor == 1`. Two rival mechanisms remain live. **Both are
hypotheses. Neither is established.**

**H-C1a — lost raising factor.** A provider-side historical conversion dropped
the real sub-sampling factor to 1 while `TotalNo` retained the raised value,
so `TotalNo / length_sum` *is* the lost factor. *What would settle it:* a
histogram of `TotalNo / length_sum` for Can-Mar whole-fish groups. Clustering
at small integers, or a smooth distribution centred well above 1, supports it;
a distribution centred near 1 with a long tail does not. A stronger check
still: if `CatCatchWgt / SubWgt` recovers a factor greater than 1 in groups
where `SubFactor` reads 1, that is close to proof.

**H-C1b — counted but not measured.** Some fish were counted into `TotalNo`
but never entered a length row: a bulk count plus a measured sub-sample, with
the residual never distributed across length classes. *What would settle it:*
regress the gap on `TotalNo` and on the number of occupied length classes
separately. H-C1a predicts the gap scales with the length sum; H-C1b predicts
it scales with `TotalNo` with no length-class term. These separate cleanly.

**What is refuted.** A per-length-class half-fish truncation does *not*
explain Can-Mar: `gap / n_length_classes` has median **1.000** (Q1 0.333,
Q3 3.333), not 0.5, and median gap 13 against median 14 occupied classes. The
quartiles are far too dispersed for any per-class mechanism, which mildly
favours H-C1a and H-C1b over anything class-based.

### Duplicated length series {#duplicated-length-series}

This is the one place where an obus grain choice measurably moves the number.

Worked case, `BITS:2017:4:DK:26HF:TVS:552:42`, aphia 127082: 38 raw HL rows —
19 under `LenMeasType` `1` and 19 under `LenMeasType` `6`, with the **same 19
`LngtClass` values, the same 19 `HLNoAtLngt` values, the same `SubFactor` 1,
the same `NoMeas` 57, the same `TotalNo` 57, and the same `CatCatchWgt`
4500**. One length series written twice under two measurement-type codes.

`dr_HL_length()` carries `LengthType` and `DevelopmentStage` as grain
dimensions and therefore sums both copies (114); the `TotalNo` route collapses
them (57). Ratio exactly 2.000.

Archive-wide, **428 of 1,925,444 groups (0.022%)** shrink under a
deduplication on `(sex, SpeciesCategory, LengthCode, LengthClass,
NumberAtLength, SubsamplingFactor)`. **426 of those 428 land in the whole-fish
gap class**, and 221 of them in the >50 band. In the units `HL_length`
actually carries — `n_haul`, haul-scaled for `DataType` C — the duplicated
copies are worth **195,059 of the 394,993 raised fish in those groups**. (In
submission units, i.e. before the `DataType` C scaling, the same excess is
228,904; the two figures must not be interchanged.)

By survey:

| Survey | Groups |
|---|---|
| BITS | 124 |
| NSSS | 97 |
| PT-IBTS | 93 |
| BTS | 80 |
| NS-IBTS | 22 |
| SCOWCGFS | 7 |
| DYFS | 4 |
| FR-CGFS | 1 |

This is why NSSS is 100% length-sum-higher in the >50 tail (69 of 69 groups
resolved) and PT-IBTS 95%.

**Hypothesis (H-C3): these are coding artefacts, not two real measurement
conventions.** `LenMeasType` codes such as `1` and `6` should be genuinely
different length definitions, so the *same fish* measured both ways would give
*different* `LngtClass` values. The observed series are identical in
`LngtClass` **and** `HLNoAtLngt` **and** `NoMeas` **and** `TotalNo` **and**
`CatCatchWgt`, which is physically impossible under a genuine dual-measurement
reading. *What would settle it:* resolving ICES vocabulary `ref=1392` and
confirming the codes are non-convertible measurement definitions, plus
confirming 100% series identity across all 428 groups.

The grain choice that exposes these 428 groups is justified by the 885 groups
that genuinely split on `LengthType`; these 428 are the subset where the two
strata are copies. obus does not deduplicate them.

### What is *not* responsible

Negative results worth recording, because each was a plausible suspect:

| Suspect | Verdict |
|---|---|
| Rows obus drops (`LengthClass` `NA`, `NumberAtLength == 0`) | Present in only 97 of 3,885 target groups, and in **zero** of them does the gap equal the raised value on the dropped rows. Dead. |
| `LngtCode` / `LngtClass` mis-binning | Enriched (2.57% against a 0.11% baseline) but mechanically inert — re-binning cannot change a sum taken over all lengths. Refuted. |
| obus's own dedup structure | Multi-sex groups are *less* common in the target bucket (4.94%) than among exactly-agreeing groups (8.69%). Multi-devstage 2.50% against 0.02%; `SpecVal ≠ 1` 1.57% against 5.53%. Not the driver. |

## Catch weight (`CatCatchWgt`)

`CatCatchWgt` (obus: `SpeciesCategoryWeight`) is an integer field in **grams**.
It has no arithmetic cross-check in the format equivalent to
`TotalNo = NoMeas × SubFactor`, so it is the less-scrutinised of the two paths.

### How obus deduplicates weight

`dr_HL_summary()` uses

```
distinct(.id, aphia, SpeciesCategory, SpeciesValidity, SpeciesCategoryWeight)
```

then sums. The key **excludes `sex` and `DevelopmentStage`** but **keeps
`SpeciesCategory`**. The reason is empirical: DATRAS demonstrably repeats one
weight across sex and devstage sub-rows regardless of what the spec says —
verified 3× inflation on a real NS-IBTS haul with three sex groups for one
category — while genuinely distinct per-sex weights (some programmes do weigh
sexes separately) are kept and summed when they differ.

**The documents cut both ways on this.** The field descriptions put `sex` in
`CatCatchWgt`'s key in **every** vintage, and add `devstage` from March 2024.
Against that, FAQ p. 19 scopes the field for `DataType` C as *"the total catch
weight per species per haul, raised to one hour of hauling"* — no sex, no
category. Whether obus's exclusion of sex costs anything where two genuinely
distinct per-sex weights happen to be numerically **equal** is
**untested**: such a pair would be collapsed by `distinct()`. The diagnostic
that would settle it is mean weight per fish, `w_haul / n_haul`, compared
against single-sex groups of the same species and year; a clean ~0.5 offset
would confirm a loss, no offset would confirm the equal weights were genuine
repeats.

### Repeated weight across categories inflates `w_haul`

Because the key keeps `SpeciesCategory`, a weight that is repeated identically
across category codes is summed once per category. Under `DataType` P this is
the documented convention: the pseudo-category memo says *"CatCatchweight
value is same in each subcategory"*, and the archive agrees — of 1,464 P main-
category blocks with more than one sub-category, **1,456 (99.5%) carry a
single identical weight**.

Measured against the shipped `HL_summary.parquet`:

| Population | Groups | Excess weight |
|---|---|---|
| `DataType` P | 1,491 | 410,122,351 g |
| `DataType` R | 211 | 18,338,213 g |
| **Combined** | **1,702** | **428,460,564 g** |

That is **1.81% of the archive's entire `w_haul` mass**, and `w_hour` is
affected identically. Because the affected groups are the large catches —
which is *why* they were sub-category-sampled in the first place — the excess
is close to 29% of all `DataType` P weight mass (29.27% measured over the
1,458 same-main-category blocks; a further 33 blocks share one weight *across*
main categories, adding 175,550 g, and are inflated in exactly the same way).

The `DataType` R instance is **undocumented**: 211 groups, concentrated
135/211 in NS-IBTS 2025 Q3, Country GB, platform 74E9, with 153 of the 211
also repeating `TotalNo` identically across categories.

Worked case: `NS-IBTS:2016:1:GB-SCT:748S:GOV:12:12`, aphia 126437 —
categories `11` and `12` both carry `CatCatchWgt = 105300`, and obus returns
`w_haul = 210,600 g` for a 105,300 g catch.

**The count path is unaffected.** For all 1,491 P duplicate-weight blocks the
underlying length frequencies **differ** (zero identical), so the mechanism
that inflates weight does not touch `n_haul`. Conversely, obus's weight key
omits `DevelopmentStage` and `LengthType`, so the
[duplicated length series](#duplicated-length-series) mechanism that inflates
`n_haul` in 428 groups cannot inflate `w_haul` there. The two affected sets
are expected to be disjoint; that intersection has **not been computed**, and
a non-empty result would be a genuinely new finding.

`w_haul` is `NA`, not `0`, for a species with no recorded weight at all
(672,065 rows archive-wide). "Not weighed" is not "weighed nothing".

### A cross-check that exists but is not stated in words

**Hypothesis (H-W0).** The December 2025 Example file establishes,
arithmetically and exactly, that `SubFactor = CatCatchWgt / SubWgt`:
343666 / 8340 = 41.206954…, reported as **41.207**, and 237 × 41.207 =
9766.059 exactly — the example reproduces only from the *rounded* factor.

The arithmetic is a fact of the file. **The generalisation is a hypothesis: no
ICES document states this identity in words anywhere.** If it holds generally,
the weight path has the exact analogue of `TotalNo = NoMeas × SubFactor`, and
becomes checkable. *What would settle it:* over all HL rows with `SubWgt`,
`CatCatchWgt` and `SubFactor` populated, compute the ratio
`CatCatchWgt / SubWgt` and compare it to the reported factor, stratified by survey × year ×
`DataType`. High agreement means the factor is weight-derived and the weight
fields are mutually consistent; agreement in some provider blocks only would
itself identify which providers derive the factor by weight and which by
number; systematic failure would mean one of the three fields is being written
with a different meaning.

The same observation may close the sub-one-fish count population
documentarily: a submitter deriving `TotalNo` from the **unrounded** factor
while writing the rounded four-decimal factor into the rows produces a forced
sub-fish gap (237 × 41.206954 = 9766.048 against the reported 9766.059). That
too is a hypothesis, testable the same way.

### Other weight hypotheses, all open

None of these is a finding. Each shares one instrument that has not been
built: **mean weight per fish, `w_haul / n_haul`, stratified by species**,
which is biologically bounded within roughly one order of magnitude, so any
mechanism that scales weight without scaling count leaves a self-normalising
signature.

| Label | Hypothesis | What would settle it |
|---|---|---|
| **H-W1** | `DataType` C weight standardisation is handled asymmetrically. HL row 22 makes `CatCatchWgt` for C a *rate* (*"For datatype C, values should be standardised to 60 minutes"*; May 2022 is blunter still: *"total species weight per haul per hour for CPUE data type"*). If `w_haul` treats C weights as per-haul while `n_haul` back-multiplies C counts by `HaulDuration/60`, then `w_haul / n_haul` for C is wrong by `60 / HaulDuration`. This would touch roughly a third of the archive rather than 1.8% of it. | For the 200 most abundant aphia codes, plot median g/fish by `DataType`, and within C regress `log(g/fish)` on `log(HaulDuration)`. Slope near +1 confirms; near 0 refutes. |
| **H-W2** | Excluding `sex` from the weight key silently halves genuinely distinct per-sex weights that happen to be equal. | Isolate groups with ≥2 sexes in one `(.id, aphia, SpeciesCategory)` and identical non-zero weights; compare their g/fish against single-sex groups of the same species and year. |
| **H-W3** | A provider wrote `SubWgt` into `CatCatchWgt` or vice versa — adjacent fields, both integer grams, only one mandatory. | H-W0's ratio test, filtered to ratio ≈ 1 and `SubFactor > 1.5`, tabulated by survey × year × country. |
| **H-W4** | Unit errors: kilograms, or 100-gram units, written into a grams field. **Partly answered already** — IBTSWG 2025 Table 3.8.1 records a realised case: CEFAS NS-IBTS Q1 from 1978, `TotalNo` "122→244", `CatCatchWgt` "2750→55", with the comments *"No per hr (\*2)"* and *"No per hour (\*2) per 100g (/100)"*. | Per species, compute the archive-wide median g/fish; per (survey, year, country, platform) compute the median ratio to that reference and look for blocks near 10⁻³, 10⁻², 10² or 10³. Discrete blocks confirm; a continuous spread is biology. |
| **H-W6** | The 33 P blocks sharing one weight *across* main categories are either genuinely equal catches or a wider repeat than the memo describes. | g/fish against the species norm for those 33, plus whether `SubWgt` also repeats across the main categories — a repeated `SubWgt` is much harder to call coincidence. |
| **H-W7** | The 59 sub-1 kg R duplicate blocks are coincidence; the 152 above 1 kg (matching to the gram, up to 1.9 tonnes, with identical `TotalNo` in 153 of 211) are not. | Model the coincidence rate: for R groups with 2+ categories, what fraction have equal weights as a function of weight magnitude? |
| **H-W8** | Groups with a missing `SubFactor` have `w_haul` populated but `n_haul` `NA`, so they have weight with no count and no g/fish diagnostic at all. | Count groups with non-`NA` `w_haul` and `NA` `n_haul`, by survey and era. Every g/fish result above is conditioned on excluding them, which must be stated. |

## Haul duration

An hourly rate needs a duration. 217 hauls report `HaulDuration == 0` — 200 of
them `DataType` C — and 2 report a negative value (both Can-Mar 2017, already
flagged `HaulValidity` `"I"`).

No ICES document states anything about a zero, negative or `-9` haul duration.
obus's treatment is a defensible choice in a documented silence:

> Where the duration is not positive, any figure that **needs** it is `NA`.

The direction depends on `DataType`, because `C` reports a rate and derives
the haul total, while everything else reports the haul total and derives the
rate:

| | non-`C` | `C` |
|---|---|---|
| hourly rate (`n_hour`, `n_totalnumber_hour`, `w_hour`) | derived — `NA` | reported — kept |
| per-haul total (`n_haul`, `n_totalnumber`, `w_haul`) | reported — kept | derived — `NA` |

Without this the arithmetic produces `Inf` on one side and a plausible-looking
`0` on the other. The `0` is the worse of the two: it looks like a real
observation.

**A documented note of caution.** The 2006 ICES workshop report on variance
estimation — circulated in draft, and marked `DRAFT` on its own cover — carries
an unresolved editorial query in its body text, beside a CPUE formula with no
duration term at all:

> [Are they assuming each haul is one hour? If not should number of hours
> fished be included in the division?]

Its 2007 successor **deleted the question and kept the formula**. ICES's own
authors were not certain whether duration belongs in the divisor.

## Valid hauls

`haulval` is a parameter of both `dr_HL_length()` and `dr_HL_summary()`; it
takes a character vector of `HaulValidity` codes to retain, and `NULL` keeps
all hauls. The filter is applied to HH before the join, in
`.dr_hh_cols_for_hl()`, so both tables stay in lockstep on which hauls are in
scope.

The only ICES statement of which codes count is survey-specific:
`Indices_Calculation_Steps_BITS` restricts to *"hauls with validity codes 'V'
and 'N'"*. The IBTS procedure states none. obus does not pick a default.

## Register of cross-document discrepancies

Every entry below was checked by opening the cited file. Where an earlier
reading overstated a discrepancy's severity or genre, the entry says so.

### Tier 1 — changes the number you compute

| # | Discrepancy | Severity |
|---|---|---|
| **D1** | `TotalNo=SUM(HLNoAtLngt)` (HL row 17) and `NoMeas=Sum(HLNoAtLngt)` (HL row 19) in the same current sheet. Cannot both hold when `SubFactor > 1`. Falsified by the workbook's own Example file (237 against 9766.059). | Contradiction |
| **D2** | `HLNoAtLngt` was **already-raised** by definition (`HLNoAtLngt=NoMeas*Subfactor`) through March 2023, then redefined as raw "measured" in March 2024. Reverses the field's meaning. Contradicted the unchanged Example file for the whole two years it stood. The `General Notes` list it only under "Improved descriptions" — but the change is minuted in IBTSWG 2024 §3.3, so it was flagged in the ICES record, just not in the spreadsheet. | Contradiction |
| **D3** | FAQ p. 18 gives **two different `TotalNo` formulas two paragraphs apart**. Line 745, R: *"TotalNo = Sum (HLNoAtLngt) x SubFactor"*. Line 764, S: *"TotalNo = Sum (HLNoAtLngt)"*, with the `SubFactor` term dropped — even though line 756, in the same S block, says *"Value is always > 1. Make sure that TotalNo = NoMeas x SubFactor"*, and the FAQ's own Example 1 is a `DataType` S case giving 50 → 800. The S block's own worked example falsifies the S block's own formula. | Contradiction |
| **D4** | `TotalNo`'s scope documented three incompatible ways: FAQ (2014→) per species × sex × category; spreadsheet (≤2023) haul × species; spreadsheet (2024→) per species × sex × devstage × category. Both submitter conventions are present in the archive; nothing marks which a row follows. | Contradiction |
| **D5** | The IBTS and BITS **index-calculation procedures never reference `TotalNo` at any step** (zero grep hits in either), while the format publishes it as a non-mandatory field. *Narrowed:* it is **not** true that ICES never consumes `TotalNo`. WKDATR13 (2013) §3.2.1.3 agrees exactly this comparison as submission check 1 of 4 — *"HL Records - Sum(Number at Length) vs. Reported Total Number for each Haul/Species/Sex/Category"*, with a worked 2,473-fish example; the DATSU check register consumes it (*"If validity code is 4 then TotalNo > 0"*); and WKABSENS step 17a uses it as a fallback. What is missing is any statement of what a *user* should do when the two disagree. | Silence, material |
| **D6** | `DataType` P has **no raising, `TotalNo`, `SubFactor` or `CatCatchWgt` rule in any published ICES document.** HH row 29 enumerates the code and stops. The FAQ does not contain the word "pseudo". The only guidance is the unpublished 2021 memo. | Material silence |
| **D7** | HH row 29 says only *"C: CPUE (per hour) data; TotalNo and CatCatchWgt are adjusted per hour trawling"* — omitting `HLNoAtLngt`, which HL row 25 and FAQ p. 19 both make per-hour for C. A reader could act on the omission. | Material |
| **D8** | `SubFactor` is described as varying *"by haul, species, sex, length"* (FAQ p. 18) while `TotalNo = NoMeas × SubFactor` presumes one scalar per category. *Downgraded from a contradiction to an ambiguity:* Annex I applies the identical parenthetical to **every** HL field, including `"Total number (TotalNo): subfact*nomeas (by haul, species, sex, category, length)"`, so the list is describing the HL **row key** the field is carried on, not asserting that the value varies. The December 2025 `SubFactor` description gives no key list at all. obus's per-row product remains the safer implementation, but this is not a documented contradiction. | Ambiguity |
| **D9** | The units specification asserts `HLNoAtLngt` is *"Number of fish"* **unconditionally**, in both v3.0 and v3.1, with no `DataType` C carve-out anywhere — contradicting the per-hour rule in the format sheet and the FAQ. `TotalNo` and `NoMeas` are likewise flat "Number of fish". | Material |

### Tier 2 — changes which rows, or which units

| # | Discrepancy | Severity |
|---|---|---|
| **D10** | `NoMeas` has `MaxWidth 3` (≤ 999) yet is defined as `Sum(HLNoAtLngt)`, where each length row is `MaxWidth 10`. The field cannot hold its own defining formula for any sub-sample of 1,000 or more. A reason not to trust `NoMeas` as an arbiter. | Material |
| **D11** | `CatCatchWgt`'s key **includes sex in every vintage**; obus's weight dedup excludes it. *Correction:* the five-part string *"cruise haul, species, sex, devstage and subsampling category identifier"* is **March 2024 onward only**; May 2022 and March 2023 read *"cruise haul, species, sex, and category identifier"*. And FAQ p. 19 cuts the other way for `DataType` C — *"the total catch weight per species per haul"*, no sex, no category — which is a documentary point in obus's favour. | Material |
| **D12** | BTS indices *"standardised by half an hour haul and not by one hour"* (2006), a caveat dropped from the 2007 successor, while the units specification labels everything "per hour of hauling". *Not latent — realised:* IBTSWG 2025 Table 3.8.1 records CEFAS NS-IBTS Q1 from 1978 submitted per half hour, `TotalNo` "122→244", plus a 100 g unit convention, `CatCatchWgt` "2750→55". | Contradiction, realised |
| **D13** | The 2006 variance-estimation report carries an **unresolved editorial query in its body text** — *"[Are they assuming each haul is one hour? …]"* — beside a formula with no duration term; the 2007 successor deleted the question and kept the formula. *Framing corrected:* the 2006 document is marked `DRAFT` on its own cover, so it is an ICES workshop report circulated in draft, not a published standard. | Contradiction |
| **D14** | *"the length distribution is raised up with the total catch in number"* (2006, BITS herring and sprat), **deleted** from the 2007 successor with no note. *Downgraded:* this sits in a survey-description section reporting BITS national practice, not in a specification, and it does not in fact conflict with the exchange mechanism — a provider who raises a length distribution to a known total would report the derived ratio as `SubFactor`. Keep the silent deletion as a maintenance-quality observation; drop the claim of a contradictory raising rule. | Maintenance |
| **D15** | *"We disregard the issue of sub-sampling the catch by length"* (2006, §5). *Corrected:* the companion 2007 quotation *"This parameter is ignored in this analysis"* is **not** about sub-sampling — it concerns the number of fish measured as a driver of index accuracy in a sampling-intensity simulation, scoped explicitly to "this analysis". Neither document is a format standard. The inference that this is *why* no rounding or reconciliation rule was ever written is a **hypothesis**, and D5's own counter-evidence (WKDATR13, 2013) shows ICES did take the reconciliation question up. | Material silence, narrowed |
| **D16** | IBTS states no `SpecVal` rule at all; BITS specifies `SpecValCode = 1 OR NULL` and treats missing as valid. Survey-specific in the documentation. | Material |
| **D17** | `LngtClass` carries **mixed units within one column**, and the format standard's `Vocab` column is populated in exactly one cell across all three unaggregated December 2025 sheets. The self-described format standard never enumerates its own code values. | Material |
| **D18** | The FAQ-mandated dummy HL record carries `SpecVal = 0` while the FAQ's own adjacent instruction is to *"report -9 for all other fields"*. These records are legitimate artefacts and must be excluded before any reconciliation. *Corrected:* the `0` falls in the `SpecVal` position, not `CatIdentifier`; the FAQ's *"cannot be reported as 0"* `CatIdentifier` rule is untouched by it, and `-9` in the remaining fields is what the `General Notes` require. | Material, narrowed |
| **D19** | Units v3.1 (18-06-2021) renames **three** fields (`GearExp`→`GearEx`, `Stratum`→`DepthStratum`, `BycSpecRecCode`→`BySpecRecCode`), **deletes the `Survey` row** from the Exchange table, gives `Country` two different vocabulary refs in one document (`ref=4` and `ref=337`), and drops two whole products — all with an **empty version note**. `CANoAtLngt` carries two incompatible units ("Number of fish" / "Catch in numbers per hour"); `Age_n` has three across products. *Corrected:* an earlier count of four renames included `Bootnumber`, which disappears with the dropped Bootstrap product, and `MaturityScale`, which is newly added. | Material |
| **D20** | BITS gear-conversion scope stated three ways in one document: cod only (*"In case of cod (Gadus morhua), conversion factors are used…"*); unconditional (*"CPUE are converted to standard trawl by multiplying with a conversion factor"*); and *"apply fishing power"*, a term that occurs twice and is never defined. | Contradiction |

### Tier 3 — minor, but evidence of maintenance quality

- Annex I indexes `TotalNo` *"(by haul, species, sex, category, length)"* — by length, which is impossible for a category total.
- The Example file's `TotalNo = 9766.059` carries three decimals against its own declared `decimal2` type, in all five vintages.
- December 2025's `General Notes` claim *"Decimal data types reduced to only one type"*, while the December 2025 sheets still carry `decimal1`, `decimal2` and `decimal4`.
- `Survey` is `Mandatory` blank in all three September 2025 sheets and `Mandatory = Yes` in all three December 2025 sheets, with no version note.
- The `-9` rule was narrowed to *"fields with no information but header"* in December 2025, unannounced.
- The two upload guides disagree: *"A valid file should contain 3 record types"* against *"A valid file should contain at least 2 record types"*.
- `Indices_Calculation_Steps_IBTS` section II step (a) is textually garbled in the published PDF.
- `Indices_Calculation_Steps_BITS` Annex 6 duplicates its five Quarter-1 lookup rows verbatim after the Quarter-4 block.
- The 2006 variance-estimation report contains `Error! Bookmark not defined.` as body text.

## What remains open

Stated plainly, because a reader is entitled to know the edges of this.

| Open item | Size |
|---|---|
| Can-Mar's whole-fish disagreement — H-C1a and H-C1b both live, with a clean separating test | 30,214 groups |
| The >50 tail after both the duplication and factor-rounding accounts | 165 `TotalNo`-higher + 43 length-higher, of 443 |
| The 6–50 bucket where neither identity reconciles | ~301 of 3,885 |
| Whether `SubFactor = CatCatchWgt / SubWgt` generalises (H-W0), which would give the weight path its first arithmetic cross-check | untested |
| `DataType` C weight standardisation (H-W1) — potentially a third of the archive, against 1.8% for the confirmed weight inflation | untested |
| Whether excluding `sex` from the weight key loses anything (H-W2 / D11) | untested |
| The intersection of the 428 duplicated-length groups with the 1,702 inflated-weight groups — predicted empty | not computed |

## An engineering note: R and SQL disagree about `NA`

Not a DATRAS matter, but the single largest source of silent defects in this
code. obus runs the same functions eagerly (data frames, in the tests) and
lazily (DuckDB, in the build), and the two backends differ in four ways.
Anyone extending these functions needs all four.

**1. Joining.** SQL treats `NULL = NULL` as unknown, so a lazy join drops
every row with an `NA` key — and `sex` is `NA` in 54% of HL, with
`SpeciesCategory`, `SpeciesValidity` and `TotalNumber` able to be `NA` too.
Every join in `dr_HL_summary()` whose key can hold `NA` carries
`na_matches = "na"`, which dbplyr renders as `IS NOT DISTINCT FROM` and DuckDB
supports natively. Without it, every `NA`-sex row fails to find its own
length-derived expectation, never reconciles, falls to the fallback, and
undercounts. Confirmed on a real case
(`BITS:1999:1:DK:26HI:TVS:017492:22`, aphia 127143): true total 4, and 3
without the flag.

**2. Aggregating over an all-`NA` group.** `sum(x, na.rm = TRUE)` is `0` in R
and `NULL` in SQL. Every aggregation here carries an explicit non-`NA` counter
(`.n_ok`, `.h_ok`, `.w_ok`, `.wh_ok`) and forces `NA` when the counter is
zero, so both backends agree. Without it the eager path silently converts a
deliberate `NA` — a missing `SubsamplingFactor`, a non-positive
`HaulDuration`, a species that was never weighed — straight back into the
confident `0` it was meant to prevent.

**3. Self-reference inside `summarise()`.** dplyr evaluates expressions in
order, so `n_ok = sum(!is.na(n_haul))` written *after* `n_haul = sum(n_haul)`
sees the new scalar; SQL resolves every aggregate against the source column.
`dr_HL_length()` renames to `.n_raw`/`.h_raw` before aggregating, so the two
agree by construction rather than by luck.

**4. String concatenation with `NA`.** On a plain data frame R's `paste(sep =)`
renders an `NA` field as the literal two-character string `"NA"`; on a
`tbl_lazy`, dbplyr translates the same call to DuckDB's `CONCAT_WS()`, which
drops `NA` fields instead. `.id` is the join key between HH, HL and CA, so it
must come out identical whichever backend built it. `.dr_concat_ws()` builds
the `NA`-skipping behaviour explicitly from `is.na()`, `case_when()` and
`paste0()`, which translate the same way on both — with one deliberate
departure from raw `CONCAT_WS()`: when every field is `NA` it returns `NA`,
not `CONCAT_WS()`'s `""`, because an empty string would collide every such row
onto one `.id`.

Each of these produced a real, silent, wrong number at some point. Keeping
both an eager test suite and a lazy build is what catches them; neither does
alone.
