---
title: "DATRAS conventions, and what obus does about them"
subtitle: "What the ICES exchange format specifies, and how obus reads it."
format:
  html:
    toc: true
    toc-depth: 3
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Working notes on the parts of the DATRAS exchange format that shape how
`dr_HL_length()` and `dr_HL_summary()` are built. Each entry states what the
format says, what the archive actually contains, and what obus does — in that
order, because the order matters.

## What counts as authoritative here

**Only ICES's own published documents.** For this package that means the
primary sources kept in `~/R/Pakkar/imbus/DATRAS/external/` — principally
`DATRAS_Field_descriptions_and_example_file_December2025.xlsx` (the field
descriptions, with its `General Notes` sheet), the SISP survey manuals, and
the ICES workshop and working-group reports.

Derived commentary — including the authored `.qmd` documents one directory up,
and everything in `obus_retired` — is a useful pointer to *where to look*, and
is not evidence on its own. Several claims checked during this work turned out
to differ from the primary source. Where this article cites a figure, it was
measured against the full archive on 2026-08-31.

## The HL record key

The field descriptions define both `TotalNumber` and `SpeciesCategoryWeight`
as totals

> per category (unique combination of cruise haul, species, sex, devstage and
> subsampling category identifier)

so an HL record is keyed by **haul × species × sex × DevelopmentStage ×
SpeciesCategory**, and the length rows add length to that.

**obus:** both tables carry the full key. `HL_length` is exactly unique at
`.id × aphia × length_mm × length_cm × accuracy × LengthType × sex ×
DevelopmentStage × SpeciesValidity`; `HL_summary` at
`.id × aphia × SpeciesValidity`. Verified: zero duplicated groups in either.

`DevelopmentStage` is easy to overlook — it is `NA` for 99.8% of rows — but
5,476 `HL_length` groups genuinely split on it, and dropping it would sum a
berried or egg-bearing record together with an unstaged one.

## `SpeciesValidity` is a record type, not a quality grade

The field descriptions call it the

> Validity of species records, e.g. to specify if species has only been
> counted, measured, weighed, etc.

The codes describe *what data is present*, not how good it is: `1` is lengths,
numbers and weights; `4` is total number only; `5` presence/absence; `6` weight
only; `7` numbers and weights but no lengths. There is no ordering, so there is
no "most valid" code to collapse to.

**obus:** `SpeciesValidity` is part of `HL_summary`'s grain. 1,219 of its
groups (0.05%) carry more than one record type; the commonest pairs are
`{1,5}` and `{4,7}`. Filtering to one code makes `.id × aphia` exactly unique.

```{r}
dr_con("HL_summary") |>
  dplyr::filter(SpeciesValidity == "1")   # the usual starting point
```

**Can-Mar is the exception.** It puts real `LengthClass`/`NumberAtLength` data
on `"5"` rows, where every other survey uses `"5"` for a length-free presence
marker. Filtering to `"1"` there discards genuine length data. Treat Can-Mar
separately; ICES documents the situation as unresolved.

## `-9` means "no information", and it is already gone

The `General Notes` sheet instructs submitters:

> For the fields with no information but header, please submit -9

So the format has no concept of "blank" — an absent value is submitted as
`-9`. The code carries a per-field meaning where the field has a vocabulary:

| field | `-9` |
|---|---|
| `SpeciesSex` | "not sampled" (distinct from `U`, "Unidentified") |
| `DevelopmentStage` | "No information" |
| `DataType` | "Invalid hauls" — a real classification |

**The archive:** zero `-9` values anywhere in HL. opus resolves them when
building the archive, nulling any whose vocabulary label means absence and
keeping the three that assert something (`Tickler`, `AgePlusGroup`,
`DataType`). That line — keep a sentinel that states a fact, null one that
states nothing — is why an `NA` `sex` in obus can be read as "not sampled".

**obus:** never re-introduces a sentinel. It cannot tell which `NA`s were
`-9`, and a sentinel placed back into a numeric column would corrupt any sum
taken over it.

## `SubsamplingFactor` by `DataType`

The field is **mandatory**, and the rule is per `DataType`:

> If DataType is C, it should be reported as 1. If DataType is S, it is always
> >1. If DataType is R, the SubFactor is = 1 or >1 for different species
> depending on whether they were subsampled.

So `1` is a specific claim — *this species was not subsampled* — and it is not
the same statement as "no factor was supplied", which the format expects as
`-9`.

**obus:** a missing `SubsamplingFactor` propagates to `NA` for every
`DataType`, including `"R"`.

This is a deliberate departure from the DATRAS R package, which treats `"R"`
plus a missing factor as 1. That convention asserts a fact the submission does
not contain: it silently converts "the raising factor is unknown" into "the
catch was fully measured", understating `n_haul` whenever the true factor was
greater than 1. Following the format instead costs little — 24,298 length rows
(0.17%), worth 0.003% of the archive's total `n_haul`, and 98% of them
Can-Mar.

## `DataType` `P` versus `R`

`R` is one sub-sampling rate per species; `P` is size-stratified, with several
`SpeciesCategory` grades each sub-sampled separately. The haul total is
computed the same way for both — sum `NumberAtLength × SubsamplingFactor` over
the rows — so the distinction is descriptive rather than arithmetic.

It is also widely got wrong, and ICES has acknowledged why: IBTSWG 2024 §3.3
records a 2023 intersessional group flagging `DataType` value **`P` as missing
from the format documentation**. Submitters cannot use a code that is not
documented, so size-stratified hauls are submitted as `R`. Measured here,
34,531 `R` groups carry more than one `SpeciesCategory` each with its own
rate — the shape `P` describes — across many surveys and back to at least
2009.

**obus:** unaffected, because it uses the row-level formula for every
`DataType` and never assumes one category per species.

## `TotalNumber` and the raised length sum are two different routes

The field descriptions define `TotalNo = SUM(HLNoAtLngt)` — so the reported
total and the length-frequency sum are meant to be the same number.

**obus keeps both, in different tables, and names them apart:**

- `HL_summary$n_totalnumber` — the reported `TotalNumber`
- `sum(HL_length$n_haul)` — the raised length frequencies

Summing `HL_length` reproduces `Σ(NumberAtLength × SubsamplingFactor)` exactly
(1,925,444 groups, zero differing), so nothing is lost by not duplicating it
into `HL_summary`:

```{r}
dr_con("HL_summary") |>
  dplyr::left_join(
    dr_con("HL_length") |>
      dplyr::group_by(.id, aphia, SpeciesValidity) |>
      dplyr::summarise(n_length = sum(n_haul), .groups = "drop"),
    by = c(".id", "aphia", "SpeciesValidity")
  )
```

`HL_summary` still needs `TotalNumber` because 16% of its rows are species
with no length data at all, where no length-derived figure exists.

The two disagree for **3.37%** of comparable groups. That is a real property of
the submitted data, not a defect: it is concentrated in `DataType == "C"`
(where the reported rate and a raised sample need not match exactly) and in
Can-Mar's documented 2021–22 historical conversion.

## Where the two totals disagree, and what is actually wrong

Two independent routes reach the same number:

- `HL_summary$n_totalnumber` — the total DATRAS *reported*
- `sum(HL_length$n_haul)` — the same total *reconstructed* from the measured
  length frequencies

The format says they are equal (`TotalNo = SUM(HLNoAtLngt)`). Where they are
not, the question is not *how big* the difference is but *what kind* of
difference it is — a magnitude threshold cannot tell arithmetic apart from a
miscount, and picking one silently mixes them.

**Compare in submission units.** For `DataType == "C"` both sides are scaled by
`HaulDuration/60`, which turns whole fish into fractions: a genuine one-fish
disagreement in a 30-minute haul appears as 0.5 and slips under any
sub-unit tolerance. Divide the scaling out first, then the gap is in fish.

### Step 1: separate arithmetic from fish

Of 1,912,556 comparable groups:

| gap, in submission units | groups | share | |
|---|---|---|---|
| exactly zero | 1,702,315 | 89.0% | reconciles |
| less than one fish | ~138,000 | 7.2% | **arithmetic — not a data issue** |
| one or more whole fish | 73,596 | **3.85%** | real |

The sub-unit population is arithmetic and nothing else: a non-integer raising
factor (2.91, say) applied to integer counts cannot give an integer, and the
reported `TotalNumber` is an integer. It carries no information about the
data and is excluded from everything below. Reporting it as part of a
"disagreement rate" only inflates the number.

The real residual is **3.85%** — gaps of at least one whole fish. Where the
raising factor is exactly 1, which is the majority, the raised sum is a plain
integer addition, so these differences cannot be arithmetic under any
reading.

### Step 2: the real residual is two populations, not one

Splitting by gap size and asking which side is larger separates them cleanly:

| gap | groups | `TotalNumber` higher |
|---|---|---|
| 1 fish | 22,559 | 48.5% |
| 2 fish | 9,475 | 53.8% |
| 3-5 | 6,995 | 62.5% |
| 6-10 | 2,442 | **86.3%** |
| 11-50 | 1,449 | **81.5%** |
| >50 | 462 | 38.5% |

(non-Can-Mar; Can-Mar is separated out below because it is a single known
event)

**Small gaps are symmetric; large gaps are directional.** A gap of one or two
fish falls either way at close to a coin flip — that is two independent counts
of the same haul disagreeing slightly, not an error in either table. A gap of
6-50 fish points one way more than four times in five — that is systematic
under-raising, a real defect. They are different phenomena that a median or a
percentage over the whole set would blend into one meaningless middle.

This is worth stating plainly because the obvious summary statistic is
actively misleading here: the overall direction split is 55/45, which looks
like noise, and it is not — it is a symmetric population and a strongly
directional one added together.

### Step 3: what each population is

| population | groups | share of real residual | status |
|---|---|---|---|
| Can-Mar: lost raising factor | 30,214 | 41% | known, provider-side |
| Small symmetric disagreement (1-2 fish) | 32,034 | 44% | not an error |
| Directional under-raising, 6-50 fish | 3,891 | 5% | **open** |
| 3-5 fish, partly directional | 6,995 | 10% | mixed |
| Large reversed tail (>50 fish) | 462 | 0.6% | **open** |

**Can-Mar (41%)** — a bulk conversion of historical data set
`SubsamplingFactor` to 1 on records whose catch had been subsampled.
`TotalNumber` kept its raised value; the length rows lost the factor. Median
gap **13 fish**, and **98%** in the one predicted direction. Trust
`n_totalnumber` here, not the length sum.

**Small symmetric disagreement (44%)** — median gap under 2 fish, direction
~50/50, and concentrated in `DataType == "C"` (36,950 groups, 56.6% one way),
where `TotalNumber` is a reported catch rate and the length table is a
separately measured sample. Two observations of a similar quantity that need
not agree exactly. There is no "true" value to converge on.

**Directional under-raising, 6-50 fish (5%)** — the same signature as Can-Mar
(80-86% `TotalNumber` higher) in surveys with no known conversion event. About
3,900 groups. This is the part genuinely worth chasing, and it is far smaller
and better defined than a headline "3-4% disagree" suggests.

**Large reversed tail (0.6%)** — above 50 fish the direction flips: 61.5% have
the *length sum* higher. NSSS (69 groups, 100% length-higher) and PT-IBTS (44,
95%) stand out. Consistent with length rows submitted twice, a duplication
pattern seen elsewhere in these data.

## `HaulDuration`

An hourly rate needs a duration. 217 hauls report `HaulDuration == 0` and 2
report a negative value.

**obus:** where the duration is not positive, any figure that needs it is `NA`.
The direction depends on `DataType`, because `"C"` reports a rate and derives
the haul total, while everything else reports the haul total and derives the
rate:

| | non-`C` | `C` |
|---|---|---|
| hourly rate | derived — `NA` | reported — kept |
| per-haul total | reported — kept | derived — `NA` |

Without this the arithmetic produces `Inf` on one side and a plausible-looking
`0` on the other.

## An engineering note: R and SQL disagree about `NA`

Not a DATRAS matter, but the single largest source of defects found while
building these tables. obus runs the same functions eagerly (data frames, in
the tests) and lazily (DuckDB, in the build), and the two backends differ in
three separate ways:

1. **Joining.** SQL treats `NULL = NULL` as unknown, so a lazy join drops
   every row with an `NA` key — `sex` is `NA` in 54% of HL. Fixed with
   `na_matches = "na"`, which renders as `IS NOT DISTINCT FROM`.
2. **Aggregating.** `sum(x, na.rm = TRUE)` over an all-`NA` group is `0` in R
   and `NULL` in SQL. Every aggregation here carries an explicit non-`NA`
   counter so both return `NA`.
3. **Self-reference in `summarise()`.** dplyr evaluates expressions in order,
   so `n_ok = sum(!is.na(n_haul))` written after `n_haul = sum(n_haul)` sees
   the new scalar; SQL resolves every aggregate against the source column.
   Renaming before aggregating removes the ambiguity.

Each of these produced a real, silent, wrong number. Keeping both an eager
test suite and a lazy build is what catches them — neither does alone.
