---
title: "Downloading DATRAS data locally"
format:
  html:
    footnotes-hover: true
execute:
  warning: false
  message: false
  eval: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(obus)
library(dplyr)
```

The processed DATRAS parquet files are hosted on a web server and can be
accessed directly without a local copy — see the
[Accessing DATRAS data](access.html) article for that workflow. This article
covers how to download a local copy and access it.

## Downloading with `dr_download()`

`dr_download()` fetches all three record types from the DATRAS API, applies
standard tidying, and writes the results as parquet files to a directory of
your choice:

```{r}
dr_download("/path/to/local/datras")
```

This writes `HH.parquet`, `HL.parquet`, and `CA.parquet` to the given
directory. By default, only the processed files are saved. To also retain the
unprocessed raw download in a `raw/` subdirectory, use `save_raw = TRUE`:

```{r}
dr_download("/path/to/local/datras", save_raw = TRUE)
```

## Connecting to locally stored files

Once files are stored locally, you have two options depending on whether you
want lazy or eager loading.

### Lazy access via `duckdbfs` (recommended for large tables)

`duckdbfs::open_dataset()` returns a lazy DuckDB-backed table. No data is read
until you call `collect()`, so you can filter before loading:

```{r}
path <- "/path/to/local/datras"

hh <- duckdbfs::open_dataset(file.path(path, "HH.parquet"))
hl <- duckdbfs::open_dataset(file.path(path, "HL.parquet"))
ca <- duckdbfs::open_dataset(file.path(path, "CA.parquet"))

# Only the filtered rows are loaded into R
hh |>
  filter(Survey == "NS-IBTS", Year >= 2020) |>
  collect()
```

### Eager loading via `nanoparquet`

`nanoparquet::read_parquet()` reads the entire file into memory as a plain
data frame. This is convenient for HH (which is relatively small), but may be
slow for the full HL or CA tables:

```{r}
hh <- nanoparquet::read_parquet(file.path(path, "HH.parquet"))
hl <- nanoparquet::read_parquet(file.path(path, "HL.parquet"))
ca <- nanoparquet::read_parquet(file.path(path, "CA.parquet"))
```

---

## Background: how the web-based parquet files were created

This section documents the process used to produce the parquet files on the
web server. Most users can ignore it.

`dr_download()` was used with `save_raw = TRUE` to also keep the unprocessed
data:

```{r}
dr_download(
  path     = "/home/hafri/einarhj/public_html/datras",
  save_raw = TRUE
)
```

After writing, files were made world-readable so they can be served over HTTP:

```{r}
system("chmod -R a+r /home/hafri/einarhj/public_html/datras")
```
