---
title: "HL record types"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning  = FALSE,
  message  = FALSE,
  comment  = "#>"
)
```

## Overview

The DATRAS length table (HL) is not a uniform table of length-frequency
observations. A substantial minority of records carry no `LengthClass` at all,
and even within the length-bearing records there are structural differences
driven by the haul-level `DataType` field and the presence of sex or
development-stage annotations. Treating all rows the same leads to silent
errors in abundance or CPUE calculations.

`dr_add_record_type()` classifies every row of a processed HL table into one
of ten mutually exclusive integer types. The companion object
`hl_record_type_lookup` is the lookup table that maps those integers to labels
and descriptions.

### The classification function

The function requires that `dr_add_n_and_cpue()` has already been called (so
`n_haul` is present). It uses `n_haul` as a validity proxy: if `n_haul` is
`NA` for a length-bearing record the haul is invalid (`DataType` of `-9` or
`NA` in the HH table).

```{r eval = FALSE}
library(tidyverse)
library(obus)

# Typical pipeline: load → join HaulDuration → compute n_haul → classify
hh <- dr_hl(survey = "NS-IBTS")   # haul header
hl <- dr_hl(survey = "NS-IBTS")   # length table (illustrative; use actual loaders)

hl_typed <- hl |>
  left_join(hh |> select(.id, DataType, HaulDuration), by = ".id") |>
  dr_add_n_and_cpue() |>
  dr_add_record_type()
```

For this article the full multi-survey dataset (HL joined with HH) was used,
covering all surveys available in the DATRAS parquet store.

### Record type lookup

```{r}
library(obus)
hl_record_type_lookup
```

## Distribution across all surveys

```{r eval = FALSE}
hl_typed |>
  count(record_type, sort = TRUE) |>
  left_join(hl_record_type_lookup |> select(record_type, label),
            by = "record_type") |>
  mutate(pct = round(100 * n / sum(n), 2)) |>
  select(record_type, label, n, pct)
```

| record_type | label | n | % |
|---|---|---:|---:|
| 1 | Length-frequency, standard | 7,450,315 | 52.9 |
| 2 | Length-frequency, sex-disaggregated | 6,226,236 | 44.2 |
| 3 | Length-frequency, with development stage | 24,705 | 0.17 |
| 4 | Length-frequency, invalid haul | 24,298 | 0.17 |
| 10 | Zero/null catch | 65,049 | 0.46 |
| 11 | Weight-only summary | 21,097 | 0.15 |
| 12 | Count (+/- weight) summary, no subsampling | 36,834 | 0.26 |
| 13 | Subsampled summary, no sex | 97,716 | 0.69 |
| 14 | Subsampled summary, sex-disaggregated | 146,374 | 1.04 |
| 15 | Sex-coded null record | 100 | <0.01 |

The great majority of records (97.3 %) carry a `LengthClass` and a
computable `n_haul`. Of those, roughly 46 % are additionally annotated by sex
(type 2) and a small fraction by development/maturity stage (type 3). The
remaining 2.7 % are summary or null records discussed in detail below.

---

## Records with `LengthClass` (types 1–4) {#sec-lc-present}

These are the rows that drive length-frequency and abundance analyses.
`n_haul` is calculated from `NumberAtLength`, `SubsamplingFactor`, and
(for `DataType == "C"`) `HaulDuration`:

| `DataType` | `n_haul` formula |
|---|---|
| `C` — CPUE | `NumberAtLength × SubsamplingFactor × HaulDuration / 60` |
| `R` — by haul | `NumberAtLength × SubsamplingFactor` |
| `P` — pseudocategory | `NumberAtLength × SubsamplingFactor` |
| `S` — subsampled | `NumberAtLength × SubsamplingFactor` |
| `-9` or `NA` | `NA` (invalid haul, type 4) |

**Type 3 note:** `DevelopmentStage` takes priority in the classification. A
record with both `DevelopmentStage` and `SpeciesSex` present is assigned
type 3 (~24 600 such records exist, mostly *Sepia officinalis* in BTS).

### Filtering for standard abundance analysis

```{r eval = FALSE}
hl_lf <- hl_typed |>
  filter(record_type %in% 1:3)   # drops invalid hauls (type 4)
```

To restrict further to unsexed/aggregated data (e.g., when sexed and unsexed
rows for the same length class would double-count):

```{r eval = FALSE}
hl_lf_standard <- hl_typed |>
  filter(record_type == 1)
```

---

## Records without `LengthClass` (types 10–15) {#sec-lc-absent}

All records in this group have `LengthClass`, `LengthCode`, and
`NumberAtLength` equal to `NA`, so `n_haul` is always `NA`. They should be
**excluded from length-frequency and abundance analyses**.

### Type 10 — Zero/null catch {#sec-type10}

These 65 049 records represent species that were explicitly recorded as absent
from a haul. Two findings support this interpretation:

1. `SpeciesValidity` is `5` ("species not found") in 93.8 % of cases and `0`
   in the remaining 6.2 %.
2. In 97 % of hauls that contain a type-10 record, at least one other species
   was measured and assigned a length-frequency record — confirming the haul
   was valid and the species was simply not caught.

```{r eval = FALSE}
# Confirm: hauls with type-10 records mostly also have length data for other species
hauls_with_zero <- hl_typed |>
  filter(record_type == 10L) |>
  distinct(.id)

hl_typed |>
  semi_join(hauls_with_zero, by = ".id") |>
  filter(record_type %in% 1:3) |>
  summarise(hauls_also_with_length = n_distinct(.id))
# 7 080 out of 7 267 (97 %)
```

Type-10 records are concentrated in BTS and NS-IBTS:

| Survey | Records | Hauls | Species |
|---|---:|---:|---:|
| BTS | 43,388 | 4,679 | 106 |
| NS-IBTS | 20,486 | 1,690 | 80 |
| DYFS | 784 | 652 | 22 |
| BITS | 155 | 54 | 16 |
| SNS | 228 | 184 | 6 |

These are standard-species-list absences — exactly the explicit zero-catch
entries needed when computing indices that require hauls with zero catch to be
included. They should **not** be confused with hauls that simply don't mention
a species (implicit zeros).

### Type 11 — Weight-only summary {#sec-type11}

These 21 097 records carry only `SpeciesCategoryWeight`; `TotalNumber`,
`SubsampledNumber`, and `SubsamplingFactor` are all `NA`. Two distinct
sub-populations exist.

**Sub-population A — Bulk bycatch (non-fish taxa):** The dominant pattern
(20 819 of 20 836 distinct haul × species combinations). Species are
invertebrates, algae, and other organisms for which individual length
measurement is not practical: sponges (*Alcyonium*, *Porifera*, *Suberites*),
hydroids, nemerteans, algae (*Fucus*, *Ulva*), tunicates, etc. These records
appear primarily in BTS (17 452 records) with smaller contributions from DYFS,
NS-IBTS, BITS, PT-IBTS, SNS, and SE-SOUND. For these entries the
`SpeciesCategoryWeight` is the total wet weight of the bulk material and is
the only meaningful quantity.

**Sub-population B — Companion weight records (*Sepia officinalis*, BTS;
*Mnemiopsis leidyi*, DYFS):** Only 17 haul × species combinations mix type-11
with length-frequency records. Inspection of those hauls reveals a consistent
four-record-type pattern per haul:

```{r eval = FALSE}
# Typical structure for one Sepia haul
hl_typed |>
  filter(.id == "BTS:2021:3:BE:45CE:BT4A:36:60", ValidAphiaID == 141444L) |>
  distinct(record_type, SpeciesCategory, TotalNumber, SubsampledNumber,
           SubsamplingFactor, SpeciesCategoryWeight) |>
  arrange(record_type)
```

| record_type | SpeciesCategory | TotalNumber | SubsampledNumber | SubsamplingFactor | SpeciesCategoryWeight |
|---|---|---:|---:|---:|---:|
| 1 | 1 | 2 | 2 | 1 | 538 |
| 3 | 1 | NA | NA | 1 | 28 |
| 11 | 1 | NA | NA | 1 | 28 |
| 13 | 1 | 2 | 2 | 1 | 538 |

The type-3 (development-stage length-frequency) and type-11 records carry
**identical** `SpeciesCategoryWeight` values within each haul. The type-11
entry is a companion weight record that reports the total weight for the
development-stage category — it is not a standalone bulk weight. The type-1
(standard length-frequency) and type-13 (subsampled summary) records carry the
overall haul weight. For *Sepia officinalis* this pattern reflects the Belgian
BTS protocol of recording both individual maturity-at-length data and a
companion total weight for that maturity group.

**Practical consequence:** For bulk bycatch (Sub-population A) the weight in
type-11 records is the only information available and can be used in bycatch
weight reporting. For *Sepia* in BTS (Sub-population B) the weight is already
captured in the co-occurring type-3 records; the type-11 record is redundant
and should be ignored to avoid double-counting.

```{r eval = FALSE}
# Bycatch weight per haul for bulk taxa (excluding Sepia companion records)
bycatch_weights <- hl_typed |>
  filter(record_type == 11L) |>
  anti_join(
    hl_typed |>
      filter(record_type %in% 1:3) |>
      distinct(.id, ValidAphiaID),
    by = c(".id", "ValidAphiaID")
  ) |>
  group_by(.id, ValidAphiaID, latin) |>
  summarise(total_weight_g = sum(SpeciesCategoryWeight, na.rm = TRUE),
            .groups = "drop")
```

### Types 12–14 — Summary records without length

These records summarise catch totals at the haul × species level but carry no
length distribution. They appear when a species was caught but either not
subsampled for length measurement (type 12) or subsampled with a factor but
without individual length classes being recorded (types 13–14).

| Type | Key variables present | Typical use |
|---|---|---|
| 12 | `TotalNumber`, optionally `SpeciesCategoryWeight` | Whole-catch totals; no subsampling |
| 13 | `TotalNumber`, `SubsamplingFactor` (± weight) | Subsampled catch aggregate, no sex |
| 14 | `TotalNumber`, `SubsamplingFactor`, `SpeciesSex` (± weight) | Sex-disaggregated catch aggregate |

These records can supply total catch numbers and weights when length
information is not needed, but they **must not be used to derive length
frequencies**. Types 13 and 14 are the no-length counterparts of the
sex-disaggregated type-10 records found in surveys such as NS-IBTS and BTS.

### Type 15 — Sex-coded null record

One hundred records with `SpeciesSex` present but all other measurement fields
absent. These are effectively sex-coded zero-catch entries. They can safely be
treated as type-10 equivalents for most purposes.

---

## Summary: recommended filters by use case

```{r eval = FALSE}
# Length-frequency and abundance / CPUE analysis
hl_for_lf <- hl_typed |>
  filter(record_type %in% 1:3)

# Unsexed only (avoid double-counting in surveys with mixed sex/unsexed rows)
hl_unsexed <- hl_typed |>
  filter(record_type == 1)

# Explicit zero-catch entries (for index calculations needing true zeros)
hl_zeros <- hl_typed |>
  filter(record_type == 10L)

# Bulk bycatch weights (non-fish taxa, excluding Sepia companion records)
hl_bycatch_wt <- hl_typed |>
  filter(record_type == 11L) |>
  anti_join(
    hl_typed |> filter(record_type %in% 1:3) |> distinct(.id, ValidAphiaID),
    by = c(".id", "ValidAphiaID")
  )
```

::: {.callout-warning}
Records of types 10–15 all have `n_haul = NA`. Any pipeline that passes these
rows to functions expecting valid abundance values will silently propagate
`NA`s or, worse, drop rows that should be explicit zeros. Filter to types 1–3
before computing length frequencies or indices.
:::
