---
title: "Parquet Files and DuckDB"
subtitle: "How {obus} serves 14 million rows in under a second"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

```{r}
#| label: setup
#| include: false
library(obus)
library(dplyr)
```

This article explains two technologies that power `{obus}` under the hood — **Parquet** and **DuckDB** — and why they matter even if you never think about them directly.

The short version: the full DATRAS exchange tables (HH, HL, CA) live on a web server as Parquet files. When you call `dr_con()`, a tiny in-process database called DuckDB connects to those files and lets you filter, join, and compute on them using ordinary `{dplyr}` code — without ever downloading the parts you don't need. You can switch to a regular R data frame at any point and **the same `{dplyr}` code keeps working unchanged**.

---

## Why not just CSV?

The traditional way to share tabular data is a CSV file. CSV is readable by nearly any tool, which is why it has outlasted every format that tried to replace it. But CSV has three weaknesses that become painful at DATRAS scale:

**No type information.** A CSV file is plain text. When you read it, the parser has to guess whether `"2026"` is a year (integer) or an ID (character). DATRAS has had persistent problems with fields like `StNo` arriving as integer in some national submissions and character in others, causing `bind_rows()` to fail. Every time you read a CSV you re-run the guessing game.

**Reads the whole file.** If you want only the `Survey`, `Year`, and `HaulNumber` columns from a 14-million-row HL file, a CSV reader still reads all 40+ columns, then discards the ones you did not ask for.

**No compression awareness.** A plain CSV of the full HL table is multiple gigabytes. The same data in Parquet fits in a fraction of that — the exact numbers are measured in the [Compression](#compression) section below.

Parquet fixes all three.

---

## Parquet: the columnar file format

### Columns, not rows

A CSV stores one row at a time:

```
Survey,Year,Quarter,Country,Platform,...
NS-IBTS,2026,1,Scotland,Scotia,...
NS-IBTS,2026,1,Scotland,Scotia,...
```

Parquet stores one **column** at a time: all `Survey` values together, then all `Year` values together, and so on. This sounds like a minor implementation detail, but the consequence is important: if you ask for only `Survey` and `Year`, Parquet reads only those two columns from disk. The other 38 columns are never touched.

### Metadata stored in the file

Every Parquet file carries a **footer** — a compact block of binary metadata at the end of the file. The footer contains:

- column names
- column types (`INT32`, `UTF8`, `DOUBLE`, …)
- number of rows
- row-group boundaries (the file is split into chunks; the footer says where each starts)
- min/max statistics per chunk, per column

When DuckDB first connects to a Parquet file it reads only the footer — a few kilobytes even for a file with millions of rows. From that it knows the schema, the row count, and which chunks it can skip entirely when a filter is applied.

### Strict types

Types are recorded in the file and enforced on read. An `INT32` column comes back as integer, a `UTF8` column comes back as character — no guessing, no surprises. When reading from the XML source (`dr_get(source = "xml")`), the parquet file is not involved at all; types are applied separately by `dr_settypes()` using the `dr_lookup_fields` table as its reference. Either way the result is a correctly typed data frame, but the two paths are independent.

### Compression

Within each column chunk, values are compressed. Parquet supports several algorithms; the obus server files use **Snappy**, a fast algorithm optimised for decompression speed. Similar values compress extremely well: a column that is the string `"NS-IBTS"` repeated 500 000 times compresses to almost nothing. Numeric columns with many repeated values (like `Quarter`, which is only 1–4) shrink similarly.

```{r}
#| label: filesize-comparison
#| echo: false
#| cache: false
#| message: false
#| eval: false

# Parquet file size: read Content-Length from HTTP HEAD — no data transferred
resp <- httr2::request("https://heima.hafro.is/~einarhj/datras/HL.parquet") |>
  httr2::req_method("HEAD") |>
  httr2::req_perform()
parquet_bytes <- as.numeric(httr2::resp_header(resp, "content-length"))

# CSV equivalent: import the full table, write to a temporary file, measure
hl_imported <- dr_get("HL")
tmp_csv <- tempfile(fileext = ".csv")
arrow::write_csv_arrow(hl_imported, tmp_csv)
csv_bytes <- file.size(tmp_csv)
file.remove(tmp_csv)

# Derived values for inline use
parquet_mb <- round(parquet_bytes / 1024^2)
csv_gb     <- round(csv_bytes     / 1024^3, 1)
ratio      <- round(csv_bytes / parquet_bytes)
hl_nrow    <- format(nrow(hl_imported), big.mark = ",")
hl_ncol    <- ncol(hl_imported)
```

The result, measured directly from the server file and a freshly written CSV of the same data: the full HL table (14,397,334 rows, 30 columns) occupies **97 MB** as Parquet. Writing the identical data as a plain CSV produces a **2.1 GB** file — 22 × larger.

---

## DuckDB: a database that lives inside R

### What it is

[DuckDB](https://duckdb.org) is a relational database engine designed for analytical queries on large tables. Unlike PostgreSQL or MySQL, DuckDB runs **in-process** — it starts up as a library inside your R session, with no separate server to install, configure, or connect to. When you install `{obus}`, DuckDB comes along as an R package dependency and is ready to use immediately.

### Connecting to a remote Parquet file

`dr_con()` does two things:

1. Sends an HTTP HEAD request to confirm the file exists.
2. Tells DuckDB to open the Parquet file at that URL.

```{r}
system.time({
  hl <- dr_con("HL")
})
```

That elapsed time is almost entirely the HEAD request. DuckDB reads the Parquet footer — a few kilobytes — and nothing else. The 14-million-row HL table has not moved.

Compare with a full import:

```{r}
system.time({
  hl_local <- dr_get("HL")
})
```

Both give you a table you can work with, but they represent very different things. `hl_local` is a plain R data frame sitting in memory. `hl` is a **lazy connection** — a query plan that has not run yet.

```{r}
class(hl)
nrow(hl)   # NA: DuckDB has not counted rows yet
```

The `NA` row count is a hint: `hl` is not data, it is a **promise** of data, contingent on a query.

---

## dbplyr: writing dplyr, getting SQL

### The translation layer

`{dbplyr}` is the package that makes the magic happen. It intercepts ordinary `{dplyr}` verbs — `filter()`, `select()`, `mutate()`, `group_by()`, `summarise()`, `left_join()` — and translates them into SQL, which DuckDB can then execute against the Parquet file.

You never have to write a line of SQL. You write `{dplyr}`:

```{r}
q <- hl |>
  filter(Survey == "NS-IBTS", Year == 2026, Quarter == 1)
```

What does `q` contain? Still a lazy query. To see the SQL that was built:

```{r}
q |> show_query()
```

Nothing has been downloaded yet. The filter is encoded as a `WHERE` clause that DuckDB will push down to the Parquet row-group level: any chunk whose min/max statistics show it contains no NS-IBTS 2026 Q1 rows is skipped entirely.

### Building up a query step by step

You can chain verbs freely. Each one adds to the SQL plan:

```{r}
q <- hl |>
  dr_add_length_cm() |>
  filter(
    Survey  == "NS-IBTS",
    Quarter == 1,
    !is.na(LengthCode)
  ) |>
  select(.id, Survey, Year, aphia, length_cm, NumberAtLength, SubsamplingFactor)

q |> show_query()
```

The `CASE WHEN` block that `dr_add_length_cm()` adds internally is translated to SQL and included in the same query. No intermediate data frame is created.

### `collect()` — the moment data moves

Only when you call `collect()` does DuckDB actually read data from the Parquet file and return it as a plain R data frame:

```{r}
system.time(
  data <- q |> collect()
)
nrow(data)
```

DuckDB reads only the columns used in the query, skips row-groups that can't match the filter, decompresses what remains, and returns the result. A query that touches a few percent of the 14-million-row file completes in under a second even over an HTTP connection.

### Joins stay lazy too

You can join multiple connections before collecting. The entire join is pushed down to DuckDB:

```{r}
hh <- dr_con("HH")
sp <- dr_con("species")

q <- hl |>
  inner_join(
    hh |> select(.id, HaulValidity, DataType, HaulDuration),
    by = ".id"
  ) |>
  inner_join(sp, by = "aphia") |>
  dr_add_length_cm() |>
  dr_add_n_and_cpue() |>
  filter(
    HaulValidity == "V",
    latin        == "Gadus morhua",
    Survey       %in% c("NS-IBTS", "SCOWCGFS"),
    Quarter      == 1
  ) |>
  select(.id, Survey, Year, latin, length_cm, n_haul, n_hour)

q |> show_query()
```

```{r}
system.time(
  cod <- q |> collect()
)
glimpse(cod)
```

Three Parquet files joined, filtered, and computed — all in DuckDB, before a single row enters R memory.

---

## Same code, two backends

Here is the key insight for the practical user: **you do not need to know which backend you are on**. Once you have a data frame or a DuckDB connection, the `{dplyr}` verbs and all `dr_add_*()` functions work identically on both.

Consider this small helper:

```{r}
summarise_cod <- function(hl_input, hh_input) {
  hl_input |>
    inner_join(hh_input |> select(.id, HaulValidity, DataType, HaulDuration),
               by = ".id") |>
    dr_add_length_cm() |>
    dr_add_n_and_cpue() |>
    filter(HaulValidity == "V", aphia == 126436L) |>
    group_by(Survey, Year, Quarter) |>
    summarise(n_haul_total = sum(n_haul, na.rm = TRUE), .groups = "drop")
}
```

Call it with **lazy DuckDB connections** — nothing downloaded until `collect()`:

```{r}
#| eval: false
summarise_cod(dr_con("HL"), dr_con("HH")) |> collect()
```

Call it with **in-memory data frames** — everything already in R:

```{r}
#| eval: false
hh_local <- dr_get("HH", surveys = "NS-IBTS", years = 2024, quarters = 1)
hl_local <- dr_get("HL", surveys = "NS-IBTS", years = 2024, quarters = 1)
summarise_cod(hl_local, hh_local)
```

Same function, no changes. The `dr_add_*()` functions in `{obus}` are written to handle both `tbl_duckdb_connection` and ordinary `data.frame` inputs; `{dplyr}` dispatches to the right method automatically.

### When to prefer each approach

| Situation | Recommended approach |
|-----------|---------------------|
| Exploring the full dataset, filtering by survey / year / species | `dr_con()` — lazy; only your subset downloads |
| Running the same filtered subset repeatedly in one session | `dr_get()` once, then work on the data frame |
| Joining across HH + HL + species without caring about the full tables | `dr_con()` — DuckDB handles the join before collect |
| Working offline or on a slow connection | `dr_get()` up front, save locally |
| Building a function that others will call either way | Write against `{dplyr}` — it works on both |

---

## What you never have to do

A few things that used to be standard practice with DATRAS data are now unnecessary:

**Maintain local copies.** There is no need to periodically download and version a local `HL.csv`. The Parquet files on the server are the authoritative copy. Connect, filter, collect — and you always get the current version.

**Manage file formats.** You do not need to know whether the underlying storage is Snappy-compressed Parquet, zstd-compressed Parquet, or anything else. DuckDB handles the format; you handle the dplyr.

**Write SQL.** `{dbplyr}` writes it for you. `show_query()` is there if you are curious, not because you need to worry about it.

---

## Summary

```
Parquet file on server
        │
        │  (dr_con reads only the footer — nearly instant)
        ▼
DuckDB connection in R
        │
        │  (dplyr verbs build an SQL plan — no data moves yet)
        ▼
Lazy query (tbl_duckdb_connection)
        │
        │  (collect() executes the query — only matching rows and columns travel)
        ▼
Plain R data frame
```

The three technologies — Parquet, DuckDB, dbplyr — are independently useful, but together they give you interactive performance on a 14-million-row dataset, with no local storage, no SQL, and code that works identically on an in-memory data frame the moment you need to switch.
