---
title: "Zero-fill expansion: a feasibility study"
date: today
format:
  html:
    toc: true
    toc-depth: 3
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Background

Length-based abundance indices require that every haul contributes a record for every species, including a zero count for hauls where the species was not observed. This **zero-filling** step is unavoidable for index calculations, but it is not something the current DATRAS HL parquet provides — the parquet contains only observed (non-zero) catch records. Users who need a zero-filled table must construct it themselves from the raw HL data.

The construction is straightforward in principle: for each Survey × Quarter combination, cross-join the set of distinct hauls with the set of species ever caught in that group, then left-join the observed catch and replace missing counts with zero. Adding a length dimension extends this: each haul × species combination should cover the complete length spectrum (minimum to maximum observed length class, in observed steps) for that species within that Survey × Quarter.

This article asks a practical question: **how large would a pre-computed, zero-filled parquet be, and is it feasible to host it as a web-accessible file?** The motivation is that downloading a pre-expanded parquet — even a large one — may be far cheaper than asking every user to run the expansion themselves against the 14 million-row raw HL.

---

## Setup

```{r}
#| label: load-packages
library(tidyverse)
library(obus)
```

```{r}
#| label: load-data
hl <- dr_con("HL") |> rename(ValidAphiaID = Valid_Aphia)
```

The raw HL parquet covers all DATRAS surveys from 1965 to the present.

```{r}
#| label: raw-size
n_hl_raw <- hl |> count() |> pull(n)
```

It contains `{r} format(n_hl_raw, big.mark = ",")` rows. The file on the server is approximately 102 MB — roughly 7 bytes per row after Snappy compression, reflecting the efficiency of parquet's columnar encoding for survey data with many repeated categorical values.

---

## Level 1: species-only expansion

The first level of expansion produces a skeleton with one row per haul × species combination, where the set of species is constrained to those observed within the same Survey × Quarter. This is sufficient for haul-level CPUE calculations (numbers per hour, total weight) but not for length-frequency indices.

```{r}
#| label: species-grid
# Distinct hauls and species per Survey x Quarter
hauls <- hl |> distinct(Survey, Quarter, .id) |> collect()
species_grid <- hl |> distinct(Survey, Quarter, ValidAphiaID) |> collect()

# Full zero-fill skeleton: cross-product within each Survey x Quarter
full_grid <- inner_join(hauls, species_grid,
                        by = c("Survey", "Quarter"),
                        relationship = "many-to-many")

# Summary counts for reporting
hauls_sq <- hauls |> count(Survey, Quarter, name = "n_hauls")
species_sq <- species_grid |> count(Survey, Quarter, name = "n_species")

grid_summary <- inner_join(hauls_sq, species_sq, by = c("Survey", "Quarter")) |>
  mutate(n_grid_species = n_hauls * n_species)

n_grid_species_total <- sum(grid_summary$n_grid_species)
```

The species-only grid contains `{r} format(n_grid_species_total, big.mark = ",")` rows — `{r} round(n_grid_species_total / n_hl_raw, 1)`× larger than the raw HL. This is a modest expansion: the average haul-to-species ratio within a Survey × Quarter is about four. At the same compression efficiency as the raw HL, a parquet with five columns (`.id`, `Survey`, `Quarter`, `ValidAphiaID`, `n_hour`) would be in the range of a few hundred MB.

---

## Level 2: species × length expansion

The second level adds the length dimension. For each Survey × Quarter × ValidAphiaID group, the full sequence of length classes from the observed minimum to the observed maximum is constructed, with the step size inferred from the smallest gap between consecutive observed lengths. The result is a skeleton with one row per haul × species × length class.

```{r}
#| label: length-sequences
raw_lengths <- hl |>
  distinct(Survey, Quarter, ValidAphiaID, LengthCode, LengthClass) |>
  collect()

length_seqs <- raw_lengths |>
  dr_add_length_mm() |>
  # Remove sentinel zeros (LengthClass = 0) before computing sequences
  filter(length_mm > 0) |>
  # Group by LengthCode so each measurement scale gets its own native step:
  # "." = 1mm bins, "0" = 5mm, "1" = 10mm, "5" = 50mm
  group_by(Survey, Quarter, ValidAphiaID, LengthCode) |>
  reframe(
    length_mm = {
      lv   <- sort(unique(length_mm))
      step <- if (length(lv) > 1) min(diff(lv)) else lv
      seq(lv[1], lv[length(lv)], by = step)
    }
  ) |>
  # Union across LengthCodes: one canonical length grid per species group
  distinct(Survey, Quarter, ValidAphiaID, length_mm)

lengths_sq_aphia <- length_seqs |>
  count(Survey, Quarter, ValidAphiaID, name = "n_lengths")

grid_len_summary <- inner_join(hauls_sq, lengths_sq_aphia,
                                by = c("Survey", "Quarter")) |>
  mutate(n_grid_len = n_hauls * n_lengths)

n_grid_len_total <- sum(grid_len_summary$n_grid_len)
avg_lengths      <- sum(lengths_sq_aphia$n_lengths) / nrow(lengths_sq_aphia)
```

The average number of length classes per Survey × Quarter × species group is `{r} round(avg_lengths, 1)`. This drives a `{r} round(n_grid_len_total / n_grid_species_total, 1)`× further expansion over the species-only grid, resulting in `{r} format(n_grid_len_total, big.mark = ",")` rows total — `{r} round(n_grid_len_total / n_hl_raw, 1)`× the size of the raw HL.

At the same ~7 bytes/row compression as the raw HL, this would correspond to a file of roughly `{r} round(n_grid_len_total * 7 / 1e9, 1)` GB. This is an **upper bound**: the expanded table would have only six columns rather than HL's 28, and the `n_hour` column would be dominated by zeros (highly compressible). In practice the file would likely be substantially smaller, but it cannot be quantified without actually writing it.

---

## Breakdown by survey

```{r}
#| label: survey-table
avg_by_survey <- lengths_sq_aphia |>
  group_by(Survey) |>
  summarise(
    avg_lengths = round(mean(n_lengths), 1),
    max_lengths = max(n_lengths),
    .groups = "drop"
  )

report_tbl <- grid_len_summary |>
  group_by(Survey) |>
  summarise(
    n_hauls_total  = sum(n_hauls),
    n_grid_len     = sum(n_grid_len),
    .groups = "drop"
  ) |>
  left_join(
    grid_summary |>
      group_by(Survey) |>
      summarise(n_grid_species = sum(n_grid_species), .groups = "drop"),
    by = "Survey"
  ) |>
  left_join(avg_by_survey, by = "Survey") |>
  mutate(pct_len = round(n_grid_len / n_grid_len_total * 100, 1)) |>
  arrange(desc(n_grid_len)) |>
  select(Survey, n_hauls_total, n_grid_species, avg_lengths, n_grid_len, pct_len)

report_tbl |>
  mutate(
    n_hauls_total  = format(n_hauls_total,  big.mark = ","),
    n_grid_species = format(n_grid_species, big.mark = ","),
    n_grid_len     = format(n_grid_len,     big.mark = ","),
    pct_len        = paste0(pct_len, "%")
  ) |>
  knitr::kable(
    col.names = c("Survey", "Hauls (rows)", "Species-only grid",
                  "Avg lengths", "Length grid", "% of total"),
    align = "lrrrrr"
  )
```

NS-IBTS alone accounts for `{r} report_tbl |> filter(Survey == "NS-IBTS") |> pull(pct_len)`% of the total length-grid rows, owing to its long time series, high species richness, and fine length resolution. The top three surveys together contribute `{r} report_tbl |> slice_head(n = 3) |> pull(pct_len) |> sum()`% of the total.

---

## Proof-of-concept: species-only parquet write

To move from row-count estimates to actual file sizes, we write the species-only grid to disk and measure directly. The `n_hour` column is set to `0` throughout — a conservative stand-in for the real file where roughly 76% of rows would be genuine zeros (absent species) and the rest non-zero observed CPUE values. Zeros compress perfectly; adding real non-zero values will increase the file size somewhat, but the majority-zero structure is real.

```{r}
#| label: write-parquet-poc
library(arrow)

grid_to_write <- full_grid |> mutate(n_hour = 0)

tmp_snappy <- tempfile(fileext = ".parquet")
tmp_zstd   <- tempfile(fileext = ".parquet")

t0 <- proc.time()
write_parquet(grid_to_write, tmp_snappy, compression = "snappy")
t_snappy <- (proc.time() - t0)["elapsed"]

t0 <- proc.time()
write_parquet(grid_to_write, tmp_zstd, compression = "zstd")
t_zstd <- (proc.time() - t0)["elapsed"]

size_snappy_mb <- file.size(tmp_snappy) / 1e6
size_zstd_mb   <- file.size(tmp_zstd)   / 1e6
```

| | Snappy | Zstd |
|---|---|---|
| File size | `{r} round(size_snappy_mb)` MB | `{r} round(size_zstd_mb)` MB |
| Write time | `{r} round(t_snappy)`s | `{r} round(t_zstd)`s |

The Zstd-compressed file comes in at `{r} round(size_zstd_mb)` MB — **smaller than the raw HL.parquet (102 MB)**, despite containing `{r} round(nrow(grid_to_write) / n_hl_raw, 1)`× as many rows. The explanation is column count: 5 vs 28, with the key categorical columns (`Survey`, `Quarter`, `ValidAphiaID`) repeating heavily across the `.id` groups and compressing very efficiently with parquet's dictionary and RLE encoding.

---

## Proof-of-concept: length-expanded parquet write

The same test applied to the length-expanded grid, using DuckDB's `COPY ... TO` to stream the 2.24 billion-row join directly to disk without loading it into R memory.

```{r}
#| label: write-length-parquet-poc
library(duckdb)

con <- dbConnect(duckdb())
duckdb_register(con, "full_grid",   full_grid)
duckdb_register(con, "length_seqs", length_seqs)

tmp_len_zstd <- tempfile(fileext = ".parquet")

t0 <- proc.time()
dbExecute(con, sprintf(
  "COPY (
    SELECT
      fg.Survey, fg.Quarter, fg.\".id\", fg.ValidAphiaID,
      ls.length_mm,
      0.0 AS n_hour
    FROM full_grid   fg
    JOIN length_seqs ls
      ON  fg.Survey       = ls.Survey
      AND fg.Quarter      = ls.Quarter
      AND fg.ValidAphiaID = ls.ValidAphiaID
  ) TO '%s' (FORMAT PARQUET, COMPRESSION ZSTD)",
  tmp_len_zstd
))
t_len_write <- (proc.time() - t0)["elapsed"]

size_len_mb <- file.size(tmp_len_zstd) / 1e6
dbDisconnect(con)
```

The file is `{r} round(size_len_mb / 1e3, 1)` GB, written in `{r} round(t_len_write)` seconds — far below the 15 GB upper bound. At `{r} round(size_len_mb * 1e6 / n_grid_len_total, 2)` bytes per row, the length dimension compresses even more aggressively than the species-only case: `Survey`, `Quarter`, `ValidAphiaID`, and `length_mm` all repeat across every `.id` in a group, giving parquet's dictionary and RLE encoding very little unique data to store per row.

---

## Summary and conclusions

```{r}
#| label: final-comparison
#| echo: false
tibble(
  Level         = c("Raw HL (baseline)", "Species-only grid", "Species × length grid"),
  Columns       = c(28L, 5L, 6L),
  Rows          = c(n_hl_raw, n_grid_species_total, n_grid_len_total),
  `Size (Zstd)` = c("102 MB",
                     paste0(round(size_zstd_mb), " MB"),
                     paste0(round(size_len_mb / 1e3, 1), " GB")),
  `Write time`  = c("—", paste0(round(t_zstd), "s"), paste0(round(t_len_write), "s"))
) |>
  mutate(Rows = format(Rows, big.mark = ",")) |>
  knitr::kable(align = "lrrrr")
```

Both expanded parquets are viable as hosted files. The species-only grid at `{r} round(size_zstd_mb)` MB is smaller than the raw HL and adds no meaningful burden to the server. The length-expanded grid at `{r} round(size_len_mb / 1e3, 1)` GB is large but manageable; at 1.27 bytes per row the compression is remarkably effective because the six key columns carry little unique information per row.

A few practical considerations for hosting:

- **Partitioning.** The length-expanded file warrants partitioning by `Survey` (and optionally `Quarter`) before upload. This allows DuckDB's HTTP range-request reader to skip irrelevant row groups entirely, making typical single-survey queries fast regardless of total file size.

- **Recomputation.** Both grids must be regenerated whenever the source HL parquet is updated. With write times of under 30 seconds even for the 2.2-billion-row grid, regeneration is not a bottleneck.

- **Zero-fill values.** The files produced here use a placeholder `n_hour = 0` throughout. The actual production parquets should be joined against observed haul-level CPUE so that present-species rows carry real values and only absent-species rows carry zero. This will increase file size modestly — the non-zero fraction is roughly `{r} round(n_hl_raw / n_grid_len_total * 100, 1)`% of the length-grid rows.
