---
title: "Catch products: from raw exchange data to survey indices"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4
)
```

Three functions cover most survey-index workflows. They are designed for
**interactive analysis**: lazy DuckDB queries, automatic zero-filling, and a
concise pipeline that goes from raw exchange data to a collected data frame in
a single chain. If instead you need a **shareable data product** — a parquet
file with unambiguous columns and protocol metadata attached — see the
[Standardized HL](standardize_hl.qmd) article.



| Function | Input | Output | Zero-fill |
|---|---|---|---|
| `dr_catch_by_length()` | raw HH + HL | one row per haul × species × length bin | no |
| `dr_catch_by_haul()` | output of above | one row per haul × species | yes |
| `dr_expand_length()` | output of above | one row per haul × species × length bin | yes |

The typical pipeline is:

```
dr_catch_by_length(hh, hl)  →  filter  →  dr_catch_by_haul(hh)
                                        ↘  dr_expand_length(hh)
```

`dr_catch_by_length()` reads the `NumberAtLength` field and applies
DataType-aware subsampling corrections. It returns only observed rows — no
zeros. The downstream functions add the zero hauls needed for unbiased CPUE
estimation.

```{r load}
library(obus)
library(dplyr)
library(ggplot2)
library(Hmisc)  # required for stat_summary(fun.data = "mean_cl_boot")

hh  <- dr_con("HH")
hl  <- dr_con("HL")

# Build the base catch table once (lazy); filter downstream
cbl <- dr_catch_by_length(hh, hl, haulval = "V")

# Ensure DuckDB can write temp files for large spill-to-disk queries
local({
  tmp <- file.path(tempdir(), "duckdb_temp")
  dir.create(tmp, showWarnings = FALSE, recursive = TRUE)
  DBI::dbExecute(duckdbfs::cached_connection(),
                 sprintf("SET temp_directory='%s'", tmp))
})
```


---

## NS-IBTS Q1 — North Sea demersal survey

### Why zero hauls matter

Species are absent from many hauls. Dropping those zeros inflates the mean CPUE
and makes trends unreliable. `dr_catch_by_haul()` adds an explicit zero row for
every haul where a species was not caught (within the same Survey/Year/Quarter
stratum).

Goatfish (*Mullus surmuletus*) is a good illustration: it appears in virtually
every year since the late 1990s but is caught in only ~13 % of hauls, and when
present it arrives in schools giving a high CPUE. Ignoring zero hauls
overstates the mean by roughly 7-fold. Atlantic cod shows the same bias, more
so in recent decades as the stock declined and hauls with no cod became the
norm.

```{r ns_ibts_zero_demo}
species_ns <- c("goatfish", "Atlantic cod")

cpue_ns <- cbl |>
  filter(Survey == "NS-IBTS", Quarter == 1, species %in% species_ns) |>
  dr_catch_by_haul(hh) |>
  collect()

bind_rows(
  cpue_ns                      |> mutate(zeros = "with zero hauls"),
  cpue_ns |> filter(n_hour > 0) |> mutate(zeros = "without zero hauls")
) |>
  ggplot(aes(Year, n_hour, colour = zeros)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", size = 0.3) +
  facet_wrap(~species, scales = "free_y") +
  scale_colour_manual(values = c("with zero hauls"    = "#2166ac",
                                 "without zero hauls" = "#d73027")) +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = NULL, colour = NULL,
       title = "NS-IBTS Q1 — effect of zero-haul inclusion on CPUE") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Spatial distribution — catch per haul

A bubble plot on haul positions shows where a species is concentrated in a
given year. Here Atlantic cod in 2015; circle area is proportional to numbers
caught per haul (zero hauls shown as small open circles).

```{r ns_ibts_spatial, fig.height=5}
cod_hauls_2015 <- cbl |>
  filter(Survey == "NS-IBTS", Quarter == 1,
         species == "Atlantic cod", Year == 2015) |>
  dr_catch_by_haul(hh) |>
  left_join(select(hh, .id, ShootLongitude, ShootLatitude), by = ".id") |>
  collect()

ggplot(cod_hauls_2015,
       aes(ShootLongitude, ShootLatitude)) +
  geom_sf(data = dr_coastline, inherit.aes = FALSE,
          fill = "grey85", colour = "grey50", linewidth = 0.2) +
  geom_point(data = ~filter(.x, n_haul == 0),
             shape = 1, size = 1, colour = "grey40") +
  geom_point(data = ~filter(.x, n_haul > 0),
             aes(size = n_haul), colour = "#2166ac", alpha = 0.6) +
  scale_size_area(max_size = 10, name = "Numbers per haul") +
  coord_sf(xlim = c(-4, 13), ylim = c(49, 62)) +
  labs(x = NULL, y = NULL,
       title = "NS-IBTS Q1 2015 — Atlantic cod catch per haul") +
  theme_bw() +
  theme(legend.position = "right")
```

### Year-class progression in length distributions

The pipeline is not limited to a single quarter. Passing both Q1 and Q3 of
NS-IBTS in one call shows how the same year-class shifts rightward within a
year as fish grow, and then again across years. Three years bracketing the
strong 2000 year-class illustrate this clearly — within each year Q3 fish are
noticeably larger than Q1 fish, and the dominant cohort advances ~10 cm per
year.

```{r ns_ibts_length, fig.height = 6}
cbl |>
  filter(Survey == "NS-IBTS", Quarter %in% c(1, 3), species == "haddock",
         Year %in% 2000:2002) |>
  dr_expand_length(hh) |>
  collect() |>
  mutate(YQ = paste0(Year, "-", Quarter)) |> 
 # mutate(Quarter = paste0("Q", Quarter)) |>
  ggplot(aes(length_cm, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "ribbon",
               aes(ymin = after_stat(ymin), ymax = after_stat(ymax)),
               alpha = 0.2, colour = NA) +
  stat_summary(fun = mean, geom = "line") +
  #scale_fill_manual(values   = c(Q1 = "#2166ac", Q3 = "#d73027")) +
  #scale_colour_manual(values = c(Q1 = "#2166ac", Q3 = "#d73027")) +
  facet_grid(YQ ~ .) +
  labs(x = "Length (cm)", y = "Numbers per hour (mean ± 95 % boot CI)",
       fill = NULL, colour = NULL,
       title = "NS-IBTS Q1 & Q3 — haddock length distribution 2000–2002") +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(xlim = c(5, 45))
```

### Juvenile index from length-structured data

Length-based indices require zero-filling at the length level before
aggregating — otherwise hauls where no small fish were caught are silently
dropped. `dr_expand_length()` ensures every haul contributes, including those
with zero juveniles.

Haddock spawn in spring (March–May). The NS-IBTS runs in both Q1
(January–March) and Q3 (July–September), giving two windows onto the same
year-classes at different ages:

- **Q3 ≤ 15 cm**: 0-group fish (~3–4 months old), representing the year-class
  spawned that same spring (year-class *Y* seen in Q3 of year *Y*).
- **Q1 ≤ 22 cm**: 1-group fish (~10 months old), representing the year-class
  spawned the previous spring (year-class *Y*−1 seen in Q1 of year *Y*).

Aligning both to year-class puts them on the same axis and provides an
independent cross-check of year-class strength.

```{r ns_ibts_juvenile}
cbl |>
  filter(Survey == "NS-IBTS", Quarter %in% c(1, 3), species == "haddock") |>
  dr_expand_length(hh) |>
  collect() |>
  filter((Quarter == 1 & length_cm <= 22) | (Quarter == 3 & length_cm <= 15)) |>
  mutate(
    year_class = if_else(Quarter == 1, Year - 1L, Year),
    quarter_label = if_else(Quarter == 1,
                            "Q1 (<= 22 cm, age ~10 months)",
                            "Q3 (<= 15 cm, age ~4 months)")
  ) |>
  group_by(year_class, quarter_label, .id) |>
  summarise(n_hour = sum(n_hour), .groups = "drop") |>
  ggplot(aes(year_class, n_hour, colour = quarter_label)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               size = 0.3, position = position_dodge(0.4)) +
  scale_colour_manual(values = c("Q1 (<= 22 cm, age ~10 months)" = "#2166ac",
                                 "Q3 (<= 15 cm, age ~4 months)"  = "#d73027")) +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = "Year-class", colour = NULL,
       title = "NS-IBTS — haddock year-class strength index") +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

## BTS Q3 — Southern North Sea flatfish survey

BTS covers the Dutch, Belgian and English coastal waters in summer. Common dab
and plaice dominate; this survey also records seahorses consistently.

```{r bts_cpue}
species_bts <- c("common dab", "European plaice", "common sole")

cbl |>
  filter(Survey == "BTS", Quarter == 3, species %in% species_bts) |>
  dr_catch_by_haul(hh) |>
  collect() |>
  ggplot(aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               colour = "#1b7837", size = 0.3) +
  facet_wrap(~species, scales = "free_y") +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)",
       x = NULL,
       title = "BTS Q3 — flatfish CPUE indices") +
  theme_bw()
```

### Seahorse — a rare bycatch indicator

Seahorses (*Hippocampus* spp.) are rarely caught but their occurrence in trawl
surveys is used as an indicator of seagrass habitat condition. BTS has the most
consistent records in DATRAS.

```{r bts_seahorse}
seahorse_aphia <- cbl |>
  filter(Survey == "BTS", grepl("Hippocampus", latin)) |>
  distinct(aphia, latin, species) |>
  collect()

cbl |>
  filter(Survey == "BTS", aphia %in% !!seahorse_aphia$aphia) |>
  dr_catch_by_haul(hh) |>
  collect() |>
  mutate(present = n_hour > 0) |>
  group_by(Year, species) |>
  summarise(prevalence = mean(present),
            mean_cpue  = mean(n_hour),
            .groups = "drop") |>
  ggplot(aes(Year)) +
  geom_col(aes(y = prevalence), fill = "#c2a5cf", alpha = 0.7) +
  geom_line(aes(y = mean_cpue / max(mean_cpue) * max(prevalence)),
            colour = "#762a83", linewidth = 0.8) +
  facet_wrap(~species) +
  scale_y_continuous(
    name = "Proportion of hauls with catch (bars)",
    sec.axis = sec_axis(~ . * 1, name = "Scaled mean CPUE (line)")
  ) +
  labs(x = NULL,
       title = "BTS Q3 — seahorse prevalence and relative CPUE") +
  theme_bw()
```

---

## How fast is it?

Ten of the most common NS-IBTS Q1 species — full CPUE pipeline from raw
exchange data through to a collected data frame, timed with `system.time()`.

```{r timing}
species_10 <- c(
  "Atlantic herring", "European sprat", "Norway pout",
  "whiting", "haddock", "common dab",
  "Atlantic cod", "grey gurnard", "Atlantic mackerel",
  "European plaice"
)

t <- system.time(
  cpue_10 <- cbl |>
    filter(Survey == "NS-IBTS", Quarter == 1, species %in% species_10) |>
    dr_catch_by_haul(hh) |>
    collect()
)

cat(sprintf("10 species, all NS-IBTS Q1 years: %.1f s\n", t["elapsed"]))
```

```{r timing_plot, fig.height=6}
cpue_10 |>
  mutate(species = factor(species, levels = species_10)) |>
  ggplot(aes(Year, n_hour)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               colour = "#2166ac", size = 0.2) +
  facet_wrap(~species, scales = "free_y", ncol = 2) +
  labs(y = "Numbers per hour (mean ± 95 % boot CI)", x = NULL,
       title = "NS-IBTS Q1 — ten species CPUE indices") +
  theme_bw()
```

---

## A note on `dr_expand_length()` and large selections

`dr_expand_length()` builds a **complete haul × species × length-bin grid**,
so output size grows as $N_\text{hauls} \times N_\text{species} \times N_\text{length bins}$.
Even so, the DuckDB backend handles surprisingly large jobs efficiently.

```{r expand_timing}
t10 <- system.time(
  res10 <- cbl |>
    filter(Survey == "NS-IBTS", Quarter == 1, species %in% species_10) |>
    dr_expand_length(hh) |>
    collect()
)

cat(sprintf("10 species: %s rows — %.1f s\n",
            format(nrow(res10), big.mark = ","), t10["elapsed"]))
```

Running the same pipeline over all species in NS-IBTS Q1 — every year, every species — takes tens of seconds and produces millions of rows. The only real limit is R memory: the lazy query is always safe, but `collect()` across multiple surveys simultaneously may exceed available RAM.

```{r expand_timing_all, eval=FALSE}
t_all <- system.time(
  res_all <- cbl |>
    filter(Survey == "NS-IBTS", Quarter == 1) |>
    dr_expand_length(hh) |>
    collect()
)

cat(sprintf("all species: %s rows — %.1f s\n",
            format(nrow(res_all), big.mark = ","), t_all["elapsed"]))
```

::: {.callout-tip}
## When to use `dr_catch_by_haul()` vs `dr_expand_length()`

Use `dr_catch_by_haul()` when you only need haul-level totals — it skips the
length grid entirely and is substantially faster. Reserve `dr_expand_length()`
for analyses that genuinely need the length structure: length-based indices,
year-class tracking, or age-length keys.
:::

