---
title: "Standardized HL: the semantic layer for DATRAS length data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4
)
```

## The problem with raw DATRAS HL

The DATRAS HL (length) table is a denormalized source of record: it contains both length-frequency measurements and haul-level bookkeeping. In addition, the value of one variable often requires knowledge of others or external context:

- **`LengthClass`** value depends on **`LengthCode`** (is it mm? 0.5 cm? cm?)
- **`NumberAtLength`** meaning depends on **`DataType`** (is it raw count? CPUE? subsampled?) and **`HaulDuration`**
- **`TotalNumber`** and **`SpeciesCategoryWeight`** are repeated on every length row, creating redundancy
- **`aphia`** is a code; users think in species names (latin or common names).
- **Supporting columns** (`LengthCode`, `SubsamplingFactor`) are implementation details once extracted

Users shouldn't need to understand these interdependencies. They should work with clean, self-contained data.

## The solution: standardized HL

`dr_standardize_hl()` creates a **semantic layer** — a parquet file that is designed
to be written once and reused across many analyses. It is also the recommended
starting point for interactive work: `filter(type == "length")` output feeds
directly into `dr_catch_by_haul()` and `dr_expand_length()` for zero-filling,
while `filter(type == "haul")` gives haul-level numbers and weights with protocol
metadata (`StandardSpeciesCode`, `BycatchSpeciesCode`) attached. See
[Catch products](catch_products.qmd) for worked zero-fill examples.

The standardized HL is a parquet file that is:

- **Self-contained**: every variable has unambiguous meaning. `length_mm` is always millimetres.
- **User-friendly**: species names pre-joined. No codes to translate.
- **Type-flagged**: distinguishes length measurements from haul bookkeeping
- **Publication-ready**: can be served over https and reused by many analyses

```{r load}
library(obus)
library(dplyr)
library(ggplot2)

hh <- dr_con("HH")
hl <- dr_con("HL")

# Generate standardized HL for a small example
hl_std <- dr_standardize_hl(
  hh |> filter(Survey == "NS-IBTS", Quarter == 1, Year == 2015, HaulValidity == "V") |> head(10),
  hl |> filter(Survey == "NS-IBTS", Quarter == 1, Year == 2015)
) |> collect()

cat("Standardized HL rows:", nrow(hl_std), "\n")
cat("Type breakdown:\n")
print(table(hl_std$type))
```

## Two row types, one table

The standardized HL has a `type` flag that separates two fundamentally different data:

```{r}
length_rows <- hl_std |> filter(type == "length") |> head(3)
cat("Type='length' rows — length-frequency measurements:\n")
print(length_rows |> select(.id, species, length_mm, n_haul, n_hour, p_females))

haul_rows <- hl_std |> filter(type == "haul") |> head(3)
cat("\nType='haul' rows — haul-level bookkeeping (deduplicated):\n")
print(haul_rows |> select(.id, species, n_haul, w_haul, p_females))
```

`p_females` encodes sex composition as a single proportion: `n_F / (n_F + n_M)`
computed from fish with recorded sex (`sex` ∈ `{"F", "M"}`). It is `NA` when
no fish in the group were sexed (unsexed catches, `"B"`, `"U"`). Males and
females can be recovered downstream:

```r
filter(type == "length") |>
  mutate(n_females = n_haul * p_females,
         n_males   = n_haul * (1 - p_females))
```

**Why two types in one table?**

1. **Simplicity**: users interact with one parquet file
2. **Flexibility**: users choose which data to use downstream
3. **Completeness**: preserves all caught species (even those not measured at length)
4. **Zero-fill**: type="length" feeds `dr_catch_by_haul()` and `dr_expand_length()` unchanged

For worked examples that build survey indices on top of these two row
types — CPUE, length distributions, sex-specific indices using
`p_females` — see [Catch products](catch_products.qmd).

## Design rationale and trade-offs

### Why this approach?

1. **Separation of concerns**: standardization (units, names, scaling) is independent of filtering/aggregation
2. **Reusability**: one semantic layer, many downstream uses
3. **Reproducibility**: everyone works from the same standardized source
4. **Self-documenting**: the type flag makes the contract explicit

### Expected mismatches

**Length rows don't sum to haul rows.** This is **not a bug**; it's revealing:

```{r}
length_total <- hl_std |>
  filter(type == "length", .id == hl_std$.id[1]) |>
  group_by(.id, species) |>
  summarise(n_haul_from_length = sum(n_haul, na.rm = TRUE), .groups = "drop")

haul_total <- hl_std |>
  filter(type == "haul", .id == hl_std$.id[1]) |>
  select(.id, species, n_haul)

comparison <- full_join(length_total, haul_total, by = c(".id", "species"))
cat("Sample comparison (may not match):\n")
print(head(comparison, 5))
```

**Why they differ:**

- **`NumberAtLength`** (length rows) = individuals actually measured at length
- **`TotalNumber`** (haul rows) = total individuals caught, regardless of length measurement
- **Not all fish are measured** — some are too small, damaged, or unmeasurable

This is **expected behavior in real DATRAS data**, not an error.

### Weaknesses

1. **Redundancy eliminated** — you can't reconstruct the original HL from type="haul" rows alone (species/sex/category grouping is lost)
2. **Inference required** — if you need to know which fish went unmeasured, you'd need to rejoin to raw HL
3. **Trade-off**: smaller, cleaner file vs. lose granular subsampling details

## When to use each type

| Need | Use | Example |
|------|-----|---------|
| Length structure, zero-filled per haul | `filter(type=="length")` → `dr_expand_length(hh)` | ALK, maturity ogive |
| Haul-level CPUE, zero-filled per species | `filter(type=="length")` → `dr_catch_by_haul(hh)` | CPUE index, stock assessment |
| Haul totals with weight | `filter(type=="haul")` | biomass index |
| Sex composition at length | `filter(type=="length")` → use `p_females` | sex-ratio at size — see [Catch products](catch_products.qmd) |
| All caught species (even unmeasured) | `filter(type=="haul")` species list | ecosystem analysis |
| Know which fish went unmeasured | rejoin to raw HL | detailed QA/QC |

## Sampling protocol flags

The output carries two haul-level flags, `StandardSpeciesCode` and `BycatchSpeciesCode`, that document what the vessel was required to record. These can substantially affect any analysis that depends on species presence or zero-hauls. See the [Sampling protocols](https://einarhjorleifsson.github.io/imbus/DATRAS/sampling_protocol.html) article in the IMBUS documentation for a detailed treatment, including examples of how ignoring these flags leads to phantom zeros and biased CPUE indices.

To exclude hauls with incomplete recording before zero-filling:

```{r eval=FALSE}
hl_std <- dr_standardize_hl(hh, hl)

complete_hauls <- hl_std |>
  filter(type == "haul",
         StandardSpeciesCode == "1",   # full standard species list
         BycatchSpeciesCode  != "0")   # some bycatch recording done
```

## Implementation

`dr_standardize_hl(hh, hl, species = NULL, haulval = NULL)` takes:
- `hh`: haul header table (with DataType, HaulDuration, StandardSpeciesCode, BycatchSpeciesCode)
- `hl`: length table (with NumberAtLength, TotalNumber, etc.)
- `species`: species lookup (defaults to `dr_con("species")`)
- `haulval`: optional haul validity filter (e.g. `"V"` only)

Output: lazy DuckDB table with 19 columns and two row types. Can be written to parquet:

```{r eval=FALSE}
dr_standardize_hl(hh, hl) |>
  dplyr::collect() |>
  arrow::write_parquet("hl_standardized.parquet")
```

## Why now?

The standardized HL solves a user experience problem that has existed since DATRAS was published. By pre-computing the hard parts (scaling, naming, deduplication), we let users focus on their actual questions: population dynamics, recruitment, gear selectivity, etc. — not on decoding DATRAS conventions.
